### 2020-10-29 paper/paper.Rmd
updates

---
title: "**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: msystems.csl
fontsize: 12pt
geometry: margin=1.0in
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
  echo = FALSE,
  eval = TRUE,
  warning = FALSE,
  cache = FALSE
)

inline_hook <- function(x){
	print(x)
	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=2, nsmall=2, scientific=FALSE))
		}

	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)
```

\begin{center}
\vspace{25mm}

Kelly L. Sovacool${^1}$, Sarah L. Westcott${^2}$, M. Brodie Mumphrey${^1}$,
Gabrielle A. Dotson${^1}$, Patrick D. Schloss${^2\dagger}$

$1$ Department of Computational Medicine and Bioinformatics,
University of Michigan, Ann Arbor, MI 48109

$2$ Department of Microbiology and Immunology,
University of Michigan, Ann Arbor, MI 48109

\vspace{30mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

\end{center}

\newpage
\linenumbers

## Abstract

Assigning amplicon sequences to Operational Taxonomic Units (OTUs) is an
important step in characterizing the composition of microbial communities across
large datasets. OptiClust, a _de novo_ OTU clustering method in the mothur
program, has been shown to produce higher quality OTU assignments than other
methods and at comparable or faster speeds [@westcott_opticlust_2017;
@schloss_introducing_2009]. A notable difference between _de novo_ clustering
and database-dependent methods is that OTU assignments clustered with _de novo_
methods are not stable when new sequences are added to a dataset
[@westcott_novo_2015]. However, in some cases one may wish to incorporate new
samples into a previously clustered dataset without performing clustering again
on all sequences, such as when deploying a machine learning model where OTUs are
features [@topcuoglu_framework_2020]. To provide an efficient and robust method
to fit amplicon sequence data to existing OTUs, we developed the OptiFit
algorithm as a new component of the mothur program. To benchmark the OptiFit
algorithm against _de novo_ clustering with the OptiClust algorithm, we used
four published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse [@schloss_stabilization_2012], and human
[@baxter_microbiota-based_2016] samples. For each dataset, a subset of sequences
was randomly selected and clustered into OTUs with OptiClust, then the remaining
sequences were fit to the existing OTUs using the OptiFit algorithm. This was
repeated with subsets of varying sizes ranging from 10 to 90% of sequences in
order to evaluate the bounds of the dataset size required for OptiFit.
Separately, all sequences were clustered with OptiClust to provide a baseline of
OTU assignment quality and runtime performance. Each of these routines was
repeated 10 times with different random seeds to produce results that are robust
to random variation. OTU quality was evaluated using the Matthews Correlation
Coefficient (MCC) with a sequence similarity threshold of 97% as described
previously [@schloss_application_2016; @westcott_novo_2015]. On average, fitting
sequences into existing OTUs with OptiFit performed 10 times faster than _de
novo_ clustering with OptiClust, while the average MCC scores produced were
nearly indistinguishable across each dataset. The OptiFit results across subset
sizes ranging from 10 to 90% of sequences were also very similar, with slightly
higher MCC scores for larger subset sizes. Thus, OptiFit is an efficient way to
fit new sequences to existing OTUs yet without sacrificing the quality of OTU
assignments.

### Importance

\newpage

## Introduction

## Results

## Discussion

## Materials and Methods

## Acknowledgements

KLS received support from the NIH Training Program in Bioinformatics (T32 GM070449).

The funders had no role in study design, data collection and interpretation, 
or the decision to submit the work for publication.

## References

### 2020-10-29 code/R/stats.R

@@ -1,0 +1,94 @@
+library(tidyverse)
+library(mosaic)
+library(reshape2)
+
+#functions adapted from Begum mbio paper
+read_files <- function(filenames){
+  for(file in filenames){
+    # Read the files generated in main.R
+    # These files have cvAUCs and testAUCs for 100 data-splits
+    data <- read.delim(file, header=T, sep=',') %>%
+      mutate(level=unlist(strsplit(file,"-"))[2])
+    
+  }
+  return(data)
+}
+
+#compare two models levels within a taxonomic level
+perm_p_value <- function(data, model_name_1, model_name_2, level_name){
+  test_subsample <- data %>%
+    select(-cv_auroc) %>%
+    filter((method == model_name_1 | method == model_name_2) & level == level_name)
+  obs <- abs(diff(mosaic::mean(test_auroc ~ method, data = test_subsample)))
+  auc.null <- do(10000) * diff(mosaic::mean(test_auroc ~ shuffle(method), data = test_subsample))
+  n <- length(auc.null[,1])
+  # r = #replications at least as extreme as observed effect
+  r <- sum(abs(auc.null[,1]) >= obs)
+  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
+  p.value=(r+1)/(n+1)
+  #plot <- ggplot(auc.null, aes(x=auc.null[,1], color=auc.null[,1]>=obs)) + geom_histogram(fill="white", alpha=0.5, position="identity") + 
+  #  scale_color_brewer(palette="Dark2") + geom_vline(xintercept = obs,color="grey",lty="dashed")
+  return(p.value)
+}
+
+#compare two taxonomic levels within a model type
+perm_p_value_model <- function(data, model_name, level_name_1, level_name_2){
+  test_subsample <- data %>%
+    select(-cv_auroc) %>%
+    filter(method == model_name & ( level == level_name_1 | level == level_name_2))
+  obs <- abs(diff(mosaic::mean(test_auroc ~ level, data = test_subsample)))
+  auc.null <- do(10000) * diff(mosaic::mean(test_auroc ~ shuffle(level), data = test_subsample))
+  n <- length(auc.null[,1])
+  # r = #replications at least as extreme as observed effect
+  r <- sum(abs(auc.null[,1]) >= obs)
+  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
[…]

### 2020-10-29 code/concat_pipeline_auc_output.py

@@ -6,14 +6,14 @@ parser = argparse.ArgumentParser(description='Provide model and taxonomy level')
 parser.add_argument("--taxonomy", action ="store", dest="tax_level",
     help="Taxonomy options: kingdom, phylum, class, order, family, genus, otu, asv")
 parser.add_argument("--model", action="store", dest="model",
-    help= "Model options: L2_Logistic_Regression, L1_Linear_SVM, L2_Linear_SVM, RBF_SVM, Decision_Tree, Random_Forest, XGBoost")
+    help= "Model options: rpart2, regLogistic, svmRadial, rf, xgbTree")
 
 args = parser.parse_args()
 tax_level = args.tax_level
 model = args.model
 
 
-search_dir = "data/temp/"+tax_level
+search_dir = "data/"+tax_level+"/temp"
 output_file = "data/process/combined-"+tax_level+"-"+model+".csv"
 
 output = open(output_file, "w+" )
@@ -21,7 +21,7 @@ seed = 1
 
 for file in os.listdir(search_dir):                             #loop over files in each taxonomy directory
     while seed <= 100:                                           #for all seed 100 files
-        input_file = "data/temp/"+tax_level+"/"+model+"."+str(seed)+".csv"
+        input_file = "data/"+tax_level+"/temp/"+model+"."+str(seed)+".csv"
         input = open(input_file,"r")
         if seed == 1:                                           #we want the header and the info
             lines = input.readlines()                           #read each line into a list

### 2020-10-29 code/datasets_process.sh

@@ -14,7 +14,7 @@ Rscript code/datasets_make_files.R $RAW
 
 
 # Run data through standard mothur pipeline through the steps prior to clustering
-mothur "#set.seed(seed=19760620);
+bin/mothur/mothur "#set.seed(seed=19760620);
 	set.dir(output=$MOTHUR);
 	make.contigs(inputdir=$RAW, file=data.files, processors=8);
 	screen.seqs(fasta=current, group=current, maxambig=0, maxlength=275, maxhomop=8);

### 2020-10-29 code/get_asv_shared.sh

@@ -6,7 +6,7 @@ FASTA=data/mothur/crc.fasta
 COUNT=data/mothur/crc.count_table
 TAXONOMY=data/mothur/crc.taxonomy
 
-mothur "#set.seed(seed=19760620);
+bin/mothur/mothur "#set.seed(seed=19760620);
 	make.shared(count=$COUNT, label=ASV, outputdir=$WORKDIR);
 	classify.otu(list=current, count=current, taxonomy=$TAXONOMY);
 	sub.sample(shared=current, size=10500)"

### 2020-10-29 code/get_otu_shared.sh

@@ -1,7 +1,7 @@
 PROCESSORSSPLIT=8
 PROCESSORSCLUSTER=1
 mkdir -p data/otu
-mothur "#set.seed(seed=19760620);
+bin/mothur/mothur "#set.seed(seed=19760620);
 	set.dir(input=data/mothur, output=data/otu);
 	cluster.split(fasta=crc.fasta, count=crc.count_table, taxonomy=crc.taxonomy, taxlevel=4, processors=$PROCESSORSSPLIT, runsensspec=FALSE, cluster=FALSE);
 	cluster.split(file=current, processors=$PROCESSORSCLUSTER, runsensspec=FALSE);

### 2020-10-29 run.sh

@@ -61,4 +61,4 @@ make data/otu/input_data_preproc.csv
 
 make data/asv/crc.shared
 make data/asv/input_data.csv
-make data/asv/input_data_preproc.csv
+#make data/asv/input_data_preproc.csv

### 2021-01-18 code/R/merge_hyperparameters.R
add hyperparamter tuning RF

@@ -1,0 +1,44 @@
+library(docopt)
+library(tidyverse)
+
+'Merge hyperparameter data.
+
+Usage:
+  merge_hyperparameters.R --level=<level> --method=<method>
+  merge_hyperparameters.R (-h | --help)
+
+Options
+  -h --help 	     Display this message.
+  --method=<method>  Model name, options
+			rf
+			glmnet
+			svmRadial
+			rpart2
+			xgbTree
+  --level=<level>    Taxonomic level.
+
+' -> doc
+
+args <- docopt(doc)
+
+level <- args$level
+method <- args$method
+
+#function from Begum mBio 2020 paper
+read_files <- function(filenames){
+  for(file in filenames){
+    data <- read.delim(file, header=T, sep=',') %>%
+	mutate(seed=unlist(strsplit(file,"\\."))[3])
+  }
+  return(data)
+}
+
+#read in all the importance files for the 100 seeds
+files <- list.files(path=paste0("./data/",level,"/temp/"),pattern=paste0("hp_",method,".*"),full.names=TRUE)
+
+#build the merged dataframe
+hp <- map_df(files, read_files) %>%
+   mutate(level=level)
+
+outfile <- paste0("./data/process/hp-",level,"-",method,".csv")
+write_csv(hp,outfile)

### 2021-01-18 code/R/merge_importance.R

@@ -1,0 +1,43 @@
+library(docopt)
+library(tidyverse)
+
+'Merge importance data.
+
+Usage:
+  merge_importance.R --level=<level> --method=<method>
+  merge_importance.R (-h | --help)
+
+Options
+  -h --help 	     Display this message.
+  --method=<method>  Model name, options
+			rf
+			glmnet
+			svmRadial
+			rpart2
+			xgbTree
+  --level=<level>    Taxonomic level.
+
+' -> doc
+
+args <- docopt(doc)
+
+level <- args$level
+method <- args$method
+
+#function from Begum mBio 2020 paper
+read_files <- function(filenames){
+  for(file in filenames){
+    data <- read.delim(file, header=T, sep=',')    
+  }
+  return(data)
+}
+
+#read in all the importance files for the 100 seeds
+files <- list.files(path=paste0("./data/",level,"/temp/"),pattern=paste0("importance_",method,".*"),full.names=TRUE)
+
+#build the merged dataframe
+importance <- map_df(files, read_files) %>%
+   mutate(level=level)
+
+outfile <- paste0("./data/process/importance-",level,"-",method,".csv")
+write_csv(importance,outfile)

### 2021-01-18 code/R/preprocess_data.R

@@ -27,10 +27,11 @@ data <- as_tibble(fread(args$data))
 # for ASV level data, filter out columns with count 1 
 # these will be removed anyway in preprocessing
 # so this speeds up the processing for large data
-if( args$taxonomy == "asv" ){
+if( args$taxonomy %in% c("otu","asv") ){
 	data <- data[,colSums(data != 0) > 1]
 }
 
-data_proc <- mikropml::preprocess_data(data, 'dx')$dat_transformed
+data_proc <- mikropml::preprocess_data(data, 'dx')
 
-write_csv(data_proc,paste0("./data/",args$taxonomy,"/input_data_preproc.csv"))
+#write_csv(as.data.frame(data_proc$removed_feats),paste0("./data/",args$taxonomy,"/input_data_removed.csv"),col_names=FALSE)
+write_csv(data_proc$dat_transformed,paste0("./data/",args$taxonomy,"/input_data_preproc.csv"))

### 2021-01-18 code/R/run_model.R

@@ -1,3 +1,5 @@
+starttime=Sys.time()
+
 library(docopt)
 library(mikropml)
 library(tidyverse)
@@ -5,7 +7,7 @@ library(tidyverse)
 'Run mikRopML pipeline.
 
 Usage:
-  run_model.R --data=<csv> --method=<name> --seed=<num> --taxonomy=<level> --outcome_colname=<colname> --outcome_value=<value>
+  run_model.R --data=<csv> --method=<name> --seed=<num> --taxonomy=<level> --outcome_colname=<colname> --outcome_value=<value> --hyperparams=<csv>
   run_model.R (-h | --help)
 
 Options
@@ -21,6 +23,7 @@ Options
   --taxonomy=<level>           taxonomic level.
   --outcome_colname=<colname>  colname in data for outcome variable.
   --outcome_value=<value>      value of interest in outcome column.
+  --hyperparams=<csv>          csv with hyperparameter values.
 
 ' -> doc
 
@@ -31,21 +34,42 @@ if( !dir.exists(outdir) ){
         dir.create(outdir,recursive=T)
 }
 
-data_proc <- read_csv(paste0("./data/",args$taxonomy,"/input_data_preproc.csv"))
+# read in preprocessed data
+data_proc <- read_csv(args$data)
+
+# get hyperparamter values for model/taxonomy
+hp <- read_csv(args$hyperparams)
+sub_hp <- hp %>% 
+	filter(model == args$method, level == args$taxonomy)
+
+# create list of hyperparams
+new_hp <- list()
+params <- unique(sub_hp$param)
+for(i in 1:length(params)){
+	p <- params[i]  
+	new_hp[[p]] <- sub_hp %>% 
+		filter(param == p) %>% 
+		pull(number)    
+}
 
 model <- mikropml::run_ml(dataset = data_proc,
                           method = args$method,
                           outcome_colname = args$outcome_colname, 
-                          outcome_value = args$outcome_value,
-                          seed = as.numeric(as.character(args$seed)) )
-model
+                          seed = as.numeric(as.character(args$seed)),
+			  find_feature_importance=TRUE,
+			  hyperparameters = new_hp )
+
+feature_importance <- model$feature_importance
+write_csv(feature_importance,paste0(outdir,"importance_",args$method,".",args$seed,".csv"))
 
 hyperparameters <- get_hp_performance(model$trained_model)$dat
[…]

### 2021-01-18 run.sh
updates

@@ -13,12 +13,12 @@
 #SBATCH --ntasks-per-node=1
 #SBATCH --cpus-per-task=1
 #SBATCH --mem-per-cpu=100GB
-#SBATCH --time=200:00:00
+#SBATCH --time=20:00:00
 # Account
-#SBATCH --account=pschloss99
+#SBATCH --account=pschloss1
 #SBATCH --partition=standard
 # Logs
-#SBATCH --mail-user=armourc@med.umich.edu
+#SBATCH --mail-user=armourc@umich.edu
 #SBATCH --mail-type=END,FAIL
 #SBATCH --output=slurm_logs/%x-%a-%j.out
 # Environment
@@ -30,35 +30,35 @@
 #                   #
 #####################
 
-source /etc/profile.d/http_proxy.sh  # required for internet on the Great Lakes cluster
-make data/references/silva.v4.%
-make data/trainset16_022016.pds.%
-make data/mothur/crc.fasta
+#source /etc/profile.d/http_proxy.sh  # required for internet on the Great Lakes cluster
+#make data/references/silva.v4.%
+#make data/trainset16_022016.pds.%
+#make data/mothur/crc.fasta
 
-make data/phylum/crc.shared
-make data/phylum/input_data.csv
-make data/phylum/input_data_preproc.csv
+#make data/otu/crc.shared
+#make data/otu/input_data.csv
+#make data/otu/input_data_preproc.csv
 
-make data/class/crc.shared
-make data/class/input_data.csv
-make data/class/input_data_preproc.csv
+#make data/phylum/crc.shared
+#make data/phylum/input_data.csv
+#make data/phylum/input_data_preproc.csv
 
-make data/order/crc.shared
-make data/order/input_data.csv
-make data/order/input_data_preproc.csv
+#make data/class/crc.shared
+#make data/class/input_data.csv
+#make data/class/input_data_preproc.csv
 
-make data/family/crc.shared
-make data/family/input_data.csv
-make data/family/input_data_preproc.csv
+#make data/order/crc.shared
+#make data/order/input_data.csv
+#make data/order/input_data_preproc.csv
 
-make data/genus/crc.shared
-make data/genus/input_data.csv
-make data/genus/input_data_preproc.csv
+#make data/family/crc.shared
+#make data/family/input_data.csv
+#make data/family/input_data_preproc.csv
 
-make data/otu/crc.shared
-make data/otu/input_data.csv
[…]

### 2021-01-19 paper/manuscript.Rmd
initial manuscript draft

---
output:
  word_document: default
  pdf_document:
    keep_tex: true
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
---

# TITLE GOES HERE

\vspace{20mm}

**Running title:** Optimal resolution

\vspace{10mm}

Courtney R. Armour, (ande? begum? etc?), Patrick D. Schloss ${^\dagger}$

\vspace{20mm}

${\dagger}$ To whom correspondence should be addressed:

\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

Department of Microbiology

University of Michigan

Ann Arbor, MI 48109

\vspace{20mm}

**(observation format - max 1200 words, 2 figures, 25 ref)**

\newpage

## Abstract (max 250 words)

## Importance (max 150 words)

\newpage

## Introduction

\newpage

## Results

\newpage

## Conclusion

\newpage

## Materials and Methods
(data availability)

\newpage

## Acknowledgements

\newpage

## Figures

### 2021-01-19 paper/manuscript.tex
initial manuscript draft

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\linenumbers
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\hypertarget{title-goes-here}{%
\section{TITLE GOES HERE}\label{title-goes-here}}

\vspace{20mm}

\textbf{Running title:} Optimal resolution

\vspace{10mm}

Courtney R. Armour, (ande? begum? etc?), Patrick D. Schloss
\({^\dagger}\)

\vspace{20mm}

\({\dagger}\) To whom correspondence should be addressed:

\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

Department of Microbiology

University of Michigan

Ann Arbor, MI 48109

\vspace{20mm}

\textbf{(observation format - max 1200 words, 2 figures, 25 ref)}

\newpage

\hypertarget{abstract-max-250-words}{%
\subsection{Abstract (max 250 words)}\label{abstract-max-250-words}}

\hypertarget{importance-max-150-words}{%
\subsection{Importance (max 150 words)}\label{importance-max-150-words}}

\newpage

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\newpage

\hypertarget{results}{%
\subsection{Results}\label{results}}

\newpage

\hypertarget{conclusion}{%
\subsection{Conclusion}\label{conclusion}}

\newpage

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

(data availability)

\newpage

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

\newpage

\hypertarget{figures}{%
\subsection{Figures}\label{figures}}

\end{document}

### 2021-01-19 paper/manuscript.Rmd
initial outline

@@
 
-**(observation format - max 1200 words, 2 figures, 25 ref)**
+**observation format - max 1200 words, 2 figures, 25 ref**
 
@@
 
+- CRC is one of the most common cancers and a leading cause of cancer related death
+- There is evidence that the microbiome has a role in CRC development/progression and could be useful for biomarker detection and diagnostics.
+- Begum et al (mBio 2020) recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
+- A common question when applying ML methods to microbiome data is which method and taxonomic level should be use.
+- This analysis utilizes the reproducible framework developed by Begum et al to quantify which ML method and taxonomic level produce the best performing classifier for CRC data.
+
 \newpage
@@
 ## Results
+- Of the five ML methods tested, Random forest was consistently the top performer (supplemental figure of all models?) at most taxonomic levels.
+  - RF might be more appropriate anyways since its more suitable for zero inflated data? (need to look into literature)
+- Within the RF model, the highest AUCs were observed for family, genus, and otu level data with no significant difference between the three. (Figure 1)
 
@@
 ## Materials and Methods
-(data availability)
+- data from prior study {baxter}
+- mikropml package
+- pvalues as previously described {begum}
 

### 2021-01-19 paper/manuscript.tex
initial outline

@@
 
+\begin{itemize}
+\tightlist
+\item
+  CRC is one of the most common cancers and a leading cause of cancer
+  related death
+\item
+  There is evidence that the microbiome has a role in CRC
+  development/progression and could be useful for biomarker detection
+  and diagnostics.
+\item
+  Begum et al (mBio 2020) recently demonstrated effective application of
+  machine learning (ML) to microbiome based classification problems and
+  developed a framework for applying ML practices in a more reproducible
+  way (mikropml).
+\item
+  A common question when applying ML methods to microbiome data is which
+  method and taxonomic level should be use.
+\item
+  This analysis utilizes the reproducible framework developed by Begum
+  et al to quantify which ML method and taxonomic level produce the best
+  performing classifier for CRC data.
+\end{itemize}
+
 \newpage
@@
 \subsection{Results}\label{results}}
+
+\begin{itemize}
+\tightlist
+\item
+  Of the five ML methods tested, Random forest was consistently the top
+  performer (supplemental figure of all models?) at most taxonomic
+  levels.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    RF might be more appropriate anyways since its more suitable for
+    zero inflated data? (need to look into literature)
+  \end{itemize}
+\item
+  Within the RF model, the highest AUCs were observed for family, genus,
+  and otu level data with no significant difference between the three.
+  (Figure 1)
+\end{itemize}
 
@@
 
-(data availability)
+\begin{itemize}
+\tightlist
+\item
+  data from prior study \{baxter\}
+\item
+  mikropml package
+\item
+  pvalues as previously described \{begum\}
+\end{itemize}
 

### 2021-01-19 code/R/run_model.R
modify option for hyperparams

@@ -37,27 +37,36 @@ if( !dir.exists(outdir) ){
 # read in preprocessed data
 data_proc <- read_csv(args$data)
 
-# get hyperparamter values for model/taxonomy
-hp <- read_csv(args$hyperparams)
-sub_hp <- hp %>% 
-	filter(model == args$method, level == args$taxonomy)
+if(args$method == "rf"){
+  # get hyperparamter values for model/taxonomy
+  hp <- read_csv(args$hyperparams)
+  sub_hp <- hp %>%
+  	filter(model == args$method, level == args$taxonomy)
 
-# create list of hyperparams
-new_hp <- list()
-params <- unique(sub_hp$param)
-for(i in 1:length(params)){
-	p <- params[i]  
-	new_hp[[p]] <- sub_hp %>% 
-		filter(param == p) %>% 
-		pull(number)    
+  # create list of hyperparams
+  new_hp <- list()
+  params <- unique(sub_hp$param)
+  for(i in 1:length(params)){
+  	p <- params[i]
+  	new_hp[[p]] <- sub_hp %>%
+  		filter(param == p) %>%
+  		pull(number)
+  }
+
+  model <- mikropml::run_ml(dataset = data_proc,
+                            method = args$method,
+                            outcome_colname = args$outcome_colname,
+                            seed = as.numeric(as.character(args$seed)),
+  			  find_feature_importance=TRUE,
+  			  hyperparameters = new_hp )
+}else {
+  model <- mikropml::run_ml(dataset = data_proc,
+                            method = args$method,
+                            outcome_colname = args$outcome_colname,
+                            seed = as.numeric(as.character(args$seed)),
+                            find_feature_importance=TRUE)
+
 }
-
-model <- mikropml::run_ml(dataset = data_proc,
-                          method = args$method,
-                          outcome_colname = args$outcome_colname, 
-                          seed = as.numeric(as.character(args$seed)),
-			  find_feature_importance=TRUE,
-			  hyperparameters = new_hp )
 
 feature_importance <- model$feature_importance
 write_csv(feature_importance,paste0(outdir,"importance_",args$method,".",args$seed,".csv"))

### 2021-01-19 paper/manuscript.Rmd
outline results

@@
 ## Results
-- Of the five ML methods tested, Random forest was consistently the top performer (supplemental figure of all models?) at most taxonomic levels.
-  - RF might be more appropriate anyways since its more suitable for zero inflated data? (need to look into literature)
-- Within the RF model, the highest AUCs were observed for family, genus, and otu level data with no significant difference between the three. (Figure 1)
+- Across the five ML methods tested, model performance tends to increase with taxnomic level usually peaking around genus level and dropping off slightly with ASVs.
+- Random forest was consistently the top performer at most taxonomic levels.
+  - RF might be more appropriate for microbiome analysis since its more suitable for zero inflated data (need to look into literature)
+- Within the RF model, the highest AUCs were observed for family, genus, and otu level data with no significant difference between the three (Figure 1).
+ASV performed significantly lower than OTU but not family and genus levels.
 

### 2021-01-19 paper/manuscript.tex
outline results

@@
 
-\textbf{(observation format - max 1200 words, 2 figures, 25 ref)}
+\textbf{observation format - max 1200 words, 2 figures, 25 ref}
 
@@
 \item
-  Of the five ML methods tested, Random forest was consistently the top
-  performer (supplemental figure of all models?) at most taxonomic
+  Across the five ML methods tested, model performance tends to increase
+  with taxnomic level usually peaking around genus level and dropping
+  off slightly with ASVs.
+\item
+  Random forest was consistently the top performer at most taxonomic
   levels.
@@
   \item
-    RF might be more appropriate anyways since its more suitable for
-    zero inflated data? (need to look into literature)
+    RF might be more appropriate for microbiome analysis since its more
+    suitable for zero inflated data (need to look into literature)
   \end{itemize}
@@
   Within the RF model, the highest AUCs were observed for family, genus,
-  and otu level data with no significant difference between the three.
-  (Figure 1)
+  and otu level data with no significant difference between the three
+  (Figure 1). ASV performed significantly lower than OTU but not family
+  and genus levels.
 \end{itemize}

### 2021-01-20 code/R/plot_hp_performance.R
add code to make hp plots

@@ -1,0 +1,36 @@
+library(tidyverse)
+
+#functions adapted from begum mbio paper
+read_files <- function(filenames){
+  for(file in filenames){
+    # Read the files generated in main.R
+    # These files have cvAUCs and testAUCs for 100 data-splits
+    data <- read.delim(file, header=T, sep=',') %>%
+      mutate(level=unlist(strsplit(file,"-"))[2])
+
+  }
+  return(data)
+}
+
+hp_files <- list.files(path= 'data/process', pattern='hp-.*-.*.csv', full.names = TRUE)
+hp.df <- map_df(hp_files, read_files)
+
+hp.df %>% 
+  group_by(mtry,level) %>%
+  summarise("mean_AUC" = mean(AUC),
+            "sd_AUC" = sd(AUC),
+            # is there a less repetitive way to do this cleanly?
+            ymin_metric = mean_AUC - sd_AUC,
+            ymax_metric = mean_AUC + sd_AUC) %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                      labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV"))) %>% 
+  ggplot(aes(x=mtry,y=mean_AUC)) +
+  geom_point(size=2,alpha=0.5) + geom_line(alpha=0.5) +
+  facet_grid(.~level,scales = "free_x") +
+  theme_bw() +
+  geom_errorbar(aes(
+    ymin = .data$ymin_metric,
+    ymax = .data$ymax_metric),
+    width = .001)
+
+ggsave("analysis/hp-rf.png")

### 2021-01-21 code/R/check_hp_performance.R
update hp plotting

@@ -1,0 +1,62 @@
+library(tidyverse)
+library(docopt)
+library(mikropml)
+library(cowplot)
+
+'check hp performance
+
+Usage:
+  check_hp_performance.R --input=<csv>
+
+Options
+  --input=<csv>  HP performance file
+
+' -> doc
+
+args <- docopt(doc)
+
+hp_in <- read_csv(args$input)
+
+method <- args$input %>%
+  basename() %>% 
+  str_replace("hp_","") %>% 
+  str_replace(".[0-9]+.csv","")
+
+
+if(method == "glmnet"){
+    min_lambda <- min(hp_in$lambda)
+
+    if(min_lambda < 0.0001) {
+	plot_hp_performance(hp_in,lambda,AUC) +
+	   scale_x_continuous(trans='log10')
+    }else{
+	plot_hp_performance(hp_in,lambda,AUC)
+    }
+    ggsave("hp_plot.png")
+}
+
+if(method == "rpart2"){
+    plot_hp_performance(hp_in,maxdepth,AUC) 
+    ggsave("hp_plot.png")
+}
+
+if(method == "rf"){
+    plot_hp_performance(hp_in,mtry,AUC)
+    ggsave("hp_plot.png")
+}
+
+if(method == "svmRadial"){
+    mod_hp <- hp_in    
+    mod_hp$sigma <- as.factor(mod_hp$sigma)
+    p1 <- ggplot(mod_hp, aes(x=C,y=AUC,color=sigma)) +
+	geom_line() + geom_point() +
+	scale_x_continuous(trans='log10')
+
+    mod_hp <- hp_in
+    mod_hp$C <- as.factor(mod_hp$C)
+    p2 <- ggplot(mod_hp, aes(x=sigma,y=AUC,color=C)) +
+	geom_line() + geom_point() +
+	scale_x_continuous(trans='log10')
+    plot_grid(p1,p2)
+    ggsave("hp_plot.png",width=10,height=5)
+}

### 2021-01-21 code/R/plot_hp_performance.R

@@ -1,4 +1,23 @@
 library(tidyverse)
+library(docopt)
+library(cowplot)
+
+'plot hp performance
+
+Usage:
+  plot_hp_performance.R --method=<str>
+
+Options:
+  --method=<str>  model name. options:
+                   rf
+                   regLogistic
+                   svmRadial
+                   rpart2
+		   xgbTree
+
+' -> doc
+
+args <- docopt(doc)
 
 #functions adapted from begum mbio paper
 read_files <- function(filenames){
@@ -12,25 +31,87 @@ read_files <- function(filenames){
   return(data)
 }
 
-hp_files <- list.files(path= 'data/process', pattern='hp-.*-.*.csv', full.names = TRUE)
+hp_files <- list.files(path= 'data/process', pattern=paste0('hp-.*-',args$method,'.csv'), full.names = TRUE)
 hp.df <- map_df(hp_files, read_files)
 
-hp.df %>% 
-  group_by(mtry,level) %>%
-  summarise("mean_AUC" = mean(AUC),
-            "sd_AUC" = sd(AUC),
-            # is there a less repetitive way to do this cleanly?
-            ymin_metric = mean_AUC - sd_AUC,
-            ymax_metric = mean_AUC + sd_AUC) %>% 
-  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
-                      labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV"))) %>% 
-  ggplot(aes(x=mtry,y=mean_AUC)) +
-  geom_point(size=2,alpha=0.5) + geom_line(alpha=0.5) +
-  facet_grid(.~level,scales = "free_x") +
-  theme_bw() +
-  geom_errorbar(aes(
-    ymin = .data$ymin_metric,
-    ymax = .data$ymax_metric),
-    width = .001)
+if(args$method == "svmRadial"){
+  param1 <- colnames(hp.df)[1]
+  param2 <- colnames(hp.df)[2]
 
-ggsave("analysis/hp-rf.png")
+
+  p1 <- hp.df %>%
+    rename(param1 := !!param1, param2 := !!param2) %>%
+    mutate(param1 = as.factor(param1)) %>%
+    group_by(param1,param2,level) %>%
+    summarise("mean_AUC" = mean(AUC),
+              "sd_AUC" = sd(AUC),
+              # is there a less repetitive way to do this cleanly?
[…]

### 2021-01-21 code/R/run_model.R

@@ -37,7 +37,7 @@ if( !dir.exists(outdir) ){
 # read in preprocessed data
 data_proc <- read_csv(args$data)
 
-if(args$method == "rf"){
+if(args$method %in% c("rf","rpart2","glmnet","svmRadial")){
   # get hyperparamter values for model/taxonomy
   hp <- read_csv(args$hyperparams)
   sub_hp <- hp %>%

### 2021-01-21 code/R/caluclate_pvalues.R
add pvalue script and rule

@@ -1,0 +1,83 @@
+library(tidyverse)
+library(mosaic)
+
+#models <- c("rf","glmnet","xgbTree","svmRadial","rpart2")
+models <- c("rf","glmnet","rpart2")
+levels <- c("phylum","class","order","family","genus","otu","asv")
+
+#functions adapted from begum mbio paper
+read_files <- function(filenames){
+  for(file in filenames){
+    # Read the files generated in main.R
+    # These files have cvAUCs and testAUCs for 100 data-splits
+    data <- read.delim(file, header=T, sep=',') %>%
+      mutate(level=unlist(strsplit(file,"-"))[2])
+
+  }
+  return(data)
+}
+
+#compare two models levels within a taxonomic level
+perm_p_value_level <- function(data, model_name_1, model_name_2, level_name){
+  test_subsample <- data %>%
+    select(AUC,method,level) %>%
+    filter((method == model_name_1 | method == model_name_2) & level == level_name)
+  obs <- abs(diff(mosaic::mean(AUC ~ method, data = test_subsample)))
+  auc.null <- do(10000) * diff(mosaic::mean(AUC ~ shuffle(method), data = test_subsample))
+  n <- length(auc.null[,1])
+  # r = #replications at least as extreme as observed effect
+  r <- sum(abs(auc.null[,1]) >= obs)
+  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
+  p.value=(r+1)/(n+1)
+  return(p.value)
+}
+
+#compare two taxonomic levels within a model type
+perm_p_value_model <- function(data, model_name, level_name_1, level_name_2){
+  test_subsample <- data %>%
+    select(AUC,method,level) %>%
+    filter(method == model_name & ( level == level_name_1 | level == level_name_2))
+  obs <- abs(diff(mosaic::mean(AUC ~ level, data = test_subsample)))
+  auc.null <- do(10000) * diff(mosaic::mean(AUC ~ shuffle(level), data = test_subsample))
+  n <- length(auc.null[,1])
+  # r = #replications at least as extreme as observed effect
+  r <- sum(abs(auc.null[,1]) >= obs)
+  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
+  p.value=(r+1)/(n+1)
+  #plot <- ggplot(auc.null, aes(x=auc.null[,1], color=auc.null[,1]>=obs)) + geom_histogram(fill="white", alpha=0.5, position="identity") +
[…]

### 2021-01-21 code/R/plot_hp_performance.R
update hp plots

@@ -84,13 +84,37 @@ if(args$method == "svmRadial"){
       width = .001)
 
   plot_grid(p1,p2)
-  ggsave(paste0("analysis/hp-",args$method,".png"),height=5,width=10)
+  ggsave(paste0("analysis/hp-",args$method,".png"),height=4,width=10)
 
+}
+if(args$method == "glmnet"){
+  hp.df <- hp.df %>%
+    select(-alpha)
+
+  param <- colnames(hp.df)[1]
+
+  hp.df %>%
+    rename(param := !!param) %>%
+    group_by(param,level) %>%
+    summarise("mean_AUC" = mean(AUC),
+              "sd_AUC" = sd(AUC),
+              # is there a less repetitive way to do this cleanly?
+              ymin_metric = mean_AUC - sd_AUC,
+              ymax_metric = mean_AUC + sd_AUC) %>%
+    mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                        labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV"))) %>%
+    ggplot(aes(x=param,y=mean_AUC)) +
+    geom_point(size=2,alpha=0.5) + geom_line(alpha=0.5) +
+    facet_grid(.~level,scales = "free_x") +
+    theme_bw() +
+    xlab(param) +
+    scale_x_continuous(trans='log10') +
+    geom_errorbar(aes(
+      ymin = .data$ymin_metric,
+      ymax = .data$ymax_metric),
+      width = .001)
+  ggsave(paste0("analysis/hp-",args$method,".png"),height=4,width=7)
 }else{
-  if(args$method == "glmnet"){
-    hp.df <- hp.df %>%
-	select(-alpha)
-  }
 
   param <- colnames(hp.df)[1]
 
@@ -113,5 +137,5 @@ if(args$method == "svmRadial"){
       ymin = .data$ymin_metric,
       ymax = .data$ymax_metric),
       width = .001)
-  ggsave(paste0("analysis/hp-",args$method,".png"),height=5,width=7)
+  ggsave(paste0("analysis/hp-",args$method,".png"),height=4,width=7)
 }

### 2021-01-21 code/R/stats.R

@@ -1,94 +1,0 @@
-library(tidyverse)
-library(mosaic)
-library(reshape2)
-
-#functions adapted from Begum mbio paper
-read_files <- function(filenames){
-  for(file in filenames){
-    # Read the files generated in main.R
-    # These files have cvAUCs and testAUCs for 100 data-splits
-    data <- read.delim(file, header=T, sep=',') %>%
-      mutate(level=unlist(strsplit(file,"-"))[2])
-    
-  }
-  return(data)
-}
-
-#compare two models levels within a taxonomic level
-perm_p_value <- function(data, model_name_1, model_name_2, level_name){
-  test_subsample <- data %>%
-    select(-cv_auroc) %>%
-    filter((method == model_name_1 | method == model_name_2) & level == level_name)
-  obs <- abs(diff(mosaic::mean(test_auroc ~ method, data = test_subsample)))
-  auc.null <- do(10000) * diff(mosaic::mean(test_auroc ~ shuffle(method), data = test_subsample))
-  n <- length(auc.null[,1])
-  # r = #replications at least as extreme as observed effect
-  r <- sum(abs(auc.null[,1]) >= obs)
-  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
-  p.value=(r+1)/(n+1)
-  #plot <- ggplot(auc.null, aes(x=auc.null[,1], color=auc.null[,1]>=obs)) + geom_histogram(fill="white", alpha=0.5, position="identity") + 
-  #  scale_color_brewer(palette="Dark2") + geom_vline(xintercept = obs,color="grey",lty="dashed")
-  return(p.value)
-}
-
-#compare two taxonomic levels within a model type
-perm_p_value_model <- function(data, model_name, level_name_1, level_name_2){
-  test_subsample <- data %>%
-    select(-cv_auroc) %>%
-    filter(method == model_name & ( level == level_name_1 | level == level_name_2))
-  obs <- abs(diff(mosaic::mean(test_auroc ~ level, data = test_subsample)))
-  auc.null <- do(10000) * diff(mosaic::mean(test_auroc ~ shuffle(level), data = test_subsample))
-  n <- length(auc.null[,1])
-  # r = #replications at least as extreme as observed effect
-  r <- sum(abs(auc.null[,1]) >= obs)
-  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
[…]

### 2021-01-22 paper/manuscript.Rmd
outline edits

@@
 
-# TITLE GOES HERE
+# Optimal Resolution for Microbiome-Based Classification of Colorectal Cancer
 
@@
 
-**Running title:** Optimal resolution
+**Running title:** Optimal Resolution
 
@@
 
-Courtney R. Armour, (ande? begum? etc?), Patrick D. Schloss ${^\dagger}$
+Courtney R. Armour, (ande? begum?), Patrick D. Schloss ${^\dagger}$
 
@@
 - Begum et al (mBio 2020) recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
-- A common question when applying ML methods to microbiome data is which method and taxonomic level should be use.
+- A common question when applying ML methods to microbiome data is which method and taxonomic level is optimal.
 - This analysis utilizes the reproducible framework developed by Begum et al to quantify which ML method and taxonomic level produce the best performing classifier for CRC data.
@@
 ## Results
-- Across the five ML methods tested, model performance tends to increase with taxnomic level usually peaking around genus level and dropping off slightly with ASVs.
+- Across the five ML methods tested, model performance tends to increase with taxonomic level usually peaking around genus/otu level and dropping off slightly with ASVs.
 - Random forest was consistently the top performer at most taxonomic levels.
@@
 - mikropml package
+  - preprocessing
+  - hp tuning
+  - n seeds
 - pvalues as previously described {begum}

### 2021-01-22 paper/manuscript.tex
outline edits

@@
 
-\hypertarget{title-goes-here}{%
-\section{TITLE GOES HERE}\label{title-goes-here}}
+\hypertarget{optimal-resolution-for-microbiome-based-classification-of-colorectal-cancer}{%
+\section{Optimal Resolution for Microbiome-Based Classification of
+Colorectal
+Cancer}\label{optimal-resolution-for-microbiome-based-classification-of-colorectal-cancer}}
 
@@
 
-\textbf{Running title:} Optimal resolution
+\textbf{Running title:} Optimal Resolution
 
@@
 
-Courtney R. Armour, (ande? begum? etc?), Patrick D. Schloss
-\({^\dagger}\)
+Courtney R. Armour, (ande? begum?), Patrick D. Schloss \({^\dagger}\)
 
@@
   A common question when applying ML methods to microbiome data is which
-  method and taxonomic level should be use.
+  method and taxonomic level is optimal.
 \item
@@
   Across the five ML methods tested, model performance tends to increase
-  with taxnomic level usually peaking around genus level and dropping
-  off slightly with ASVs.
+  with taxonomic level usually peaking around genus/otu level and
+  dropping off slightly with ASVs.
 \item

### 2021-02-11 paper/manuscript.Rmd
paper update

@@
 
-**Running title:** Optimal Resolution
+**Running title:** ASV level resolution is unnecessary for prediction of SRNs from microbiome data
 
@@
 
-- CRC is one of the most common cancers and a leading cause of cancer related death
+- CRC is one of the most common cancers and a leading cause of cancer related death in the United States.
 - There is evidence that the microbiome has a role in CRC development/progression and could be useful for biomarker detection and diagnostics.
@@
 - A common question when applying ML methods to microbiome data is which method and taxonomic level is optimal.
-- This analysis utilizes the reproducible framework developed by Begum et al to quantify which ML method and taxonomic level produce the best performing classifier for CRC data.
+- This analysis utilizes the reproducible framework developed by Begum et al to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.
 
@@
 ## Results
-- Across the five ML methods tested, model performance tends to increase with taxonomic level usually peaking around genus/otu level and dropping off slightly with ASVs.
+- Across the five ML methods tested, model performance tends to increase with taxonomic level usually peaking around genus/OTU level and dropping off slightly with ASVs.
 - Random forest was consistently the top performer at most taxonomic levels.
   - RF might be more appropriate for microbiome analysis since its more suitable for zero inflated data (need to look into literature)
-- Within the RF model, the highest AUCs were observed for family, genus, and otu level data with no significant difference between the three (Figure 1).
+- Within the RF model, the highest AUCs were observed for family, genus, and OTU level data with no significant difference between the three (Figure 1).
 ASV performed significantly lower than OTU but not family and genus levels.

### 2021-02-11 paper/manuscript.tex
paper update

@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-]{article}
-\usepackage{amsmath,amssymb}
+\documentclass[]{article}
+\usepackage[T1]{fontenc}
 \usepackage{lmodern}
+\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
   \fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use microtype if available
+\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
 \usepackage[margin=1.0in]{geometry}
 \usepackage{graphicx}
+% Redefine \includegraphics so that, unless explicit options are
+% given, the image width will not exceed the width of the page.
+% Images get their normal width if they fit onto the page, but
+% are scaled down if they would overflow the margins.
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\def\ScaleIfNeeded{%
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
 \makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
+\let\Oldincludegraphics\includegraphics
+{%
+ \catcode`\@=11\relax%
+ \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
+}%
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{0}
+
+\author{}
+\date{\vspace{-2.5em}}
 \usepackage{helvet}
@@
 \linenumbers
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-
-\author{}
-\date{\vspace{-2.5em}}
 
@@
 
-\hypertarget{optimal-resolution-for-microbiome-based-classification-of-colorectal-cancer}{%
 \section{Optimal Resolution for Microbiome-Based Classification of
 Colorectal
-Cancer}\label{optimal-resolution-for-microbiome-based-classification-of-colorectal-cancer}}
+Cancer}\label{optimal-resolution-for-microbiome-based-classification-of-colorectal-cancer}
 
@@
 
-Courtney R. Armour, (ande? begum?), Patrick D. Schloss \({^\dagger}\)
+Courtney R. Armour, (ande? begum?), Patrick D. Schloss ${^\dagger}$
 
@@
 
-\({\dagger}\) To whom correspondence should be addressed:
+${\dagger}$ To whom correspondence should be addressed:
 
@@
 
-\hypertarget{abstract-max-250-words}{%
-\subsection{Abstract (max 250 words)}\label{abstract-max-250-words}}
+\subsection{Abstract (max 250 words)}\label{abstract-max-250-words}
 
-\hypertarget{importance-max-150-words}{%
-\subsection{Importance (max 150 words)}\label{importance-max-150-words}}
+\subsection{Importance (max 150 words)}\label{importance-max-150-words}
 
@@
 
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
+\subsection{Introduction}\label{introduction}
 
 \begin{itemize}
-\tightlist
+\itemsep1pt\parskip0pt\parsep0pt
 \item
@@
   et al to quantify which ML method and taxonomic level produce the best
-  performing classifier for CRC data.
+  performing classifier for detecting SRNs from microbiome data.
 \end{itemize}
@@
 
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
+\subsection{Results}\label{results}
 
 \begin{itemize}
-\tightlist
+\itemsep1pt\parskip0pt\parsep0pt
 \item
@@
   levels.
-
-  \begin{itemize}
-  \tightlist
 \item
@@
   suitable for zero inflated data (need to look into literature)
-  \end{itemize}
 \item
@@
 
-\hypertarget{conclusion}{%
-\subsection{Conclusion}\label{conclusion}}
+\subsection{Conclusion}\label{conclusion}
 
@@
 
-\hypertarget{materials-and-methods}{%
-\subsection{Materials and Methods}\label{materials-and-methods}}
+\subsection{Materials and Methods}\label{materials-and-methods}
 
 \begin{itemize}
-\tightlist
+\itemsep1pt\parskip0pt\parsep0pt
 \item
@@
   mikropml package
+\item
+  preprocessing
+\item
+  hp tuning
+\item
+  n seeds
 \item
@@
 
-\hypertarget{acknowledgements}{%
-\subsection{Acknowledgements}\label{acknowledgements}}
+\subsection{Acknowledgements}\label{acknowledgements}
 
@@
 
-\hypertarget{figures}{%
-\subsection{Figures}\label{figures}}
+\subsection{Figures}\label{figures}
 

### 2021-02-18 exploratory/exploratory.Rmd
add exploratory rmd

---
title: "Exploratory: Optimal Resolution"
author: "Courtney R Armour"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup_environment, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(knitr)
library(kableExtra)
library(ggpubr)
library(nationalparkcolors)
library(data.table)
library(cowplot)

models <- c("rf","glmnet","xgbTree","svmRadial","rpart2")
levels <- c("phylum","class","order","family","genus","otu","asv")
levels_names <- c("Phylum","Class","Order","Family","Genus","OTU","ASV")
pal <- park_palette("Everglades",length(models))
names(pal) <- models
pal2 <- park_palette("Everglades",length(models))
names(pal2) <- c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")

full_df <- NULL
auc_df <- NULL
cv_table <- tibble(level=c("phylum","class","order","family","genus","otu","asv"))
t_table <- tibble(level=c("phylum","class","order","family","genus","otu","asv"))

for( m in models ){
  phylum <- read_csv(paste0("./data/final/process/combined-phylum-",m,".csv")) %>% 
    mutate(level="phylum")
  class <- read_csv(paste0("./data/final/process/combined-class-",m,".csv")) %>% 
    mutate(level="class")
  order <- read_csv(paste0("./data/final/process/combined-order-",m,".csv")) %>% 
    mutate(level="order")
  family <- read_csv(paste0("./data/final/process/combined-family-",m,".csv")) %>% 
    mutate(level="family")
  genus <- read_csv(paste0("./data/final/process/combined-genus-",m,".csv")) %>%
    mutate(level="genus")
  otu <- read_csv(paste0("./data/final/process/combined-otu-",m,".csv")) %>%
    mutate(level="otu")
  asv <- read_csv(paste0("./data/final/process/combined-asv-",m,".csv")) %>%
    mutate(level="asv")

  model_df <- bind_rows(phylum,class,order,family,genus,otu,asv)

  full_df <- bind_rows(full_df,model_df)

  m_auc.df <- select(model_df,cv_metric_AUC,AUC,method,level) %>% 
    rename(test_AUC=AUC) %>% 
    pivot_longer(cols = c("cv_metric_AUC","test_AUC"),names_to="auc_type",values_to="AUC")

  m_auc.df$level <- factor(m_auc.df$level,levels=c("phylum","class","order","family","genus","otu","asv"))

  auc_df <- bind_rows(auc_df,m_auc.df)

  cv_med <- m_auc.df %>%
    filter(auc_type == "cv_metric_AUC") %>%
    group_by(level) %>% 
    summarise(!!m := median(AUC))

  t_med <- m_auc.df %>% 
    filter(auc_type == "test_AUC") %>%
    group_by(level) %>% 
    summarise(!!m := median(AUC))

  cv_table <- inner_join(cv_table,cv_med,by="level")
  t_table <- inner_join(t_table,t_med,by="level")
}

#compare two models levels within a taxonomic level
p.table.level <- read_csv("./data/analysis/exploratory3/pvalues_by_level.csv")
#compare two taxonomic levels within a model type
p.table.model <- read_csv("./data/analysis/exploratory3/pvalues_by_model.csv")

```

### 2021-02-18 code/R/quantify_input_values.R
add rule to summarize input data

@@ -1,0 +1,20 @@
+library(tidyverse)
+library(data.table)
+
+levels <- c("phylum","class","order","family","genus","otu","asv")
+
+input <- NULL
+
+for(l in levels){
+  in_data <- as.data.frame(fread(paste0("./data/",l,"/input_data.csv")))
+  in_data_preproc <- as.data.frame(fread(paste0("./data/",l,"/input_data_preproc.csv")))
+
+  n_samp <- nrow(in_data)
+  n_feat <- ncol(in_data) - 1
+  n_feat_preproc <- ncol(in_data_preproc) - 1
+
+  input <- input %>%
+    bind_rows(c(level=l,n_samples=n_samp,n_features=n_feat,n_features_preproc=n_feat_preproc))
+}
+
+write_csv(input,"./analysis/input_values.csv")

### 2021-02-19 code/R/plot_hp_performance.R
update svmRadial hp plot

@@ -38,7 +38,6 @@ if(args$method == "svmRadial"){
   param1 <- colnames(hp.df)[1]
   param2 <- colnames(hp.df)[2]
 
-
   p1 <- hp.df %>%
     rename(param1 := !!param1, param2 := !!param2) %>%
     mutate(param1 = as.factor(param1)) %>%
@@ -52,10 +51,10 @@ if(args$method == "svmRadial"){
                         labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV"))) %>%
     ggplot(aes(x=param2,y=mean_AUC,color=param1)) +
     geom_point(size=2,alpha=0.5) + geom_line(alpha=0.5) +
-    facet_grid(.~level,scales = "free_x") +
+    facet_wrap(.~level,scales = "free_x",ncol=1) +
     theme_bw() +
     xlab(param2) +
-    scale_x_continuous(trans='log10')
+    scale_x_continuous(trans='log10') +
     geom_errorbar(aes(
       ymin = .data$ymin_metric,
       ymax = .data$ymax_metric),
@@ -74,19 +73,19 @@ if(args$method == "svmRadial"){
                         labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV"))) %>%
     ggplot(aes(x=param1,y=mean_AUC,color=param2)) +
     geom_point(size=2,alpha=0.5) + geom_line(alpha=0.5) +
-    facet_grid(.~level,scales = "free_x") +
+    facet_wrap(.~level,scales = "free_x",ncol=1) +
     theme_bw() +
     xlab(param1) +
-    scale_x_continuous(trans='log10')
+    scale_x_continuous(trans='log10') +
     geom_errorbar(aes(
       ymin = .data$ymin_metric,
       ymax = .data$ymax_metric),
       width = .001)
 
-  plot_grid(p1,p2)
-  ggsave(paste0("analysis/hp-",args$method,".png"),height=4,width=10)
-
+  plot <- plot_grid(p1,p2,ncol=2)
+  save_plot(filename=paste0("analysis/hp-",args$method,".png"),plot,base_height=10,base_width=8)
 }
+
 if(args$method == "glmnet"){
   hp.df <- hp.df %>%
     select(-alpha)
@@ -114,7 +113,9 @@ if(args$method == "glmnet"){
       ymax = .data$ymax_metric),
       width = .001)
[…]

### 2021-02-22 paper/manuscript.Rmd
add figures to outline

--- 2021-01-19 paper/manuscript.Rmd
@@
 output:
-  word_document: default
   pdf_document:
     keep_tex: true
+  word_document: default
 geometry: margin=1.0in
@@
 
-# TITLE GOES HERE
+```{r setup_environment, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE,message=FALSE)
+options(scipen=999)
+library(tidyverse)
+```
+
+# Optimal Taxonomic Resolution for Microbiome-Based Classification of Colorectal Cancer
 
@@
 
-**Running title:** Optimal resolution
+**Other title options:**  
+ASV level resolution is unnecessary for prediction of SRNs from microbiome data  
+OTUs are the Optimal Taxonomic Level for Microbiome-Based Classification of Colorectal Cancer
 
@@
 
-- CRC is one of the most common cancers and a leading cause of cancer related death
-- There is evidence that the microbiome has a role in CRC development/progression and could be useful for biomarker detection and diagnostics.
-- Begum et al (mBio 2020) recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
-- A common question when applying ML methods to microbiome data is which method and taxonomic level should be use.
-- This analysis utilizes the reproducible framework developed by Begum et al to quantify which ML method and taxonomic level produce the best performing classifier for CRC data.
-
-\newpage
+- Colorectal cancer (CRC) is one of the most common cancers in men and women and a leading cause of cancer related death in the United States {}. 
+- Early detection and treatment are essential to increase survival rates, but for a variety of reasons including the invasiveness and high cost of screening (i.e.
+colonoscopy), many people do not comply with recommended screening guidelines {}.
+- Less invasive and more affordable screening methods (e.g.
+fecal immunochemical test) are available, however these tests are less sensitive than colonoscopies, especially for detecting early stage adenomas.
+- A growing body of research points to the gut microbiome as having a role in tumor development and progression {}.
+For example, studies find that Fusobacterium nucleatum and enterotoxigenic Bacteriodes fragilis tend to be enriched in the gut microbiome of subjects with CRC relative to healthy controls {}, while potentially protective bacteria such as members of Lachnospiraceae tend to be depleted {}.
+- The gut microbiome shows promise as diagnostic {}
+- Efforts to realize the diagnostic potential of the gut microbiome in detecting screen relevant neoplasias (SRNs) have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on the amplicon sequencing of the v4 region of the 16S rRNA gene {}.
+However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown.
+  - Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs.
+However, whether the additional resolution provided by ASVs is useful for ML classification is unclear.
+    - Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups. 
+- Topçuoğlu *et al* (mBio 2020) recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
+- This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.
 
 ## Results
-- Across the five ML methods tested, model performance tends to increase with taxnomic level usually peaking around genus level and dropping off slightly with ASVs.
+
+```{r median_AUC}
+table <- read_csv("../exploratory/figures/median_test_auc.csv")
+
+rf_med_auc <- table %>% 
+  pull(rf)
+
+names(rf_med_auc) <- table %>% pull(level)
+```
+
+```{r input_table}
+input <- read_csv("../analysis/input_values.csv") %>% 
+  mutate(pct_kept = round(n_features_preproc/n_features,digits=3)*100)
+```
+
+- Across the five ML methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs.
+[(Figure)](#model-performance-across-taxonomy)
 - Random forest was consistently the top performer at most taxonomic levels.
-  - RF might be more appropriate for microbiome analysis since its more suitable for zero inflated data (need to look into literature)
-- Within the RF model, the highest AUCs were observed for family, genus, and otu level data with no significant difference between the three (Figure 1).
-ASV performed significantly lower than OTU but not family and genus levels.
+  - RF might be more appropriate for microbiome analysis since its more suitable for zero inflated data 
+  - TODO: dig into literature on this
+  - TODO: address lower L2 logistic performance that prior results?
+- Within the RF model, the highest AUCs were observed for family (median AUC: `r round(rf_med_auc["family"],digits=3)`), genus(median AUC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three.
+[(Figure)](#random-forest-model-performance-with-significance)
+  - ASV (median AUC: `r round(rf_med_auc["asv"],digits=3)`) performed significantly lower than OTU and genus (p < 0.05) but not family level P = 0.06).
+- One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g.
+phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls.
+  - As you reach genus/OTU level data there is enough data and variation but at the ASV level, the data is too specific to individuals and doesn't overlap enough. 
+  - Examination of the prevalence of taxa in samples at each level supports this idea.
+A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels.
+The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples.
+[(Figure)](#prevalence-of-taxa-in-samples)
+- Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods).
+As part of this step, features are removed that wont provide useful information to build the model.
+For example, strongly correlated features provide the same information to build the model and thus can be collapsed.
+Additionally, features with zero or near-zero variance will not help the model differentiate groups and thus can be removed.
+  - Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing.
+At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing.
+[(Table)](#summary-of-features)
+  - TODO: check why ASVs were removed in preprocessing (nzv?)
+  - (depending on above can discuss why features were removed - e.g. only in one or a few samples)
+- OTU level data provides an idea balance of overlap and distinction to classify samples. 
+- Overall, the fine resolution of ASVs is unnecessary for ML classification with microbiome data since the vast majority of the ASVs are removed during preprocessing and the model performance is diminshed compared to other taxonomic levels.
 
-\newpage
+other items that could be addressed:
 
-## Conclusion
+- what are important taxa, patterns along levels (e.g fuso always in top )
+- abundance of important taxa: many of the top important features are low abund [(Figure)](#relative-abundance-of-important-features)
 
-\newpage
+## Conclusions
 
 ## Materials and Methods
-- data from prior study {baxter}
-- mikropml package
-- pvalues as previously described {begum}
-
-\newpage
+- 16S rRNA data from 490 subjects {baxter}
+  - 261 controls
+  - 229 cases
+- processed with mothur v1.44.3
+  - SILVA v132 reference
+- ML with mikropml package
+  - preprocessing
+    -normalize values
+    - what is removed and why
+      - near zero variance & zero variance removed
+      - correlated collapsed
+  - hp tuning
+  - n seeds
+  - for more detail see mikropml package{}
+- pvalues calculated as previously described {begum}
+- prevalence = number of samples with non-zero abundance / total number of samples
 
@@
 
-\newpage
+## Figures
 
-## Figures
+### Model Performance across Taxonomy  
+![](../exploratory/figures/all_model_level.png)
+
+### Random Forest Model Performance with Significance  
+![](../exploratory/figures/rf_with_stats.png){width=75%}
+
+### Prevalence of Taxa in Samples  
+![](../exploratory/figures/bin_prevalence.png){height=50%}
+
+### Summary of Features
+```{r feats}
+knitr::kable(input)
+```
+
+### Relative Abundance of Important Features
+![](../exploratory/figures/relabund_top10.png){height=95%}

### 2021-02-22 paper/manuscript.tex
add figures to outline

--- 2021-01-19 paper/manuscript.tex
@@
 \usepackage[margin=1.0in]{geometry}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 
-\hypertarget{title-goes-here}{%
-\section{TITLE GOES HERE}\label{title-goes-here}}
+\hypertarget{optimal-taxonomic-resolution-for-microbiome-based-classification-of-colorectal-cancer}{%
+\section{Optimal Taxonomic Resolution for Microbiome-Based
+Classification of Colorectal
+Cancer}\label{optimal-taxonomic-resolution-for-microbiome-based-classification-of-colorectal-cancer}}
 
@@
 
-\textbf{Running title:} Optimal resolution
+\textbf{Other title options:}\\
+ASV level resolution is unnecessary for prediction of SRNs from
+microbiome data\\
+OTUs are the Optimal Taxonomic Level for Microbiome-Based Classification
+of Colorectal Cancer
 
@@
 \item
-  CRC is one of the most common cancers and a leading cause of cancer
-  related death
+  Colorectal cancer (CRC) is one of the most common cancers in men and
+  women and a leading cause of cancer related death in the United States
+  \{\}.
 \item
-  There is evidence that the microbiome has a role in CRC
-  development/progression and could be useful for biomarker detection
-  and diagnostics.
+  Early detection and treatment are essential to increase survival
+  rates, but for a variety of reasons including the invasiveness and
+  high cost of screening (i.e.~colonoscopy), many people do not comply
+  with recommended screening guidelines \{\}.
 \item
-  Begum et al (mBio 2020) recently demonstrated effective application of
-  machine learning (ML) to microbiome based classification problems and
-  developed a framework for applying ML practices in a more reproducible
-  way (mikropml).
+  Less invasive and more affordable screening methods (e.g.~fecal
+  immunochemical test) are available, however these tests are less
+  sensitive than colonoscopies, especially for detecting early stage
+  adenomas.
 \item
-  A common question when applying ML methods to microbiome data is which
-  method and taxonomic level should be use.
+  A growing body of research points to the gut microbiome as having a
+  role in tumor development and progression \{\}. For example, studies
+  find that Fusobacterium nucleatum and enterotoxigenic Bacteriodes
+  fragilis tend to be enriched in the gut microbiome of subjects with
+  CRC relative to healthy controls \{\}, while potentially protective
+  bacteria such as members of Lachnospiraceae tend to be depleted \{\}.
 \item
-  This analysis utilizes the reproducible framework developed by Begum
-  et al to quantify which ML method and taxonomic level produce the best
-  performing classifier for CRC data.
+  The gut microbiome shows promise as diagnostic \{\}
+\item
+  Efforts to realize the diagnostic potential of the gut microbiome in
+  detecting screen relevant neoplasias (SRNs) have focused on machine
+  learning (ML) methods using abundances from operational taxonomic unit
+  (OTU) classifications based on the amplicon sequencing of the v4
+  region of the 16S rRNA gene \{\}. However, whether this is the optimal
+  taxonomic resolution for classifying SRNs from microbiome data is
+  unknown.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Additionally, recent work has pushed for the use of amplicon
+    sequence variants (ASVs) to replace OTUs for marker-gene analysis
+    because of the improved resolution with ASVs. However, whether the
+    additional resolution provided by ASVs is useful for ML
+    classification is unclear.
+
+    \begin{itemize}
+    \tightlist
+    \item
+      Since ML classification relies on consistent differences between
+      groups, its possible that the resolution at the ASV level is too
+      individualized to accurately differentiate groups.
     \end{itemize}
-
-\newpage
+  \end{itemize}
+\item
+  Topçuoğlu \emph{et al} (mBio 2020) recently demonstrated effective
+  application of machine learning (ML) to microbiome based
+  classification problems and developed a framework for applying ML
+  practices in a more reproducible way (mikropml).
+\item
+  This analysis utilizes the reproducible framework developed by
+  Topçuoğlu \emph{et al} to quantify which ML method and taxonomic level
+  produce the best performing classifier for detecting SRNs from
+  microbiome data.
+\end{itemize}
 
@@
 \item
-  Across the five ML methods tested, model performance tends to increase
-  with taxnomic level usually peaking around genus level and dropping
-  off slightly with ASVs.
+  Across the five ML methods tested, model performance tended to
+  increase with taxonomic level usually peaking around genus/OTU level
+  before dropping off slightly with ASVs.
+  \protect\hyperlink{model-performance-across-taxonomy}{(Figure)}
 \item
@@
     RF might be more appropriate for microbiome analysis since its more
-    suitable for zero inflated data (need to look into literature)
+    suitable for zero inflated data
+  \item
+    TODO: dig into literature on this
+  \item
+    TODO: address lower L2 logistic performance that prior results?
   \end{itemize}
 \item
-  Within the RF model, the highest AUCs were observed for family, genus,
-  and otu level data with no significant difference between the three
-  (Figure 1). ASV performed significantly lower than OTU but not family
-  and genus levels.
+  Within the RF model, the highest AUCs were observed for family (median
+  AUC: 0.687), genus(median AUC: 0.686), and OTU (median AUC: 0.698)
+  level data with no significant difference between the three.
+  \protect\hyperlink{random-forest-model-performance-with-significance}{(Figure)}
+
+  \begin{itemize}
+  \tightlist
+  \item
+    ASV (median AUC: 0.676) performed significantly lower than OTU and
+    genus (p \textless{} 0.05) but not family level P = 0.06).
+  \end{itemize}
+\item
+  One hypothesis for the observation that model performance increases
+  from phylum to OTU level then drops slightly at ASV level is that at
+  higher taxonomic levels (e.g.~phylum) there are too few taxa and too
+  much overlap to reliably differentiate between cases and controls.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    As you reach genus/OTU level data there is enough data and variation
+    but at the ASV level, the data is too specific to individuals and
+    doesn't overlap enough.
+  \item
+    Examination of the prevalence of taxa in samples at each level
+    supports this idea. A majority of taxa are present in greater than
+    75\% of samples at the phylum (67\% of taxa) and class (63\% of
+    taxa) levels. The opposite is observed at the OTU and ASV level
+    where 60\% and 53\% of taxa respectively are only present in less
+    than 25\% of samples.
+    \protect\hyperlink{prevalence-of-taxa-in-samples}{(Figure)}
+  \end{itemize}
+\item
+  Of note, the ML pipeline includes a pre-processing step that occurs
+  prior to training and classifying the ML models (methods). As part of
+  this step, features are removed that wont provide useful information
+  to build the model. For example, strongly correlated features provide
+  the same information to build the model and thus can be collapsed.
+  Additionally, features with zero or near-zero variance will not help
+  the model differentiate groups and thus can be removed.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Interestingly, despite starting with 104106 features at the ASV
+    level, only 478 (0.5\%) remained after pre-processing. At the OTU
+    level, 20079 of the 705 features (3.5\%) remained after
+    preprocessing. \protect\hyperlink{summary-of-features}{(Table)}
+  \item
+    TODO: check why ASVs were removed in preprocessing (nzv?)
+  \item
+    (depending on above can discuss why features were removed -
+    e.g.~only in one or a few samples)
+  \end{itemize}
+\item
+  OTU level data provides an idea balance of overlap and distinction to
+  classify samples.
+\item
+  Overall, the fine resolution of ASVs is unnecessary for ML
+  classification with microbiome data since the vast majority of the
+  ASVs are removed during preprocessing and the model performance is
+  diminshed compared to other taxonomic levels.
 \end{itemize}
 
-\newpage
+other items that could be addressed:
 
-\hypertarget{conclusion}{%
-\subsection{Conclusion}\label{conclusion}}
+\begin{itemize}
+\tightlist
+\item
+  what are important taxa, patterns along levels (e.g fuso always in top
+  )
+\item
+  abundance of important taxa: many of the top important features are
+  low abund
+  \protect\hyperlink{relative-abundance-of-important-features}{(Figure)}
+\end{itemize}
 
-\newpage
+\hypertarget{conclusions}{%
+\subsection{Conclusions}\label{conclusions}}
 
@@
 \item
-  data from prior study \{baxter\}
+  16S rRNA data from 490 subjects \{baxter\}
+
+  \begin{itemize}
+  \tightlist
   \item
-  mikropml package
+    261 controls
   \item
-  pvalues as previously described \{begum\}
+    229 cases
   \end{itemize}
+\item
+  processed with mothur v1.44.3
 
-\newpage
+  \begin{itemize}
+  \tightlist
+  \item
+    SILVA v132 reference
+  \end{itemize}
+\item
+  ML with mikropml package
+
+  \begin{itemize}
+  \tightlist
+  \item
+    preprocessing -normalize values
+
+    \begin{itemize}
+    \tightlist
+    \item
+      what is removed and why
+
+      \begin{itemize}
+      \tightlist
+      \item
+        near zero variance \& zero variance removed
+      \item
+        correlated collapsed
+      \end{itemize}
+    \end{itemize}
+  \item
+    hp tuning
+  \item
+    n seeds
+  \item
+    for more detail see mikropml package\{\}
+  \end{itemize}
+\item
+  pvalues calculated as previously described \{begum\}
+\item
+  prevalence = number of samples with non-zero abundance / total number
+  of samples
+\end{itemize}
 
@@
 
-\newpage
-
 \hypertarget{figures}{%
@@
 
+\hypertarget{model-performance-across-taxonomy}{%
+\subsubsection{Model Performance across
+Taxonomy}\label{model-performance-across-taxonomy}}
+
+\includegraphics{../exploratory/figures/all_model_level.png}
+
+\hypertarget{random-forest-model-performance-with-significance}{%
+\subsubsection{Random Forest Model Performance with
+Significance}\label{random-forest-model-performance-with-significance}}
+
+\includegraphics[width=0.75\textwidth,height=\textheight]{../exploratory/figures/rf_with_stats.png}
+
+\hypertarget{prevalence-of-taxa-in-samples}{%
+\subsubsection{Prevalence of Taxa in
+Samples}\label{prevalence-of-taxa-in-samples}}
+
+\includegraphics[width=\textwidth,height=0.5\textheight]{../exploratory/figures/bin_prevalence.png}
+
+\hypertarget{summary-of-features}{%
+\subsubsection{Summary of Features}\label{summary-of-features}}
+
+\begin{longtable}[]{@{}lrrrr@{}}
+\toprule
+level & n\_samples & n\_features & n\_features\_preproc &
+pct\_kept \\ \addlinespace
+\midrule
+\endhead
+phylum & 490 & 19 & 9 & 47.4 \\ \addlinespace
+class & 490 & 36 & 19 & 52.8 \\ \addlinespace
+order & 490 & 65 & 28 & 43.1 \\ \addlinespace
+family & 490 & 124 & 54 & 43.5 \\ \addlinespace
+genus & 490 & 316 & 115 & 36.4 \\ \addlinespace
+otu & 490 & 20079 & 705 & 3.5 \\ \addlinespace
+asv & 490 & 104106 & 478 & 0.5 \\ \addlinespace
+\bottomrule
+\end{longtable}
+
+\hypertarget{relative-abundance-of-important-features}{%
+\subsubsection{Relative Abundance of Important
+Features}\label{relative-abundance-of-important-features}}
+
+\includegraphics[width=\textwidth,height=0.95\textheight]{../exploratory/figures/relabund_top10.png}
+
 \end{document}

### 2021-03-25 exploratory/exploratory.Rmd

@@
 ```{r setup_environment, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE,message=FALSE)
 
@@
 
+# Variables
 models <- c("rf","glmnet","xgbTree","svmRadial","rpart2")
@@
 
+# Generate dataframe of all results and tables of the median cv and test AUCs
 full_df <- NULL
@@
 for( m in models ){
-  phylum <- read_csv(paste0("./data/final/process/combined-phylum-",m,".csv")) %>% 
+  phylum <- read_csv(paste0("../data/process/combined-phylum-",m,".csv")) %>% 
     mutate(level="phylum")
-  class <- read_csv(paste0("./data/final/process/combined-class-",m,".csv")) %>% 
+  class <- read_csv(paste0("../data/process/combined-class-",m,".csv")) %>% 
     mutate(level="class")
-  order <- read_csv(paste0("./data/final/process/combined-order-",m,".csv")) %>% 
+  order <- read_csv(paste0("../data/process/combined-order-",m,".csv")) %>% 
     mutate(level="order")
-  family <- read_csv(paste0("./data/final/process/combined-family-",m,".csv")) %>% 
+  family <- read_csv(paste0("../data/process/combined-family-",m,".csv")) %>% 
     mutate(level="family")
-  genus <- read_csv(paste0("./data/final/process/combined-genus-",m,".csv")) %>%
+  genus <- read_csv(paste0("../data/process/combined-genus-",m,".csv")) %>%
     mutate(level="genus")
-  otu <- read_csv(paste0("./data/final/process/combined-otu-",m,".csv")) %>%
+  otu <- read_csv(paste0("../data/process/combined-otu-",m,".csv")) %>%
     mutate(level="otu")
-  asv <- read_csv(paste0("./data/final/process/combined-asv-",m,".csv")) %>%
+  asv <- read_csv(paste0("../data/process/combined-asv-",m,".csv")) %>%
     mutate(level="asv")
@@
 
+write_csv(t_table,"./figures/median_test_auc.csv")
+
 #compare two models levels within a taxonomic level
-p.table.level <- read_csv("./data/analysis/exploratory3/pvalues_by_level.csv")
+p.table.level <- read_csv("../analysis/pvalues_by_level.csv")
 #compare two taxonomic levels within a model type
-p.table.model <- read_csv("./data/analysis/exploratory3/pvalues_by_model.csv")
+p.table.model <- read_csv("../analysis/pvalues_by_model.csv")
 
 ```
+
+## Input data overview
+
+To total number of samples and features before and after preprocessing for each taxnomic level. 
+```{r input}
+
+input <- read_csv("../analysis/input_values.csv")
+
+kable(input)
+```
+
+## Hyperparameter Performance
+
+Plots of HP performance for each model/taxonomic level
+
+### Random Forest
+The default hyperparameter (mtry) selection for RF is:
+
+1. sqrt_features / 2
+1. sqrt_features
+1. sqrt_features * 2
+
+Based on previous results, I added mtry=1 for Phylum through Genus levels and mtry=100 for OTU and ASV levels. This set of hyperparameters seems to cover the optimal range for each taxonomic level.
+
+![](../analysis/hp-rf.png)
+
+### Logistic Regression
+
+Two metrics: alpha and lambda  
+By default alpha is set to zero (for L2 regularization) and lambda values are: 1e-04, 1e-03, 1e-02, 1e-01, 1e+00, 1e+01
+I added 1e-05 as a lambda value for phylum and class, and higher values of lambda for the rest
+
+![](../analysis/hp-glmnet.png)
+
+### Decision Tree
+
+The default hyperparameters were retained for decision tree (maxdepth = 1  2  4  8 16 30). It would not allow larger maxdepth than 30.
+
+![](../analysis/hp-rpart2.png)
+
+### XGBoost
+
+(needs adjustment)
+
+### SVM Radial
+
+![](../analysis/hp-svmRadial.png)
+
+## Model Performance
+
+```{r plot_by_method, echo=F,message=F,fig.height = 3.5, fig.width = 10, fig.align = "center"}
+
+df_by_method <- full_df %>% 
+  select(cv_metric_AUC,AUC,method,level) %>% 
+  rename(test_AUC=AUC,cv_AUC=cv_metric_AUC) %>% 
+  pivot_longer(cols=c("cv_AUC","test_AUC"),names_to="auc_type",values_to="AUC")
+
+df_by_method$method <- factor(df_by_method$method,
+                        levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                        labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))
+
+ggplot(df_by_method,aes(x=factor(level,
+                            levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+                   y=AUC, fill=auc_type)) + 
+    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+    geom_boxplot(notch = TRUE) + 
+    facet_grid(.~method) + 
+    theme_bw() + xlab("") + ylab("AUROC") +
+    theme(legend.position ="top",
+          axis.text.x = element_text(angle = 45,hjust = 1)) 
+ggsave("./figures/performance_by_method.png")
+```
+
+```{r plot_by_level, echo=FALSE, message=FALSE,fig.height = 6, fig.width = 10, fig.align = "center"}
+df_by_level <- full_df %>% 
+  select(cv_metric_AUC,AUC,method,level) %>% 
+  rename(test_AUC=AUC,cv_AUC=cv_metric_AUC) %>% 
+  pivot_longer(cols=c("cv_AUC","test_AUC"),names_to="auc_type",values_to="AUC") %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")))
+
+ggplot(df_by_level,aes(x=method, y=AUC, fill=auc_type)) + 
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+  geom_boxplot(notch = TRUE) + 
+  facet_wrap(.~level,nrow = 2) + 
+  theme_bw() + xlab("") + ylab("AUROC") + 
+  theme(legend.position ="top",
+        axis.text.x = element_text(angle = 45,hjust = 1)) 
+ggsave("./figures/performance_by_tax.png")
+
+```
+
+```{r plot_all, echo=F, message=F, out.width="100%",fig.height=4}
+df_all <- full_df %>% 
+  select(cv_metric_AUC,AUC,method,level) %>% 
+  rename(test_AUC=AUC,cv_AUC=cv_metric_AUC) %>% 
+  pivot_longer(cols=c("cv_AUC","test_AUC"),names_to="auc_type",values_to="AUC") %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")))
+
+df_all %>% 
+  filter(auc_type == "test_AUC") %>% 
+  ggplot(aes(x=level,y=AUC,fill=method)) + 
+    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+    geom_boxplot(notch = TRUE) + 
+    theme_bw() + xlab("") + ylab("AUROC") + 
+    theme(legend.position ="top",
+          axis.text.x = element_text(angle = 45,hjust = 1)) +
+    scale_fill_manual(values=pal2,name="Model")
+ggsave("./figures/all_model_level.png")
+```
+
+### DADA2 Comparison
+
+```{r dada2}
+
+dada2_results <- read
+
+```
+
+### Random Forest Model Performance with Significance
+
+Random Forest on OTU level data yields the highest median AUC, followed by Family, Genus, and ASV.
+While the OTU AUC is not significantly higher than that of Family or Genus level, it is significantly higher than ASV.
+
+<div class = "row">
+<div class = "col-md-6">
+```{r rf_table,echo=FALSE}
+t_table %>%
+  select(level,rf) %>% 
+  mutate(rf=round(rf,digits = 3)) %>% 
+  rename("Median AUC"=rf,"Level"=level) %>% 
+  kable() %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, position="center")
+```
+
+</div>
+
+<div class = "col-md-6">
+```{r rf_plot,echo=F,message=F}
+m <- "Random Forest"
+
+sub_df <- auc_df %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                       labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(method==m & auc_type=="test_AUC")
+
+sub_stats <- p.table.model %>% 
+   mutate(level1=factor(level1,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+          level2=factor(level2,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         model=factor(model,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(model==m) %>% 
+  mutate(sig=case_when(p_value >= 0.05 ~ "NS",
+                       TRUE ~ "*")) %>% 
+  filter(sig=="NS")
+
+sub_stats <- sub_stats %>% 
+  mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
+  rename(group1=level1,group2=level2)
+
+ggplot(sub_df,aes(x=level,y=AUC)) + 
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+  geom_boxplot(aes(fill=method),notch = T) +
+  scale_fill_manual(values = pal2[m]) + 
+  scale_y_continuous(limits=c(0.2,1.05)) +
+  theme_bw() + xlab("") + ylab("AUROC") +
+  theme(legend.position ="none",
+        axis.text.x = element_text(angle = 45,hjust = 1),
+        panel.grid.major.x = element_blank(),
+        axis.text = element_text(size=14)) +
+  ggtitle(m) +
+  stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
+  stat_pvalue_manual(sub_stats,label="sig")
+ggsave("./figures/rf_with_stats.png",width = 5,height=4,units="in")
+
+```
+</div>
+</div>
+
+### Logistic Regression Model Performance with Significance
+
+<div class = "row">
+<div class = "col-md-6">
+```{r glmnet_table,echo=FALSE}
+t_table %>%
+  select(level,glmnet) %>% 
+  mutate(glmnet=round(glmnet,digits = 3)) %>% 
+  rename("Median AUC"=glmnet,"Level"=level) %>% 
+  kable() %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, position="center")
+```
+</div>
+
+<div class = "col-md-6">
+```{r glmnet_plot,echo=F,message=F}
+## glmnet Model Performance
+m <- "Logistic Regression"
+
+sub_df <- auc_df %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                       labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(method==m & auc_type=="test_AUC")
+
+sub_stats <- p.table.model %>% 
+   mutate(level1=factor(level1,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+          level2=factor(level2,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         model=factor(model,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(model==m) %>% 
+  mutate(sig=case_when(p_value >= 0.05 ~ "NS",
+                       TRUE ~ "*")) %>% 
+  filter(sig=="NS")
+
+sub_stats <- sub_stats %>% 
+  mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
+  rename(group1=level1,group2=level2)
+
+ggplot(sub_df,aes(x=level,y=AUC)) + 
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+  geom_boxplot(aes(fill=method),notch = T) +
+  scale_fill_manual(values = pal2[m]) + 
+  scale_y_continuous(limits=c(0.2,1.05)) +
+  theme_bw() + xlab("") + ylab("AUROC") +
+  theme(legend.position ="none",
+        axis.text.x = element_text(angle = 45,hjust = 1),
+        panel.grid.major.x = element_blank(),
+        axis.text = element_text(size=14)) +
+  ggtitle(m) +
+  stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
+  stat_pvalue_manual(sub_stats,label="sig")
+ggsave("./figures/glmnet_with_stats.png",width = 5,height=4,units="in")
+
+```
+</div>
+</div>
+
+### Decision Tree Model Performance with Significance
+
+<div class = "row">
+<div class = "col-md-6">
+```{r rpart2_table,echo=FALSE}
+t_table %>%
+  select(level,rpart2) %>% 
+  mutate(rpart2=round(rpart2,digits = 3)) %>% 
+  rename("Median AUC"=rpart2,"Level"=level) %>% 
+  kable() %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, position="center")
+```
+</div>
+
+<div class = "col-md-6">
+```{r rpart2_plot,echo=F,message=F}
+## rpart2 Model Performance
+m <- "Decision Tree"
+
+sub_df <- auc_df %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                       labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(method==m & auc_type=="test_AUC")
+
+sub_stats <- p.table.model %>% 
+   mutate(level1=factor(level1,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+          level2=factor(level2,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         model=factor(model,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(model==m) %>% 
+  mutate(sig=case_when(p_value >= 0.05 ~ "NS",
+                       TRUE ~ "*")) %>% 
+  filter(sig=="NS")
+
+sub_stats <- sub_stats %>% 
+  mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
+  rename(group1=level1,group2=level2)
+
+ggplot(sub_df,aes(x=level,y=AUC)) + 
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+  geom_boxplot(aes(fill=method),notch = T) +
+  scale_fill_manual(values = pal2[m]) + 
+  scale_y_continuous(limits=c(0.2,1.05)) +
+  theme_bw() + xlab("") + ylab("AUROC") +
+  theme(legend.position ="none",
+        axis.text.x = element_text(angle = 45,hjust = 1),
+        panel.grid.major.x = element_blank(),
+        axis.text = element_text(size=14)) +
+  ggtitle(m) +
+  stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
+  stat_pvalue_manual(sub_stats,label="sig")
+ggsave("./figures/rpart2_with_stats.png",width = 5,height=4,units="in")
+
+```
+</div>
+</div>
+
+### SVM Radial
+
+### XGBTree
+
+## Feature Importance
+The mikropml packages includes an option for finding [feature importance](http://www.schlosslab.org/mikropml/reference/get_feature_importance.html) using a permutation method.
+This significantly increases the time to run the models.
+The function permutes each model feature (e.g.
+genus) and recalulates the AUC with that feature permuted.
+The outputs are the permuted AUC value and the difference between this AUC value and the actual AUC value.
+Larger positive differences can be interpreted as being more important since the AUC value is smaller when this feature is permuted.
+
+The importance values are all from the Random Forest model.

### 2021-03-25 paper/manuscript.Rmd

@@
 
-Courtney R. Armour, (ande? begum?), Patrick D. Schloss ${^\dagger}$
+Courtney R. Armour, Begüm D.Topçuoğlu, Andrea Garretto, Patrick D. Schloss ${^\dagger}$
 
@@
   - TODO: dig into literature on this
-  - TODO: address lower L2 logistic performance that prior results?
 - Within the RF model, the highest AUCs were observed for family (median AUC: `r round(rf_med_auc["family"],digits=3)`), genus(median AUC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three.
@@
   - ASV (median AUC: `r round(rf_med_auc["asv"],digits=3)`) performed significantly lower than OTU and genus (p < 0.05) but not family level P = 0.06).
+  - TODO: compare dada2 performance
 - One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g.
@@
   - 261 controls
-  - 229 cases
+  - 229 SRNs (describe how SRN defined)
 - processed with mothur v1.44.3

### 2021-03-25 code/R/caluclate_pvalues.R
cleanup

@@ -1,83 +1,0 @@
-library(tidyverse)
-library(mosaic)
-
-#models <- c("rf","glmnet","xgbTree","svmRadial","rpart2")
-models <- c("rf","glmnet","rpart2")
-levels <- c("phylum","class","order","family","genus","otu","asv")
-
-#functions adapted from begum mbio paper
-read_files <- function(filenames){
-  for(file in filenames){
-    # Read the files generated in main.R
-    # These files have cvAUCs and testAUCs for 100 data-splits
-    data <- read.delim(file, header=T, sep=',') %>%
-      mutate(level=unlist(strsplit(file,"-"))[2])
-
-  }
-  return(data)
-}
-
-#compare two models levels within a taxonomic level
-perm_p_value_level <- function(data, model_name_1, model_name_2, level_name){
-  test_subsample <- data %>%
-    select(AUC,method,level) %>%
-    filter((method == model_name_1 | method == model_name_2) & level == level_name)
-  obs <- abs(diff(mosaic::mean(AUC ~ method, data = test_subsample)))
-  auc.null <- do(10000) * diff(mosaic::mean(AUC ~ shuffle(method), data = test_subsample))
-  n <- length(auc.null[,1])
-  # r = #replications at least as extreme as observed effect
-  r <- sum(abs(auc.null[,1]) >= obs)
-  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
-  p.value=(r+1)/(n+1)
-  return(p.value)
-}
-
-#compare two taxonomic levels within a model type
-perm_p_value_model <- function(data, model_name, level_name_1, level_name_2){
-  test_subsample <- data %>%
-    select(AUC,method,level) %>%
-    filter(method == model_name & ( level == level_name_1 | level == level_name_2))
-  obs <- abs(diff(mosaic::mean(AUC ~ level, data = test_subsample)))
-  auc.null <- do(10000) * diff(mosaic::mean(AUC ~ shuffle(level), data = test_subsample))
-  n <- length(auc.null[,1])
-  # r = #replications at least as extreme as observed effect
-  r <- sum(abs(auc.null[,1]) >= obs)
-  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
-  p.value=(r+1)/(n+1)
-  #plot <- ggplot(auc.null, aes(x=auc.null[,1], color=auc.null[,1]>=obs)) + geom_histogram(fill="white", alpha=0.5, position="identity") +
[…]

### 2021-03-25 code/concat_pipeline_auc_output.py

@@ -6,7 +6,7 @@ parser = argparse.ArgumentParser(description='Provide model and taxonomy level')
 parser.add_argument("--taxonomy", action ="store", dest="tax_level",
     help="Taxonomy options: kingdom, phylum, class, order, family, genus, otu, asv")
 parser.add_argument("--model", action="store", dest="model",
-    help= "Model options: rpart2, regLogistic, svmRadial, rf, xgbTree")
+    help= "Model options: rpart2, glmnet, svmRadial, rf, xgbTree")
 
 args = parser.parse_args()
 tax_level = args.tax_level

### 2021-03-26 run.sh
rename

@@ -5,14 +5,14 @@
 #                             #
 ###############################
 # Name
-#SBATCH --job-name=prepdata
+#SBATCH --job-name=dada2
 # Resources
 # For MPI, increase ntasks-per-node
 # For multithreading, increase cpus-per-task
 #SBATCH --nodes=1
 #SBATCH --ntasks-per-node=1
-#SBATCH --cpus-per-task=1
-#SBATCH --mem-per-cpu=100GB
+#SBATCH --cpus-per-task=12
+#SBATCH --mem-per-cpu=10GB
 #SBATCH --time=20:00:00
 # Account
 #SBATCH --account=pschloss1
@@ -29,6 +29,14 @@
 #  2) Job Commands  #
 #                   #
 #####################
+
+# DADA2 analysis
+source /etc/profile.d/http_proxy.sh  # required for internet on the Great Lakes cluster
+#make data/raw/merged
+#make data/dada2/silva_%
+#make data/dada2/asv_table.tsv
+#make data/dada2/input_data.csv
+#make data/dada2/input_data_preproc.csv
 
 #source /etc/profile.d/http_proxy.sh  # required for internet on the Great Lakes cluster
 #make data/references/silva.v4.%
@@ -61,4 +69,4 @@
 
 #make data/asv/crc.shared
 #make data/asv/input_data.csv
-make data/asv/input_data_preproc.csv
+#make data/asv/input_data_preproc.csv

### 2021-03-26 exploratory/exploratory.Rmd
rename

@@
 author: "Courtney R Armour"
-date: '`r Sys.Date()`'
 output: html_document
 ---
-
+updated: `r Sys.Date()`
 ```{r setup_environment, include=FALSE}
@@
 ```
+
+## Table of Contents
+- [Input data overview](#input-data-overview)
+- [Hyperparameter Performance](#hyperparameter-performance)
+- [Model Performance](#model-performance)
+- [DADA2 Comparison](#dada2-comparison)
 
@@
 
-```{r dada2}
+```{r dada2_input}
 
-dada2_results <- read
+#mothur and dada2 input values
+input_data <- bind_rows(read_csv("../analysis/input_values.csv"),
+                        read_csv("../data/dada2/process/summary_input_values.csv"))
 
+kable(input_data) 
 ```

### 2021-03-26 paper/manuscript.Rmd
rename

@@
 geometry: margin=1.0in
+font-size: 11pt---
+output:
+  pdf_document:
+    keep_tex: true
+  word_document: default
+geometry: margin=1.0in
 font-size: 11pt
+header-includes:
+  - \usepackage{helvet}
+  - \renewcommand*\familydefault{\sfdefault}
+  - \usepackage{setspace}
+  - \doublespacing
+  - \usepackage[left]{lineno}
+  - \linenumbers
+---
+
+```{r setup_environment, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE,message=FALSE)
+options(scipen=999)
+library(tidyverse)
+```
+
+# Optimal Taxonomic Resolution for Microbiome-Based Classification of Colorectal Cancer
+
+\vspace{20mm}
+
+**Other title options:**  
+ASV level resolution is unnecessary for prediction of SRNs from microbiome data  
+Taxonomic Resolution Matters for Microbiome-Based Classification of Colorectal Cancer
+
+\vspace{10mm}
+
+Courtney R. Armour, Begüm D.Topçuoğlu, Andrea Garretto, Patrick D. Schloss ${^\dagger}$
+
+\vspace{20mm}
+
+${\dagger}$ To whom correspondence should be addressed:
+
+\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
+
+Department of Microbiology
+
+University of Michigan
+
+Ann Arbor, MI 48109
+
+\vspace{20mm}
+
+**observation format - max 1200 words, 2 figures, 25 ref**
+
+\newpage
+
+## Abstract (max 250 words)
+
+## Importance (max 150 words)
+
+\newpage
+
+## Introduction
+
+Efforts to realize the diagnostic potential of the gut microbiome in detecting screen relevant neoplasias (SRNs) have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on the amplicon sequencing of the v4 region of the 16S rRNA gene {}.
+However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown.
+Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs.
+However, whether the additional resolution provided by ASVs is useful for ML classification is unclear.
+Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups.
+Topçuoğlu *et al* (mBio 2020) recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
+This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.
+
+## Results
+
+```{r median_AUC}
+table <- read_csv("../exploratory/figures/median_test_auc.csv")
+
+rf_med_auc <- table %>% 
+  pull(rf)
+
+names(rf_med_auc) <- table %>% pull(level)
+```
+
+```{r input_table}
+input <- read_csv("../analysis/input_values.csv") %>% 
+  mutate(pct_kept = round(n_features_preproc/n_features,digits=3)*100)
+```
+
+(one-two sentance summary of methods) Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs.
+[(Figure)](#model-performance-across-taxonomy).
+Of the five ML methods, random forest (RF) was consistently the top performer at most taxonomic levels.
+[**TODO**: dig into literature on this: RF might be more appropriate for microbiome analysis since its more suitable for zero inflated data].
+Within the RF model, the highest AUCs were observed for family (median AUC: `r round(rf_med_auc["family"],digits=3)`), genus(median AUC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three.
+[(Figure)](#random-forest-model-performance-with-significance) Perfomance with ASVs (median AUC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU and genus (p < 0.05) but not family level P = 0.06).
+These results were consistent with ASVs generated with DADA2 (median AUC: <fill in>) {}.
+
+  One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g.
+phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls.
+As you reach genus/OTU level data there is enough data and variation but at the ASV level, the data is too specific to individuals and doesn't overlap enough.
+Examination of the prevalence of taxa in samples at each level supports this idea.
+A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels.
+The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples.
+[(Figure)](#prevalence-of-taxa-in-samples)
+
+  Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods).
+As part of this step, features are removed that wont provide useful information to build the model.
+For example, strongly correlated features provide the same information to build the model and thus can be collapsed.
+Additionally, features with zero or near-zero variance will not help the model differentiate groups and thus can be removed.
+Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing.
+At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing.
+[(Table)](#summary-of-features)
+
+- TODO: check why ASVs were removed in preprocessing (nzv?)
+- (depending on above can discuss why features were removed - e.g. only in one or a few samples)
+- OTU level data provides an idea balance of overlap and distinction to classify samples. 
+- Overall, the fine resolution of ASVs is unnecessary for ML classification with microbiome data since the vast majority of the ASVs are removed during preprocessing and the model performance is diminshed compared to other taxonomic levels.
+
+other items that could be addressed:
+
+- what are important taxa, patterns along levels (e.g fuso always in top )
+- abundance of important taxa: many of the top important features are low abund [(Figure)](#relative-abundance-of-important-features)
+
+## Conclusions
+
+## Materials and Methods
+Raw sequence data generated from a prior anaoysis {Baxter} was downlaoded from SRA (accession #).
+This data contains stool samples from 490 subjects.
+Based on the available metadata, samples categorized as normal, high risk normal, or adenoma were labeled "normal" for this analysis and samples categorized as advanced adenoma or carcinoma were labeled as "screen relevant neoplasia" (SRN).
+This resulted in a totla of 261 "normal" samples and 229 "SRN" samples.
+Sequence data was processed with Mothur (1.44.3) using the SILVA reference database (v132) to produce count tables for phylum, class, order, family, genus, OTU, and ASV.
+
+Machine learning models were run with mikropml (version) {}.
+Data was preprocessed to normalize values (scale/center?
+range?), remove values with zero or near-zero variance, and collapse collinear features using default parameters.
+Each taxnomic level/model combination was run with 100 different seeds.
+
+- pvalues calculated as previously described {begum}
+- prevalence = number of samples with non-zero abundance / total number of samples
+
+## Acknowledgements
+
+## Figures
+
+### Model Performance across Taxonomy  
+![](../exploratory/figures/all_model_level.png)
+
+### Random Forest Model Performance with Significance  
+![](../exploratory/figures/rf_with_stats.png){width=75%}
+
+### Prevalence of Taxa in Samples  
+![](../exploratory/figures/bin_prevalence.png){height=50%}
+
+### Summary of Features
+```{r feats}
+knitr::kable(input)
+```
+
+### Relative Abundance of Important Features
+![](../exploratory/figures/relabund_top10.png){height=95%}
+
 header-includes:

### 2021-04-06 exploratory/exploratory.Rmd
pages upload

@@
 knitr::opts_chunk$set(echo = FALSE,message=FALSE)
+options(scipen=999)
 
@@
 
+#given a number, adds correct commas
+set_commas <- function(num){
+
+}
+```
+
+```{r data_setup}
 # Variables
@@
 
-kable(input_data) 
+```
+
+Since there are some that believe ASVs generated by Mothur are not as good as ASVs generated by DADA2, I ran a comparison.
+I used DADA2 (v1.18.0) to generate ASVs and ran them through the same mikropml model pipeline.
+I ran dada(..., pool=T) to caputure more rare ASVs and subsampled to XX reads after processing.
+
+Below is a summary table of the number of features at each level.
+Interestingly there are only `r formatC(input_data %>% filter(level=="dada2") %>% pull(n_features),big.mark = ",",format="f",digits=0)` ASVs identified with DADA2 compared to `r formatC(input_data %>% filter(level=="asv") %>% pull(n_features),big.mark = ",",format="f",digits=0)` with Mothur.
+After preprocessing there are just `r input_data %>% filter(level=="dada2") %>% pull(n_features_preproc)` ASVs, about half of what we find with Mothur.
+
+```{r dada2_table}
+kable(input_data) %>% 
+  kable_styling(bootstrap_options = "striped")
+```
+
+```{r dada2_auc}
+
+mothur_df <- full_df %>% 
+  filter(level=="asv") %>% 
+  mutate(type="Mothur")
+
+mothur_auc_df <- auc_df %>% 
+  filter(level=="asv") %>% 
+  mutate(type="Mothur")
+
+dada2_df <- NULL
+dada2_auc_df <- NULL
+for( m in models ){
+  asv <- read_csv(paste0("../data/dada2/process/combined-dada2-",m,".csv")) %>%
+    mutate(type="DADA2")
+  dada2_df <- bind_rows(dada2_df,asv)
+
+  m_auc.df <- select(asv,cv_metric_AUC,AUC,method,type) %>% 
+    rename(test_AUC=AUC) %>% 
+    pivot_longer(cols = c("cv_metric_AUC","test_AUC"),names_to="auc_type",values_to="AUC")
+
+  dada2_auc_df <- bind_rows(dada2_auc_df,m_auc.df)
+}
+
+md_df <- bind_rows(mothur_df,dada2_df)
+
+md_auc_df <- bind_rows(mothur_auc_df,dada2_auc_df)
+
+```
+
+```{r dada2_stats}
+dada2_p_level <- read.csv("../data/dada2/analysis/dada2_pvalues_by_level.csv")
+dada2_p_model <- read.csv("../data/dada2/analysis/dada2_pvalues_by_model.csv")
+```
+
+```{r dada2_plot_auc,out.width="50%"}
+mod_md_df <- md_df %>% 
+  select(cv_metric_AUC,AUC,method,level,type) %>% 
+  rename(test_AUC=AUC,cv_AUC=cv_metric_AUC) %>% 
+  pivot_longer(cols=c("cv_AUC","test_AUC"),names_to="auc_type",values_to="AUC") %>% 
+  mutate(method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")))
+
+mod_md_df %>% 
+  filter(type=="DADA2") %>% 
+  ggplot(aes(x=method, y=AUC, fill=auc_type)) + 
+    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+    geom_boxplot(notch = TRUE) + 
+    theme_bw() + xlab("") + ylab("AUROC") + 
+    theme(legend.position ="top",
+          axis.text.x = element_text(angle = 45,hjust = 1)) +
+    ggtitle("DADA2 ASV Model Performance across Models")
+
+mod_md_df %>% 
+  ggplot(aes(x=type, y=AUC, fill=auc_type)) + 
+    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+    geom_boxplot(notch = TRUE) + 
+    facet_grid(.~method) +
+    theme_bw() + xlab("") + ylab("AUROC") + 
+    theme(legend.position ="top",
+          axis.text.x = element_text(angle = 45,hjust = 1)) +
+    ggtitle("DADA2 vs Mothur ASV Model Performance")
+
+```
+
+```{r dada2_plot_auc_wstat,out.width="50%"}
+
+#subset to just test data
+mod_md_auc_df <- md_auc_df %>% 
+  filter(auc_type=="test_AUC") %>% 
+  mutate(method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")))
+
+#compare different models with DADA2 ASVs
+sub_stats_level <- dada2_p_level %>% 
+  mutate(level=factor(level,levels=c("asv","dada2"),
+                            labels=c("Mothur","DADA2")),
+         model1=factor(model1,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")),
+         model2=factor(model2,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  mutate(sig=case_when(p_value >= 0.05 ~ "NS",
+                       TRUE ~ "*")) %>% 
+  mutate(y.position=seq.int(0.85,1.05,length.out = nrow(.))) %>% 
+  rename(group1=model1,group2=model2) %>% 
+  filter(sig=="NS")
+
+mod_md_auc_df %>% 
+  filter(type=="DADA2") %>% 
+  ggplot(aes(x=method,y=AUC)) + 
+    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+    geom_boxplot(aes(fill=method),notch = T) +
+    scale_fill_manual(values = pal2) + 
+    scale_y_continuous(limits=c(0.2,1.05)) +
+    theme_bw() + xlab("") + ylab("AUROC") +
+    theme(legend.position ="none",
+          axis.text.x = element_text(angle = 45,hjust = 1),
+          panel.grid.major.x = element_blank(),
+          axis.text = element_text(size=14)) +
+    ggtitle("DADA2") +
+    stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
+    stat_pvalue_manual(sub_stats_level,label="sig")
+
+#compare Mother and DADA2 ASVs across models
+sub_stats_model <- dada2_p_model %>% 
+   mutate(level1=factor(level1,levels=c("asv","dada2"),
+                            labels=c("Mothur","DADA2")),
+          level2=factor(level2,levels=c("asv","dada2"),
+                            labels=c("Mothur","DADA2")),
+         model=factor(model,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  mutate(sig=case_when(p_value >= 0.05 ~ "NS",
+                       TRUE ~ "*")) %>% 
+  #filter(sig=="NS") %>% 
+  mutate(y.position=rep(0.85,length.out = nrow(.))) %>% 
+  rename(group1=level1,group2=level2) %>% 
+  mutate(p_value=case_when(p_value < 0.001 ~ round(p_value,digits=4),
+                           TRUE ~ round(p_value,digits=2)))
+
+mod_md_auc_df %>%
+  rename(group=type,model=method) %>% 
+  ggplot(aes(x=group,y=AUC,color=group)) +
+    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+    geom_jitter() +
+    stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+    #geom_boxplot(aes(fill=model),notch = T) +
+    scale_y_continuous(limits=c(0.4,0.9)) +
+    theme_bw() + xlab("") + ylab("AUROC") +
+    theme(legend.position ="none",
+          axis.text.x = element_text(angle = 45,hjust = 1),
+          panel.grid.major.x = element_blank(),
+          axis.text = element_text(size=14)) +
+    ggtitle("DADA2 v Mothur ASVs") +
+    facet_grid(~model) +
+    stat_pvalue_manual(sub_stats_model,label="p_value")
+
+```
+
+```{r dada2_all, echo=F, message=F, out.width="100%",fig.height=3}
+dada2_auc_df %>% 
+  rename(level=type) %>% 
+  bind_rows(auc_df) %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv","DADA2"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV","DADA2")),
+         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(auc_type == "test_AUC") %>% 
+  ggplot(aes(x=level,y=AUC,fill=method)) + 
+    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+    geom_boxplot(notch = TRUE) + 
+    theme_bw() + xlab("") + ylab("AUROC") + 
+    theme(legend.position ="top",
+          axis.text.x = element_text(angle = 45,hjust = 1)) +
+    scale_fill_manual(values=pal2,name="Model")
+
 ```

### 2021-04-11 paper/manuscript.Rmd

@@
 ## Materials and Methods
+Raw 16S rRNA gene amplicon sequence data isolated from human gut samples {Baxter} was downloaded from SRA (accession #). The sequence data was processed using mothur (v1.)
+
+sequences were processed using mothur according to the Schloss Lab MiSeq SOP as described in the mothur wiki and accompanying study by Kozich et al.
+(13, 14).
+These steps included trimming and filtering for quality, aligning to the SILVA reference alignment (11), discarding sequences that aligned outside the V4 region, removing chimeric reads with UCHIME (15), and calculating distances between all pairs of sequences within each dataset prior to clustering.
+
 Raw sequence data generated from a prior anaoysis {Baxter} was downlaoded from SRA (accession #).

### 2021-04-11 code/R/calculate_pvalues.R
update pvalues

@@ -1,8 +1,8 @@
 library(tidyverse)
 library(mosaic)
 
-#models <- c("rf","glmnet","xgbTree","svmRadial","rpart2")
-models <- c("rf","glmnet","rpart2")
+models <- c("rf","glmnet","xgbTree","svmRadial","rpart2")
+#models <- c("rf","glmnet","rpart2")
 levels <- c("phylum","class","order","family","genus","otu","asv")
 
 #functions adapted from begum mbio paper

### 2021-04-29 code/R/summarize_prevalence.R
add prevalence code

@@ -1,0 +1,54 @@
+# This script reads in the input data and preprocessed input data for
+# each taxonomic level and quantifies the prevalence of each taxa
+# as the number of samples with count > 1 / number of samples
+
+library(tidyverse)
+library(data.table)
+
+#function to read input files and calculate prevalence
+get_prevalence_preprocess <- function(file,file_proc,level){
+  data <- as.data.frame(fread(file))
+  data_proc <- as.data.frame(fread(file_proc))
+  if(nrow(data) != nrow(data_proc)){
+    stop("Why is there a different number of samples in preprocessed data!")
+  }
+  n_samp <- nrow(data)
+  keep_names <- str_replace(colnames(data_proc)[-1],"_1","")
+  prev <- data %>%
+    select(-dx) %>%
+    #select(data,keep_names) %>% #only keep taxa in preprocessed data
+    pivot_longer(everything(),names_to="name",values_to="count") %>%
+    filter(count > 0) %>%
+    group_by(name) %>%
+    tally() %>%
+    mutate(prevalence=n/n_samp) %>%
+    mutate(level := !!level) %>%
+    select(level,name,prevalence) %>%
+    mutate(kept=case_when(name %in% keep_names ~ 1,
+                          TRUE ~ 0))
+  return(prev)
+}
+
+
+prev_df <- get_prevalence_preprocess("./data/phylum/input_data.csv",
+                                     "./data/phylum/input_data_preproc.csv",
+                                     "phylum") %>%
+  bind_rows(get_prevalence_preprocess("./data/class/input_data.csv",
+                                      "./data/class/input_data_preproc.csv",
+                                      "class")) %>%
+  bind_rows(get_prevalence_preprocess("./data/order/input_data.csv",
+                                      "./data/order/input_data_preproc.csv",
+                                      "order")) %>%
+  bind_rows(get_prevalence_preprocess("./data/family/input_data.csv",
+                                      "./data/family/input_data_preproc.csv",
+                                      "family")) %>%
+  bind_rows(get_prevalence_preprocess("./data/genus/input_data.csv",
+                                      "./data/genus/input_data_preproc.csv",
+                                      "genus")) %>%
+  bind_rows(get_prevalence_preprocess("./data/otu/input_data.csv",
+                                      "./data/otu/input_data_preproc.csv",
+                                      "otu")) %>%
+  bind_rows(get_prevalence_preprocess("./data/asv/input_data.csv",
[…]

### 2021-04-29 exploratory/exploratory.Rmd
ms updates

@@
 for( m in models ){
-  phylum <- read_csv(paste0("../data/process/combined-phylum-",m,".csv")) %>% 
+  phylum <- read_csv(paste0("../data/process/combined-phylum-",m,".csv"),
+                     col_types = c(method = col_character(),.default=col_double())) %>% 
     mutate(level="phylum")
-  class <- read_csv(paste0("../data/process/combined-class-",m,".csv")) %>% 
+  class <- read_csv(paste0("../data/process/combined-class-",m,".csv"),
+                     col_types = c(method = col_character(),.default=col_double())) %>% 
     mutate(level="class")
-  order <- read_csv(paste0("../data/process/combined-order-",m,".csv")) %>% 
+  order <- read_csv(paste0("../data/process/combined-order-",m,".csv"),
+                     col_types = c(method = col_character(),.default=col_double())) %>% 
     mutate(level="order")
-  family <- read_csv(paste0("../data/process/combined-family-",m,".csv")) %>% 
+  family <- read_csv(paste0("../data/process/combined-family-",m,".csv"),
+                     col_types = c(method = col_character(),.default=col_double())) %>% 
     mutate(level="family")
-  genus <- read_csv(paste0("../data/process/combined-genus-",m,".csv")) %>%
+  genus <- read_csv(paste0("../data/process/combined-genus-",m,".csv"),
+                     col_types = c(method = col_character(),.default=col_double())) %>%
     mutate(level="genus")
-  otu <- read_csv(paste0("../data/process/combined-otu-",m,".csv")) %>%
+  otu <- read_csv(paste0("../data/process/combined-otu-",m,".csv"),
+                     col_types = c(method = col_character(),.default=col_double())) %>%
     mutate(level="otu")
-  asv <- read_csv(paste0("../data/process/combined-asv-",m,".csv")) %>%
+  asv <- read_csv(paste0("../data/process/combined-asv-",m,".csv"),
+                     col_types = c(method = col_character(),.default=col_double())) %>%
     mutate(level="asv")
@@
     group_by(level) %>% 
-    summarise(!!m := median(AUC))
+    summarise(!!m := median(AUC),.groups = 'drop')
 
@@
 #compare two models levels within a taxonomic level
-p.table.level <- read_csv("../analysis/pvalues_by_level.csv")
+p.table.level <- read_csv("../analysis/pvalues_by_level.csv",
+                     col_types = c(p_value = col_double(),.default=col_character()))
 #compare two taxonomic levels within a model type
-p.table.model <- read_csv("../analysis/pvalues_by_model.csv")
+p.table.model <- read_csv("../analysis/pvalues_by_model.csv",
+                     col_types = c(p_value = col_double(),.default=col_character()))
 
@@
 - [Model Performance](#model-performance)
+  - [Random Forest](#random-forest-model-performance-with-significance)
 - [DADA2 Comparison](#dada2-comparison)
+- [Feature Importance](#feature-importance)
 
@@
 
-input <- read_csv("../analysis/input_values.csv")
+input <- read_csv("../analysis/input_values.csv",
+                  col_types = c(level = col_character(), .default = col_double()))
 
@@
     geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
-    geom_boxplot(notch = TRUE) + 
+    geom_boxplot() + 
     facet_grid(.~method) + 
@@
   geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
-  geom_boxplot(notch = TRUE) + 
+  geom_boxplot() + 
   facet_wrap(.~level,nrow = 2) + 
@@
     geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
-    geom_boxplot(notch = TRUE) + 
+    geom_boxplot(alpha=0.9) + 
     theme_bw() + xlab("") + ylab("AUROC") + 
     theme(legend.position ="top",
-          axis.text.x = element_text(angle = 45,hjust = 1)) +
-    scale_fill_manual(values=pal2,name="Model")
-ggsave("./figures/all_model_level.png")
+          axis.text.x = element_text(angle = 0),
+          axis.text = element_text(size=14)) +
+    scale_fill_manual(values=pal2,name="Model") #+
+  #guides(fill=guide_legend(nrow=2))
+ggsave("./figures/all_model_level.png",width = 7,height=4,units="in")
 ```
@@
 #mothur and dada2 input values
-input_data <- bind_rows(read_csv("../analysis/input_values.csv"),
-                        read_csv("../data/dada2/process/summary_input_values.csv"))
+input_data <- bind_rows(read_csv("../analysis/input_values.csv",
+                                 col_types = c(level = col_character(), .default = col_double())),
+                        read_csv("../data/dada2/process/summary_input_values.csv",
+                                 col_types = c(level = col_character(), .default = col_double())))
 
@@
 for( m in models ){
-  asv <- read_csv(paste0("../data/dada2/process/combined-dada2-",m,".csv")) %>%
+  asv <- read_csv(paste0("../data/dada2/process/combined-dada2-",m,".csv"),
+                     col_types = c(method = col_character(),.default=col_double())) %>%
     mutate(type="DADA2")
@@
     geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
-    geom_boxplot(notch = TRUE) + 
+    geom_boxplot() + 
     theme_bw() + xlab("") + ylab("AUROC") + 
@@
     geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
-    geom_boxplot(notch = TRUE) + 
+    geom_boxplot() + 
     facet_grid(.~method) +
@@
     geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
-    geom_boxplot(aes(fill=method),notch = T) +
+    geom_boxplot(aes(fill=method)) +
     scale_fill_manual(values = pal2) + 
@@
     stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
-    #geom_boxplot(aes(fill=model),notch = T) +
+    #geom_boxplot(aes(fill=model)) +
     scale_y_continuous(limits=c(0.4,0.9)) +
@@
     scale_fill_manual(values=pal2,name="Model")
+
+```
+
+```{r dada2_rf_wstat}
+
+# #subset to just test data
+# mod_md_auc_df <- md_auc_df %>% 
+#   filter(auc_type=="test_AUC",method=="rf") %>% 
+#   mutate(method=factor(method,levels=c("rf"),
+#                        labels=c("Random Forest")))
+# 
+# #compare different models with DADA2 ASVs
+# sub_stats_dada2_rf <- dada2_p_model %>% 
+#   filter(model=="rf") %>%    
+#   mutate(sig=case_when(p_value >= 0.05 ~ "NS",
+#                        TRUE ~ "*")) %>% 
+#   mutate(y.position=seq.int(0.85,1.05,length.out = nrow(.))) %>% 
+#   rename(group1=level1,group2=level2) %>% 
+#   filter(sig=="NS")
+# 
+# mod_md_auc_df %>% 
+#   filter(type=="DADA2") %>% 
+#   ggplot(aes(x=method,y=AUC)) + 
+#     geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+#     geom_boxplot(aes(fill=method)) +
+#     scale_fill_manual(values = pal2) + 
+#     scale_y_continuous(limits=c(0.2,1.05)) +
+#     theme_bw() + xlab("") + ylab("AUROC") +
+#     theme(legend.position ="none",
+#           axis.text.x = element_text(angle = 45,hjust = 1),
+#           panel.grid.major.x = element_blank(),
+#           axis.text = element_text(size=14)) +
+#     ggtitle("DADA2") +
+#     stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
+#     stat_pvalue_manual(sub_stats_level,label="sig")
+# 
+# 
+# #compare Mother and DADA2 ASVs across models
+# sub_stats_model <- dada2_p_model %>% 
+#    mutate(level1=factor(level1,levels=c("asv","dada2"),
+#                             labels=c("Mothur","DADA2")),
+#           level2=factor(level2,levels=c("asv","dada2"),
+#                             labels=c("Mothur","DADA2")),
+#          model=factor(model,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+#                        labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+#   mutate(sig=case_when(p_value >= 0.05 ~ "NS",
+#                        TRUE ~ "*")) %>% 
+#   #filter(sig=="NS") %>% 
+#   mutate(y.position=rep(0.85,length.out = nrow(.))) %>% 
+#   rename(group1=level1,group2=level2) %>% 
+#   mutate(p_value=case_when(p_value < 0.001 ~ round(p_value,digits=4),
+#                            TRUE ~ round(p_value,digits=2)))
+# 
+# mod_md_auc_df %>%
+#   rename(group=type,model=method) %>% 
+#   ggplot(aes(x=group,y=AUC,color=group)) +
+#     geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+#     geom_jitter() +
+#     stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+#     #geom_boxplot(aes(fill=model)) +
+#     scale_y_continuous(limits=c(0.4,0.9)) +
+#     theme_bw() + xlab("") + ylab("AUROC") +
+#     theme(legend.position ="none",
+#           axis.text.x = element_text(angle = 45,hjust = 1),
+#           panel.grid.major.x = element_blank(),
+#           axis.text = element_text(size=14)) +
+#     ggtitle("DADA2 v Mothur ASVs") +
+#     facet_grid(~model) +
+#     stat_pvalue_manual(sub_stats_model,label="p_value")
 
@@
   geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
-  geom_boxplot(aes(fill=method),notch = T) +
+  geom_boxplot(aes(fill=method)) +
   scale_fill_manual(values = pal2[m]) + 
@@
   ggtitle(m) +
-  stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
+  #stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
   stat_pvalue_manual(sub_stats,label="sig")
@@
 </div>
+
+```{r rf_wstat_side,include=F}
+sub_stats <- sub_stats %>% 
+  #mutate(y.position=seq.int(0.85,.95,length.out = nrow(sub_stats)))
+  mutate(y.position=c(0.85,0.95,0.9,.95))
+
+ggplot(sub_df,aes(x=factor(level,levels=c("ASV", "OTU","Genus","Family","Order","Class", "Phylum")),
+                           y=AUC)) + 
+  #geom_jitter(aes(color=method,width=0,height=0.2)) +
+  geom_boxplot(aes(fill=method,alpha=0.8)) +
+  scale_fill_manual(values = pal2[m]) +
+  #scale_color_manual(values = pal2[m]) +
+  #scale_x_continuous(limits=c(0.4,1)) +
+  theme_bw() + xlab("") + ylab("AUROC") +
+  theme(legend.position ="none",
+        #axis.text.x = element_text(angle = 45,hjust = 1),
+        panel.grid.major.x = element_blank(),
+        axis.text = element_text(size=14)) +
+  ggtitle(m) +
+  stat_pvalue_manual(sub_stats,label="sig",coord.flip = T,angle = 0,hjust=0) +
+  coord_flip() +
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed") 
+ggsave("./figures/rf_with_stats_sideways.png",width = 5,height=4,units="in")
+```
 
@@
 
+<div class = "row">
+<div class = "col-md-6">
+```{r svmRadial_table,echo=FALSE}
+t_table %>%
+  select(level,svmRadial) %>% 
+  mutate(svmRadial=round(svmRadial,digits = 3)) %>% 
+  rename("Median AUC"=svmRadial,"Level"=level) %>% 
+  kable() %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, position="center")
+```
+</div>
+
+<div class = "col-md-6">
+```{r svmRadial_plot,echo=F,message=F}
+## svmRadial Model Performance
+m <- "SVM Radial"
+
+sub_df <- auc_df %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                       labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(method==m & auc_type=="test_AUC")
+
+sub_stats <- p.table.model %>% 
+   mutate(level1=factor(level1,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+          level2=factor(level2,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         model=factor(model,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(model==m) %>% 
+  mutate(sig=case_when(p_value >= 0.05 ~ "NS",
+                       TRUE ~ "*")) %>% 
+  filter(sig=="NS")
+
+sub_stats <- sub_stats %>% 
+  mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
+  rename(group1=level1,group2=level2)
+
+ggplot(sub_df,aes(x=level,y=AUC)) + 
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+  geom_boxplot(aes(fill=method)) +
+  scale_fill_manual(values = pal2[m]) + 
+  scale_y_continuous(limits=c(0.2,1.05)) +
+  theme_bw() + xlab("") + ylab("AUROC") +
+  theme(legend.position ="none",
+        axis.text.x = element_text(angle = 45,hjust = 1),
+        panel.grid.major.x = element_blank(),
+        axis.text = element_text(size=14)) +
+  ggtitle(m) +
+  stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
+  stat_pvalue_manual(sub_stats,label="sig")
+ggsave("./figures/svmRadial_with_stats.png",width = 5,height=4,units="in")
+
+```
+</div>
+</div>
+
 ### XGBTree
+
+<div class = "row">
+<div class = "col-md-6">
+```{r xgbTree_table,echo=FALSE}
+t_table %>%
+  select(level,xgbTree) %>% 
+  mutate(xgbTree=round(xgbTree,digits = 3)) %>% 
+  rename("Median AUC"=xgbTree,"Level"=level) %>% 
+  kable() %>%
+  kable_styling(bootstrap_options = "striped", full_width = FALSE, position="center")
+```
+</div>
+
+<div class = "col-md-6">
+```{r xgbTree_plot,echo=F,message=F}
+## XGBoost Model Performance
+m <- "XGBoost"
+
+sub_df <- auc_df %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                       labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(method==m & auc_type=="test_AUC")
+
+sub_stats <- p.table.model %>% 
+   mutate(level1=factor(level1,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+          level2=factor(level2,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         model=factor(model,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(model==m) %>% 
+  mutate(sig=case_when(p_value >= 0.05 ~ "NS",
+                       TRUE ~ "*")) %>% 
+  filter(sig=="NS")
+
+sub_stats <- sub_stats %>% 
+  mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
+  rename(group1=level1,group2=level2)
+
+ggplot(sub_df,aes(x=level,y=AUC)) + 
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+  geom_boxplot(aes(fill=method)) +
+  scale_fill_manual(values = pal2[m]) + 
+  scale_y_continuous(limits=c(0.2,1.05)) +
+  theme_bw() + xlab("") + ylab("AUROC") +
+  theme(legend.position ="none",
+        axis.text.x = element_text(angle = 45,hjust = 1),
+        panel.grid.major.x = element_blank(),
+        axis.text = element_text(size=14)) +
+  ggtitle(m) +
+  stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
+  stat_pvalue_manual(sub_stats,label="sig")
+ggsave("./figures/xgbTree_with_stats.png",width = 5,height=4,units="in")
+
+```
+</div>
+</div>
 
@@
 The function permutes each model feature (e.g.
-genus) and recalulates the AUC with that feature permuted.
+genus) and recalculates the AUC with that feature permuted.
 The outputs are the permuted AUC value and the difference between this AUC value and the actual AUC value.
@@
 The importance values are all from the Random Forest model.
+
+```{r imp_plots}
+imp_plots <- list()
+m="rf"
+for(l in levels){
+  test <- read_csv(paste0("../data/process/importance-",l,"-",m,".csv"),
+                   col_types = c(perf_metric = col_double(),
+                                 perf_metric_diff = col_double(),
+                                 seed = col_double(),
+                                 .default = col_character())) %>% 
+    mutate(level := !!l)
+  test_results <- read_csv(paste0("../data/process/combined-",l,"-",m,".csv"),
+                           col_types = c(method = col_character(), .default = col_double())) %>% 
+    mutate(level := !!l)
+
+  test_tax <- read_tsv(paste0("../data/",l,"/crc.taxonomy"),
+                   col_types = c(Size = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level := !!l)
+
+  test_merge <- test %>%
+    select(perf_metric,perf_metric_diff,names,method,seed,level) %>% 
+    rename(AUC=perf_metric,OTU=names,AUC_diff=perf_metric_diff) %>% 
+    inner_join(test_tax,by=c("level","OTU")) %>% 
+    mutate(name=paste0(Taxonomy," (",OTU,")"))
+
+  test_quartiles <- quantile(pull(test_results,AUC))
+  names(test_quartiles) <- c("q0","q1","q2","q3","q4")
+
+  test_median <- median(pull(test_results,AUC))
+
+  if(l %in% c("otu","asv")){
+    imp_plots[[paste0(m,"-",l,"-1")]] <- test_merge %>% 
+      mutate(OTU=reorder(OTU,desc(AUC),.fun="median")) %>% 
+      ggplot(aes(x=AUC,y=OTU)) + 
+        geom_point(alpha=0.5,color="lightblue") +
+        stat_summary(fun = mean,geom = "point", shape = 18, size = 3,aes(color="Mean")) +
+        stat_summary(fun = median,geom = "point", shape = 18, size = 3,aes(color="Median")) +
+        scale_color_manual(name="",values=c("red","green")) +
+        theme_bw() +
+        theme(legend.position = "bottom",
+              axis.text.y = element_blank()) +
+        geom_vline(xintercept = test_quartiles["q1"]) +
+        geom_vline(xintercept = test_quartiles["q3"]) + 
+        geom_vline(xintercept = test_median,lty="dashed") +
+        ggtitle(paste0(m,"-",l,": AUC ordered")) +
+        xlab("Permuted AUC") + ylab("Feature")
+    #ggsave(paste0("./data/analysis/exploratory3/importance_metric_",m,"_",l,".png"))
+
+  imp_plots[[paste0(m,"-",l,"-2")]] <- test_merge %>% 
+    mutate(OTU=reorder(OTU,AUC_diff,.fun="median")) %>% 
+    ggplot(aes(x=AUC_diff,y=OTU)) + 
+      geom_point(alpha=0.5,color="lightblue") +
+      stat_summary(fun = mean,geom = "point", shape = 18, size = 3,aes(color="Mean")) +
+      stat_summary(fun = median,geom = "point", shape = 18, size = 3,aes(color="Median")) +
+      scale_color_manual(name="",values=c("red","green")) +
+      geom_vline(xintercept = 0,lty="dashed",color="black") +
+      theme_bw() +
+      theme(legend.position = "bottom",
+            axis.text.y = element_blank()) +
+      ggtitle(paste0(m,"-",l,": AUC difference sorted")) +
+      ylab("Feature") + xlab("AUC Difference (actual - permuted)")
+    #ggsave(paste0("./data/analysis/exploratory3/importance_diff_",m,"_",l,".png"))
+
+  }else{
+
+    imp_plots[[paste0(m,"-",l,"-1")]] <- test_merge %>% 
+      mutate(name=reorder(name,desc(AUC),.fun="median")) %>% 
+      ggplot(aes(x=AUC,y=name)) + 
+        geom_point(alpha=0.5,color="lightblue") +
+        stat_summary(fun = mean,geom = "point", shape = 18, size = 3,aes(color="Mean")) +
+        stat_summary(fun = median,geom = "point", shape = 18, size = 3,aes(color="Median")) +
+        scale_color_manual(name="",values=c("red","green")) +
+        theme_bw() +
+        theme(legend.position = "bottom") +
+        geom_vline(xintercept = test_quartiles["q1"]) +
+        geom_vline(xintercept = test_quartiles["q3"]) + 
+        geom_vline(xintercept = test_median,lty="dashed") +
+        ggtitle(paste0(m,"-",l,": AUC ordered")) +
+        xlab("Permuted AUC") + ylab("Feature")
+    #ggsave(paste0("./data/analysis/exploratory3/importance_metric_",m,"_",l,".png"))
+
+    imp_plots[[paste0(m,"-",l,"-2")]] <- test_merge %>% 
+      mutate(name=reorder(name,AUC_diff,.fun="median")) %>% 
+      ggplot(aes(x=AUC_diff,y=name)) + 
+        geom_point(alpha=0.5,color="lightblue") +
+        stat_summary(fun = mean,geom = "point", shape = 18, size = 3,aes(color="Mean")) +
+        stat_summary(fun = median,geom = "point", shape = 18, size = 3,aes(color="Median")) +
+        scale_color_manual(name="",values=c("red","green")) +
+        geom_vline(xintercept = 0,lty="dashed",color="black") +
+        theme_bw() +
+        theme(legend.position = "bottom") +
+        ggtitle(paste0(m,"-",l,": AUC difference sorted")) +
+        ylab("Feature") + xlab("AUC Difference (actual - permuted)")
+    #ggsave(paste0("./data/analysis/exploratory3/importance_diff_",m,"_",l,".png"))
+
+  }
+}
+
+```
+
+```{r rf_importance_plots,echo=F,message=F,fig.show="hold",out.width="50%"}
+imp_plots[["rf-phylum-1"]]
+imp_plots[["rf-phylum-2"]]
+
+imp_plots[["rf-class-1"]]
+imp_plots[["rf-class-2"]]
+
+imp_plots[["rf-order-1"]]
+imp_plots[["rf-order-2"]]
+
+imp_plots[["rf-family-1"]]
+imp_plots[["rf-family-2"]]
+
+imp_plots[["rf-genus-1"]]
+imp_plots[["rf-genus-2"]]
+
+```
+
+### Top 10 Important features
+```{r importance_rank, message=F,echo=F}
+
+rf_imp <- read_csv("../data/process/importance-phylum-rf.csv",
+                   col_types = c(perf_metric = col_double(),
+                                 perf_metric_diff = col_double(),
+                                 seed = col_double(),
+                                 .default = col_character())) %>%
+  mutate(level="phylum") %>%
+  bind_rows(read_csv("../data/process/importance-class-rf.csv",
+                   col_types = c(perf_metric = col_double(),
+                                 perf_metric_diff = col_double(),
+                                 seed = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="class")) %>%
+  bind_rows(read_csv("../data/process/importance-order-rf.csv",
+                   col_types = c(perf_metric = col_double(),
+                                 perf_metric_diff = col_double(),
+                                 seed = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="order")) %>%
+  bind_rows(read_csv("../data/process/importance-family-rf.csv",
+                   col_types = c(perf_metric = col_double(),
+                                 perf_metric_diff = col_double(),
+                                 seed = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="family")) %>%
+  bind_rows(read_csv("../data/process/importance-genus-rf.csv",
+                   col_types = c(perf_metric = col_double(),
+                                 perf_metric_diff = col_double(),
+                                 seed = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="genus")) %>%
+  bind_rows(read_csv("../data/process/importance-otu-rf.csv",
+                   col_types = c(perf_metric = col_double(),
+                                 perf_metric_diff = col_double(),
+                                 seed = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="otu")) %>% 
+  bind_rows(read_csv("../data/process/importance-asv-rf.csv",
+                   col_types = c(perf_metric = col_double(),
+                                 perf_metric_diff = col_double(),
+                                 seed = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="asv"))
+
+rf_tax <- read_tsv("../data/phylum/crc.taxonomy",
+                   col_types = c(Size = col_double(),
+                                 .default = col_character())) %>%
+  mutate(level="phylum") %>%
+  bind_rows(read_tsv("../data/class/crc.taxonomy",
+                   col_types = c(Size = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="class")) %>%
+  bind_rows(read_tsv("../data/order/crc.taxonomy",
+                   col_types = c(Size = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="order")) %>%
+  bind_rows(read_tsv("../data/family/crc.taxonomy",
+                   col_types = c(Size = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="family")) %>%
+  bind_rows(read_tsv("../data/genus/crc.taxonomy",
+                   col_types = c(Size = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="genus")) %>%
+  bind_rows(read_tsv("../data/otu/crc.taxonomy",
+                   col_types = c(Size = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="otu")) %>% 
+  bind_rows(read_tsv("../data/asv//crc.taxonomy",
+                   col_types = c(Size = col_double(),
+                                 .default = col_character())) %>%
+    mutate(level="asv"))
+
+rf_imp_tax <- rf_imp %>%
+  select(perf_metric,perf_metric_diff,names,method,seed,level) %>%
+  rename(AUC=perf_metric,OTU=names,AUC_diff=perf_metric_diff) %>%
+  inner_join(rf_tax,by=c("level","OTU"))
+
+rank_top_tax <- function(data,tax){
+  sub_data <- data %>%
+    filter(level==tax) %>%
+    group_by(level,Taxonomy,OTU) %>% 
+    summarise(mean_AUC_diff = mean(AUC_diff)) %>% 
+    ungroup() %>% 
+    arrange(desc(mean_AUC_diff)) %>% 
+    mutate(rank=seq(1,nrow(.)))
+
+  return(sub_data)
+}
+
+ranked_tax <- bind_rows(rank_top_tax(rf_imp_tax,"phylum"),
+                        rank_top_tax(rf_imp_tax,"class"),
+                        rank_top_tax(rf_imp_tax,"order"),
+                        rank_top_tax(rf_imp_tax,"family"),
+                        rank_top_tax(rf_imp_tax,"genus"),
+                        rank_top_tax(rf_imp_tax,"otu"),
+                        rank_top_tax(rf_imp_tax,"asv"))
+
+top_10_tax <- ranked_tax %>% 
+  filter(rank <= 10) %>% 
+  select(level,Taxonomy,rank) %>% 
+  pivot_wider(names_from = level,values_from = Taxonomy)
+
+kable(top_10_tax,format="html",booktabs=T,align="clllllll",
+      col.names = c("Rank",levels_names)) %>% 
+   kable_material(c("striped", "hover","condensed"))
+```
+
+## Prevalence
+
+```{r prevalence_preprocess,echo=F,message=F}
+get_prevalence_preprocess <- function(file,file_proc,level){
+  data <- as.data.frame(fread(file))
+  data_proc <- as.data.frame(fread(file_proc))
+  keep_names <- str_replace(colnames(data_proc)[-1],"_1","")
+  n_samp <- nrow(data_proc)
+  prev <- select(data,keep_names) %>% 
+    pivot_longer(everything(),names_to="name",values_to="count") %>% 
+    filter(count > 0) %>% 
+    group_by(name) %>% 
+    tally() %>% 
+    mutate(prevalence=n/n_samp) %>% 
+    mutate(level := !!level) %>% 
+    select(level,name,prevalence)
+  return(prev)
+}
+
+# prev_df <- get_prevalence_preprocess("./data/final/phylum/input_data.csv",
+#                                      "./data/final/phylum/input_data_preproc.csv",
+#                                      "phylum") %>%
+#   bind_rows(get_prevalence_preprocess("./data/final/class/input_data.csv",
+#                                       "./data/final/class/input_data_preproc.csv",
+#                                       "class")) %>%
+#   bind_rows(get_prevalence_preprocess("./data/final/order/input_data.csv",
+#                                       "./data/final/order/input_data_preproc.csv",
+#                                       "order")) %>%
+#   bind_rows(get_prevalence_preprocess("./data/final/family/input_data.csv",
+#                                       "./data/final/family/input_data_preproc.csv",
+#                                       "family")) %>%
+#   bind_rows(get_prevalence_preprocess("./data/final/genus/input_data.csv",
+#                                       "./data/final/genus/input_data_preproc.csv",
+#                                       "genus")) %>%
+#   bind_rows(get_prevalence_preprocess("./data/final/otu/input_data.csv",
+#                                       "./data/final/otu/input_data_preproc.csv",
+#                                       "otu")) %>%
+#   bind_rows(get_prevalence_preprocess("./data/final/asv/input_data.csv",
+#                                       "./data/final/asv/input_data_preproc.csv",
+#                                       "asv"))
+# write_csv(prev_df,"./data/analysis/exploratory3/prevalence_by_level_preprocess.csv")
+
+prevalence_prepro <- read_csv("./data/analysis/exploratory3/prevalence_by_level_preprocess.csv")
+
+prevalence_prepro$level <- factor(prevalence_prepro$level,levels=levels)
+
+ggplot(prevalence_prepro,aes(x=level,y=prevalence,fill=level)) + 
+  #geom_jitter() +
+  geom_boxplot() +
+  stat_summary(fun = mean, color="red",geom = "point", shape = 18, size = 4,show.legend = FALSE) +
+  theme_bw() + xlab("") +
+  theme(legend.position = "none",
+        axis.text = element_text(size=14),
+        axis.title.y = element_text(size=14))
+ggsave("./data/analysis/exploratory3/prevalence_preprocess.png",width=5,height=4,units = "in")
+
+ggplot(prevalence_prepro,aes(x=prevalence,fill=level)) +
+  geom_histogram(binwidth = 0.05) +
+  facet_wrap(.~level,scales = "free") +
+  theme_bw() +
+  scale_y_continuous(expand=c(0,0)) +
+  scale_x_continuous(expand=c(0,0)) +
+  theme(legend.position = "none") 
+ggsave("./data/analysis/exploratory3/prevalence_histogram.png",width=5,height=4,units = "in")
+
+ggplot(prevalence_prepro,)
+
+tax_count <- prevalence_prepro %>% 
+  group_by(level) %>% 
+  mutate(numOTU = n_distinct(name)) %>% 
+  select(level,numOTU) %>% 
+  distinct()
+
+# prevalence_prepro %>% 
+#   group_by(level) %>% 
+#   count(prevalence) %>% 
+#   inner_join(tax_count,by="level") %>% 
+#   mutate(pct_OTU = n/numOTU) %>% 
+#   mutate(prevalence = round(prevalence,digits=2)) %>% 
+#   ggplot(aes(x=prevalence,y=pct_OTU)) +
+#     geom_line() +
+#     facet_wrap(~level,scales = "free")
+  #ggplot(aes(x=prevalence,y=n)) +
+  #  geom_bar(stat="identity") +
+  #  facet_grid(~level)
+
+pct_bins <- NULL
+level_nOTU <- NULL
+for(l in levels){
+  sub_data <- prevalence_prepro %>% 
+    filter(level==l)
+  hist <- hist(sub_data$prevalence,breaks = seq(0,1,0.25))
+
+  nOTU <- sub_data %>%
+    select(name) %>% 
+    n_distinct()
+
+  level_nOTU <- bind_rows(level_nOTU,
+                          data.frame(l,nOTU))
+
+  pct_bins <- pct_bins %>% 
+    bind_rows(data.frame(level=l,
+                         prev=hist$breaks[-1],
+                         count=hist$counts,
+                         pct=hist$counts/nOTU))
+}
+
+level_nOTU <- level_nOTU %>% 
+  mutate(level_name = levels_names) 
+labs <- paste0(level_nOTU$level_name," (",level_nOTU$nOTU,")")
+pct_bins %>%
+  mutate(level=factor(level,levels=levels,labels=labs)) %>% 
+  ggplot(aes(x=prev,y=pct,fill=level)) +
+    geom_bar(stat = "identity") +
+    facet_wrap(~level) +
+    scale_y_continuous(expand=c(0,0),limits=c(0,0.8),labels = scales::percent_format()) +
+    scale_x_continuous(expand=c(0,0),breaks = c(0,.25,0.50,.75,1)) +
+    xlab("Prevalence in Samples") +
+    ylab("Percent of Taxa") +
+    theme_bw() +
+    theme(legend.position = "none") +
+  geom_text(aes(label = paste0((round(pct,digits=2)*100),"%")),nudge_y=0.05)
+ggsave("./data/analysis/exploratory3/bin_prevalence.png")
+
+# 
+# ggplot(prevalence_prepro,aes(x=prevalence,fill=level)) +
+#   geom_histogram(binwidth = 0.25) +
+#   facet_wrap(.~level,scales = "free") +
+#   theme_bw() +
+#   scale_y_continuous(expand=c(0,0)) +
+#   scale_x_continuous(expand=c(0,0)) +
+#   theme(legend.position = "none")
+
+```

### 2021-04-29 paper/manuscript.Rmd
ms updates

--- 2021-03-25 paper/manuscript.Rmd
@@
 library(tidyverse)
+library(kableExtra)
 ```
 
-# Optimal Taxonomic Resolution for Microbiome-Based Classification of Colorectal Cancer
+# Taxonomic Resolution Matters for Microbiome-Based Classification of Colorectal Cancer
 
-\vspace{20mm}
+\vspace{10mm}
 
 **Other title options:**  
+Optimal Taxonomic Resolution for Microbiome-Based Classification of Colorectal Cancer
 ASV level resolution is unnecessary for prediction of SRNs from microbiome data 
-OTUs are the Optimal Taxonomic Level for Microbiome-Based Classification of Colorectal Cancer
+...
 
@@
 
-## Introduction
+## Introduction (~250 words)
 
-- Colorectal cancer (CRC) is one of the most common cancers in men and women and a leading cause of cancer related death in the United States {}. 
-- Early detection and treatment are essential to increase survival rates, but for a variety of reasons including the invasiveness and high cost of screening (i.e.
-colonoscopy), many people do not comply with recommended screening guidelines {}.
-- Less invasive and more affordable screening methods (e.g.
-fecal immunochemical test) are available, however these tests are less sensitive than colonoscopies, especially for detecting early stage adenomas.
-- A growing body of research points to the gut microbiome as having a role in tumor development and progression {}.
-For example, studies find that Fusobacterium nucleatum and enterotoxigenic Bacteriodes fragilis tend to be enriched in the gut microbiome of subjects with CRC relative to healthy controls {}, while potentially protective bacteria such as members of Lachnospiraceae tend to be depleted {}.
-- The gut microbiome shows promise as diagnostic {}
-- Efforts to realize the diagnostic potential of the gut microbiome in detecting screen relevant neoplasias (SRNs) have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on the amplicon sequencing of the v4 region of the 16S rRNA gene {}.
+Colorectal cancer is one of the most common cancers in men and women and a leading cause of cancer related deaths in the United States {}.
+Early detection and treatment are essential to increase survival rates, but for a variety of reasons including the invasiveness and high cost of screening (i.e.
+colonoscopy), many people do not comply with recommended screening guidelines {} prompting a need for low cost, non-invasive detection methods.
+A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasias (neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e.
+advanced adenomas) and cancer.
+Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on the amplicon sequencing of the 16S rRNA gene {}.
 However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown.
-  - Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs.
+Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs {}.
 However, whether the additional resolution provided by ASVs is useful for ML classification is unclear.
-    - Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups. 
-- Topçuoğlu *et al* (mBio 2020) recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
-- This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.
+Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups.
+Topçuoğlu *et al* {mBio 2020} recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
+This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.
 
-## Results
+## Results (~700 words)
 
 ```{r median_AUC}
-table <- read_csv("../exploratory/figures/median_test_auc.csv")
+table <- read_csv("../exploratory/figures/median_test_auc.csv",
+                  col_types = cols(level=col_character(),
+                                   .default=col_double()))
 
@@
 ```{r input_table}
-input <- read_csv("../analysis/input_values.csv") %>% 
+input <- read_csv("../analysis/input_values.csv",
+                  col_types = cols(level=col_character(),
+                                   .default = col_double())) %>% 
   mutate(pct_kept = round(n_features_preproc/n_features,digits=3)*100)
@@
 
-- Across the five ML methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs.
-[(Figure)](#model-performance-across-taxonomy)
-- Random forest was consistently the top performer at most taxonomic levels.
-  - RF might be more appropriate for microbiome analysis since its more suitable for zero inflated data 
-  - TODO: dig into literature on this
-- Within the RF model, the highest AUCs were observed for family (median AUC: `r round(rf_med_auc["family"],digits=3)`), genus(median AUC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three.
-[(Figure)](#random-forest-model-performance-with-significance)
-  - ASV (median AUC: `r round(rf_med_auc["asv"],digits=3)`) performed significantly lower than OTU and genus (p < 0.05) but not family level P = 0.06).
-  - TODO: compare dada2 performance
-- One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g.
+```{r dada2}
+dada2_results <- read_csv("../data/dada2/process/combined-dada2-rf.csv",
+                          col_types = cols(method = col_character(),
+                                           .default = col_double()))
+dada2_med_auc <- dada2_results %>% 
+  pull(AUC) %>% 
+  median()
+
+```
+
+Utilizing publicly available 16S rRNA sequence data from patients with cancer and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels.
+Using the taxonomic abundance data, we quanitfied how accurately samples could be classified as "normal" or "SRN" (i.e.
+advanced adenoma or cancer) using machine learning (methods).
+Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1).
+Random forest (RF) was consistently the top performer at most taxonomic levels.
+[**TODO**: dig into literature on this: RF might be more appropriate for microbiome analysis since its more suitable for zero inflated data].
+Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1).
+Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU and genus (p < 0.05) but not family level P = 0.06).
+These results were consistent with ASVs generated with DADA2 {} (median AUROC: `r round(dada2_med_auc,digits=3)`).
+These results suggest that finer taxonomic resolution is not necessarily better for differentiating a persons cancer status from the microbiome.
+
+One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g.
 phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls.
-  - As you reach genus/OTU level data there is enough data and variation but at the ASV level, the data is too specific to individuals and doesn't overlap enough. 
-  - Examination of the prevalence of taxa in samples at each level supports this idea.
-A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels.
-The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples.
-[(Figure)](#prevalence-of-taxa-in-samples)
-- Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods).
-As part of this step, features are removed that wont provide useful information to build the model.
+At the level of genus/OTU data there is enough data and variation but at the ASV level, the data is too specific to individuals and doesn't overlap enough.
+Examination of the prevalence of taxa in samples at each level supports this idea.
+A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels).
+The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 2).
+**SUMMARY SENTANCE**
+
+|   **TODO: what are important taxa, patterns along levels (e.g fuso always in top )**
+
+Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods).
+As part of this step, features are removed that will not provide useful information to build the model.
 For example, strongly correlated features provide the same information to build the model and thus can be collapsed.
 Additionally, features with zero or near-zero variance will not help the model differentiate groups and thus can be removed.
-  - Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing.
-At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing.
-[(Table)](#summary-of-features)
-  - TODO: check why ASVs were removed in preprocessing (nzv?)
-  - (depending on above can discuss why features were removed - e.g. only in one or a few samples)
-- OTU level data provides an idea balance of overlap and distinction to classify samples. 
+Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing.
+At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).
+
+|   **TODO: check why ASVs were removed in preprocessing (nzv?)**
+|   (depending on answer can discuss why features were removed - e.g. only in one or a few samples)
+
+## Discussion/Conclusions (~250 words)
+
+These results demonstrate that consideration of the appropriate taxonomic resolution for utilizing the microbiome as a predictive tool.
+In general, we found that finer resolution (e.g.
+ASVs) does not add additional sensitivity to model predicition and at the ASV level actually impedes model performance due to the sparsity of shared taxa.
+Additionally, utilizing genus or family level data could allow for merging data generated from different 16S regions or sequencing platforms.
+
 - Overall, the fine resolution of ASVs is unnecessary for ML classification with microbiome data since the vast majority of the ASVs are removed during preprocessing and the model performance is diminshed compared to other taxonomic levels.
-
-other items that could be addressed:
-
-- what are important taxa, patterns along levels (e.g fuso always in top )
-- abundance of important taxa: many of the top important features are low abund [(Figure)](#relative-abundance-of-important-features)
-
-## Conclusions
+- since family/genus level is just as good as otu could we use data from different 16S regions?
 
 ## Materials and Methods
-- 16S rRNA data from 490 subjects {baxter}
-  - 261 controls
-  - 229 SRNs (describe how SRN defined)
-- processed with mothur v1.44.3
-  - SILVA v132 reference
-- ML with mikropml package
-  - preprocessing
-    -normalize values
-    - what is removed and why
-      - near zero variance & zero variance removed
-      - correlated collapsed
-  - hp tuning
-  - n seeds
-  - for more detail see mikropml package{}
-- pvalues calculated as previously described {begum}
-- prevalence = number of samples with non-zero abundance / total number of samples
+
+***Dataset.*** Raw 16S rRNA gene amplicon sequence data isolated from human gut samples {Baxter} was downloaded from NCBI SRA (accession #).
+This dataset contains stool samples from 490 subjects.
+Based on the available metadata, samples categorized as normal, high risk normal, or adenoma were labeled "normal" for this analysis and samples categorized as advanced adenoma or carcinoma were labeled as "screen relevant neoplasia" (SRN).
+This resulted in a total of 261 "normal" samples and 229 "SRN" samples.
+
+***Data processing.*** Sequence data was processed with Mothur (1.44.3) using the SILVA reference database (v132) to produce count tables for phylum, class, order, family, genus, OTU, and ASV following the Schloss Lab MiSeq SOP described on the Mothur website {}.
+ASV level data was also produced using DADA2 to ensure consistent results with a different pipeline.
+Data was processed following the DADA2 pipeline, but setting pool=TRUE to infer ASVs from the whole dataset rather than per sample.
+The resulting ASV table was subsampled for consistency with the Mothur data.
+
+***Machine Learning.*** Machine learning models were run with the R package mikropml (v0.0.2) {} to predict the diagnosis category (normal vs SRN) of each sample.
+Data was preprocessed to normalize values (scale/center?
+range?), remove values with zero or near-zero variance, and collapse collinear features using default parameters.
+Initially the models were run with default hyperparameters, but were expanded if the peak performance was at the edge of the hyperparameter range.
+Each taxonomic model taxonomic level combination (e.g.
+Random Forest on genus) was run with 100 different seeds (1-100).
+Each seed split the data into a training (80%) and testing (20%) set, and output performance of the training and testing as area under the receiver operating curve (AUROC).
+
+To compare performance between taxonomic levels and models, P values were calculated as previously described {begum}.
+Another output from the mikropml package is the permuted feature importance which is quantified by iteratively permuting each feature in the model and assessing the change in model performance.
+Features are presumed to be important if the performance of the model, measured by the AUROC, decreases when that feature is permuted.
+Ranking of feature importance was determined by ordering the features based on the change in AUROC where features with a larger decrease in AUROC are ranked higher in importance.
+
+To quantify prevalence of the features, the number of samples with non-zero abundance was divided by the total number of samples resulting in values ranging from 0 to 1 where 0 indicates the feature is not found in any samples, 0.5 indicates the feature is found in half of the samples, an 1 indicating the feature is found in all of the samples.
+
+All code is available at: **TODO: link to code**
 
@@
 
+\newpage
+
 ## Figures
 
-### Model Performance across Taxonomy  
+![](../exploratory/figures/rf_with_stats_sideways.png){width=75%}
+
+**Figure 1: Random Forest Model Performance with Significance.** Boxplots of AUROC values on the test dataset for 100 seeds predicting SRNs using a Random Forest model.
+Dashed line denotes AUROC of 0.5 which is equivalent to random classification.
+Differences in AUROC values between taxonomic levels are significant unless otherwise specified with "NS" (not significant)
+
+\newpage
+## Tables
+
+```{r feats}
+
+input %>%
+  select(-n_samples) %>%
+  mutate(pct_kept=paste0(pct_kept," %"),
+         level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV"))) %>%
+  knitr::kable(format="markdown",booktabs=T,escape=F,align="lrrc",
+               linesep="",
+               col.names = linebreak(c("Taxonomic Level",
+                                 "Number of Features",
+                                 "Number of Features \n After Preprocessing",
+                                 "Percent of Features Kept \n After Preprocessing"),align="c"))
+
+```
+
+**Table 1: Summary of Features.** Overview of the number of features at each taxonomic level before and after preprocessing as described in the methods.
+
+\newpage
+
+## Supplemental Figures
+
 ![](../exploratory/figures/all_model_level.png)
 
-### Random Forest Model Performance with Significance  
-![](../exploratory/figures/rf_with_stats.png){width=75%}
+**Supplemental Figure 1: Model Performance across Taxonomy.** Boxplot of AUROC values on the test dataset for 100 seeds for each model type across all taxonomic levels.
 
-### Prevalence of Taxa in Samples  
 ![](../exploratory/figures/bin_prevalence.png){height=50%}  
-
-### Summary of Features
-```{r feats}
-knitr::kable(input)
-```
-
-### Relative Abundance of Important Features
-![](../exploratory/figures/relabund_top10.png){height=95%}
+**Supplemental Figure 2: Prevalence of Taxa in Samples.** Distribution of taxa prevalence in samples at each taxonomic level.

### 2021-04-29 paper/manuscript.tex
ms updates

--- 2021-01-22 paper/manuscript.tex
@@
 \linenumbers
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
 \ifluatex
@@
 
-\hypertarget{optimal-resolution-for-microbiome-based-classification-of-colorectal-cancer}{%
-\section{Optimal Resolution for Microbiome-Based Classification of
-Colorectal
-Cancer}\label{optimal-resolution-for-microbiome-based-classification-of-colorectal-cancer}}
-
-\vspace{20mm}
-
-\textbf{Running title:} Optimal Resolution
+\hypertarget{taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}{%
+\section{Taxonomic Resolution Matters for Microbiome-Based
+Classification of Colorectal
+Cancer}\label{taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}}
 
@@
 
-Courtney R. Armour, (ande? begum?), Patrick D. Schloss \({^\dagger}\)
+\textbf{Other title options:}\\
+Optimal Taxonomic Resolution for Microbiome-Based Classification of
+Colorectal Cancer ASV level resolution is unnecessary for prediction of
+SRNs from microbiome data \ldots{}
+
+\vspace{10mm}
+
+Courtney R. Armour, Begüm D.Topçuoğlu, Andrea Garretto, Patrick D.
+Schloss \({^\dagger}\)
 
@@
 
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
+\hypertarget{introduction-250-words}{%
+\subsection{Introduction (\textasciitilde250
+words)}\label{introduction-250-words}}
+
+Colorectal cancer is one of the most common cancers in men and women and
+a leading cause of cancer related deaths in the United States \{\}.
+Early detection and treatment are essential to increase survival rates,
+but for a variety of reasons including the invasiveness and high cost of
+screening (i.e.~colonoscopy), many people do not comply with recommended
+screening guidelines \{\} prompting a need for low cost, non-invasive
+detection methods. A growing body of research points to the gut
+microbiome as a promising target for non-invasive screening to detect
+screen relevant neoplasias (neoplasms?) (SRNs) consisting of
+pre-cancerous polyps (i.e.~advanced adenomas) and cancer. Efforts to
+realize the diagnostic potential of the gut microbiome in detecting SRNs
+have focused on machine learning (ML) methods using abundances from
+operational taxonomic unit (OTU) classifications based on the amplicon
+sequencing of the 16S rRNA gene \{\}. However, whether this is the
+optimal taxonomic resolution for classifying SRNs from microbiome data
+is unknown. Additionally, recent work has pushed for the use of amplicon
+sequence variants (ASVs) to replace OTUs for marker-gene analysis
+because of the improved resolution with ASVs \{\}. However, whether the
+additional resolution provided by ASVs is useful for ML classification
+is unclear. Since ML classification relies on consistent differences
+between groups, its possible that the resolution at the ASV level is too
+individualized to accurately differentiate groups. Topçuoğlu \emph{et
+al} \{mBio 2020\} recently demonstrated effective application of machine
+learning (ML) to microbiome based classification problems and developed
+a framework for applying ML practices in a more reproducible way
+(mikropml). This analysis utilizes the reproducible framework developed
+by Topçuoğlu \emph{et al} to quantify which ML method and taxonomic
+level produce the best performing classifier for detecting SRNs from
+microbiome data.
+
+\hypertarget{results-700-words}{%
+\subsection{Results (\textasciitilde700
+words)}\label{results-700-words}}
+
+Utilizing publicly available 16S rRNA sequence data from patients with
+cancer and healthy controls, we generated abundance tables with Mothur
+\{\} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV
+levels. Using the taxonomic abundance data, we quanitfied how accurately
+samples could be classified as ``normal'' or ``SRN'' (i.e.~advanced
+adenoma or cancer) using machine learning (methods). Across the five
+machine learning methods tested, model performance tended to increase
+with taxonomic level usually peaking around genus/OTU level before
+dropping off slightly with ASVs (Supplemental Figure 1). Random forest
+(RF) was consistently the top performer at most taxonomic levels.
+{[}\textbf{TODO}: dig into literature on this: RF might be more
+appropriate for microbiome analysis since its more suitable for zero
+inflated data{]}. Within the RF model, the highest AUROCs were observed
+for family (median AUROC: 0.687), genus (median AUROC: 0.686), and OTU
+(median AUROC: 0.698) level data with no significant difference between
+the three (Figure 1). Performance with ASVs (median AUROC: 0.676) was
+significantly lower than OTU and genus (p \textless{} 0.05) but not
+family level P = 0.06). These results were consistent with ASVs
+generated with DADA2 \{\} (median AUROC: 0.66). These results suggest
+that finer taxonomic resolution is not necessarily better for
+differentiating a persons cancer status from the microbiome.
+
+One hypothesis for the observation that model performance increases from
+phylum to OTU level then drops slightly at ASV level is that at higher
+taxonomic levels (e.g.~phylum) there are too few taxa and too much
+overlap to reliably differentiate between cases and controls. At the
+level of genus/OTU data there is enough data and variation but at the
+ASV level, the data is too specific to individuals and doesn't overlap
+enough. Examination of the prevalence of taxa in samples at each level
+supports this idea. A majority of taxa are present in greater than 75\%
+of samples at the phylum (67\% of taxa) and class (63\% of taxa)
+levels). The opposite is observed at the OTU and ASV level where 60\%
+and 53\% of taxa respectively are only present in less than 25\% of
+samples (Supplemental Figure 2). \textbf{SUMMARY SENTANCE}
+
+~~\textbf{TODO: what are important taxa, patterns along levels (e.g fuso
+always in top )}
+
+Of note, the ML pipeline includes a pre-processing step that occurs
+prior to training and classifying the ML models (methods). As part of
+this step, features are removed that will not provide useful information
+to build the model. For example, strongly correlated features provide
+the same information to build the model and thus can be collapsed.
+Additionally, features with zero or near-zero variance will not help the
+model differentiate groups and thus can be removed. Interestingly,
+despite starting with 104106 features at the ASV level, only 478 (0.5\%)
+remained after pre-processing. At the OTU level, 20079 of the 705
+features (3.5\%) remained after preprocessing (Table 1).
+
+~~\textbf{TODO: check why ASVs were removed in preprocessing (nzv?)}\\
+\hspace*{0.333em}\hspace*{0.333em}(depending on answer can discuss why
+features were removed - e.g.~only in one or a few samples)
+
+\hypertarget{discussionconclusions-250-words}{%
+\subsection{Discussion/Conclusions (\textasciitilde250
+words)}\label{discussionconclusions-250-words}}
+
+These results demonstrate that consideration of the appropriate
+taxonomic resolution for utilizing the microbiome as a predictive tool.
+In general, we found that finer resolution (e.g.~ASVs) does not add
+additional sensitivity to model predicition and at the ASV level
+actually impedes model performance due to the sparsity of shared taxa.
+Additionally, utilizing genus or family level data could allow for
+merging data generated from different 16S regions or sequencing
+platforms.
 
@@
 \item
-  CRC is one of the most common cancers and a leading cause of cancer
-  related death
+  Overall, the fine resolution of ASVs is unnecessary for ML
+  classification with microbiome data since the vast majority of the
+  ASVs are removed during preprocessing and the model performance is
+  diminshed compared to other taxonomic levels.
 \item
-  There is evidence that the microbiome has a role in CRC
-  development/progression and could be useful for biomarker detection
-  and diagnostics.
-\item
-  Begum et al (mBio 2020) recently demonstrated effective application of
-  machine learning (ML) to microbiome based classification problems and
-  developed a framework for applying ML practices in a more reproducible
-  way (mikropml).
-\item
-  A common question when applying ML methods to microbiome data is which
-  method and taxonomic level is optimal.
-\item
-  This analysis utilizes the reproducible framework developed by Begum
-  et al to quantify which ML method and taxonomic level produce the best
-  performing classifier for CRC data.
+  since family/genus level is just as good as otu could we use data from
+  different 16S regions?
 \end{itemize}
-
-\newpage
-
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
-
-\begin{itemize}
-\tightlist
-\item
-  Across the five ML methods tested, model performance tends to increase
-  with taxonomic level usually peaking around genus/otu level and
-  dropping off slightly with ASVs.
-\item
-  Random forest was consistently the top performer at most taxonomic
-  levels.
-
-  \begin{itemize}
-  \tightlist
-  \item
-    RF might be more appropriate for microbiome analysis since its more
-    suitable for zero inflated data (need to look into literature)
-  \end{itemize}
-\item
-  Within the RF model, the highest AUCs were observed for family, genus,
-  and otu level data with no significant difference between the three
-  (Figure 1). ASV performed significantly lower than OTU but not family
-  and genus levels.
-\end{itemize}
-
-\newpage
-
-\hypertarget{conclusion}{%
-\subsection{Conclusion}\label{conclusion}}
-
-\newpage
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  data from prior study \{baxter\}
-\item
-  mikropml package
-\item
-  pvalues as previously described \{begum\}
-\end{itemize}
+\textbf{\emph{Dataset.}} Raw 16S rRNA gene amplicon sequence data
+isolated from human gut samples \{Baxter\} was downloaded from NCBI SRA
+(accession \#). This dataset contains stool samples from 490 subjects.
+Based on the available metadata, samples categorized as normal, high
+risk normal, or adenoma were labeled ``normal'' for this analysis and
+samples categorized as advanced adenoma or carcinoma were labeled as
+``screen relevant neoplasia'' (SRN). This resulted in a total of 261
+``normal'' samples and 229 ``SRN'' samples.
 
-\newpage
+\textbf{\emph{Data processing.}} Sequence data was processed with Mothur
+(1.44.3) using the SILVA reference database (v132) to produce count
+tables for phylum, class, order, family, genus, OTU, and ASV following
+the Schloss Lab MiSeq SOP described on the Mothur website \{\}. ASV
+level data was also produced using DADA2 to ensure consistent results
+with a different pipeline. Data was processed following the DADA2
+pipeline, but setting pool=TRUE to infer ASVs from the whole dataset
+rather than per sample. The resulting ASV table was subsampled for
+consistency with the Mothur data.
+
+\textbf{\emph{Machine Learning.}} Machine learning models were run with
+the R package mikropml (v0.0.2) \{\} to predict the diagnosis category
+(normal vs SRN) of each sample. Data was preprocessed to normalize
+values (scale/center? range?), remove values with zero or near-zero
+variance, and collapse collinear features using default parameters.
+Initially the models were run with default hyperparameters, but were
+expanded if the peak performance was at the edge of the hyperparameter
+range. Each taxonomic model taxonomic level combination (e.g.~Random
+Forest on genus) was run with 100 different seeds (1-100). Each seed
+split the data into a training (80\%) and testing (20\%) set, and output
+performance of the training and testing as area under the receiver
+operating curve (AUROC).
+
+To compare performance between taxonomic levels and models, P values
+were calculated as previously described \{begum\}. Another output from
+the mikropml package is the permuted feature importance which is
+quantified by iteratively permuting each feature in the model and
+assessing the change in model performance. Features are presumed to be
+important if the performance of the model, measured by the AUROC,
+decreases when that feature is permuted. Ranking of feature importance
+was determined by ordering the features based on the change in AUROC
+where features with a larger decrease in AUROC are ranked higher in
+importance.
+
+To quantify prevalence of the features, the number of samples with
+non-zero abundance was divided by the total number of samples resulting
+in values ranging from 0 to 1 where 0 indicates the feature is not found
+in any samples, 0.5 indicates the feature is found in half of the
+samples, an 1 indicating the feature is found in all of the samples.
+
+All code is available at: \textbf{TODO: link to code}
 
@@
 
+\includegraphics[width=0.75\textwidth,height=\textheight]{../exploratory/figures/rf_with_stats_sideways.png}
+
+\textbf{Figure 1: Random Forest Model Performance with Significance.}
+Boxplots of AUROC values on the test dataset for 100 seeds predicting
+SRNs using a Random Forest model. Dashed line denotes AUROC of 0.5 which
+is equivalent to random classification. Differences in AUROC values
+between taxonomic levels are significant unless otherwise specified with
+``NS'' (not significant)
+
+\newpage
+
+\hypertarget{tables}{%
+\subsection{Tables}\label{tables}}
+
+\begin{longtable}[]{@{}lrrc@{}}
+\toprule
+Taxonomic Level & Number of Features &
+\makecell[c]{Number of Features \\ After Preprocessing} &
+\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\ \addlinespace
+\midrule
+\endhead
+Phylum & 19 & 9 & 47.4 \% \\ \addlinespace
+Class & 36 & 19 & 52.8 \% \\ \addlinespace
+Order & 65 & 28 & 43.1 \% \\ \addlinespace
+Family & 124 & 54 & 43.5 \% \\ \addlinespace
+Genus & 316 & 115 & 36.4 \% \\ \addlinespace
+OTU & 20079 & 705 & 3.5 \% \\ \addlinespace
+ASV & 104106 & 478 & 0.5 \% \\ \addlinespace
+\bottomrule
+\end{longtable}
+
+\textbf{Table 1: Summary of Features.} Overview of the number of
+features at each taxonomic level before and after preprocessing as
+described in the methods.
+
+\newpage
+
+\hypertarget{supplemental-figures}{%
+\subsection{Supplemental Figures}\label{supplemental-figures}}
+
+\includegraphics{../exploratory/figures/all_model_level.png}
+
+\textbf{Supplemental Figure 1: Model Performance across Taxonomy.}
+Boxplot of AUROC values on the test dataset for 100 seeds for each model
+type across all taxonomic levels.
+
+\includegraphics[width=\textwidth,height=0.5\textheight]{../exploratory/figures/bin_prevalence.png}\\
+\textbf{Supplemental Figure 2: Prevalence of Taxa in Samples.}
+Distribution of taxa prevalence in samples at each taxonomic level.
+
 \end{document}

### 2021-05-08 exploratory/exploratory.Rmd
ms updates

@@
 ---
+
 updated: `r Sys.Date()`
+
 ```{r setup_environment, include=FALSE}
@@
 ## Table of Contents
+
 -   [Input data overview](#input-data-overview)
+
 -   [Hyperparameter Performance](#hyperparameter-performance)
+
 -   [Model Performance](#model-performance)
+
     -   [Random Forest](#random-forest-model-performance-with-significance)
+
 -   [DADA2 Comparison](#dada2-comparison)
+
 -   [Feature Importance](#feature-importance)
 
-## Input data overview
+-   [Prevalence](#prevalence)
+
+## Input data overview {#input-data-overview}
 
 To total number of samples and features before and after preprocessing for each taxnomic level.
+
 ```{r input}
@@
 
-## Hyperparameter Performance
+## Hyperparameter Performance {#hyperparameter-performance}
 
@@
 ### Random Forest
+
 The default hyperparameter (mtry) selection for RF is:
@@
 1.  sqrt_features / 2
-1. sqrt_features
-1. sqrt_features * 2
+2.  sqrt_features
+3.  sqrt_features \* 2
 
@@
 
-Two metrics: alpha and lambda  
-By default alpha is set to zero (for L2 regularization) and lambda values are: 1e-04, 1e-03, 1e-02, 1e-01, 1e+00, 1e+01
-I added 1e-05 as a lambda value for phylum and class, and higher values of lambda for the rest
+Two metrics: alpha and lambda\
+By default alpha is set to zero (for L2 regularization) and lambda values are: 1e-04, 1e-03, 1e-02, 1e-01, 1e+00, 1e+01 I added 1e-05 as a lambda value for phylum and class, and higher values of lambda for the rest
 
@@
 
-## Model Performance
+## Model Performance {#model-performance}
 
@@
 
-### DADA2 Comparison
+### DADA2 Comparison {#dada2-comparison}
 
@@
 
-### Random Forest Model Performance with Significance
+### Random Forest Model Performance with Significance {#random-forest-model-performance-with-significance}
 
@@
 
-<div class = "row">
-<div class = "col-md-6">
+::: {.row}
+::: {.col-md-6}
 ```{r rf_table,echo=FALSE}
@@
 ```
-
-</div>
-
-<div class = "col-md-6">
+:::
+
+::: {.col-md-6}
 ```{r rf_plot,echo=F,message=F}
+m <- "Random Forest"
+
+sub_df <- auc_df %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                       labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(method==m & auc_type=="test_AUC")
+
+# sub_stats <- p.table.model %>% 
+#    mutate(level1=factor(level1,levels=c("phylum","class","order","family","genus","otu","asv"),
+#                             labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+#           level2=factor(level2,levels=c("phylum","class","order","family","genus","otu","asv"),
+#                             labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+#          model=factor(model,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+#                        labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+#   filter(model==m) %>% 
+#   select(-model) %>% 
+#   pivot_wider(names_from=level2,values_from=p_value) %>% 
+#   column_to_rownames(var="level1")
+# sub_stats <- sub_stats[levels_names[1:6],levels_names[2:7]]
+
+# sub_stats <- sub_stats %>% 
+#   mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
+#   rename(group1=level1,group2=level2)
+
+ggplot(sub_df,aes(x=factor(level,levels=c("ASV", "OTU","Genus","Family","Order","Class", "Phylum")), y=AUC)) + 
+#ggplot(sub_df,aes(x=level, y=AUC)) + 
+  geom_jitter(aes(fill=method,color=method),height = 0, width = 0.2,shape=21) +
+  stat_summary(fun.data = median_hilow,fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+  scale_fill_manual(values = alpha(pal2[m],0.6)) +
+  scale_color_manual(values = pal2[m]) +
+  scale_y_continuous(limits=c(0.45,0.87),breaks = seq(0.5,0.8,0.1),expand=c(0,0)) +
+  theme_bw() + xlab("") + ylab("AUROC") +
+  theme(legend.position ="none",
+        #axis.text.x = element_text(angle = 45,hjust = 1),
+        panel.grid.major.y = element_blank(),
+        axis.text = element_text(size=14),
+        axis.title = element_text(size=14)) +
+  coord_flip() +
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+  annotate("text",x=levels_names,y=rep(0.85,length(levels_names)),label=c("A","B","C","D,E","D","D","E"))
+ggsave("./figures/rf_with_stats.png",width = 5,height=4,units="in")
+
+```
+:::
+:::
+
+```{r rf_wstat_options,include=F}
+
+# OPTION 1 ------------
 m <- "Random Forest"
@@
   geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
-  geom_boxplot(aes(fill=method)) +
+  geom_boxplot(aes(fill=method,alpha=0.8)) +
   scale_fill_manual(values = pal2[m]) + 
-  scale_y_continuous(limits=c(0.2,1.05)) +
+  scale_y_continuous(limits=c(0.4,1.05)) +
   theme_bw() + xlab("") + ylab("AUROC") +
@@
         axis.text = element_text(size=14)) +
-  ggtitle(m) +
+  #ggtitle(m) +
   #stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
   stat_pvalue_manual(sub_stats,label="sig")
-ggsave("./figures/rf_with_stats.png",width = 5,height=4,units="in")
-
-```
-</div>
-</div>
-
-```{r rf_wstat_side,include=F}
-sub_stats <- sub_stats %>% 
-  #mutate(y.position=seq.int(0.85,.95,length.out = nrow(sub_stats)))
-  mutate(y.position=c(0.85,0.95,0.9,.95))
+
+# OPTION 2 ------------
 
@@
   #geom_jitter(aes(color=method,width=0,height=0.2)) +
-  geom_boxplot(aes(fill=method,alpha=0.8)) +
-  scale_fill_manual(values = pal2[m]) +
+  geom_boxplot(aes(fill=level,alpha=0.8)) +
+  #scale_fill_manual(values = pal2[m]) +
   #scale_color_manual(values = pal2[m]) +
@@
         axis.text = element_text(size=14)) +
-  ggtitle(m) +
+  #ggtitle(m) +
   stat_pvalue_manual(sub_stats,label="sig",coord.flip = T,angle = 0,hjust=0) +
@@
   geom_hline(yintercept = 0.5,color="grey",lty="dashed")
-ggsave("./figures/rf_with_stats_sideways.png",width = 5,height=4,units="in")
+
+# OPTION 3 ------------
+
+sub_df <- auc_df %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                       labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(method==m & auc_type=="test_AUC")
+
+sub_stats <- p.table.model %>% 
+   mutate(level1=factor(level1,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+          level2=factor(level2,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         model=factor(model,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(model==m) %>% 
+  mutate(sig=case_when(p_value >= 0.05 ~ "NS",
+                       TRUE ~ "*")) %>% 
+  filter(sig=="NS")
+
+sub_stats <- sub_stats %>% 
+  #mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
+  mutate(y.position=c(0.85,0.95,0.9,.95)) %>%
+  rename(group1=level1,group2=level2)
+
+ggplot(sub_df,aes(x=level,y=AUC)) + 
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+  #geom_boxplot(aes(fill=method)) +
+  geom_jitter(aes(fill=method),height = 0, width = 0.2,alpha=0.8,shape=21) +
+  stat_summary(fun.data = median_hilow,fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+  scale_fill_manual(values = pal2[m]) + 
+  scale_y_continuous(limits=c(0.4,1),breaks = seq(0.4,1,0.1)) +
+  theme_bw() + xlab("") + ylab("AUROC") +
+  theme(legend.position ="none",
+        #axis.text.x = element_text(angle = 45,hjust = 1),
+        panel.grid.major.x = element_blank(),
+        axis.text = element_text(size=14)) +
+  #ggtitle(m) +
+  #stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
+  stat_pvalue_manual(sub_stats,label="sig")
+
+# OPTION 4 ------------
+
+sub_stats <- sub_stats %>% 
+  #mutate(y.position=seq.int(0.85,.95,length.out = nrow(sub_stats)))
+  mutate(y.position=c(0.85,0.95,0.9,.95))
+
+ggplot(sub_df,aes(x=factor(level,levels=c("ASV", "OTU","Genus","Family","Order","Class", "Phylum")),
+                           y=AUC)) + 
+  geom_jitter(aes(fill=level),height = 0, width = 0.2,alpha=0.8,shape=21) +
+  stat_summary(fun.data = median_hilow,fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+  #scale_fill_manual(values = pal2[m]) +
+  #scale_color_manual(values = pal2[m]) +
+  #scale_x_continuous(limits=c(0.4,1)) +
+  theme_bw() + xlab("") + ylab("AUROC") +
+  theme(legend.position ="none",
+        #axis.text.x = element_text(angle = 45,hjust = 1),
+        panel.grid.major.x = element_blank(),
+        axis.text = element_text(size=14)) +
+  scale_y_continuous(position = "right") +
+  #ggtitle(m) +
+  stat_pvalue_manual(sub_stats,label="sig",coord.flip = T,angle = 0,hjust=0) +
+  coord_flip() +
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed")
+
+# OPTION 5 ------------
+
+sub_df <- auc_df %>% 
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                       labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(method==m & auc_type=="test_AUC")
+
+sub_stats <- p.table.model %>% 
+   mutate(level1=factor(level1,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+          level2=factor(level2,levels=c("phylum","class","order","family","genus","otu","asv"),
+                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         model=factor(model,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
+                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
+  filter(model==m) %>% 
+  mutate(sig=case_when(p_value >= 0.05 ~ "NS",
+                       TRUE ~ "*")) %>% 
+  filter(sig!="NS") %>% 
+  mutate(p_value_mod=case_when(p_value < 0.0001 ~ round(p_value,digits=5),
+                               p_value < 0.001 ~ round(p_value,digits=4),
+                               p_value < 0.01 ~ round(p_value,digits=3),
+                               TRUE ~ round(p_value,digits=2)))
+
+sub_stats <- sub_stats %>% 
+  #mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
+  mutate(y.position=c(1,1,1,1,1,
+                      0.9,
+                      0.95,0.95,0.95,0.95,
+                      0.85,
+                      1.05,1.05,1.05,1.05,1.05,1.05)) %>% 
+  #mutate(y.position=c(0.85,0.95,0.9,.95)) %>%
+  rename(group1=level1,group2=level2)
+
+ggplot(sub_df,aes(x=level,y=AUC)) + 
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+  #geom_boxplot(aes(fill=method)) +
+  geom_jitter(aes(fill=method),height = 0, width = 0.2,alpha=0.8,shape=21) +
+  stat_summary(fun.data = median_hilow,fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+  scale_fill_manual(values = pal2[m]) + 
+  scale_y_continuous(limits=c(0.4,1.05),breaks = seq(0.4,1.05,0.1)) +
+  theme_bw() + xlab("") + ylab("AUROC") +
+  theme(legend.position ="none",
+        #axis.text.x = element_text(angle = 45,hjust = 1),
+        panel.grid.major.x = element_blank(),
+        axis.text = element_text(size=14)) +
+  #ggtitle(m) +
+  #stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
+  #stat_pvalue_manual(sub_stats,label="p_value_mod") 
+  stat_pvalue_manual(sub_stats,label="sig")
 ```
@@
 
-<div class = "row">
-<div class = "col-md-6">
+::: {.row}
+::: {.col-md-6}
 ```{r glmnet_table,echo=FALSE}
@@
 ```
-</div>
-
-<div class = "col-md-6">
+:::
+
+::: {.col-md-6}
 ```{r glmnet_plot,echo=F,message=F}
@@
 ```
-</div>
-</div>
+:::
+:::
 
@@
 
-<div class = "row">
-<div class = "col-md-6">
+::: {.row}
+::: {.col-md-6}
 ```{r rpart2_table,echo=FALSE}
@@
 ```
-</div>
-
-<div class = "col-md-6">
+:::
+
+::: {.col-md-6}
 ```{r rpart2_plot,echo=F,message=F}
@@
 
-<div class = "row">
-<div class = "col-md-6">
+::: {.row}
+::: {.col-md-6}
 ```{r svmRadial_table,echo=FALSE}
@@
 ```
-</div>
-
-<div class = "col-md-6">
+:::
+
+::: {.col-md-6}
 ```{r svmRadial_plot,echo=F,message=F}
@@
 
-<div class = "row">
-<div class = "col-md-6">
+::: {.row}
+::: {.col-md-6}
 ```{r xgbTree_table,echo=FALSE}
@@
 ```
-</div>
-
-<div class = "col-md-6">
+:::
+
+::: {.col-md-6}
 ```{r xgbTree_plot,echo=F,message=F}
@@
 ```
-</div>
-</div>
-
-## Feature Importance
+:::
+:::
+
+## Feature Importance {#feature-importance}
+
 The mikropml packages includes an option for finding [feature importance](http://www.schlosslab.org/mikropml/reference/get_feature_importance.html) using a permutation method.
@@
 ### Top 10 Important features
+
 ```{r importance_rank, message=F,echo=F}
@@
   filter(rank <= 10) %>% 
+  mutate(Taxonomy=case_when(level=="otu" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
+                            level=="asv" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
+                            TRUE ~ Taxonomy)) %>% 
+  mutate(Taxonomy=sapply(strsplit(Taxonomy,"\\("), `[`, 1)) %>% 
+  mutate(Taxonomy=paste0(Taxonomy," (",OTU,")"))  %>% 
   select(level,Taxonomy,rank) %>% 
   pivot_wider(names_from = level,values_from = Taxonomy) 
+colnames(top_10_tax) <- c("Rank","Phylum","Class","Order","Family","Genus","OTU","ASV")
+
+ranked_tax %>% 
+  filter(rank <= 10) %>% 
+  mutate(Taxonomy=case_when(level=="otu" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
+                            level=="asv" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
+                            TRUE ~ Taxonomy)) %>% 
+  mutate(Taxonomy=sapply(strsplit(Taxonomy,"\\("), `[`, 1)) %>% 
+  mutate(Taxonomy=paste0(Taxonomy," (",OTU,")"))
 
@@
       col.names = c("Rank",levels_names)) %>% 
-   kable_material(c("striped", "hover","condensed"))
-```
-
-## Prevalence
-
-```{r prevalence_preprocess,echo=F,message=F}
-get_prevalence_preprocess <- function(file,file_proc,level){
-  data <- as.data.frame(fread(file))
-  data_proc <- as.data.frame(fread(file_proc))
-  keep_names <- str_replace(colnames(data_proc)[-1],"_1","")
-  n_samp <- nrow(data_proc)
-  prev <- select(data,keep_names) %>% 
-    pivot_longer(everything(),names_to="name",values_to="count") %>% 
-    filter(count > 0) %>% 
-    group_by(name) %>% 
-    tally() %>% 
-    mutate(prevalence=n/n_samp) %>% 
-    mutate(level := !!level) %>% 
-    select(level,name,prevalence)
-  return(prev)
-}
-
-# prev_df <- get_prevalence_preprocess("./data/final/phylum/input_data.csv",
-#                                      "./data/final/phylum/input_data_preproc.csv",
-#                                      "phylum") %>%
-#   bind_rows(get_prevalence_preprocess("./data/final/class/input_data.csv",
-#                                       "./data/final/class/input_data_preproc.csv",
-#                                       "class")) %>%
-#   bind_rows(get_prevalence_preprocess("./data/final/order/input_data.csv",
-#                                       "./data/final/order/input_data_preproc.csv",
-#                                       "order")) %>%
-#   bind_rows(get_prevalence_preprocess("./data/final/family/input_data.csv",
-#                                       "./data/final/family/input_data_preproc.csv",
-#                                       "family")) %>%
-#   bind_rows(get_prevalence_preprocess("./data/final/genus/input_data.csv",
-#                                       "./data/final/genus/input_data_preproc.csv",
-#                                       "genus")) %>%
-#   bind_rows(get_prevalence_preprocess("./data/final/otu/input_data.csv",
-#                                       "./data/final/otu/input_data_preproc.csv",
-#                                       "otu")) %>%
-#   bind_rows(get_prevalence_preprocess("./data/final/asv/input_data.csv",
-#                                       "./data/final/asv/input_data_preproc.csv",
-#                                       "asv"))
-# write_csv(prev_df,"./data/analysis/exploratory3/prevalence_by_level_preprocess.csv")
-
-prevalence_prepro <- read_csv("./data/analysis/exploratory3/prevalence_by_level_preprocess.csv")
-
-prevalence_prepro$level <- factor(prevalence_prepro$level,levels=levels)
-
-ggplot(prevalence_prepro,aes(x=level,y=prevalence,fill=level)) + 
-  #geom_jitter() +
-  geom_boxplot() +
-  stat_summary(fun = mean, color="red",geom = "point", shape = 18, size = 4,show.legend = FALSE) +
-  theme_bw() + xlab("") +
-  theme(legend.position = "none",
-        axis.text = element_text(size=14),
-        axis.title.y = element_text(size=14))
-ggsave("./data/analysis/exploratory3/prevalence_preprocess.png",width=5,height=4,units = "in")
-
-ggplot(prevalence_prepro,aes(x=prevalence,fill=level)) +
-  geom_histogram(binwidth = 0.05) +
-  facet_wrap(.~level,scales = "free") +
-  theme_bw() +
-  scale_y_continuous(expand=c(0,0)) +
-  scale_x_continuous(expand=c(0,0)) +
-  theme(legend.position = "none") 
-ggsave("./data/analysis/exploratory3/prevalence_histogram.png",width=5,height=4,units = "in")
-
-ggplot(prevalence_prepro,)
-
-tax_count <- prevalence_prepro %>% 
+   kable_material(c("hover","condensed"))
+
+### EXAMPLE plot
+library(glue)
+library(ggtext)
+
+data <- tibble(bactname = c("Staphylococcaceae", "Moraxella", "Streptococcus", "Acinetobacter"),
+               OTUname = c("OTU 1", "OTU 2", "OTU 3", "OTU 4"),
+               value = c(-0.5, 0.5, 2, 3))
+
+data %>% mutate(
+  name = glue("<i>{bactname}</i> ({OTUname})"),
+  name = fct_reorder(name, value))  %>%
+  ggplot(aes(value, name)) + 
+    geom_col(alpha = 0.5) + 
+    scale_fill_identity() +
+    theme(
+      axis.text.y = element_markdown(),
+      plot.caption = element_markdown(lineheight = 1.2))
+
+```
+
+## Prevalence {#prevalence}
+
+```{r prevalence,echo=F,message=F}
+
+prevalence <- read_csv("../analysis/prevalence_by_level.csv")
+
+prevalence$level <- factor(prevalence$level,levels=levels)
+
+tax_count <- prevalence %>% 
+  filter(kept==1) %>% 
   group_by(level) %>% 
@@
 
-# prevalence_prepro %>% 
-#   group_by(level) %>% 
-#   count(prevalence) %>% 
-#   inner_join(tax_count,by="level") %>% 
-#   mutate(pct_OTU = n/numOTU) %>% 
-#   mutate(prevalence = round(prevalence,digits=2)) %>% 
-#   ggplot(aes(x=prevalence,y=pct_OTU)) +
-#     geom_line() +
-#     facet_wrap(~level,scales = "free")
-  #ggplot(aes(x=prevalence,y=n)) +
-  #  geom_bar(stat="identity") +
-  #  facet_grid(~level)
-
 pct_bins <- NULL
@@
 for(l in levels){
-  sub_data <- prevalence_prepro %>% 
-    filter(level==l)
-  hist <- hist(sub_data$prevalence,breaks = seq(0,1,0.25))
+  sub_data <- prevalence %>% 
+    filter(level==l,kept==1)
+  hist <- hist(sub_data$prevalence,breaks = seq(0,1,0.25),plot=FALSE)
 
@@
     theme_bw() +
-    theme(legend.position = "none") +
+    theme(legend.position = "none",
+          axis.text = element_text(size=12),
+          axis.title = element_text(size=14)) +
   geom_text(aes(label = paste0((round(pct,digits=2)*100),"%")),nudge_y=0.05)
-ggsave("./data/analysis/exploratory3/bin_prevalence.png")
-
-# 
-# ggplot(prevalence_prepro,aes(x=prevalence,fill=level)) +
-#   geom_histogram(binwidth = 0.25) +
-#   facet_wrap(.~level,scales = "free") +
-#   theme_bw() +
-#   scale_y_continuous(expand=c(0,0)) +
-#   scale_x_continuous(expand=c(0,0)) +
-#   theme(legend.position = "none")
-
-```
+ggsave("./figures/bin_prevalence.png")
+```

### 2021-05-08 paper/manuscript.Rmd
ms updates

@@
 ---
+bibliography: references.bib
+link-citations: true
 output:
@@
 
-**Other title options:**  
-Optimal Taxonomic Resolution for Microbiome-Based Classification of Colorectal Cancer
-ASV level resolution is unnecessary for prediction of SRNs from microbiome data 
+**Other title options:**\
+Optimal Taxonomic Resolution for Microbiome-Based Classification of Colorectal Cancer\
 ...
@@
 
-## Introduction (~250 words)
+## Introduction (\~250 words)
 
@@
 A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasias (neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e.
-advanced adenomas) and cancer.
-Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on the amplicon sequencing of the 16S rRNA gene {}.
+advanced adenomas) and carcinomas.
+Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on amplicon sequencing of the 16S rRNA gene {}.
 However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown.
@@
 Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups.
-Topçuoğlu *et al* {mBio 2020} recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
+Topçuoğlu *et al* [@topçuolu2020] recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
 This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.
 
-## Results (~700 words)
+## Results (\~700 words)
 
@@
 Using the taxonomic abundance data, we quanitfied how accurately samples could be classified as "normal" or "SRN" (i.e.
-advanced adenoma or cancer) using machine learning (methods).
+advanced adenoma or cancer) using five machine learning methods with the Mikropml R package (methods).
 Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1).
 Random forest (RF) was consistently the top performer at most taxonomic levels.
-[**TODO**: dig into literature on this: RF might be more appropriate for microbiome analysis since its more suitable for zero inflated data].
 Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1).
-Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU and genus (p < 0.05) but not family level P = 0.06).
-These results were consistent with ASVs generated with DADA2 {} (median AUROC: `r round(dada2_med_auc,digits=3)`).
+Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU and genus (p \< 0.05) but not family level (p = 0.06).
+These results were consistent for ASVs generated with DADA2 {} (median AUROC: `r round(dada2_med_auc,digits=3)`).
 These results suggest that finer taxonomic resolution is not necessarily better for differentiating a persons cancer status from the microbiome.
@@
 The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 2).
-**SUMMARY SENTANCE**
+While the resolution provided by ASVs is useful in
 
@@
 
-Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods).
+**is this paragraph useful???** *Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods).
 As part of this step, features are removed that will not provide useful information to build the model.
@@
 Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing.
-At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).
+At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).*
 
@@
 
-## Discussion/Conclusions (~250 words)
+## Discussion/Conclusions (\~250 words)
 
@@
 
-***Dataset.*** Raw 16S rRNA gene amplicon sequence data isolated from human gut samples {Baxter} was downloaded from NCBI SRA (accession #).
+***Dataset.*** Raw 16S rRNA gene amplicon sequence data isolated from human gut samples {Baxter} was downloaded from NCBI SRA (accession \#).
 This dataset contains stool samples from 490 subjects.
@@
 ***Machine Learning.*** Machine learning models were run with the R package mikropml (v0.0.2) {} to predict the diagnosis category (normal vs SRN) of each sample.
-Data was preprocessed to normalize values (scale/center?
-range?), remove values with zero or near-zero variance, and collapse collinear features using default parameters.
+Data was preprocessed to normalize values (scale/center), remove values with zero or near-zero variance, and collapse collinear features using default parameters.
 Initially the models were run with default hyperparameters, but were expanded if the peak performance was at the edge of the hyperparameter range.
@@
 
-![](../exploratory/figures/rf_with_stats_sideways.png){width=75%}
+![](../exploratory/figures/rf_with_stats.png){width="75%"}
 
@@
 Dashed line denotes AUROC of 0.5 which is equivalent to random classification.
-Differences in AUROC values between taxonomic levels are significant unless otherwise specified with "NS" (not significant)
+Significance between taxonomic levels was quantified by comparing the difference in mean AUROC and is denoted by letters A through E on the right side of the plot; taxonomic levels with the same letter are in the same significance group and are not significantly different from one another.
 
 \newpage
+
 ## Tables
@@
 
-![](../exploratory/figures/bin_prevalence.png){height=50%}  
+![](../exploratory/figures/bin_prevalence.png){height="50%"}\
 **Supplemental Figure 2: Prevalence of Taxa in Samples.** Distribution of taxa prevalence in samples at each taxonomic level.
+
+## References

### 2021-05-08 paper/manuscript.tex
ms updates

@@
 \fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 Optimal Taxonomic Resolution for Microbiome-Based Classification of
-Colorectal Cancer ASV level resolution is unnecessary for prediction of
-SRNs from microbiome data \ldots{}
+Colorectal Cancer\\
+\ldots{}
 
@@
 screen relevant neoplasias (neoplasms?) (SRNs) consisting of
-pre-cancerous polyps (i.e.~advanced adenomas) and cancer. Efforts to
+pre-cancerous polyps (i.e.~advanced adenomas) and carcinomas. Efforts to
 realize the diagnostic potential of the gut microbiome in detecting SRNs
 have focused on machine learning (ML) methods using abundances from
-operational taxonomic unit (OTU) classifications based on the amplicon
+operational taxonomic unit (OTU) classifications based on amplicon
 sequencing of the 16S rRNA gene \{\}. However, whether this is the
@@
 individualized to accurately differentiate groups. Topçuoğlu \emph{et
-al} \{mBio 2020\} recently demonstrated effective application of machine
-learning (ML) to microbiome based classification problems and developed
-a framework for applying ML practices in a more reproducible way
-(mikropml). This analysis utilizes the reproducible framework developed
-by Topçuoğlu \emph{et al} to quantify which ML method and taxonomic
-level produce the best performing classifier for detecting SRNs from
-microbiome data.
+al} (\protect\hyperlink{ref-topuxe7uolu2020}{Topçuoğlu et al. 2020})
+recently demonstrated effective application of machine learning (ML) to
+microbiome based classification problems and developed a framework for
+applying ML practices in a more reproducible way (mikropml). This
+analysis utilizes the reproducible framework developed by Topçuoğlu
+\emph{et al} to quantify which ML method and taxonomic level produce the
+best performing classifier for detecting SRNs from microbiome data.
 
@@
 samples could be classified as ``normal'' or ``SRN'' (i.e.~advanced
-adenoma or cancer) using machine learning (methods). Across the five
-machine learning methods tested, model performance tended to increase
-with taxonomic level usually peaking around genus/OTU level before
-dropping off slightly with ASVs (Supplemental Figure 1). Random forest
-(RF) was consistently the top performer at most taxonomic levels.
-{[}\textbf{TODO}: dig into literature on this: RF might be more
-appropriate for microbiome analysis since its more suitable for zero
-inflated data{]}. Within the RF model, the highest AUROCs were observed
-for family (median AUROC: 0.687), genus (median AUROC: 0.686), and OTU
-(median AUROC: 0.698) level data with no significant difference between
-the three (Figure 1). Performance with ASVs (median AUROC: 0.676) was
-significantly lower than OTU and genus (p \textless{} 0.05) but not
-family level P = 0.06). These results were consistent with ASVs
-generated with DADA2 \{\} (median AUROC: 0.66). These results suggest
-that finer taxonomic resolution is not necessarily better for
-differentiating a persons cancer status from the microbiome.
+adenoma or cancer) using five machine learning methods with the Mikropml
+R package (methods). Across the five machine learning methods tested,
+model performance tended to increase with taxonomic level usually
+peaking around genus/OTU level before dropping off slightly with ASVs
+(Supplemental Figure 1). Random forest (RF) was consistently the top
+performer at most taxonomic levels. Within the RF model, the highest
+AUROCs were observed for family (median AUROC: 0.687), genus (median
+AUROC: 0.686), and OTU (median AUROC: 0.698) level data with no
+significant difference between the three (Figure 1). Performance with
+ASVs (median AUROC: 0.676) was significantly lower than OTU and genus (p
+\textless{} 0.05) but not family level (p = 0.06). These results were
+consistent for ASVs generated with DADA2 \{\} (median AUROC: 0.66).
+These results suggest that finer taxonomic resolution is not necessarily
+better for differentiating a persons cancer status from the microbiome.
 
@@
 and 53\% of taxa respectively are only present in less than 25\% of
-samples (Supplemental Figure 2). \textbf{SUMMARY SENTANCE}
+samples (Supplemental Figure 2). While the resolution provided by ASVs
+is useful in
 
@@
 
-Of note, the ML pipeline includes a pre-processing step that occurs
-prior to training and classifying the ML models (methods). As part of
-this step, features are removed that will not provide useful information
-to build the model. For example, strongly correlated features provide
-the same information to build the model and thus can be collapsed.
-Additionally, features with zero or near-zero variance will not help the
-model differentiate groups and thus can be removed. Interestingly,
-despite starting with 104106 features at the ASV level, only 478 (0.5\%)
-remained after pre-processing. At the OTU level, 20079 of the 705
-features (3.5\%) remained after preprocessing (Table 1).
+\textbf{is this paragraph useful???} \emph{Of note, the ML pipeline
+includes a pre-processing step that occurs prior to training and
+classifying the ML models (methods). As part of this step, features are
+removed that will not provide useful information to build the model. For
+example, strongly correlated features provide the same information to
+build the model and thus can be collapsed. Additionally, features with
+zero or near-zero variance will not help the model differentiate groups
+and thus can be removed. Interestingly, despite starting with 104106
+features at the ASV level, only 478 (0.5\%) remained after
+pre-processing. At the OTU level, 20079 of the 705 features (3.5\%)
+remained after preprocessing (Table 1).}
 
@@
 (normal vs SRN) of each sample. Data was preprocessed to normalize
-values (scale/center? range?), remove values with zero or near-zero
-variance, and collapse collinear features using default parameters.
-Initially the models were run with default hyperparameters, but were
-expanded if the peak performance was at the edge of the hyperparameter
-range. Each taxonomic model taxonomic level combination (e.g.~Random
-Forest on genus) was run with 100 different seeds (1-100). Each seed
-split the data into a training (80\%) and testing (20\%) set, and output
+values (scale/center), remove values with zero or near-zero variance,
+and collapse collinear features using default parameters. Initially the
+models were run with default hyperparameters, but were expanded if the
+peak performance was at the edge of the hyperparameter range. Each
+taxonomic model taxonomic level combination (e.g.~Random Forest on
+genus) was run with 100 different seeds (1-100). Each seed split the
+data into a training (80\%) and testing (20\%) set, and output
 performance of the training and testing as area under the receiver
@@
 
-\includegraphics[width=0.75\textwidth,height=\textheight]{../exploratory/figures/rf_with_stats_sideways.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{../exploratory/figures/rf_with_stats.png}
 
@@
 SRNs using a Random Forest model. Dashed line denotes AUROC of 0.5 which
-is equivalent to random classification. Differences in AUROC values
-between taxonomic levels are significant unless otherwise specified with
-``NS'' (not significant)
+is equivalent to random classification. Significance between taxonomic
+levels was quantified by comparing the difference in mean AUROC and is
+denoted by letters A through E on the right side of the plot; taxonomic
+levels with the same letter are in the same significance group and are
+not significantly different from one another.
 
@@
 \makecell[c]{Number of Features \\ After Preprocessing} &
-\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\ \addlinespace
+\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\
 \midrule
 \endhead
-Phylum & 19 & 9 & 47.4 \% \\ \addlinespace
-Class & 36 & 19 & 52.8 \% \\ \addlinespace
-Order & 65 & 28 & 43.1 \% \\ \addlinespace
-Family & 124 & 54 & 43.5 \% \\ \addlinespace
-Genus & 316 & 115 & 36.4 \% \\ \addlinespace
-OTU & 20079 & 705 & 3.5 \% \\ \addlinespace
-ASV & 104106 & 478 & 0.5 \% \\ \addlinespace
+Phylum & 19 & 9 & 47.4 \% \\
+Class & 36 & 19 & 52.8 \% \\
+Order & 65 & 28 & 43.1 \% \\
+Family & 124 & 54 & 43.5 \% \\
+Genus & 316 & 115 & 36.4 \% \\
+OTU & 20079 & 705 & 3.5 \% \\
+ASV & 104106 & 478 & 0.5 \% \\
 \bottomrule
@@
 
+\hypertarget{references}{%
+\subsection*{References}\label{references}}
+\addcontentsline{toc}{subsection}{References}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-topuxe7uolu2020}{}%
+Topçuoğlu, Begüm D., Nicholas A. Lesniak, Mack T. Ruffin, Jenna Wiens,
+and Patrick D. Schloss. 2020. {``A Framework for Effective Application
+of Machine Learning to Microbiome-Based Classification Problems.''}
+\emph{mBio} 11 (3). \url{https://doi.org/10.1128/mBio.00434-20}.
+
+\end{CSLReferences}
+
 \end{document}

### 2021-08-04 run_rf_merge.sh
add .sh to run result merge

@@ -1,0 +1,79 @@
+#!/bin/bash
+###############################
+#                             #
+#  1) Job Submission Options  #
+#                             #
+###############################
+# Name
+#SBATCH --job-name=rf_merge
+# Resources
+# For MPI, increase ntasks-per-node
+# For multithreading, increase cpus-per-task
+#SBATCH --nodes=1
+#SBATCH --ntasks-per-node=1
+#SBATCH --cpus-per-task=12
+#SBATCH --mem-per-cpu=5GB
+#SBATCH --time=20:00:00
+# Account
+#SBATCH --account=pschloss1
+#SBATCH --partition=standard
+# Logs
+#SBATCH --mail-user=armourc@umich.edu
+#SBATCH --mail-type=FAIL
+#SBATCH --output=slurm_logs/%x-%a-%j.out
+# Environment
+#SBATCH --export=ALL
+
+#####################
+#                   #
+#  2) Job Commands  #
+#                   #
+#####################
+
+make data/process/combined-phylum-rf.csv
+make data/process/combined-class-rf.csv
+make data/process/combined-order-rf.csv
+make data/process/combined-family-rf.csv
+make data/process/combined-genus-rf.csv
+make data/process/combined-otu-rf.csv
+make data/process/combined-asv-rf.csv
+
+make data/process/hp-phylum-rf.csv
+make data/process/hp-class-rf.csv
+make data/process/hp-order-rf.csv
+make data/process/hp-family-rf.csv
+make data/process/hp-genus-rf.csv
+make data/process/hp-otu-rf.csv
+make data/process/hp-asv-rf.csv
+
+make data/process/importance-phylum-rf.csv
+make data/process/importance-class-rf.csv
+make data/process/importance-order-rf.csv
+make data/process/importance-family-rf.csv
+make data/process/importance-genus-rf.csv
+make data/process/importance-otu-rf.csv
+make data/process/importance-asv-rf.csv
+
+make data/process/importance90-phylum-rf.csv
+make data/process/importance90-class-rf.csv
+make data/process/importance90-order-rf.csv
+make data/process/importance90-family-rf.csv
+make data/process/importance90-genus-rf.csv
+make data/process/importance90-otu-rf.csv
+make data/process/importance90-asv-rf.csv
+
+make data/process/probability-phylum-rf.csv
+make data/process/probability-class-rf.csv
+make data/process/probability-order-rf.csv
+make data/process/probability-family-rf.csv
[…]

### 2021-08-05 paper/manuscript.Rmd
final edits of first draft

@@
 # Taxonomic Resolution Matters for Microbiome-Based Classification of Colorectal Cancer
-
-\vspace{10mm}
-
-**Other title options:**\
-Optimal Taxonomic Resolution for Microbiome-Based Classification of Colorectal Cancer\
-...
 
@@
 
-Colorectal cancer is one of the most common cancers in men and women and a leading cause of cancer related deaths in the United States {}.
+Colorectal cancer is one of the most common cancers in men and women and a leading cause of cancer related deaths in the United States [@siegel2020].
 Early detection and treatment are essential to increase survival rates, but for a variety of reasons including the invasiveness and high cost of screening (i.e.
-colonoscopy), many people do not comply with recommended screening guidelines {} prompting a need for low cost, non-invasive detection methods.
+colonoscopy), many people do not comply with recommended screening guidelines [@garcía2011a] prompting a need for low cost, non-invasive detection methods.
 A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasias (neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e.
@@
 However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown.
-Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs {}.
-However, whether the additional resolution provided by ASVs is useful for ML classification is unclear.
+Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs [@callahan2017].
+However, whether the additional resolution provided by ASVs is useful for ML classification is unclear {pat preprint?}.
 Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups.
@@
 
-Utilizing publicly available 16S rRNA sequence data from patients with cancer and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels.
-Using the taxonomic abundance data, we quanitfied how accurately samples could be classified as "normal" or "SRN" (i.e.
-advanced adenoma or cancer) using five machine learning methods with the Mikropml R package (methods).
+Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels.
+Using the taxonomic abundance data, we quantified how accurately samples could be classified as "normal" or "SRN" (i.e.
+advanced adenoma or carcinoma) using five machine learning methods with the Mikropml R package (methods).
 Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1).
@@
 
-**is this paragraph useful???** *Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods).
+Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods).
 As part of this step, features are removed that will not provide useful information to build the model.
@@
 Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing.
-At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).*
+At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).
 
@@
 ASVs) does not add additional sensitivity to model predicition and at the ASV level actually impedes model performance due to the sparsity of shared taxa.
-Additionally, utilizing genus or family level data could allow for merging data generated from different 16S regions or sequencing platforms.
 
--   Overall, the fine resolution of ASVs is unnecessary for ML classification with microbiome data since the vast majority of the ASVs are removed during preprocessing and the model performance is diminshed compared to other taxonomic levels.
--   since family/genus level is just as good as otu could we use data from different 16S regions?
+-   Utilizing genus or family level data could allow for merging data generated from different 16S regions or sequencing platforms. (is this true?)
+
+While predictive models utilizing microbiome data alone are not that great, microbiome in combination with FIT results are better than either alone {}.
+This analysis wanted to quantify difference between
 
@@
 
-```{r feats}
+```{r feature_counts}
 

### 2021-08-05 paper/manuscript.tex
final edits of first draft

@@
 
-\textbf{Other title options:}\\
-Optimal Taxonomic Resolution for Microbiome-Based Classification of
-Colorectal Cancer\\
-\ldots{}
-
-\vspace{10mm}
-
 Courtney R. Armour, Begüm D.Topçuoğlu, Andrea Garretto, Patrick D.
@@
 Colorectal cancer is one of the most common cancers in men and women and
-a leading cause of cancer related deaths in the United States \{\}.
-Early detection and treatment are essential to increase survival rates,
-but for a variety of reasons including the invasiveness and high cost of
+a leading cause of cancer related deaths in the United States
+(\protect\hyperlink{ref-siegel2020}{Siegel et al. 2020}). Early
+detection and treatment are essential to increase survival rates, but
+for a variety of reasons including the invasiveness and high cost of
 screening (i.e.~colonoscopy), many people do not comply with recommended
-screening guidelines \{\} prompting a need for low cost, non-invasive
-detection methods. A growing body of research points to the gut
-microbiome as a promising target for non-invasive screening to detect
-screen relevant neoplasias (neoplasms?) (SRNs) consisting of
-pre-cancerous polyps (i.e.~advanced adenomas) and carcinomas. Efforts to
-realize the diagnostic potential of the gut microbiome in detecting SRNs
-have focused on machine learning (ML) methods using abundances from
-operational taxonomic unit (OTU) classifications based on amplicon
-sequencing of the 16S rRNA gene \{\}. However, whether this is the
-optimal taxonomic resolution for classifying SRNs from microbiome data
-is unknown. Additionally, recent work has pushed for the use of amplicon
-sequence variants (ASVs) to replace OTUs for marker-gene analysis
-because of the improved resolution with ASVs \{\}. However, whether the
-additional resolution provided by ASVs is useful for ML classification
-is unclear. Since ML classification relies on consistent differences
-between groups, its possible that the resolution at the ASV level is too
-individualized to accurately differentiate groups. Topçuoğlu \emph{et
-al} (\protect\hyperlink{ref-topuxe7uolu2020}{Topçuoğlu et al. 2020})
+screening guidelines (\protect\hyperlink{ref-garcuxeda2011a}{García and
+Z 2011}) prompting a need for low cost, non-invasive detection methods.
+A growing body of research points to the gut microbiome as a promising
+target for non-invasive screening to detect screen relevant neoplasias
+(neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e.~advanced
+adenomas) and carcinomas. Efforts to realize the diagnostic potential of
+the gut microbiome in detecting SRNs have focused on machine learning
+(ML) methods using abundances from operational taxonomic unit (OTU)
+classifications based on amplicon sequencing of the 16S rRNA gene \{\}.
+However, whether this is the optimal taxonomic resolution for
+classifying SRNs from microbiome data is unknown. Additionally, recent
+work has pushed for the use of amplicon sequence variants (ASVs) to
+replace OTUs for marker-gene analysis because of the improved resolution
+with ASVs (\protect\hyperlink{ref-callahan2017}{Callahan, McMurdie, and
+Holmes 2017}). However, whether the additional resolution provided by
+ASVs is useful for ML classification is unclear \{pat preprint?\}. Since
+ML classification relies on consistent differences between groups, its
+possible that the resolution at the ASV level is too individualized to
+accurately differentiate groups. Topçuoğlu \emph{et al}
+(\protect\hyperlink{ref-topuxe7uolu2020}{Topçuoğlu et al. 2020})
 recently demonstrated effective application of machine learning (ML) to
@@
 
-Utilizing publicly available 16S rRNA sequence data from patients with
-cancer and healthy controls, we generated abundance tables with Mothur
-\{\} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV
-levels. Using the taxonomic abundance data, we quanitfied how accurately
-samples could be classified as ``normal'' or ``SRN'' (i.e.~advanced
-adenoma or cancer) using five machine learning methods with the Mikropml
-R package (methods). Across the five machine learning methods tested,
-model performance tended to increase with taxonomic level usually
-peaking around genus/OTU level before dropping off slightly with ASVs
-(Supplemental Figure 1). Random forest (RF) was consistently the top
-performer at most taxonomic levels. Within the RF model, the highest
-AUROCs were observed for family (median AUROC: 0.687), genus (median
-AUROC: 0.686), and OTU (median AUROC: 0.698) level data with no
-significant difference between the three (Figure 1). Performance with
-ASVs (median AUROC: 0.676) was significantly lower than OTU and genus (p
-\textless{} 0.05) but not family level (p = 0.06). These results were
-consistent for ASVs generated with DADA2 \{\} (median AUROC: 0.66).
-These results suggest that finer taxonomic resolution is not necessarily
-better for differentiating a persons cancer status from the microbiome.
+Utilizing publicly available 16S rRNA sequence data from stool of
+patients with SRNs and healthy controls, we generated abundance tables
+with Mothur \{\} annotated to Phylum, Class, Order, Family, Genus, OTU
+and ASV levels. Using the taxonomic abundance data, we quantified how
+accurately samples could be classified as ``normal'' or ``SRN''
+(i.e.~advanced adenoma or carcinoma) using five machine learning methods
+with the Mikropml R package (methods). Across the five machine learning
+methods tested, model performance tended to increase with taxonomic
+level usually peaking around genus/OTU level before dropping off
+slightly with ASVs (Supplemental Figure 1). Random forest (RF) was
+consistently the top performer at most taxonomic levels. Within the RF
+model, the highest AUROCs were observed for family (median AUROC:
+0.687), genus (median AUROC: 0.686), and OTU (median AUROC: 0.698) level
+data with no significant difference between the three (Figure 1).
+Performance with ASVs (median AUROC: 0.676) was significantly lower than
+OTU and genus (p \textless{} 0.05) but not family level (p = 0.06).
+These results were consistent for ASVs generated with DADA2 \{\} (median
+AUROC: 0.66). These results suggest that finer taxonomic resolution is
+not necessarily better for differentiating a persons cancer status from
+the microbiome.
 
@@
 
-\textbf{is this paragraph useful???} \emph{Of note, the ML pipeline
-includes a pre-processing step that occurs prior to training and
-classifying the ML models (methods). As part of this step, features are
-removed that will not provide useful information to build the model. For
-example, strongly correlated features provide the same information to
-build the model and thus can be collapsed. Additionally, features with
-zero or near-zero variance will not help the model differentiate groups
-and thus can be removed. Interestingly, despite starting with 104106
-features at the ASV level, only 478 (0.5\%) remained after
-pre-processing. At the OTU level, 20079 of the 705 features (3.5\%)
-remained after preprocessing (Table 1).}
+Of note, the ML pipeline includes a pre-processing step that occurs
+prior to training and classifying the ML models (methods). As part of
+this step, features are removed that will not provide useful information
+to build the model. For example, strongly correlated features provide
+the same information to build the model and thus can be collapsed.
+Additionally, features with zero or near-zero variance will not help the
+model differentiate groups and thus can be removed. Interestingly,
+despite starting with 104106 features at the ASV level, only 478 (0.5\%)
+remained after pre-processing. At the OTU level, 20079 of the 705
+features (3.5\%) remained after preprocessing (Table 1).
 
@@
 actually impedes model performance due to the sparsity of shared taxa.
-Additionally, utilizing genus or family level data could allow for
-merging data generated from different 16S regions or sequencing
-platforms.
 
@@
 \item
-  Overall, the fine resolution of ASVs is unnecessary for ML
-  classification with microbiome data since the vast majority of the
-  ASVs are removed during preprocessing and the model performance is
-  diminshed compared to other taxonomic levels.
-\item
-  since family/genus level is just as good as otu could we use data from
-  different 16S regions?
+  Utilizing genus or family level data could allow for merging data
+  generated from different 16S regions or sequencing platforms. (is this
+  true?)
 \end{itemize}
+
+While predictive models utilizing microbiome data alone are not that
+great, microbiome in combination with FIT results are better than either
+alone \{\}. This analysis wanted to quantify difference between
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-callahan2017}{}%
+Callahan, Benjamin J., Paul J. McMurdie, and Susan P. Holmes. 2017.
+{``Exact Sequence Variants Should Replace Operational Taxonomic Units in
+Marker-Gene Data Analysis.''} \emph{The ISME Journal} 11 (12): 2639--43.
+\url{https://doi.org/10.1038/ismej.2017.119}.
+
+\leavevmode\hypertarget{ref-garcuxeda2011a}{}%
+García, Gimeno, and Antonio Z. 2011. {``Factors Influencing Colorectal
+Cancer Screening Participation.''} \emph{Gastroenterology Research and
+Practice} 2012 (December): e483417.
+\url{https://doi.org/10.1155/2012/483417}.
+
+\leavevmode\hypertarget{ref-siegel2020}{}%
+Siegel, Rebecca L., Kimberly D. Miller, Ann Goding Sauer, Stacey A.
+Fedewa, Lynn F. Butterly, Joseph C. Anderson, Andrea Cercek, Robert A.
+Smith, and Ahmedin Jemal. 2020. {``Colorectal Cancer Statistics,
+2020.''} \emph{CA: A Cancer Journal for Clinicians} 70 (3): 145--64.
+https://doi.org/\url{https://doi.org/10.3322/caac.21601}.
+
 \leavevmode\hypertarget{ref-topuxe7uolu2020}{}%

### 2021-08-05 paper/manuscript.Rmd
update figure colors,layout,order

@@
 link-citations: true
+csl: msystems.csl
 output:
@@
   - \linenumbers
+  - \usepackage{multirow}
 ---
@@
 
+dada2_rf_pvals <- read_csv("../data/dada2/analysis/dada2_pvalues_by_model.csv") %>% 
+  filter(model == "rf")
+
+```
+
+```{r pvals}
+rf_pvals <- read_csv("../analysis/pvalues_by_model.csv",
+                     col_types = cols(.default = col_character(),
+                                      p_value = col_double())) %>% 
+  filter(model == "rf") %>% 
+  select(-model) %>% 
+  pivot_wider(names_from = "level2",values_from = "p_value") %>% 
+  column_to_rownames("level1")
+
+rf_notsig <- read_csv("../analysis/pvalues_by_model.csv",
+                     col_types = cols(.default = col_character(),
+                                      p_value = col_double())) %>% 
+  filter(model == "rf") %>% 
+  filter(p_value >= 0.05)
 ```
@@
 Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1).
-Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU and genus (p \< 0.05) but not family level (p = 0.06).
-These results were consistent for ASVs generated with DADA2 {} (median AUROC: `r round(dada2_med_auc,digits=3)`).
+Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU (p \< 0.01), but borderline equivalent to family (p = `r round(rf_pvals["family","asv"],digits=2)`) and genus (p = `r round(rf_pvals["genus","asv"],digits=2)`) levels.
+These results were consistent for ASVs generated with DADA2 [@callahan2016] (median AUROC: `r round(dada2_med_auc,digits=3)`).
 These results suggest that finer taxonomic resolution is not necessarily better for differentiating a persons cancer status from the microbiome.
+
+While AUROC values between models is a useful way to compare the overall model performance across all thresholds, they can be misleading [@lobo2008].
+Depending on the intended implementation of the model, one may want to optimize the true-positive rate(or sensitivity) over the false-postivite rate (or specificity).
+To further compare the model performance across taxonomic levels we compared the sensitivity of the models
+
+-   on average, how many positives are missed with ASV?
 
@@
 Examination of the prevalence of taxa in samples at each level supports this idea.
-A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels).
-The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 2).
+A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels.
+The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 3).
 While the resolution provided by ASVs is useful in
@@
 
+Mikropml{}
+
 ## Discussion/Conclusions (\~250 words)
@@
 In general, we found that finer resolution (e.g.
-ASVs) does not add additional sensitivity to model predicition and at the ASV level actually impedes model performance due to the sparsity of shared taxa.
+ASVs) does not add additional sensitivity to model prediction and at the ASV level actually impedes model performance due to the sparsity of shared taxa.
 
@@
 
-![](../exploratory/figures/rf_with_stats.png){width="75%"}
+```{r fig1}
+#![](../exploratory/figures/rf_with_stats.png){width="50%"} ![](../exploratory/figures/rf_with_stats.png){width="50%"}
+# library(cowplot)
+# 
+# p1 <- ggdraw + draw_image("../exploratory/figures/rf_with_stats.png", width = 0.5) 
+# p2 <- ggdraw + draw_image("../exploratory/figures/rf_with_stats.png", width = 0.5, x = 0.5)
+# 
+# plot_grid(p1,p1,labels = c('A', 'B'), label_size = 12)
 
-**Figure 1: Random Forest Model Performance with Significance.** Boxplots of AUROC values on the test dataset for 100 seeds predicting SRNs using a Random Forest model.
+library(png)
+library(grid)
+library(gridExtra)
+img1 <-  rasterGrob(as.raster(readPNG("../exploratory/figures/rf_with_stats.png")), interpolate = FALSE)
+img2 <-  rasterGrob(as.raster(readPNG("../exploratory/figures/sens_90spec.png")), interpolate = FALSE)
+grid.arrange(arrangeGrob(img1, left = textGrob("A)", x = unit(1, "npc"), y = unit(0.85, "npc"), gp = gpar(fontface="bold"))),
+             arrangeGrob(img2, left = textGrob("B)", x = unit(1, "npc"), y = unit(0.85, "npc"), gp = gpar(fontface="bold"))), 
+             ncol = 2)
+```
+
+**Figure 1: Random Forest Model Performance.** **A)** Boxplots with points of area under the receiver operating characteristic curve (AUROC) values on the test dataset for 100 seeds predicting SRNs using a Random Forest model.
 Dashed line denotes AUROC of 0.5 which is equivalent to random classification.
 Significance between taxonomic levels was quantified by comparing the difference in mean AUROC and is denoted by letters A through E on the right side of the plot; taxonomic levels with the same letter are in the same significance group and are not significantly different from one another.
+**B)** Strip plot of the sensitivity at a specificity of 90% across the 100 model iterations.
+Black points denote the median and the lines denote the IQR.
+The letters W through Z denote the significance groups.
 
@@
 
-![](../exploratory/figures/bin_prevalence.png){height="50%"}\
-**Supplemental Figure 2: Prevalence of Taxa in Samples.** Distribution of taxa prevalence in samples at each taxonomic level.
+![](../exploratory/figures/roc_by_level.png){height="50%"}
+
+**Supplemental Figure 2: Averaged ROC curves** ROC curves averaged across the 100 iterations of the model. The shaded region represents the standard deviation form the mean.
+
+![](../exploratory/figures/bin_prevalence.png){height="45%"}\
+**Supplemental Figure 3: Prevalence of Taxa in Samples.** Distribution of taxa prevalence in samples at each taxonomic level.
+
+\newpage
+
+![](../exploratory/figures/top10_imp.png){height="85%"}\
+**Supplemental Figure 4: Top 10 important taxa at each taxonomic level.** Summary of the 10 most important taxa for the random forest models at each taxonomic level based on the average decrease in AUC when the feature is permuted.
+
+\newpage
 

### 2021-08-05 paper/manuscript.tex
update figure colors,layout,order

@@
 \linenumbers
+\usepackage{multirow}
 \usepackage{booktabs}
@@
 \usepackage{makecell}
+\usepackage{xcolor}
 \ifluatex
@@
 a leading cause of cancer related deaths in the United States
-(\protect\hyperlink{ref-siegel2020}{Siegel et al. 2020}). Early
-detection and treatment are essential to increase survival rates, but
-for a variety of reasons including the invasiveness and high cost of
-screening (i.e.~colonoscopy), many people do not comply with recommended
-screening guidelines (\protect\hyperlink{ref-garcuxeda2011a}{García and
-Z 2011}) prompting a need for low cost, non-invasive detection methods.
-A growing body of research points to the gut microbiome as a promising
-target for non-invasive screening to detect screen relevant neoplasias
-(neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e.~advanced
-adenomas) and carcinomas. Efforts to realize the diagnostic potential of
-the gut microbiome in detecting SRNs have focused on machine learning
-(ML) methods using abundances from operational taxonomic unit (OTU)
+(\protect\hyperlink{ref-siegel2020}{1}). Early detection and treatment
+are essential to increase survival rates, but for a variety of reasons
+including the invasiveness and high cost of screening
+(i.e.~colonoscopy), many people do not comply with recommended screening
+guidelines (\protect\hyperlink{ref-garcuxeda2011a}{2}) prompting a need
+for low cost, non-invasive detection methods. A growing body of research
+points to the gut microbiome as a promising target for non-invasive
+screening to detect screen relevant neoplasias (neoplasms?) (SRNs)
+consisting of pre-cancerous polyps (i.e.~advanced adenomas) and
+carcinomas. Efforts to realize the diagnostic potential of the gut
+microbiome in detecting SRNs have focused on machine learning (ML)
+methods using abundances from operational taxonomic unit (OTU)
 classifications based on amplicon sequencing of the 16S rRNA gene \{\}.
@@
 replace OTUs for marker-gene analysis because of the improved resolution
-with ASVs (\protect\hyperlink{ref-callahan2017}{Callahan, McMurdie, and
-Holmes 2017}). However, whether the additional resolution provided by
-ASVs is useful for ML classification is unclear \{pat preprint?\}. Since
-ML classification relies on consistent differences between groups, its
-possible that the resolution at the ASV level is too individualized to
-accurately differentiate groups. Topçuoğlu \emph{et al}
-(\protect\hyperlink{ref-topuxe7uolu2020}{Topçuoğlu et al. 2020})
-recently demonstrated effective application of machine learning (ML) to
-microbiome based classification problems and developed a framework for
-applying ML practices in a more reproducible way (mikropml). This
-analysis utilizes the reproducible framework developed by Topçuoğlu
-\emph{et al} to quantify which ML method and taxonomic level produce the
-best performing classifier for detecting SRNs from microbiome data.
+with ASVs (\protect\hyperlink{ref-callahan2017}{3}). However, whether
+the additional resolution provided by ASVs is useful for ML
+classification is unclear \{pat preprint?\}. Since ML classification
+relies on consistent differences between groups, its possible that the
+resolution at the ASV level is too individualized to accurately
+differentiate groups. Topçuoğlu \emph{et al}
+(\protect\hyperlink{ref-topuxe7uolu2020}{4}) recently demonstrated
+effective application of machine learning (ML) to microbiome based
+classification problems and developed a framework for applying ML
+practices in a more reproducible way (mikropml). This analysis utilizes
+the reproducible framework developed by Topçuoğlu \emph{et al} to
+quantify which ML method and taxonomic level produce the best performing
+classifier for detecting SRNs from microbiome data.
 
@@
 Performance with ASVs (median AUROC: 0.676) was significantly lower than
-OTU and genus (p \textless{} 0.05) but not family level (p = 0.06).
-These results were consistent for ASVs generated with DADA2 \{\} (median
+OTU (p \textless{} 0.01), but borderline equivalent to family (p = 0.06)
+and genus (p = 0.05) levels. These results were consistent for ASVs
+generated with DADA2 (\protect\hyperlink{ref-callahan2016}{5}) (median
 AUROC: 0.66). These results suggest that finer taxonomic resolution is
@@
 the microbiome.
+
+While AUROC values between models is a useful way to compare the overall
+model performance across all thresholds, they can be misleading
+(\protect\hyperlink{ref-lobo2008}{6}). Depending on the intended
+implementation of the model, one may want to optimize the true-positive
+rate(or sensitivity) over the false-postivite rate (or specificity). To
+further compare the model performance across taxonomic levels we
+compared the sensitivity of the models
+
+\begin{itemize}
+\tightlist
+\item
+  on average, how many positives are missed with ASV?
+\end{itemize}
 
@@
 supports this idea. A majority of taxa are present in greater than 75\%
-of samples at the phylum (67\% of taxa) and class (63\% of taxa)
-levels). The opposite is observed at the OTU and ASV level where 60\%
-and 53\% of taxa respectively are only present in less than 25\% of
-samples (Supplemental Figure 2). While the resolution provided by ASVs
-is useful in
+of samples at the phylum (67\% of taxa) and class (63\% of taxa) levels.
+The opposite is observed at the OTU and ASV level where 60\% and 53\% of
+taxa respectively are only present in less than 25\% of samples
+(Supplemental Figure 3). While the resolution provided by ASVs is useful
+in
 
@@
 
+Mikropml\{\}
+
 \hypertarget{discussionconclusions-250-words}{%
@@
 In general, we found that finer resolution (e.g.~ASVs) does not add
-additional sensitivity to model predicition and at the ASV level
-actually impedes model performance due to the sparsity of shared taxa.
+additional sensitivity to model prediction and at the ASV level actually
+impedes model performance due to the sparsity of shared taxa.
 
@@
 
-\includegraphics[width=0.75\textwidth,height=\textheight]{../exploratory/figures/rf_with_stats.png}
+\includegraphics{manuscript_files/figure-latex/fig1-1.pdf}
 
-\textbf{Figure 1: Random Forest Model Performance with Significance.}
-Boxplots of AUROC values on the test dataset for 100 seeds predicting
-SRNs using a Random Forest model. Dashed line denotes AUROC of 0.5 which
-is equivalent to random classification. Significance between taxonomic
+\textbf{Figure 1: Random Forest Model Performance.} \textbf{A)} Boxplots
+with points of area under the receiver operating characteristic curve
+(AUROC) values on the test dataset for 100 seeds predicting SRNs using a
+Random Forest model. Dashed line denotes AUROC of 0.5 which is
+equivalent to random classification. Significance between taxonomic
 levels was quantified by comparing the difference in mean AUROC and is
@@
 levels with the same letter are in the same significance group and are
-not significantly different from one another.
+not significantly different from one another. \textbf{B)} Strip plot of
+the sensitivity at a specificity of 90\% across the 100 model
+iterations. Black points denote the median and the lines denote the IQR.
+The letters W through Z denote the significance groups.
 
@@
 \makecell[c]{Number of Features \\ After Preprocessing} &
-\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\
+\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\ \addlinespace
 \midrule
 \endhead
-Phylum & 19 & 9 & 47.4 \% \\
-Class & 36 & 19 & 52.8 \% \\
-Order & 65 & 28 & 43.1 \% \\
-Family & 124 & 54 & 43.5 \% \\
-Genus & 316 & 115 & 36.4 \% \\
-OTU & 20079 & 705 & 3.5 \% \\
-ASV & 104106 & 478 & 0.5 \% \\
+Phylum & 19 & 9 & 47.4 \% \\ \addlinespace
+Class & 36 & 19 & 52.8 \% \\ \addlinespace
+Order & 65 & 28 & 43.1 \% \\ \addlinespace
+Family & 124 & 54 & 43.5 \% \\ \addlinespace
+Genus & 316 & 115 & 36.4 \% \\ \addlinespace
+OTU & 20079 & 705 & 3.5 \% \\ \addlinespace
+ASV & 104106 & 478 & 0.5 \% \\ \addlinespace
 \bottomrule
@@
 
-\includegraphics[width=\textwidth,height=0.5\textheight]{../exploratory/figures/bin_prevalence.png}\\
-\textbf{Supplemental Figure 2: Prevalence of Taxa in Samples.}
+\includegraphics[width=\textwidth,height=0.5\textheight]{../exploratory/figures/roc_by_level.png}
+
+\textbf{Supplemental Figure 2: Averaged ROC curves} ROC curves averaged
+across the 100 iterations of the model. The shaded region represents the
+standard deviation form the mean.
+
+\includegraphics[width=\textwidth,height=0.45\textheight]{../exploratory/figures/bin_prevalence.png}\\
+\textbf{Supplemental Figure 3: Prevalence of Taxa in Samples.}
 Distribution of taxa prevalence in samples at each taxonomic level.
+
+\newpage
+
+\includegraphics[width=\textwidth,height=0.85\textheight]{../exploratory/figures/top10_imp.png}\\
+\textbf{Supplemental Figure 4: Top 10 important taxa at each taxonomic
+level.} Summary of the 10 most important taxa for the random forest
+models at each taxonomic level based on the average decrease in AUC when
+the feature is permuted.
+
+\newpage
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-callahan2017}{}%
-Callahan, Benjamin J., Paul J. McMurdie, and Susan P. Holmes. 2017.
-{``Exact Sequence Variants Should Replace Operational Taxonomic Units in
-Marker-Gene Data Analysis.''} \emph{The ISME Journal} 11 (12): 2639--43.
-\url{https://doi.org/10.1038/ismej.2017.119}.
+\begin{CSLReferences}{0}{1}
+\leavevmode\hypertarget{ref-siegel2020}{}%
+\CSLLeftMargin{1. }
+\CSLRightInline{\textbf{Siegel RL}, \textbf{Miller KD}, \textbf{Sauer
+AG}, \textbf{Fedewa SA}, \textbf{Butterly LF}, \textbf{Anderson JC},
+\textbf{Cercek A}, \textbf{Smith RA}, \textbf{Jemal A}. 2020. Colorectal
+cancer statistics, 2020. CA: A Cancer Journal for Clinicians
+\textbf{70}:145--164. doi:\url{https://doi.org/10.3322/caac.21601}.}
 
 \leavevmode\hypertarget{ref-garcuxeda2011a}{}%
-García, Gimeno, and Antonio Z. 2011. {``Factors Influencing Colorectal
-Cancer Screening Participation.''} \emph{Gastroenterology Research and
-Practice} 2012 (December): e483417.
-\url{https://doi.org/10.1155/2012/483417}.
+\CSLLeftMargin{2. }
+\CSLRightInline{\textbf{García G}, \textbf{Z A}. 2011. Factors
+Influencing Colorectal Cancer Screening Participation. Gastroenterology
+Research and Practice \textbf{2012}:e483417.
+doi:\href{https://doi.org/10.1155/2012/483417}{10.1155/2012/483417}.}
 
-\leavevmode\hypertarget{ref-siegel2020}{}%
-Siegel, Rebecca L., Kimberly D. Miller, Ann Goding Sauer, Stacey A.
-Fedewa, Lynn F. Butterly, Joseph C. Anderson, Andrea Cercek, Robert A.
-Smith, and Ahmedin Jemal. 2020. {``Colorectal Cancer Statistics,
-2020.''} \emph{CA: A Cancer Journal for Clinicians} 70 (3): 145--64.
-https://doi.org/\url{https://doi.org/10.3322/caac.21601}.
+\leavevmode\hypertarget{ref-callahan2017}{}%
+\CSLLeftMargin{3. }
+\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
+\textbf{Holmes SP}. 2017. Exact sequence variants should replace
+operational taxonomic units in marker-gene data analysis. The ISME
+Journal \textbf{11}:2639--2643.
+doi:\href{https://doi.org/10.1038/ismej.2017.119}{10.1038/ismej.2017.119}.}
 
 \leavevmode\hypertarget{ref-topuxe7uolu2020}{}%
-Topçuoğlu, Begüm D., Nicholas A. Lesniak, Mack T. Ruffin, Jenna Wiens,
-and Patrick D. Schloss. 2020. {``A Framework for Effective Application
-of Machine Learning to Microbiome-Based Classification Problems.''}
-\emph{mBio} 11 (3). \url{https://doi.org/10.1128/mBio.00434-20}.
+\CSLLeftMargin{4. }
+\CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+\textbf{Ruffin MT}, \textbf{Wiens J}, \textbf{Schloss PD}. 2020. A
+Framework for Effective Application of Machine Learning to
+Microbiome-Based Classification Problems. mBio \textbf{11}.
+doi:\href{https://doi.org/10.1128/mBio.00434-20}{10.1128/mBio.00434-20}.}
+
+\leavevmode\hypertarget{ref-callahan2016}{}%
+\CSLLeftMargin{5. }
+\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
+\textbf{Rosen MJ}, \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes
+SP}. 2016. DADA2: High-resolution sample inference from Illumina
+amplicon data. Nature Methods \textbf{13}:581--583.
+doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.}
+
+\leavevmode\hypertarget{ref-lobo2008}{}%
+\CSLLeftMargin{6. }
+\CSLRightInline{\textbf{Lobo JM}, \textbf{Jiménez-Valverde A},
+\textbf{Real R}. 2008. AUC: a misleading measure of the performance of
+predictive distribution models. Global Ecology and Biogeography
+\textbf{17}:145--151.
+doi:\href{https://doi.org/10.1111/j.1466-8238.2007.00358.x}{10.1111/j.1466-8238.2007.00358.x}.}
 

### 2021-08-05 exploratory/exploratory.Rmd
color,avg roc,top10 imp

@@
 library(cowplot)
+library(pROC)
 
@@
 ```{r data_setup}
+
 # Variables
@@
 
+-   [Feature Importance - threshold 0.9](#importance90)
+
 -   [Prevalence](#prevalence)
+
+-   [Prevalence VS Difference in Median Abundance](#prev-diff-abund)
+
+-   [Prevalence by Group](#prevalence-by-group)
+
+-   [Median Non-Zero Relative Abundance by Group](#med-relabund-group)
+
+-   [Sensitivity at 90% Specificity](#sens-90spec)
+
+-   [Averaged ROC curves](#avg-roc)
 
@@
 kable(input)
+```
+
+## Color Palette
+
+```{r color-pal,include=F}
+#color palette
+#colors <- c("#7E362A","#D04C44","#E6A444","#6C887A","#83BCC3","#5B6B86","#754A60")
+colors <- c("#D04C44","#E7834E","#E6A444","#829E88","#83BCC3","#7989A4","#B29ABA")
+names(colors) <- levels
+
+colors2 <- colors
+names(colors2) <- levels_names
+pie(rep(1,length(colors2)),col=colors2,labels = names(colors2))
 ```
@@
 
-```{r plot_by_method, echo=F,message=F,fig.height = 3.5, fig.width = 10, fig.align = "center"}
+```{r plot_by_method,fig.height = 3.5, fig.width = 10, fig.align = "center"}
 
@@
 
-```{r plot_by_level, echo=FALSE, message=FALSE,fig.height = 6, fig.width = 10, fig.align = "center"}
+```{r plot_by_level, fig.height = 6, fig.width = 10, fig.align = "center"}
 df_by_level <- full_df %>% 
@@
 
-```{r plot_all, echo=F, message=F, out.width="100%",fig.height=4}
+```{r plot_all, out.width="100%",fig.height=4}
 df_all <- full_df %>% 
@@
 
-```{r dada2_all, echo=F, message=F, out.width="100%",fig.height=3}
+```{r dada2_all, out.width="100%",fig.height=3}
 dada2_auc_df %>% 
@@
 ::: {.col-md-6}
-```{r rf_plot,echo=F,message=F}
+```{r rf_plot}
 m <- "Random Forest"
@@
 
-ggplot(sub_df,aes(x=factor(level,levels=c("ASV", "OTU","Genus","Family","Order","Class", "Phylum")), y=AUC)) + 
-#ggplot(sub_df,aes(x=level, y=AUC)) + 
-  geom_jitter(aes(fill=method,color=method),height = 0, width = 0.2,shape=21) +
+ggplot(sub_df,aes(x=factor(level,levels=c("ASV", "OTU","Genus","Family","Order","Class", "Phylum")),y=AUC)) + 
+  geom_jitter(aes(fill=level,color=level),height = 0, width = 0.2,shape=21,size=2) +
   stat_summary(fun.data = median_hilow,fun.args=0.5, geom="crossbar", width=0.5, color="black") +
-  scale_fill_manual(values = alpha(pal2[m],0.6)) +
-  scale_color_manual(values = pal2[m]) +
+  #scale_fill_manual(values = alpha(pal2[m],0.6)) +
+  #scale_color_manual(values = pal2[m]) +
+  scale_fill_manual(values = alpha(colors2,0.5)) +
+  scale_color_manual(values = colors2) +
   scale_y_continuous(limits=c(0.45,0.87),breaks = seq(0.5,0.8,0.1),expand=c(0,0)) +
@@
         panel.grid.major.y = element_blank(),
-        axis.text = element_text(size=14),
-        axis.title = element_text(size=14)) +
+        axis.text = element_text(size=12),
+        axis.title = element_text(size=12)) +
   coord_flip() +
-  geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
-  annotate("text",x=levels_names,y=rep(0.85,length(levels_names)),label=c("A","B","C","D,E","D","D","E"))
-ggsave("./figures/rf_with_stats.png",width = 5,height=4,units="in")
-
-```
-:::
-:::
+  annotate("text",x=levels_names,y=rep(0.85,length(levels_names)),label=c("A","B","C","D,E","D,E","D","E")) +
+  geom_hline(yintercept = 0.5,color="grey",lty="dashed") 
+ggsave("./figures/rf_with_stats.png",width = 4.5,height=5,units="in")
+
+```
 
@@
 sub_stats <- sub_stats %>% 
-  #mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
-  mutate(y.position=c(0.85,0.95,0.9,.95)) %>%
+  mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
+  #mutate(y.position=c(0.85,0.95,0.9,.95)) %>%
   rename(group1=level1,group2=level2)
@@
   scale_fill_manual(values = pal2[m]) + 
-  scale_y_continuous(limits=c(0.4,1),breaks = seq(0.4,1,0.1)) +
+  scale_y_continuous(limits=c(0.4,1.05),breaks = seq(0.4,1,0.1)) +
   theme_bw() + xlab("") + ylab("AUROC") +
@@
 sub_stats <- sub_stats %>% 
-  #mutate(y.position=seq.int(0.85,.95,length.out = nrow(sub_stats)))
-  mutate(y.position=c(0.85,0.95,0.9,.95))
+  mutate(y.position=seq.int(0.85,.95,length.out = nrow(sub_stats)))
+  #mutate(y.position=c(0.85,0.95,0.9,.95))
 
@@
 
-sub_stats <- sub_stats %>% 
-  #mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
-  mutate(y.position=c(1,1,1,1,1,
-                      0.9,
-                      0.95,0.95,0.95,0.95,
-                      0.85,
-                      1.05,1.05,1.05,1.05,1.05,1.05)) %>% 
-  #mutate(y.position=c(0.85,0.95,0.9,.95)) %>%
-  rename(group1=level1,group2=level2)
-
-ggplot(sub_df,aes(x=level,y=AUC)) + 
-  geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
-  #geom_boxplot(aes(fill=method)) +
-  geom_jitter(aes(fill=method),height = 0, width = 0.2,alpha=0.8,shape=21) +
-  stat_summary(fun.data = median_hilow,fun.args=0.5, geom="crossbar", width=0.5, color="black") +
-  scale_fill_manual(values = pal2[m]) + 
-  scale_y_continuous(limits=c(0.4,1.05),breaks = seq(0.4,1.05,0.1)) +
-  theme_bw() + xlab("") + ylab("AUROC") +
-  theme(legend.position ="none",
-        #axis.text.x = element_text(angle = 45,hjust = 1),
-        panel.grid.major.x = element_blank(),
-        axis.text = element_text(size=14)) +
-  #ggtitle(m) +
-  #stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
-  #stat_pvalue_manual(sub_stats,label="p_value_mod") 
-  stat_pvalue_manual(sub_stats,label="sig")
+# sub_stats <- sub_stats %>% 
+#   #mutate(y.position=seq.int(0.85,1.05,length.out = nrow(sub_stats))) %>% 
+#   mutate(y.position=c(1,1,1,1,1,
+#                       0.9,
+#                       0.95,0.95,0.95,0.95,
+#                       0.85,
+#                       1.05,1.05,1.05,1.05,1.05,1.05)) %>% 
+#   #mutate(y.position=c(0.85,0.95,0.9,.95)) %>%
+#   rename(group1=level1,group2=level2)
+
+# ggplot(sub_df,aes(x=level,y=AUC)) + 
+#   geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
+#   #geom_boxplot(aes(fill=method)) +
+#   geom_jitter(aes(fill=method),height = 0, width = 0.2,alpha=0.8,shape=21) +
+#   stat_summary(fun.data = median_hilow,fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+#   scale_fill_manual(values = pal2[m]) + 
+#   scale_y_continuous(limits=c(0.4,1.05),breaks = seq(0.4,1.05,0.1)) +
+#   theme_bw() + xlab("") + ylab("AUROC") +
+#   theme(legend.position ="none",
+#         #axis.text.x = element_text(angle = 45,hjust = 1),
+#         panel.grid.major.x = element_blank(),
+#         axis.text = element_text(size=14)) +
+#   #ggtitle(m) +
+#   #stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
+#   #stat_pvalue_manual(sub_stats,label="p_value_mod") 
+#   stat_pvalue_manual(sub_stats,label="sig")
 ```
@@
 ::: {.col-md-6}
-```{r glmnet_table,echo=FALSE}
+```{r glmnet_table}
 t_table %>%
@@
 ::: {.col-md-6}
-```{r glmnet_plot,echo=F,message=F}
+```{r glmnet_plot}
 ## glmnet Model Performance
@@
 
-```{r rf_importance_plots,echo=F,message=F,fig.show="hold",out.width="50%"}
+```{r rf_importance_plots,echo=F,message=F,fig.show="hold",out.width="50%",include=FALSE}
 imp_plots[["rf-phylum-1"]]
@@
 
-```{r importance_rank, message=F,echo=F}
+```{r importance_rank}
 
@@
 
-rank_top_tax <- function(data,tax){
+rank_tax <- function(data,tax){
   sub_data <- data %>%
@@
     group_by(level,Taxonomy,OTU) %>% 
-    summarise(mean_AUC_diff = mean(AUC_diff)) %>% 
+    summarise(mean_AUC_diff = median(AUC_diff)) %>% 
     ungroup() %>% 
@@
 
-ranked_tax <- bind_rows(rank_top_tax(rf_imp_tax,"phylum"),
-                        rank_top_tax(rf_imp_tax,"class"),
-                        rank_top_tax(rf_imp_tax,"order"),
-                        rank_top_tax(rf_imp_tax,"family"),
-                        rank_top_tax(rf_imp_tax,"genus"),
-                        rank_top_tax(rf_imp_tax,"otu"),
-                        rank_top_tax(rf_imp_tax,"asv"))
+ranked_tax <- bind_rows(rank_tax(rf_imp_tax,"phylum"),
+                        rank_tax(rf_imp_tax,"class"),
+                        rank_tax(rf_imp_tax,"order"),
+                        rank_tax(rf_imp_tax,"family"),
+                        rank_tax(rf_imp_tax,"genus"),
+                        rank_tax(rf_imp_tax,"otu"),
+                        rank_tax(rf_imp_tax,"asv"))
 
 top_10_tax <- ranked_tax %>% 
+  filter(rank <= 10)
+
+top_10_tax_table <- ranked_tax %>% 
   filter(rank <= 10) %>% 
@@
   pivot_wider(names_from = level,values_from = Taxonomy) 
-colnames(top_10_tax) <- c("Rank","Phylum","Class","Order","Family","Genus","OTU","ASV")
-
-ranked_tax %>% 
-  filter(rank <= 10) %>% 
-  mutate(Taxonomy=case_when(level=="otu" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
-                            level=="asv" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
-                            TRUE ~ Taxonomy)) %>% 
-  mutate(Taxonomy=sapply(strsplit(Taxonomy,"\\("), `[`, 1)) %>% 
-  mutate(Taxonomy=paste0(Taxonomy," (",OTU,")"))
-
-kable(top_10_tax,format="html",booktabs=T,align="clllllll",
+colnames(top_10_tax_table) <- c("Rank","Phylum","Class","Order","Family","Genus","OTU","ASV")
+
+# ranked_tax %>% 
+#   filter(rank <= 10) %>% 
+#   mutate(Taxonomy=case_when(level=="otu" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
+#                             level=="asv" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
+#                             TRUE ~ Taxonomy)) %>% 
+#   mutate(Taxonomy=sapply(strsplit(Taxonomy,"\\("), `[`, 1)) %>% 
+#   mutate(Taxonomy=paste0(Taxonomy," (",OTU,")"))
+
+kable(top_10_tax_table,format="html",booktabs=T,align="clllllll",
       col.names = c("Rank",levels_names)) %>% 
@@
 
-data <- tibble(bactname = c("Staphylococcaceae", "Moraxella", "Streptococcus", "Acinetobacter"),
-               OTUname = c("OTU 1", "OTU 2", "OTU 3", "OTU 4"),
-               value = c(-0.5, 0.5, 2, 3))
-
-data %>% mutate(
-  name = glue("<i>{bactname}</i> ({OTUname})"),
-  name = fct_reorder(name, value))  %>%
-  ggplot(aes(value, name)) + 
-    geom_col(alpha = 0.5) + 
-    scale_fill_identity() +
-    theme(
+# data <- tibble(bactname = c("Staphylococcaceae", "Moraxella", "Streptococcus", "Acinetobacter"),
+#                OTUname = c("OTU 1", "OTU 2", "OTU 3", "OTU 4"),
+#                value = c(-0.5, 0.5, 2, 3))
+# ranked_tax %>% 
+#   filter(rank <= 10) %>% 
+#   mutate(Taxonomy=case_when(level=="otu" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
+#                             level=="asv" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
+#                             TRUE ~ Taxonomy)) %>% 
+#   mutate(Taxonomy=sapply(strsplit(Taxonomy,"\\("), `[`, 1)) %>% 
+#   mutate(name = glue("<i>{Taxonomy}</i> ({OTU})")) %>% 
+#   ggplot(aes(mean_AUC_diff,name)) +
+#     geom_col(alpha = 0.5) + 
+#     scale_fill_identity() +
+#     theme(
+#       axis.text.y = element_markdown(),
+#       plot.caption = element_markdown(lineheight = 1.2))
+# 
+# data %>% mutate(
+#   name = glue("<i>{bactname}</i> ({OTUname})"),
+#   name = fct_reorder(name, value))  %>%
+#   ggplot(aes(value, name)) + 
+#     geom_col(alpha = 0.5) + 
+#     scale_fill_identity() +
+#     theme(
+#       axis.text.y = element_markdown(),
+#       plot.caption = element_markdown(lineheight = 1.2))
+# 
+# 
+# ggplot(data = diamonds) + stat_summary(
+#   mapping = aes(x = depth, y = cut),
+#   fun.min = function(z) { quantile(z,0.25) },
+#   fun.max = function(z) { quantile(z,0.75) },
+#   fun = median)
+
+top_10_data <- inner_join(rf_imp_tax,top_10_tax,by=c("level","OTU","Taxonomy")) %>% 
+  mutate(level = factor(level,levels=levels,labels=levels_names)) %>% 
+  select(level,seed,OTU,Taxonomy,AUC_diff,rank) %>% 
+  mutate(Taxonomy=case_when(level=="OTU" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
+                            level=="ASV" ~ sapply(strsplit(Taxonomy,"();"), `[`, 6),
+                            TRUE ~ Taxonomy)) %>% 
+  mutate(Taxonomy=sapply(strsplit(Taxonomy,"\\("), `[`, 1)) %>% 
+  mutate(Taxonomy=gsub("_"," ",Taxonomy)) %>% 
+  mutate(Taxonomy=gsub("unclassified","unclass.",Taxonomy)) %>% 
+  mutate(name = glue("<i>{Taxonomy}</i> ({OTU})"))
+
+plot_top_imp <- function(top_data,tax_level){
+  top_data %>% 
+    filter(level == tax_level) %>% 
+    arrange(rank) %>% 
+    ggplot(aes(x=AUC_diff,y=reorder(name,desc(rank)))) +
+      #geom_jitter(color=colors2[tax_level],fill=alpha(colors2[tax_level],0.6)) +
+      stat_summary(fun.data=median_hilow, fun.args=0.5, geom="pointrange", color=colors2[tax_level]) +
+      theme_bw() +
+      theme(axis.title.y = element_blank(),
         axis.text.y = element_markdown(),
-      plot.caption = element_markdown(lineheight = 1.2))
+        plot.caption = element_markdown(lineheight = 1.2)) +
+      ggtitle(tax_level) +
+      xlab("Decrease in AUC")
+}
+
+plotlist <- list()
+plotlist[[1]] <- plot_top_imp(top_10_data,"Phylum")
+plotlist[[2]] <- plot_top_imp(top_10_data,"Class")
+plotlist[[3]] <- plot_top_imp(top_10_data,"Order")
+plotlist[[4]] <- plot_top_imp(top_10_data,"Family")
+plotlist[[5]] <- plot_top_imp(top_10_data,"Genus")
+plotlist[[6]] <- plot_top_imp(top_10_data,"OTU")
+plotlist[[7]] <- plot_top_imp(top_10_data,"ASV")
+
+plot_grid(plotlist = plotlist,ncol = 2,align = "v")
+ggsave("./figures/top10_imp.png",height = 10)
+
+# gp1<- ggplot_gtable(ggplot_build(plotlist[[1]]))
+# gp2<- ggplot_gtable(ggplot_build(plotlist[[2]]))
+# gp3<- ggplot_gtable(ggplot_build(plotlist[[3]]))
+# gp4<- ggplot_gtable(ggplot_build(plotlist[[4]]))
+# gp5<- ggplot_gtable(ggplot_build(plotlist[[5]]))
+# gp6<- ggplot_gtable(ggplot_build(plotlist[[6]]))
+# gp7<- ggplot_gtable(ggplot_build(plotlist[[7]]))
+# 
+# maxWidth = unit.pmax(gp1$widths[2:3], 
+#                      gp2$widths[2:3],
+#                      gp3$widths[2:3],
+#                      gp4$widths[2:3],
+#                      gp5$widths[2:3],
+#                      gp6$widths[2:3],
+#                      gp7$widths[2:3])
+# 
+# gp1$widths[2:3] <- maxWidth
+# gp2$widths[2:3] <- maxWidth
+# gp3$widths[2:3] <- maxWidth
+# gp4$widths[2:3] <- maxWidth
+# gp5$widths[2:3] <- maxWidth
+# gp6$widths[2:3] <- maxWidth
+# gp7$widths[2:3] <- maxWidth
+# 
+# grid.arrange(gp1,gp2,gp3,gp4,gp5,gp6,gp7)
+# ggsave("./figures/top10_imp.png")
+# 
+# ggarrange(plotlist = plotlist,ncol=2)
+
+# grid.arrange(plotlist[[1]],
+#              plotlist[[2]],
+#              plotlist[[3]],
+#              plotlist[[4]],
+#              plotlist[[5]],
+#              plotlist[[6]],
+#              plotlist[[7]])
+
+```
+## Correlated Features
+Highly correlated model features should be collapsed (i.e. permuted together) when quantifying permutation importance.
+First checking if any features are correlated in this dataset.
+
+```{r corr_feats}
+
+get_annotated_corr <- function(tax_level){
+  l_cor <- read_csv(paste0("../analysis/taxa_correlations_",tax_level,".csv"),
+                    col_types = cols(.default = col_character(),
+                                     corr = col_double(), 
+                                     proc_corr = col_double())) 
+  l_tax <- read_tsv(paste0("../data/",tax_level,"/crc.taxonomy"),
+                     col_types = c(Size = col_double(),
+                                   .default = col_character())) %>%
+    select(-Size)
+
+  merge <- inner_join(l_cor,l_tax,by=c("taxa1"="OTU")) %>% 
+    rename(taxa1_name=Taxonomy) %>% 
+    inner_join(.,l_tax,by=c("taxa2"="OTU")) %>% 
+    rename(taxa2_name=Taxonomy) %>% 
+    select(level,taxa1,taxa2,taxa1_name,taxa2_name,proc_corr) %>% 
+    rename(corr=proc_corr)
+
+  return(merge)
+}
+
+annotated_feature_correlations <- get_annotated_corr("phylum") %>% 
+  bind_rows(get_annotated_corr("class")) %>% 
+  bind_rows(get_annotated_corr("order")) %>% 
+  bind_rows(get_annotated_corr("family")) %>% 
+  bind_rows(get_annotated_corr("genus")) %>% 
+  bind_rows(get_annotated_corr("otu")) %>% 
+  bind_rows(get_annotated_corr("asv"))
+
+maxmin_corr <- annotated_feature_correlations %>% 
+  mutate(level=factor(level,levels=levels)) %>% 
+  group_by(level) %>% 
+  summarise(max_corr = max(corr),
+            min_corr = min(corr))
+
+corr90 <- annotated_feature_correlations %>% 
+  filter(abs(corr) > 0.9)
+
+annotated_feature_correlations %>% 
+  filter(level == "genus") %>% 
+  filter(abs(corr) > 0.85)
+
+# View(annotated_feature_correlations %>% 
+#        filter(level == "otu") %>%
+#        filter(abs(corr) < 0.9) %>% 
+#        filter(abs(corr) > 0.8))
+
+#c("Otu00229","Otu00328","Otu00331", "Otu00361", "Otu00413", "Otu00449")
+```
+
+## Importance - threshold 0.9 {#importance90}
+
+```{r importance90,echo=F}
 
@@
     filter(level==l,kept==1)
-  hist <- hist(sub_data$prevalence,breaks = seq(0,1,0.25),plot=FALSE)
+  hist <- hist(sub_data$total_prevalence,breaks = seq(0,1,0.1),plot=FALSE)
 
@@
 labs <- paste0(level_nOTU$level_name," (",level_nOTU$nOTU,")")
+pal <- colors2
+names(pal) <- labs
+
 pct_bins %>%
@@
     geom_bar(stat = "identity") +
-    facet_wrap(~level) +
+    facet_wrap(~level,ncol=2) +
     scale_y_continuous(expand=c(0,0),limits=c(0,0.8),labels = scales::percent_format()) +
-    scale_x_continuous(expand=c(0,0),breaks = c(0,.25,0.50,.75,1)) +
+    scale_x_continuous(expand=c(0,0),breaks = seq(0,1,0.1)) +
     xlab("Prevalence in Samples") +
@@
           axis.text = element_text(size=12),
-          axis.title = element_text(size=14)) +
-  geom_text(aes(label = paste0((round(pct,digits=2)*100),"%")),nudge_y=0.05)
+          axis.title = element_text(size=14),
+          panel.grid.major.x = element_blank(),
+          panel.grid.minor.x = element_blank()) +
+  geom_text(aes(label = paste0((round(pct,digits=2)*100),"%")),nudge_y=0.08) +
+  scale_fill_manual(values=pal)
 ggsave("./figures/bin_prevalence.png")
 ```
+
+## Prevalence VS Difference in Median Abundance {#prev-diff-abund}
+
+Taxa prevalence across all samples compared to the difference in median relative abundance between cancer and normal samples.
+Top 10 taxa based on importance rank are red, rank 11-20 are blue, and the rest are grey.
+
+```{r prev_abund,echo=F}
+
+library(plotly)
+
+metadata <- read_csv("../data/metadata/metadata.csv",
+                     col_types = cols(.default = col_double(),
+                                      sample = col_character(),
+                                      Dx_Bin = col_character(),
+                                      dx = col_character(),
+                                      Height = col_character(),
+                                      Gender = col_character(),
+                                      Weight = col_character(),
+                                      BMI = col_character()))
+
+get_abund_prev <- function(tax_level,file,metadata,prev){
+  sub_meta <- metadata %>%
+    mutate(dx=case_when(Dx_Bin =="Adenoma" ~ "normal",
+                        Dx_Bin== "Normal" ~ "normal",
+                        Dx_Bin== "High Risk Normal" ~ "normal",
+                        Dx_Bin== "adv Adenoma" ~ "cancer",
+                        Dx_Bin== "Cancer" ~ "cancer")) %>% 
+    select(sample,dx) %>% 
+    drop_na() %>% 
+    rename(Group=sample)
+  sub_prev <- prev %>% 
+    filter(level == tax_level & kept == 1) %>% 
+    select(name,total_prevalence)
+  abund <- as.data.frame(fread(file)) %>% 
+    mutate(Group=as.character(Group)) %>% 
+    select(-label,-numOtus)
+
+  abund_prev <- inner_join(abund,sub_meta,by="Group") %>% 
+    select(Group,dx,everything()) %>% 
+    pivot_longer(cols=starts_with("Otu"),names_to = "name",values_to = "relabun") %>% 
+    group_by(name,dx) %>% 
+    summarize(median_relabund = median(relabun)) %>% 
+    ungroup() %>% 
+    pivot_wider(id_cols = "name",names_from="dx",values_from = "median_relabund") %>% 
+    mutate(median_diff = normal - cancer) %>% 
+    inner_join(.,sub_prev,by="name")
+
+  return(abund_prev)
+
+}
+
+plot_abund_prev <- function(abund_prev,ranked_tax,tax_level){
+  level_ranked_tax <- ranked_tax %>% 
+    filter(level == tax_level) %>% 
+    rename(name=OTU) %>% 
+    select(-level,-mean_AUC_diff)
+
+  ##non-interactive plot
+
+  # plot <- abund_prev %>% 
+  #   inner_join(.,level_ranked_tax,by="name") %>% 
+  #   mutate(rank_category = case_when(rank <= 10 ~ "top10",
+  #                                    rank <= 20 ~ "top20",
+  #                                    TRUE ~ "other")) %>% 
+  #   ggplot(aes(x=prevalence,y=median_diff,color=rank_category)) +
+  #     geom_point(alpha=0.3,size=3) +
+  #     theme_bw() +
+  #     #theme(legend.position = "none") +
+  #     xlab("Prevalence") +
+  #     scale_color_manual(name="Rank",
+  #                        values=c("red","cornflowerblue","grey"),
+  #                        breaks=c("top10","top20","other")) +
+  #     ylab("Difference in median relative abundance (cancer-normal)") +
+  #     ggtitle(tax_level)
+
+  # interactive plot
+
+  data <- abund_prev %>% 
+    inner_join(.,level_ranked_tax,by="name") %>% 
+    mutate(rank_category = case_when(rank <= 10 ~ "top10",
+                                     rank <= 20 ~ "top20",
+                                     TRUE ~ "other"))
+
+  if(tax_level %in% c("otu","asv")){
+    data <- data %>% 
+      mutate(Taxonomy=sapply(strsplit(Taxonomy,"();"), `[`, 6)) %>% 
+      mutate(Taxonomy=sapply(strsplit(Taxonomy,"\\("), `[`, 1))
+  }
+  # plot <- ggplot(data,aes(x=prevalence,y=median_diff,color=rank_category)) +
+  #   geom_point(alpha=0.3,size=3) +
+  #   theme_bw() +
+  #   #theme(legend.position = "none") +
+  #   xlab("Prevalence") +
+  #   scale_color_manual(name="Rank",
+  #                      values=c("red","cornflowerblue","grey"),
+  #                      breaks=c("top10","top20","other")) +
+  #   ylab("Difference in median relative abundance (cancer-normal)") +
+  #   ggtitle(tax_level)
+  # ggplotly(plot) %>% 
+  #   style(hoverinfo = "name")
+  # return(plot)
+
+  plotly <- plot_ly(data) %>% 
+    add_trace(type="scatter",
+              mode="markers",
+              x= ~total_prevalence,
+              y = ~median_diff, 
+              color= ~factor(rank_category,levels=c("top10","top20","other")),
+              hoverinfo="text",
+              text = ~paste("</br>",Taxonomy,"(",name,")",
+                            "</br> Prevalence: ",round(total_prevalence,digits=2),
+                            "</br> Diff median relabund: ",round(median_diff,digits=5)),
+              colors=c("red","cornflowerblue","grey"),
+              opacity=0.3,
+              marker = list(size = 10))
+  final_plotly <- plotly %>% layout(title=tax_level,
+                                    xaxis = list(title = "Overall Prevalence",autorange=TRUE,showline=TRUE), 
+                                    yaxis = list(title = "Differene in median relative abundance (cancer - normal)",
+                                                 autorange=TRUE,showline=TRUE,zerolinecolor = toRGB("grey70")))
+  return(final_plotly)
+}
+
+phylum_abund_prev <- get_abund_prev("phylum","../data/phylum/crc.relabund",metadata,prevalence)
+plot_abund_prev(phylum_abund_prev,ranked_tax,"phylum")
+
+class_abund_prev <- get_abund_prev("class","../data/class/crc.relabund",metadata,prevalence)
+plot_abund_prev(class_abund_prev,ranked_tax,"class")
+
+order_abund_prev <- get_abund_prev("order","../data/order/crc.relabund",metadata,prevalence)
+plot_abund_prev(order_abund_prev,ranked_tax,"order")
+
+family_abund_prev <- get_abund_prev("family","../data/family/crc.relabund",metadata,prevalence)
+plot_abund_prev(family_abund_prev,ranked_tax,"family")
+
+genus_abund_prev <- get_abund_prev("genus","../data/genus/crc.relabund",metadata,prevalence)
+plot_abund_prev(genus_abund_prev,ranked_tax,"genus")
+
+otu_abund_prev <- get_abund_prev("otu","../data/otu/crc.relabund",metadata,prevalence)
+plot_abund_prev(otu_abund_prev,ranked_tax,"otu")
+
+asv_abund_prev <- get_abund_prev("asv","../data/asv/crc.relabund",metadata,prevalence)
+plot_abund_prev(asv_abund_prev,ranked_tax,"asv")
+
+```
+
+## Prevalence by Group {#prevalence-by-group}
+```{r prev_group,echo=F}
+plot_prev <- function(prevalence,ranked_tax,tax_level){
+  level_ranked_tax <- ranked_tax %>% 
+    filter(level == tax_level) %>% 
+    rename(name=OTU) %>% 
+    select(-level,-mean_AUC_diff)
+
+  data <- prevalence %>% 
+    filter(level == tax_level) %>% 
+    inner_join(.,level_ranked_tax,by="name") %>% 
+    mutate(rank_category = case_when(rank <= 10 ~ "top10",
+                                     rank <= 20 ~ "top20",
+                                     TRUE ~ "other"))
+
+  if(tax_level %in% c("otu","asv")){
+    data <- data %>% 
+      mutate(Taxonomy=sapply(strsplit(Taxonomy,"();"), `[`, 6)) %>% 
+      mutate(Taxonomy=sapply(strsplit(Taxonomy,"\\("), `[`, 1))
+  }
+
+  d <- data.frame(x=seq(0,1,0.1),y=seq(0,1,0.1))
+
+  plotly <- plot_ly(data) %>% 
+    add_trace(type="scatter",
+              mode="markers",
+              x= ~normal_prevalence,
+              y = ~srn_prevalence, 
+              color= ~factor(rank_category,levels=c("top10","top20","other")),
+              hoverinfo="text",
+              text = ~paste("</br>",Taxonomy,"(",name,")",
+                            "</br> SRN Prevalence: ",round(srn_prevalence,digits=2),
+                            "</br> Normal Prevalence: ",round(normal_prevalence,digits=2)),
+              colors=c("red","cornflowerblue","grey"),
+              opacity=0.3,
+              marker = list(size = 10))
+  final_plotly <- plotly %>% layout(title=tax_level,
+                                    xaxis = list(title = "Prevalence in Normal Samples",range=c(0,1.05),showline=TRUE,
+                                                 zerolinecolor = toRGB("grey70")), 
+                                    yaxis = list(title = "Prevalence in SRN Samples",range=c(0,1.05),showline=TRUE,
+                                                 zerolinecolor = toRGB("grey70"))) %>% 
+        add_trace(data=d,x=~x,y=~y,type="scatter",mode="lines",color=I('grey'),name="y=x line")
+  return(final_plotly)
+}
+
+plot_prev(prevalence,ranked_tax,"phylum")
+plot_prev(prevalence,ranked_tax,"class")
+plot_prev(prevalence,ranked_tax,"order")
+plot_prev(prevalence,ranked_tax,"family")
+plot_prev(prevalence,ranked_tax,"genus")
+plot_prev(prevalence,ranked_tax,"otu")
+plot_prev(prevalence,ranked_tax,"asv")
+
+```
+
+## Median Non-Zero Relative Abundance by Group {#med-relabund-group}
+```{r relabund,echo=F}
+metadata <- read_csv("../data/metadata/metadata.csv",
+                     col_types = cols(.default = col_double(),
+                                      sample = col_character(),
+                                      Dx_Bin = col_character(),
+                                      dx = col_character(),
+                                      Gender = col_character(),
+                                      Height = col_character(),
+                                      Weight = col_character(),
+                                      BMI = col_character()))
+
+get_abund <- function(tax_level,file,metadata){
+  sub_meta <- metadata %>%
+    mutate(dx=case_when(Dx_Bin =="Adenoma" ~ "normal",
+                        Dx_Bin== "Normal" ~ "normal",
+                        Dx_Bin== "High Risk Normal" ~ "normal",
+                        Dx_Bin== "adv Adenoma" ~ "cancer",
+                        Dx_Bin== "Cancer" ~ "cancer")) %>% 
+    select(sample,dx) %>% 
+    drop_na() %>% 
+    rename(Group=sample)
+  abund <- as.data.frame(fread(file)) %>% 
+    mutate(Group=as.character(Group)) %>% 
+    select(-label,-numOtus)
+
+  data <- inner_join(abund,sub_meta,by="Group") %>% 
+    select(Group,dx,everything()) %>% 
+    pivot_longer(cols=starts_with("Otu"),names_to = "name",values_to = "relabun") %>% 
+    filter(relabun > 0) %>% 
+    group_by(name,dx) %>% 
+    summarize(median_relabund = median(relabun)) %>% 
+    ungroup() %>% 
+    pivot_wider(id_cols = "name",names_from="dx",values_from = "median_relabund",values_fill = 0) 
+  return(data)
+}
+
+plot_abund <- function(data,tax_level,ranked_tax){
+  plot_data <- ranked_tax %>% 
+    filter(level == tax_level) %>% 
+    select(-mean_AUC_diff) %>% 
+    rename(name=OTU) %>% 
+    inner_join(.,data,by="name") %>% 
+    mutate(rank_category = case_when(rank <= 10 ~ "top10",
+                                     rank <= 20 ~ "top20",
+                                     TRUE ~ "other"))
+
+  if(tax_level %in% c("otu","asv")){
+    plot_data <- plot_data %>% 
+      mutate(Taxonomy=sapply(strsplit(Taxonomy,"();"), `[`, 6)) %>% 
+      mutate(Taxonomy=sapply(strsplit(Taxonomy,"\\("), `[`, 1))
+  }
+
+  d <- data.frame(x=seq(0,1,0.1),y=seq(0,1,0.1))
+
+  max_relabun <- max(max(plot_data$cancer),max(plot_data$normal))
+  #axis_max <- round(max_relabun + 0.05,digits=1)
+  axis_max <- max_relabun + 0.01
+
+  plotly <- plot_ly(plot_data) %>% 
+    add_trace(type="scatter",
+              mode="markers",
+              x= ~normal,
+              y = ~cancer, 
+              color= ~factor(rank_category,levels=c("top10","top20","other")),
+              hoverinfo="text",
+              text = ~paste("</br>",Taxonomy,"(",name,")",
+                            "</br> SRN RelAbund: ",round(cancer,digits=2),
+                            "</br> Normal RelAbund: ",round(normal,digits=2)),
+              colors=c("red","cornflowerblue","grey"),
+              opacity=0.3,
+              marker = list(size = 10))
+  final_plotly <- plotly %>% layout(title=tax_level,
+                                    xaxis = list(title = "Median Non-Zero Relative Abundance in Normal Samples",range=c(0,axis_max),showline=TRUE,
+                                                 zerolinecolor = toRGB("grey70")), 
+                                    yaxis = list(title = "Median Non-Zero Relative Abundance in SRN Samples",range=c(0,axis_max),showline=TRUE,
+                                                 zerolinecolor = toRGB("grey70"))) %>% 
+        add_trace(data=d,x=~x,y=~y,type="scatter",mode="lines",color=I('grey'),name="y=x line")
+  return(final_plotly)
+}
+
+phylum_abund <- get_abund("phylum","../data/phylum/crc.relabund",metadata)
+plot_abund(phylum_abund,"phylum",ranked_tax)
+
+class_abund <- get_abund("class","../data/class/crc.relabund",metadata)
+plot_abund(class_abund,"class",ranked_tax)
+
+order_abund <- get_abund("order","../data/order/crc.relabund",metadata)
+plot_abund(order_abund,"order",ranked_tax)
+
+family_abund <- get_abund("family","../data/family/crc.relabund",metadata)
+plot_abund(family_abund,"family",ranked_tax)
+
+genus_abund <- get_abund("genus","../data/genus/crc.relabund",metadata)
+plot_abund(genus_abund,"genus",ranked_tax)
+
+otu_abund <- get_abund("otu","../data/otu/crc.relabund",metadata)
+plot_abund(otu_abund,"otu",ranked_tax)
+
+asv_abund <- get_abund("asv","../data/asv/crc.relabund",metadata)
+plot_abund(asv_abund,"asv",ranked_tax)
+```
+
+## Sensitivity at 90% Specificity {#sens-90spec}
+
+```{r senspec,echo=F}
+
+# tax_level <- "phylum"
+# senspec <- read_csv(paste0("../data/process/senspec-",tax_level,"-rf.csv"),
+#                     col_types = cols(level = col_character(),.default = col_double()))
+# 
+# # senspec %>% 
+# #   ggplot(aes(x=threshold,y=sensitivity)) +
+# #     stat_summary(fun.y=mean,
+# #                  fun.ymin=function(x) mean(x) - sd(x),
+# #                  fun.ymax=function(x) mean(x) + sd(x),
+# #                  geom="pointrange")
+# # senspec %>% 
+# #   ggplot(aes(x=threshold)) +
+# #     stat_summary(aes(y=sensitivity),
+# #                  fun.data=median_hilow,
+# #                  geom="pointrange",size=0.2,color="blue") +
+# #     stat_summary(aes(y=specificity),
+# #                  fun.data=median_hilow,
+# #                  geom="pointrange",size=0.2,color="red")
+# #     theme_bw()
+#     
+# get_sens_vals <- function(tax_level,spec){
+#   data <-  read_csv(paste0("../data/process/senspec-",tax_level,"-rf.csv"),
+#                     col_types = cols(level = col_character(),.default = col_double()))
+#   filtered_data <- NULL
+#   for(i in 1:100){
+#     idata <- data %>% 
+#       filter(seed == i) %>% 
+#       mutate(specificity=round(specificity,digits=2)) %>% 
+#       filter(specificity >= spec) 
+#     filtered_data <- bind_rows(filtered_data,idata[1,])
+#   }
+#   if(nrow(filtered_data) < 100){
+#     stop("missing data")
+#   }
+#   return(filtered_data)
+# }
+# 
+# sens_vals <- get_sens_vals("phylum",0.9) %>% 
+#   bind_rows(.,get_sens_vals("class",0.9)) %>% 
+#   bind_rows(.,get_sens_vals("order",0.9)) %>% 
+#   bind_rows(.,get_sens_vals("family",0.9)) %>% 
+#   bind_rows(.,get_sens_vals("genus",0.9)) %>% 
+#   bind_rows(.,get_sens_vals("otu",0.9)) %>% 
+#   bind_rows(.,get_sens_vals("asv",0.9)) %>% 
+#   mutate(level=factor(level,levels = levels,labels=levels_names))
+# 
+# 
+# ggplot(sens_vals,aes(x=level,y=sensitivity,color=level)) +
+#   geom_jitter(width=0.2,alpha=0.5) +
+#   stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+#   theme_bw() +
+#   xlab("") + ylab("Sensitivity") +
+#   theme(axis.text = element_text(size=14),
+#         axis.title = element_text(size=14),
+#         legend.position = "none",
+#         panel.grid.major.x = element_blank()) +
+#   scale_y_continuous(breaks=seq(0,1,0.1),labels = scales::percent_format(accuracy = 1)) +
+#   ggtitle("Sensitivity at 90% Specificity")
+# 
+# ggplot(sens_vals,aes(x=level,y=sensitivity,color=level)) +
+#   geom_jitter(width=0.2,alpha=0.5) +
+#   stat_summary(fun.data=median_hilow, fun.args=0.5, geom="pointrange", color="black") +
+#   theme_bw() +
+#   xlab("") + ylab("Sensitivity") +
+#   theme(axis.text = element_text(size=14),
+#         axis.title = element_text(size=14),
+#         legend.position = "none",
+#         panel.grid.major.x = element_blank()) +
+#   scale_y_continuous(breaks=seq(0,1,0.1),labels = scales::percent_format(accuracy = 1)) +
+#   ggtitle("Sensitivity at 90% Specificity")
+
+# all_senspec <- read_csv(paste0("../data/process/senspec-phylum-rf.csv"),
+#                         col_types = cols(level = col_character(),.default = col_double())) %>% 
+#   bind_rows(read_csv(paste0("../data/process/senspec-class-rf.csv"),
+#                      col_types = cols(level = col_character(),.default = col_double()))) %>% 
+#   bind_rows(read_csv(paste0("../data/process/senspec-order-rf.csv"),
+#                      col_types = cols(level = col_character(),.default = col_double()))) %>% 
+#   bind_rows(read_csv(paste0("../data/process/senspec-family-rf.csv"),
+#                      col_types = cols(level = col_character(),.default = col_double()))) %>% 
+#   bind_rows(read_csv(paste0("../data/process/senspec-genus-rf.csv"),
+#                      col_types = cols(level = col_character(),.default = col_double()))) %>% 
+#   bind_rows(read_csv(paste0("../data/process/senspec-otu-rf.csv"),
+#                      col_types = cols(level = col_character(),.default = col_double()))) %>% 
+#   bind_rows(read_csv(paste0("../data/process/senspec-asv-rf.csv"),
+#                      col_types = cols(level = col_character(),.default = col_double())))
+# all_senspec %>% 
+#   group_by(threshold,level) %>% 
+#   summarise(mean_sensitivity=mean(sensitivity),
+#             sd_sensitivity=sd(sensitivity),
+#             mean_specificity=mean(specificity),
+#             sd_specificty=sd(specificity))
+#   ggplot(aes(x=1-specificity,y=sensitivity,color=level)) +
+#     geom_path()
+
+get_level_senspec <- function(tax_level,spec_val){
+  l_probs <- read_csv(paste0("../data/process/probability-",tax_level,"-rf.csv"),
+                    col_types = cols( actual = col_character(),
+                                      level = col_character(),
+                                      .default = col_double()))
+
+  l_info <- NULL
+  aucs <- NULL
+  for(i in 1:100){
+    seed_probs <- l_probs %>% 
+      filter(seed == i)
+
+    response <- seed_probs$actual
+    predictor <- seed_probs$cancer
+
+    # make roc curve to get sensitivity/specificities
+    seed_roc <- roc(response,predictor,levels=c("normal","cancer"),direction="<")
+    #plot(seed_roc)
+
+    # record sensitivity/specificity values
+    # round specificities so we can find desired spec val
+    seed_spec <- round(seed_roc$specificities,digits=2)
+    seed_sens <- seed_roc$sensitivities
+
+    # make a table of all values
+    seed_info <- tibble(seed=i,
+                        sensitivity=seed_sens,
+                        specificity=seed_spec)
+
+    # get just sensitivity at desired specificity
+    spec_info <- seed_info %>%
+      filter(specificity == spec_val)
+
+    #check if there are no values at the cutoff
+    if(nrow(spec_info) == 0){
+      above <- seed_info %>%
+        filter(specificity > spec_val) %>%
+        top_n(-1,wt=specificity) %>%
+        top_n(1,wt=sensitivity)
+
+      below <- seed_info %>%
+        filter(specificity < spec_val) %>%
+        top_n(1,wt=specificity) %>%
+        top_n(-1,wt=sensitivity)
+
+      if(above$sensitivity == below$sensitivity){
+        # since the sensitivity above and below the target are the same,
+        # then assign the same sensitivity value to the target
+        spec_info <- tibble(seed=i,
+                            sensitivity=above$sensitivity,
+                            specificity=spec_val)
+
+      }else{
+        #find the slope of the line between the two points and the y-intercept
+        slope = ((above$sensitivity-below$sensitivity)/(above$specificity-below$specificity))
+        y_int = below$sensitivity - slope*below$specificity
+
+        # calculate the sensitivity for the target between the points 
+        # based on the calculated slope and intercept
+        sens_val = slope*spec_val + y_int
+
+        spec_info <- tibble(seed=i,
+                            sensitivity=sens_val,
+                            specificity=spec_val)
+      }
+    }else{
+      # get max sensitivity value for the cutoff
+      sub_info <- spec_info %>% 
+        top_n(n=1,wt=sensitivity)
+    }
+
+    l_info <- bind_rows(l_info,sub_info) %>% 
+      mutate(level = tax_level)
+
+  }
+
+  return(l_info)
+
+}
+
+library(mosaic)
+perm_p_value_senspec <- function(data, level1, level2){
+  # subset results to select AUC and condition columns and
+  # filter to only the two conditions of interest
+  test_subsample <- data %>%
+    mutate(level = as.character(level)) %>% 
+    select(level,sensitivity) %>%
+    filter((level == level1 | level == level2))
+  # quantify the absolute value of the difference in mean AUC between the two conditions (observed difference)
+  obs <- abs(diff(mosaic::mean(sensitivity ~ level, data = test_subsample)))
+  # use mosaic to shuffle condition labels and calculate difference in mean AUC - repeat 10,000 times
+  null <- do(10000) * diff(mosaic::mean(sensitivity ~ shuffle(level), data = test_subsample))
+  n <- length(null[,1]) #number of shuffled calculations
+  # r = #replications at least as extreme as observed effect
+  r <- sum(abs(null[,1]) >= obs) #count number of shuffled differences as big or bigger than observed difference
+  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
+  p.value=(r+1)/(n+1)
+  return(p.value)
+}
+
+all_sensspec90<- get_level_senspec("phylum",0.9) %>% 
+  bind_rows(get_level_senspec("class",0.9)) %>% 
+  bind_rows(get_level_senspec("order",0.9)) %>% 
+  bind_rows(get_level_senspec("family",0.9)) %>% 
+  bind_rows(get_level_senspec("genus",0.9)) %>% 
+  bind_rows(get_level_senspec("otu",0.9)) %>% 
+  bind_rows(get_level_senspec("asv",0.9)) %>% 
+  mutate(level = factor(level,levels=rev(levels),labels=rev(levels_names)))
+
+senspec_pvals <- expand_grid(x=1:length(levels_names),
+                                 y=1:length(levels_names),
+                                 ) %>%
+  filter(x < y) %>%
+  mutate(level1 = levels_names[x], 
+         level2 = levels_names[y]) %>%
+  select(-x, -y) %>%
+  group_by(level1, level2) %>%
+  summarize(p_value = perm_p_value_senspec(all_sensspec90,level1,level2),
+            .groups="drop")
+detach("package:mosaic", unload=TRUE)
+
+#testing
+# spec_val = 0.9
+# vals <- tibble(y=rev(c(0.1,0.15,0.2,0.25,0.3)),
+#                x=c(0.84,0.88,0.88,0.92,0.94))
+# vals %>% 
+#   mutate(closeness=abs(spec_val - x)) %>% 
+#   filter(closeness == min(closeness))
+# 
+# 
+# above <- vals %>%
+#   filter(x > spec_val) %>%
+#   top_n(-1,wt=x) %>%
+#   top_n(1,wt=y)
+# 
+# below <- vals %>%
+#   filter(x < spec_val) %>%
+#   top_n(1,wt=x) %>%
+#   top_n(-1,wt=y)
+#       
+# slope = ((above$y-below$y)/(above$x-below$x))
+# y_int = below$y - slope*below$x
+# 
+# slope*spec_val + y_int
+
+#interesting examples
+#tax_level="class"
+#i=13
+#tax_level="family"
+#i=89
+
+#visuals
+# plot(seed_roc)
+# abline(v=below$specificity,col="grey")
+# abline(v=above$specificity,col="grey")
+# abline(v=spec_val,col="red")
+# abline(h=sens_val,col="blue")
+
+# ggplot(all_sensspec90,aes(x=level,y=sensitivity,color=level)) +
+#   geom_jitter(width=0.2,alpha=0.5) +
+#   stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+#   theme_bw() +
+#   xlab("") + ylab("Sensitivity") +
+#   theme(axis.text = element_text(size=14),
+#         axis.title = element_text(size=14),
+#         legend.position = "none",
+#         panel.grid.major.x = element_blank()) +
+#   scale_y_continuous(breaks=seq(0,1,0.1),labels = scales::percent_format(accuracy = 1)) +
+#   ggtitle("Sensitivity at 90% Specificity")
+
+ggplot(all_sensspec90,aes(x=level,y=sensitivity,color=level)) +
+  geom_jitter(width=0.2,alpha=0.5,size=2,height = 0) +
+  stat_summary(fun.data=median_hilow, fun.args=0.5, geom="pointrange", color="black") +
+  theme_bw() +
+  xlab("") + ylab("Sensitivity") +
+  theme(axis.text = element_text(size=12),
+        axis.title = element_text(size=12),
+        legend.position = "none",
+        panel.grid.major.y = element_blank()) +
+        #panel.grid.major.x = element_blank()) +
+  coord_flip() +
+  scale_y_continuous(breaks=seq(0,1,0.1),labels = scales::percent_format(accuracy = 1)) +
+  #ggtitle("Sensitivity at 90% Specificity")
+  scale_color_manual(values=rev(colors2)) + 
+  annotate("text",x=levels_names,y=rep((max(all_sensspec90$sensitivity)+0.05),length(levels_names)),label=c("W","W","X","Y","Y","Y","Z")) 
+ggsave("./figures/sens_90spec.png",width = 4.5,height=5,units="in")
+
+```
+
+## Averaged ROC Curve {#avg-roc}
+
+```{r avg_roc}
+
+# args <- data.frame(data="./data/otu/input_data_preproc.csv",
+#           method="rf",
+#           seed=1,
+#           taxonomy="otu",
+#           outcome_colname="dx",
+#           outcome_value="cancer",
+#           hyperparams="./data/hyperparameters/rf_hyperparameters.csv")
+
+# for each seed, over a range of specified specificities, 
+# find the sensitivities
+# us ROC curve to calculate values. 
+get_level_senspec_avg <- function(tax_level){
+  l_probs <- read_csv(paste0("../data/process/probability-",tax_level,"-rf.csv"),
+                    col_types = cols( actual = col_character(),
+                                      level = col_character(),
+                                      .default = col_double()))
+
+  l_info <- NULL
+  aucs <- NULL
+  for(i in 1:100){
+    seed_probs <- l_probs %>% 
+      filter(seed == i)
+
+    response <- seed_probs$actual
+    predictor <- seed_probs$cancer
+
+    seed_roc <- roc(response,predictor,levels=c("normal","cancer"),direction="<")
+
+    seed_spec <- round(seed_roc$specificities,digits=2)
+    seed_sens <- seed_roc$sensitivities
+
+    #plot(seed_roc)
+
+    seed_info <- tibble(seed=i,
+                        sensitivity=seed_sens,
+                        specificity=seed_spec)
+
+    seed_auc <- tibble(seed=i,
+                       auc=as.numeric(auc(seed_roc)))
+
+    l_info <- bind_rows(l_info,seed_info)
+
+    aucs <- bind_rows(aucs,seed_auc)
+  }
+
+  l_info <- l_info %>% 
+    group_by(specificity) %>%
+    summarise(mean_sensitivity = mean(sensitivity),
+              sd_sensitivity = sd(sensitivity)) %>%
+    mutate(upper_sens = mean_sensitivity + sd_sensitivity,
+           lower_sens = mean_sensitivity - sd_sensitivity) %>% 
+    mutate(level = tax_level,
+           avg_auc = mean(aucs$auc))
+
+  return(l_info)
+
+}
+
+all_info <- get_level_senspec_avg("phylum") %>% 
+  bind_rows(get_level_senspec_avg("class")) %>% 
+  bind_rows(get_level_senspec_avg("order")) %>% 
+  bind_rows(get_level_senspec_avg("family")) %>% 
+  bind_rows(get_level_senspec_avg("genus")) %>% 
+  bind_rows(get_level_senspec_avg("otu")) %>% 
+  bind_rows(get_level_senspec_avg("asv")) %>% 
+  mutate(level=factor(level,levels=levels,labels=levels_names)) %>% 
+  mutate(upper_sens = case_when(upper_sens > 1 ~ 1,
+                                TRUE ~ upper_sens),
+         lower_sens = case_when(upper_sens < 0 ~ 0,
+                                TRUE ~ lower_sens))
+
+all_info %>% 
+  ggplot(aes(x=specificity,y=mean_sensitivity,
+             ymin=lower_sens,ymax=upper_sens)) +
+    geom_line(aes(color=level),alpha=0.7) +
+    geom_ribbon(aes(fill=level),alpha=0.2) +
+    coord_equal() +
+    geom_abline(intercept = 1,lty="dashed",color="grey50") +
+    #scale_x_continuous(expand = c(0,0),labels = scales::percent_format(accuracy = 1)) +
+    scale_y_continuous(expand = c(0,0),labels = scales::percent_format(accuracy = 1)) +
+    xlab("Specificity") +
+    ylab("Average Sensitivity") +
+    #facet_wrap(~level,ncol=3) +  
+    scale_x_reverse(expand = c(0,0),labels = scales::percent_format(accuracy = 1)) +
+    theme_bw() +
+    theme(legend.position = "top",
+          axis.text = element_text(size=11),
+          axis.title = element_text(size=12),
+          axis.text.x = element_text(vjust = -0.5),
+          axis.title.x = element_text(vjust= -0.5)) +
+    guides(color = guide_legend(nrow = 1,title=""),
+           fill = guide_legend(nrow = 1,title="")) +
+    scale_fill_manual(values=colors2) +
+    scale_color_manual(values=colors2)
+#ggsave("./figures/roc_all.png",width=6)
+
+ggplot(all_info,aes(x=specificity*100,y=mean_sensitivity*100,
+                    ymin=lower_sens*100,ymax=upper_sens*100)) +
+    geom_line(aes(color=level),alpha=0.8,lwd=1) +
+    geom_ribbon(aes(fill=level),alpha=0.2) +
+    coord_equal() +
+    geom_abline(intercept = 1,lty="dashed",color="grey50") +
+    #scale_x_continuous(expand = c(0,0)) +
+    scale_y_continuous(expand = c(0,0)) + #,labels = scales::percent_format(accuracy = 1)) +
+    xlab("Specificity (%)") +
+    ylab("Average Sensitivity (%)") +
+    facet_wrap(~level,nrow=2) +  
+    scale_x_reverse(expand = c(0,0)) + #,labels = scales::percent_format(accuracy = 1)) +  
+    theme_bw() +
+    theme(legend.position = "top",
+          axis.text.x = element_text(angle = -45,vjust=0.5,hjust=0),
+          axis.text = element_text(size=9),
+          axis.title = element_text(size=12)) +
+    guides(color = guide_legend(nrow = 1,title=""),
+           fill = guide_legend(nrow = 1,title="")) +
+    geom_text(aes(x = 25, y = 25, label = paste0("mean AUC\n",round(avg_auc,digits=3)))) + 
+    scale_fill_manual(values=colors2) +
+    scale_color_manual(values=colors2)
+ggsave("./figures/roc_by_level.png",width=7,height=5)
+
+```
+
+```{r misc_roc, include=F}
+# get_avg_senspec <- function(tax_level){
+#   l_senspec <- read_csv(paste0("../data/process/senspec-",tax_level,"-rf.csv"),
+#                       col_types = cols(level = col_character(),.default = col_double()))
+#   
+#   l_avg_senspec <- l_senspec %>% 
+#     group_by(threshold) %>% 
+#     summarise(mean_sensitivity = mean(sensitivity),
+#               sd_sensitivity = sd(sensitivity),
+#               mean_specificity = mean(specificity),
+#               sd_specificity = sd(specificity))  %>% 
+#     mutate(level = tax_level,
+#            sens_upper = mean_sensitivity + sd_sensitivity,
+#            sens_lower = mean_sensitivity - sd_sensitivity) 
+#   return(l_avg_senspec)
+# }
+#  
+# all_avg_senspec <- get_avg_senspec("phylum") %>% 
+#   bind_rows(get_avg_senspec("class")) %>% 
+#   bind_rows(get_avg_senspec("order")) %>% 
+#   bind_rows(get_avg_senspec("family")) %>% 
+#   bind_rows(get_avg_senspec("genus")) %>%
+#   bind_rows(get_avg_senspec("otu")) %>%
+#   bind_rows(get_avg_senspec("asv")) %>% 
+#   mutate(level=factor(level,levels=levels,labels=levels_names))
+# 
+# all_avg_senspec %>% 
+#   ggplot(aes(x=1-mean_specificity,y=mean_sensitivity,
+#              ymin=sens_lower,ymax=sens_upper)) +
+#     geom_line(aes(color=level),alpha=0.7) +
+#     geom_ribbon(aes(fill=level),alpha=0.2) +
+#     coord_equal() +
+#     geom_abline(slope=1,intercept = 0,lty="dashed",color="grey50") +
+#     scale_x_continuous(expand = c(0,0)) +
+#     scale_y_continuous(expand = c(0,0)) +
+#     xlab("1 - Averaged Specificity") +
+#     ylab("Averaged Sensitivity") +
+#     #facet_wrap(~level,ncol=3) +
+#     theme_bw() 
+# 
+# all_avg_senspec %>% 
+#   #filter(level %in% c("Family","Genus","OTU","ASV")) %>% 
+#   ggplot(aes(x=1-mean_specificity,y=mean_sensitivity,
+#              ymin=sens_lower,ymax=sens_upper)) +
+#     geom_line(aes(color=level),alpha=0.7) +
+#     geom_ribbon(aes(fill=level),alpha=0.2) +
+#     coord_equal() +
+#     geom_abline(slope=1,intercept = 0,lty="dashed",color="grey50") +
+#     scale_x_continuous(expand = c(0,0)) +
+#     scale_y_continuous(expand = c(0,0)) +
+#     xlab("1 - Averaged Specificity") +
+#     ylab("Averaged Sensitivity") +
+#     #facet_wrap(~level,ncol=3) +
+#     theme_bw()
+
+# all_avg_senspec %>% 
+#   ggplot(aes(x=1-mean_specificity,y=mean_sensitivity)) +
+#     geom_line(aes(color=level),lwd=1.2,alpha=0.7) +
+#     coord_equal() +
+#     geom_abline(slope=1,intercept = 0,lty="dashed",color="grey50") +
+#     scale_x_continuous(expand = c(0,0)) +
+#     scale_y_continuous(expand = c(0,0)) +
+#     xlab("1 - Averaged Specificity") +
+#     ylab("Averaged Sensitivity") +
+#     theme_bw()
+
+```

### 2021-08-13 paper/manuscript.Rmd
first full draft

@@
     keep_tex: true
-  word_document: default
+  word_document: 
+    reference_docx: word_formatting.docx
 geometry: margin=1.0in
@@
 
-## Introduction (\~250 words)
-
 Colorectal cancer is one of the most common cancers in men and women and a leading cause of cancer related deaths in the United States [@siegel2020].
@@
 Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs [@callahan2017].
-However, whether the additional resolution provided by ASVs is useful for ML classification is unclear {pat preprint?}.
+However, whether the additional resolution provided by ASVs is useful for ML classification is unclear {}.
 Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups.
@@
 This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.
-
-## Results (\~700 words)
 
@@
 Random forest (RF) was consistently the top performer at most taxonomic levels.
-Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1).
-Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU (p \< 0.01), but borderline equivalent to family (p = `r round(rf_pvals["family","asv"],digits=2)`) and genus (p = `r round(rf_pvals["genus","asv"],digits=2)`) levels.
+Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1A, Supplemental Figure 2).
+Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTUs (p \< 0.01), but borderline equivalent to family (p = `r round(rf_pvals["family","asv"],digits=2)`) and genus (p = `r round(rf_pvals["genus","asv"],digits=2)`) levels (Figure 1A).
 These results were consistent for ASVs generated with DADA2 [@callahan2016] (median AUROC: `r round(dada2_med_auc,digits=3)`).
-These results suggest that finer taxonomic resolution is not necessarily better for differentiating a persons cancer status from the microbiome.
+These results suggest that finer taxonomic resolution is not necessarily better for identifying individuals with SRNs based on microbiome composition.
 
-While AUROC values between models is a useful way to compare the overall model performance across all thresholds, they can be misleading [@lobo2008].
-Depending on the intended implementation of the model, one may want to optimize the true-positive rate(or sensitivity) over the false-postivite rate (or specificity).
-To further compare the model performance across taxonomic levels we compared the sensitivity of the models
-
--   on average, how many positives are missed with ASV?
+While comparing AUROC values between models is a useful way to assess the overall model performance, they summarize the performance across all thresholds and can be misleading since models with the same AUC can have different ROC curve shapes [@lobo2008].
+Depending on the intended implementation of the model, one may want to optimize the true-positive rate (or sensitivity) over the false-positive rate (or 1-specificity), or vice versa.
+In this case, the optimal model will detect as many true positives (people with SRNs) as possible.
+To further compare the model performance across taxonomic levels we compared the sensitivity of the models at a specificity of 90%.
+The highest sensitivity values were observed for Family, Genus, and OTU level data (Figure 1B), consistent with the AUROC results.
+Phylum, Class, Order, and ASV sensitivity values were all significantly lower than Family, Genus, and OTU sensitivity values (Figure 1B).
+This analysis further supports that finer resolution is does not improve SRN detection.
 
@@
 The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 3).
-While the resolution provided by ASVs is useful in
-
-|   **TODO: what are important taxa, patterns along levels (e.g fuso always in top )**
-
 Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods).
 As part of this step, features are removed that will not provide useful information to build the model.
-For example, strongly correlated features provide the same information to build the model and thus can be collapsed.
+For example, perfectly correlated features provide the same information to build the model and thus can be collapsed.
 Additionally, features with zero or near-zero variance will not help the model differentiate groups and thus can be removed.
@@
 At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).
+While the resolution provided by ASVs is useful in certain contexts {}, these results suggest that the resolution is too fine for use in machine learning classification of SRNs based on microbiome composition.
 
-|   **TODO: check why ASVs were removed in preprocessing (nzv?)**
-|   (depending on answer can discuss why features were removed - e.g. only in one or a few samples)
+A look into the most important taxa at each level for classifying samples reveals some nesting where several genera and their higher taxonomic classifications are in the top 10 most important taxa (Supplemental Figure 4).
+For example, the genus *Gemella* is an important taxa at the genus and OTU levels and its higher classifications are also important (*Firmicutes* > *Bacilli* > *Bacillales* > *Bacillales Incertae Sedis XI* > *Gemella*).
+*Fusobacterium* displays a similar pattern, except that the family level classification (*Fusobacteriaceae*) importance is ranked 16th out of 54 families, a little beyond the top 10.
+In the case of unclassified *Lachnospiraceae*, there are several OTUs with this label that are in the top 10, however at the Genus level this taxon is ranked lower in importance (21st out of 115 genera) suggesting there may be some benefit to separating different taxonomic groupings within *Lachnospiraceae*.
 
-Mikropml{}
-
-## Discussion/Conclusions (\~250 words)
-
-These results demonstrate that consideration of the appropriate taxonomic resolution for utilizing the microbiome as a predictive tool.
-In general, we found that finer resolution (e.g.
-ASVs) does not add additional sensitivity to model prediction and at the ASV level actually impedes model performance due to the sparsity of shared taxa.
-
--   Utilizing genus or family level data could allow for merging data generated from different 16S regions or sequencing platforms. (is this true?)
-
-While predictive models utilizing microbiome data alone are not that great, microbiome in combination with FIT results are better than either alone {}.
-This analysis wanted to quantify difference between
+These results demonstrate that consideration of the appropriate taxonomic resolution for utilizing the microbiome as a predictive tool is warranted.
+In general, we found that finer taxonomic resolution (e.g.
+OTU and ASV) does not add additional sensitivity to predicting SRNs based on microbiome composition.
+Additionally, at the ASV level the fine resolution actually impedes model performance due to the sparsity of shared taxa and leads to decreased model performance.
+The tendancy for ASV level annotation to split single bacterial genomes into multiple taxa {Pat mSphere 2021} could also be a contributing factor to the sparsity of shared taxa.
+Overall, either Family, Genus, or OTU level taxonomy appear to perform equally for predicting subjects with SRNs based on the composition of the gut microbiome.
+A potential benefit of utilizing genus or family level data could be that it may allow for merging data generated from different 16S regions or sequencing platforms.
 
@@
 To compare performance between taxonomic levels and models, P values were calculated as previously described {begum}.
-Another output from the mikropml package is the permuted feature importance which is quantified by iteratively permuting each feature in the model and assessing the change in model performance.
+To compare sensitivity at 90% specificity, probabilities on the test dataset were collected for each seed and used to construct ROC curves (R pROC::roc).
+From the ROC curves The sensitivity at a specificity of 90% was pulled for each seed.
+An optional output from the mikropml package is the permuted feature importance which is quantified by iteratively permuting each feature in the model and assessing the change in model performance.
 Features are presumed to be important if the performance of the model, measured by the AUROC, decreases when that feature is permuted.
@@
 
+\newpage
+
 ![](../exploratory/figures/roc_by_level.png){height="50%"}
@@
 **Supplemental Figure 2: Averaged ROC curves** ROC curves averaged across the 100 iterations of the model. The shaded region represents the standard deviation form the mean.
+
+\newpage
 

### 2021-08-13 paper/manuscript.tex
first full draft

@@
 
-\hypertarget{introduction-250-words}{%
-\subsection{Introduction (\textasciitilde250
-words)}\label{introduction-250-words}}
-
 Colorectal cancer is one of the most common cancers in men and women and
@@
 the additional resolution provided by ASVs is useful for ML
-classification is unclear \{pat preprint?\}. Since ML classification
-relies on consistent differences between groups, its possible that the
-resolution at the ASV level is too individualized to accurately
-differentiate groups. Topçuoğlu \emph{et al}
+classification is unclear \{\}. Since ML classification relies on
+consistent differences between groups, its possible that the resolution
+at the ASV level is too individualized to accurately differentiate
+groups. Topçuoğlu \emph{et al}
 (\protect\hyperlink{ref-topuxe7uolu2020}{4}) recently demonstrated
@@
 classifier for detecting SRNs from microbiome data.
-
-\hypertarget{results-700-words}{%
-\subsection{Results (\textasciitilde700
-words)}\label{results-700-words}}
 
@@
 0.687), genus (median AUROC: 0.686), and OTU (median AUROC: 0.698) level
-data with no significant difference between the three (Figure 1).
-Performance with ASVs (median AUROC: 0.676) was significantly lower than
-OTU (p \textless{} 0.01), but borderline equivalent to family (p = 0.06)
-and genus (p = 0.05) levels. These results were consistent for ASVs
-generated with DADA2 (\protect\hyperlink{ref-callahan2016}{5}) (median
-AUROC: 0.66). These results suggest that finer taxonomic resolution is
-not necessarily better for differentiating a persons cancer status from
-the microbiome.
+data with no significant difference between the three (Figure 1A,
+Supplemental Figure 2). Performance with ASVs (median AUROC: 0.676) was
+significantly lower than OTUs (p \textless{} 0.01), but borderline
+equivalent to family (p = 0.06) and genus (p = 0.05) levels (Figure 1A).
+These results were consistent for ASVs generated with DADA2
+(\protect\hyperlink{ref-callahan2016}{5}) (median AUROC: 0.66). These
+results suggest that finer taxonomic resolution is not necessarily
+better for identifying individuals with SRNs based on microbiome
+composition.
 
-While AUROC values between models is a useful way to compare the overall
-model performance across all thresholds, they can be misleading
-(\protect\hyperlink{ref-lobo2008}{6}). Depending on the intended
-implementation of the model, one may want to optimize the true-positive
-rate(or sensitivity) over the false-postivite rate (or specificity). To
+While comparing AUROC values between models is a useful way to assess
+the overall model performance, they summarize the performance across all
+thresholds and can be misleading since models with the same AUC can have
+different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{6}).
+Depending on the intended implementation of the model, one may want to
+optimize the true-positive rate (or sensitivity) over the false-positive
+rate (or 1-specificity), or vice versa. In this case, the optimal model
+will detect as many true positives (people with SRNs) as possible. To
 further compare the model performance across taxonomic levels we
-compared the sensitivity of the models
-
-\begin{itemize}
-\tightlist
-\item
-  on average, how many positives are missed with ASV?
-\end{itemize}
+compared the sensitivity of the models at a specificity of 90\%. The
+highest sensitivity values were observed for Family, Genus, and OTU
+level data (Figure 1B), consistent with the AUROC results. Phylum,
+Class, Order, and ASV sensitivity values were all significantly lower
+than Family, Genus, and OTU sensitivity values (Figure 1B). This
+analysis further supports that finer resolution is does not improve SRN
+detection.
 
@@
 taxa respectively are only present in less than 25\% of samples
-(Supplemental Figure 3). While the resolution provided by ASVs is useful
-in
+(Supplemental Figure 3). Of note, the ML pipeline includes a
+pre-processing step that occurs prior to training and classifying the ML
+models (methods). As part of this step, features are removed that will
+not provide useful information to build the model. For example,
+perfectly correlated features provide the same information to build the
+model and thus can be collapsed. Additionally, features with zero or
+near-zero variance will not help the model differentiate groups and thus
+can be removed. Interestingly, despite starting with 104106 features at
+the ASV level, only 478 (0.5\%) remained after pre-processing. At the
+OTU level, 20079 of the 705 features (3.5\%) remained after
+preprocessing (Table 1). While the resolution provided by ASVs is useful
+in certain contexts \{\}, these results suggest that the resolution is
+too fine for use in machine learning classification of SRNs based on
+microbiome composition.
 
-~~\textbf{TODO: what are important taxa, patterns along levels (e.g fuso
-always in top )}
-
-Of note, the ML pipeline includes a pre-processing step that occurs
-prior to training and classifying the ML models (methods). As part of
-this step, features are removed that will not provide useful information
-to build the model. For example, strongly correlated features provide
-the same information to build the model and thus can be collapsed.
-Additionally, features with zero or near-zero variance will not help the
-model differentiate groups and thus can be removed. Interestingly,
-despite starting with 104106 features at the ASV level, only 478 (0.5\%)
-remained after pre-processing. At the OTU level, 20079 of the 705
-features (3.5\%) remained after preprocessing (Table 1).
-
-~~\textbf{TODO: check why ASVs were removed in preprocessing (nzv?)}\\
-\hspace*{0.333em}\hspace*{0.333em}(depending on answer can discuss why
-features were removed - e.g.~only in one or a few samples)
-
-Mikropml\{\}
-
-\hypertarget{discussionconclusions-250-words}{%
-\subsection{Discussion/Conclusions (\textasciitilde250
-words)}\label{discussionconclusions-250-words}}
+A look into the most important taxa at each level for classifying
+samples reveals some nesting where several genera and their higher
+taxonomic classifications are in the top 10 most important taxa
+(Supplemental Figure 4). For example, the genus \emph{Gemella} is an
+important taxa at the genus and OTU levels and its higher
+classifications are also important (\emph{Firmicutes} \textgreater{}
+\emph{Bacilli} \textgreater{} \emph{Bacillales} \textgreater{}
+\emph{Bacillales Incertae Sedis XI} \textgreater{} \emph{Gemella}).
+\emph{Fusobacterium} displays a similar pattern, except that the family
+level classification (\emph{Fusobacteriaceae}) importance is ranked 16th
+out of 54 families, a little beyond the top 10. In the case of
+unclassified \emph{Lachnospiraceae}, there are several OTUs with this
+label that are in the top 10, however at the Genus level this taxon is
+ranked lower in importance (21st out of 115 genera) suggesting there may
+be some benefit to separating different taxonomic groupings within
+\emph{Lachnospiraceae}.
 
 These results demonstrate that consideration of the appropriate
-taxonomic resolution for utilizing the microbiome as a predictive tool.
-In general, we found that finer resolution (e.g.~ASVs) does not add
-additional sensitivity to model prediction and at the ASV level actually
-impedes model performance due to the sparsity of shared taxa.
-
-\begin{itemize}
-\tightlist
-\item
-  Utilizing genus or family level data could allow for merging data
-  generated from different 16S regions or sequencing platforms. (is this
-  true?)
-\end{itemize}
-
-While predictive models utilizing microbiome data alone are not that
-great, microbiome in combination with FIT results are better than either
-alone \{\}. This analysis wanted to quantify difference between
+taxonomic resolution for utilizing the microbiome as a predictive tool
+is warranted. In general, we found that finer taxonomic resolution
+(e.g.~OTU and ASV) does not add additional sensitivity to predicting
+SRNs based on microbiome composition. Additionally, at the ASV level the
+fine resolution actually impedes model performance due to the sparsity
+of shared taxa and leads to decreased model performance. The tendancy
+for ASV level annotation to split single bacterial genomes into multiple
+taxa \{Pat mSphere 2021\} could also be a contributing factor to the
+sparsity of shared taxa. Overall, either Family, Genus, or OTU level
+taxonomy appear to perform equally for predicting subjects with SRNs
+based on the composition of the gut microbiome. A potential benefit of
+utilizing genus or family level data could be that it may allow for
+merging data generated from different 16S regions or sequencing
+platforms.
 
@@
 To compare performance between taxonomic levels and models, P values
-were calculated as previously described \{begum\}. Another output from
-the mikropml package is the permuted feature importance which is
-quantified by iteratively permuting each feature in the model and
-assessing the change in model performance. Features are presumed to be
-important if the performance of the model, measured by the AUROC,
-decreases when that feature is permuted. Ranking of feature importance
-was determined by ordering the features based on the change in AUROC
-where features with a larger decrease in AUROC are ranked higher in
-importance.
+were calculated as previously described \{begum\}. To compare
+sensitivity at 90\% specificity, probabilities on the test dataset were
+collected for each seed and used to construct ROC curves (R pROC::roc).
+From the ROC curves The sensitivity at a specificity of 90\% was pulled
+for each seed. An optional output from the mikropml package is the
+permuted feature importance which is quantified by iteratively permuting
+each feature in the model and assessing the change in model performance.
+Features are presumed to be important if the performance of the model,
+measured by the AUROC, decreases when that feature is permuted. Ranking
+of feature importance was determined by ordering the features based on
+the change in AUROC where features with a larger decrease in AUROC are
+ranked higher in importance.
 
@@
 
+\newpage
+
 \includegraphics[width=\textwidth,height=0.5\textheight]{../exploratory/figures/roc_by_level.png}
@@
 standard deviation form the mean.
+
+\newpage
 

### 2021-08-25 code/R/figures/AUC_all_model_all_level.R
all model level auc figure

@@ -1,0 +1,82 @@
+library(tidyverse)
+
+### FUNCTIONS ########
+read_combined <- function(tax_level,model){
+  read_csv(paste0("data/process/combined-",tax_level,"-",model,".csv"),
+           col_types = c(method = col_character(),.default=col_double()),
+           na = c("","NA","NaN")) %>% #some of the Pos/Neg Pred Values were NaN
+    mutate(level=tax_level) 
+}
+
+### VARIABLES ########
+models <- c("rf","glmnet","xgbTree","svmRadial","rpart2")
+levels <- c("phylum","class","order","family","genus","otu","asv")
+levels_names <- c("Phylum","Class","Order","Family","Genus","OTU","ASV")
+pal <- park_palette("Everglades",length(models))
+names(pal) <- models
+pal2 <- park_palette("Everglades",length(models))
+names(pal2) <- c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")
+
+# Generate dataframe of all results and tables of the median cv and test AUCs
+full_df <- NULL
+auc_df <- NULL
+
+
+
+for( m in models ){
+  model_df <- map_dfr(levels, read_combined, model = m)
+  
+  full_df <- bind_rows(full_df,model_df)
+  
+  m_auc.df <- select(model_df,cv_metric_AUC,AUC,method,level) %>% 
+    rename(test_AUC=AUC) %>% 
+    pivot_longer(cols = c("cv_metric_AUC","test_AUC"),names_to="auc_type",values_to="AUC")
+  
+  m_auc.df$level <- factor(m_auc.df$level,levels=c("phylum","class","order","family","genus","otu","asv"))
+  
+  auc_df <- bind_rows(auc_df,m_auc.df)
+  
+  cv_med <- m_auc.df %>%
+    filter(auc_type == "cv_metric_AUC") %>%
+    group_by(level) %>% 
+    summarise(!!m := median(AUC),.groups = 'drop') 
+  
+  t_med <- m_auc.df %>% 
+    filter(auc_type == "test_AUC") %>%
+    group_by(level) %>% 
+    summarise(!!m := median(AUC),.groups = 'drop') 
+  
+  cv_table <- inner_join(cv_table,cv_med,by="level")
+  t_table <- inner_join(t_table,t_med,by="level")
+}
+
+write_csv(t_table,"./figures/median_test_auc.csv")
+
+#compare two models levels within a taxonomic level
[…]

### 2021-08-25 code/R/figures/AUC_all_model_all_level.R
fix errors-all model level auc figure

@@ -1,11 +1,12 @@
 library(tidyverse)
+library(nationalparkcolors)
 
 ### FUNCTIONS ########
 read_combined <- function(tax_level,model){
   read_csv(paste0("data/process/combined-",tax_level,"-",model,".csv"),
            col_types = c(method = col_character(),.default=col_double()),
            na = c("","NA","NaN")) %>% #some of the Pos/Neg Pred Values were NaN
-    mutate(level=tax_level) 
+    mutate(level=tax_level)
 }
 
 ### VARIABLES ########
@@ -17,66 +18,37 @@ names(pal) <- models
 pal2 <- park_palette("Everglades",length(models))
 names(pal2) <- c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")
 
+### MAIN ########
+
 # Generate dataframe of all results and tables of the median cv and test AUCs
 full_df <- NULL
 auc_df <- NULL
 
-
-
 for( m in models ){
   model_df <- map_dfr(levels, read_combined, model = m)
-  
+
   full_df <- bind_rows(full_df,model_df)
-  
-  m_auc.df <- select(model_df,cv_metric_AUC,AUC,method,level) %>% 
-    rename(test_AUC=AUC) %>% 
-    pivot_longer(cols = c("cv_metric_AUC","test_AUC"),names_to="auc_type",values_to="AUC")
-  
-  m_auc.df$level <- factor(m_auc.df$level,levels=c("phylum","class","order","family","genus","otu","asv"))
-  
-  auc_df <- bind_rows(auc_df,m_auc.df)
-  
-  cv_med <- m_auc.df %>%
-    filter(auc_type == "cv_metric_AUC") %>%
-    group_by(level) %>% 
-    summarise(!!m := median(AUC),.groups = 'drop') 
-  
-  t_med <- m_auc.df %>% 
-    filter(auc_type == "test_AUC") %>%
-    group_by(level) %>% 
-    summarise(!!m := median(AUC),.groups = 'drop') 
-  
-  cv_table <- inner_join(cv_table,cv_med,by="level")
-  t_table <- inner_join(t_table,t_med,by="level")
 }
 
-write_csv(t_table,"./figures/median_test_auc.csv")
-
-#compare two models levels within a taxonomic level
-p.table.level <- read_csv("../analysis/pvalues_by_level.csv",
-                          col_types = c(p_value = col_double(),.default=col_character()))
[…]

### 2021-08-25 code/R/figures/prevalence.R
prevalence figure code

@@ -1,0 +1,70 @@
+library(tidyverse)
+
+### VARIABLES ########
+levels <- c("phylum","class","order","family","genus","otu","asv")
+levels_names <- c("Phylum","Class","Order","Family","Genus","OTU","ASV")
+
+colors <- c("#D04C44","#E7834E","#E6A444","#829E88","#83BCC3","#7989A4","#B29ABA")
+names(colors) <- levels
+
+colors2 <- colors
+names(colors2) <- levels_names
+
+### MAIN ########
+
+#read in prevalence table
+prevalence <- read_csv("analysis/prevalence_by_level.csv") %>%
+  mutate(level = factor(level,levels=levels))
+
+tax_count <- prevalence %>%
+  filter(kept==1) %>%
+  group_by(level) %>%
+  mutate(numOTU = n_distinct(name)) %>%
+  select(level,numOTU) %>%
+  distinct()
+
+pct_bins <- NULL
+level_nOTU <- NULL
+for(l in levels){
+  sub_data <- prevalence %>%
+    filter(level==l,kept==1)
+  hist <- hist(sub_data$total_prevalence,breaks = seq(0,1,0.1),plot=FALSE)
+
+  nOTU <- sub_data %>%
+    select(name) %>%
+    n_distinct()
+
+  level_nOTU <- bind_rows(level_nOTU,
+                          data.frame(l,nOTU))
+
+  pct_bins <- pct_bins %>%
+    bind_rows(data.frame(level=l,
+                         prev=hist$breaks[-1],
+                         count=hist$counts,
+                         pct=hist$counts/nOTU))
+}
+
+level_nOTU <- level_nOTU %>%
+  mutate(level_name = levels_names)
+labs <- paste0(level_nOTU$level_name," (",level_nOTU$nOTU,")")
+pal <- colors2
+names(pal) <- labs
+
+pct_bins %>%
+  mutate(level=factor(level,levels=levels,labels=labs)) %>%
+  ggplot(aes(x=prev*100,y=pct,fill=level)) +
+  geom_bar(stat = "identity") +
+  facet_wrap(~level,ncol=2) +
+  scale_y_continuous(expand=c(0,0),limits=c(0,0.85),labels = scales::percent_format()) +
+  scale_x_continuous(expand=c(0,0),breaks = seq(0,100,10)) +
+  xlab("Percent of Samples") +
+  ylab("Percent of Taxa") +
+  theme_bw() +
+  theme(legend.position = "none",
[…]

### 2021-08-25 code/R/figures/rf_AUC_all_level.R
rf AUC figure

@@ -1,0 +1,67 @@
+library(tidyverse)
+
+### FUNCTIONS ########
+read_combined <- function(tax_level,model){
+  read_csv(paste0("data/process/combined-",tax_level,"-",model,".csv"),
+           col_types = c(method = col_character(),.default=col_double()),
+           na = c("","NA","NaN")) %>% #some of the Pos/Neg Pred Values were NaN
+    mutate(level=tax_level)
+}
+
+### VARIABLES ########
+
+levels <- c("phylum","class","order","family","genus","otu","asv")
+levels_names <- c("Phylum","Class","Order","Family","Genus","OTU","ASV")
+
+colors <- c("#D04C44","#E7834E","#E6A444","#829E88","#83BCC3","#7989A4","#B29ABA")
+names(colors) <- levels
+
+colors2 <- colors
+names(colors2) <- levels_names
+
+### MAIN ########
+
+rf_auc_df <- map_dfr(levels, read_combined, model = "rf") %>%
+  select(cv_metric_AUC,AUC,method,level) %>%
+  rename(test_AUC=AUC) %>%
+  pivot_longer(cols = c("cv_metric_AUC","test_AUC"),names_to="auc_type",values_to="AUC") %>%
+  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+                      labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV"))) %>%
+  filter(auc_type=="test_AUC")
+
+#compare two taxonomic levels within a model type
+rf_stats <- read_csv("analysis/pvalues_by_model.csv",
+                          col_types = c(p_value = col_double(),.default=col_character())) %>%
+  mutate(level1=factor(level1,levels=c("phylum","class","order","family","genus","otu","asv"),
+                       labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV")),
+         level2=factor(level2,levels=c("phylum","class","order","family","genus","otu","asv"),
+                       labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV"))) %>%
+  filter(model == "rf") %>%
+  select(-model) %>%
+  filter(p_value > 0.05)
+if( nrow(rf_stats) != 5){ stop("problem with stats)")}
+
+# create plot
+rf_auc_df %>%
+  ggplot(aes(x=factor(level,levels=c("ASV", "OTU","Genus","Family","Order","Class", "Phylum")),y=AUC)) +
[…]

### 2021-08-25 code/R/figures/top_10_important_tax.R
top 10 imp figure

@@ -1,0 +1,94 @@
+library(tidyverse)
+library(glue)
+library(ggtext)
+library(cowplot)
+
+### FUNCTIONS ########
+
+read_rf_imp <- function(tax_level){
+  read_csv(paste0("data/process/importance-",tax_level,"-rf.csv"),
+           col_types = c(perf_metric = col_double(),
+                         perf_metric_diff = col_double(),
+                         seed = col_double(),
+                         .default = col_character())) %>%
+  mutate(level=tax_level)
+}
+
+read_tax <- function(tax_level){
+  read_tsv(paste0("data/",tax_level,"/crc.taxonomy"),
+                  col_types = c(Size = col_double(),
+                                .default = col_character())) %>%
+    mutate(level=tax_level)
+}
+
+rank_tax <- function(data,tax){
+  sub_data <- data %>%
+    filter(level==tax) %>%
+    group_by(level,Taxonomy,OTU) %>%
+    summarise(mean_AUC_diff = median(AUC_diff)) %>%
+    ungroup() %>%
+    arrange(desc(mean_AUC_diff)) %>%
+    mutate(rank=seq(1,nrow(.)))
+
+  return(sub_data)
+}
+
+plot_top_imp <- function(top_data,tax_level){
+  top_data %>%
+    filter(level == tax_level) %>%
+    arrange(rank) %>%
+    ggplot(aes(x=AUC_diff,y=reorder(name,desc(rank)))) +
+    stat_summary(fun.data=median_hilow, fun.args=0.5, geom="pointrange", color=colors2[tax_level]) +
+    # stat_summary(fun = mean,geom = "pointrange",  color=colors2[tax_level],
+    #              fun.max = function(x) mean(x) + sd(x),
+    #              fun.min = function(x) mean(x) - sd(x)) +
+    theme_bw() +
+    theme(axis.title.y = element_blank(),
+          axis.text.y = element_markdown(),
+          plot.caption = element_markdown(lineheight = 1.2)) +
+    ggtitle(tax_level) +
+    xlab("Decrease in AUC")
+}
+
+### VARIABLES ########
+
+levels <- c("phylum","class","order","family","genus","otu","asv")
[…]

### 2021-08-25 code/R/figures/AUC_all_model_all_level.R
all model AUC figure

@@ -51,4 +51,4 @@ df_all %>%
         axis.text.x = element_text(angle = 0),
         axis.text = element_text(size=14)) +
   scale_fill_manual(values=pal2,name="Model")
-ggsave("results/figures/AUC_all_model_all_level.png",width = 7,height=4,units="in")
+ggsave("analysis/figures/AUC_all_model_all_level.png",width = 7,height=4,units="in")

### 2021-08-26 code/R/figures/top_10_important_tax.R
top 10 importance plot code

@@ -11,13 +11,13 @@ read_rf_imp <- function(tax_level){
                          perf_metric_diff = col_double(),
                          seed = col_double(),
                          .default = col_character())) %>%
-  mutate(level=tax_level)
+    mutate(level=tax_level)
 }
 
 read_tax <- function(tax_level){
   read_tsv(paste0("data/",tax_level,"/crc.taxonomy"),
-                  col_types = c(Size = col_double(),
-                                .default = col_character())) %>%
+           col_types = c(Size = col_double(),
+                         .default = col_character())) %>%
     mutate(level=tax_level)
 }
 
@@ -34,20 +34,39 @@ rank_tax <- function(data,tax){
 }
 
 plot_top_imp <- function(top_data,tax_level){
-  top_data %>%
-    filter(level == tax_level) %>%
-    arrange(rank) %>%
-    ggplot(aes(x=AUC_diff,y=reorder(name,desc(rank)))) +
-    stat_summary(fun.data=median_hilow, fun.args=0.5, geom="pointrange", color=colors2[tax_level]) +
-    # stat_summary(fun = mean,geom = "pointrange",  color=colors2[tax_level],
-    #              fun.max = function(x) mean(x) + sd(x),
-    #              fun.min = function(x) mean(x) - sd(x)) +
-    theme_bw() +
-    theme(axis.title.y = element_blank(),
-          axis.text.y = element_markdown(),
-          plot.caption = element_markdown(lineheight = 1.2)) +
-    ggtitle(tax_level) +
-    xlab("Decrease in AUC")
+  if(tax_level == "Phylum"){
+    top_data %>%
+      filter(level == tax_level) %>%
+      arrange(rank) %>%
+      ggplot(aes(x=AUC_diff,y=reorder(name,desc(rank)))) +
+      stat_summary(fun.data=median_hilow, fun.args=0.5, geom="pointrange", color=colors2[tax_level]) +
+      # stat_summary(fun = mean,geom = "pointrange",  color=colors2[tax_level],
+      #              fun.max = function(x) mean(x) + sd(x),
+      #              fun.min = function(x) mean(x) - sd(x)) +
+      scale_x_continuous(breaks=seq(-0.02,0.04,0.02)) +
+      theme_bw() +
[…]

### 2021-08-27 code/R/get_sens_spec.R
code to make sens/spec file

@@ -1,0 +1,64 @@
+library(tidyverse)
+
+### FUNCTIONS ############################
+
+get_file_list <- function(tax_level){
+  list.files(path=paste0("data/",tax_level,"/temp/data"),
+             pattern="rf_model.*.Rdata",
+             full.names=T)
+}
+
+get_sensitivity <- function(lookup, x){
+  if(x >= max(lookup$specificity)){
+    tibble(specificity=x,sensitivity=0)
+  } else {
+    lookup %>%
+      filter(specificity - x > 0) %>%
+      top_n(sensitivity, n=1) %>%
+      summarize(specificity=x,
+                sensitivity = unique(sensitivity))
+  }
+}
+
+pool_sens_spec <- function(file_name,specificities){
+  load(file_name)
+
+  prob <- predict(model$trained_model, model$test_data, type="prob")
+
+  prob_obs <- bind_cols(prob_srn = prob$cancer,
+                        observed=model$test_data$dx)
+
+  total <- count(prob_obs, observed) %>%
+    pivot_wider(names_from="observed", values_from="n")
+
+  lookup <- prob_obs %>%
+    arrange(desc(prob_srn)) %>%
+    mutate(is_srn = observed == "cancer") %>%
+    mutate(tp = cumsum(is_srn),
+           fp = cumsum(!is_srn),
+           sensitivity = tp / total$cancer,
+           fpr = fp / total$normal) %>%
+    mutate(specificity = 1- fpr) %>%
+    select(sensitivity, specificity, fpr)
+
+  map_dfr(specificities,get_sensitivity,lookup=lookup) %>%
+    mutate(model = str_replace(file_name,
+                               "data/(.*)/temp/data/rf_model.\\d*.Rdata", "\\1"),
+           seed = str_replace(file_name,
+                              "data/(.*)/temp/data/rf_model.(\\d*).Rdata", "\\2"))
+}
+
+loop_sens_spec <- function(tax_level,specificities){
+  get_file_list(tax_level) %>%
+    map_dfr(pool_sens_spec,specificities)
+}
+
+### VARIABLES ###########################
+
+levels <- c("phylum","class","order","family","genus","otu","asv")
+specificities <- seq(0, 1, 0.01)
+
+### MAIN ################################
+
+map_dfr(levels,loop_sens_spec,specificities) %>%
+  write_csv("analysis/sens_spec_by_level.csv")

### 2021-08-27 code/R/get_sens_spec.R
fix table header

@@ -42,7 +42,7 @@ pool_sens_spec <- function(file_name,specificities){
     select(sensitivity, specificity, fpr)
 
   map_dfr(specificities,get_sensitivity,lookup=lookup) %>%
-    mutate(model = str_replace(file_name,
+    mutate(level = str_replace(file_name,
                                "data/(.*)/temp/data/rf_model.\\d*.Rdata", "\\1"),
            seed = str_replace(file_name,
                               "data/(.*)/temp/data/rf_model.(\\d*).Rdata", "\\2"))

### 2021-08-27 code/R/figures/averaged_ROC_curve.R
code for avg roc curve

@@ -1,0 +1,94 @@
+library(tidyverse)
+library(DescTools)
+
+### VARIABLES #####################
+levels <- c("phylum","class","order","family","genus","otu","asv")
+levels_names <- c("Phylum","Class","Order","Family","Genus","OTU","ASV")
+colors <- c("#D04C44","#E7834E","#E6A444","#829E88","#83BCC3","#7989A4","#B29ABA")
+names(colors) <- levels
+
+colors2 <- colors
+names(colors2) <- levels_names
+
+sens_spec <- read_csv("analysis/sens_spec_by_level.csv")
+
+### PLOTS #####################
+
+# calculate average/sd sensitivity
+avg_sens <- sens_spec %>%
+  group_by(specificity,level) %>%
+  summarise(mean_sensitivity = mean(sensitivity),
+            sd_sensitivity = sd(sensitivity),
+            .groups = "drop") %>%
+  mutate(upper_sens = mean_sensitivity + sd_sensitivity,
+         lower_sens = mean_sensitivity - sd_sensitivity) %>%
+  mutate(level=factor(level,levels=levels,labels=levels_names)) %>%
+  mutate(upper_sens = case_when(upper_sens > 1 ~ 1,
+                                TRUE ~ upper_sens),
+         lower_sens = case_when(upper_sens < 0 ~ 0,
+                                TRUE ~ lower_sens)) %>%
+  mutate(fpr = 1-specificity)
+
+# plot all levels together
+avg_sens %>%
+  ggplot(aes(x=fpr,y=mean_sensitivity,
+             ymin=lower_sens,ymax=upper_sens)) +
+  geom_line(aes(color=level),alpha=0.7) +
+  geom_ribbon(aes(fill=level),alpha=0.2) +
+  coord_equal() +
+  geom_abline(intercept = 0,lty="dashed",color="grey50") +
+  scale_x_continuous(expand = c(0,0)) +
+  scale_y_continuous(expand = c(0,0)) +
+  xlab("False Positive Rate") +
+  ylab("Average True Positive Rate") +
+  #facet_wrap(~level,ncol=3) +
+  #scale_x_reverse(expand = c(0,0),labels = scales::percent_format(accuracy = 1)) +
+  theme_classic() +
+  theme(legend.position = "top",
+        axis.text = element_text(size=11),
+        axis.title = element_text(size=12),
+        axis.text.x = element_text(vjust = -0.5),
[…]

### 2021-08-27 code/R/calculate_pvalues_senspec.R
plot sens at 90 spec with stats

@@ -1,0 +1,42 @@
+library(tidyverse)
+library(mosaic)
+
+### FUNCTIONS ##################
+perm_p_value_senspec <- function(data, level1, level2){
+  # subset results to select AUC and condition columns and
+  # filter to only the two conditions of interest
+  test_subsample <- data %>%
+    mutate(level = as.character(level)) %>%
+    select(level,sensitivity) %>%
+    filter((level == level1 | level == level2))
+  # quantify the absolute value of the difference in mean AUC between the two conditions (observed difference)
+  obs <- abs(diff(mosaic::mean(sensitivity ~ level, data = test_subsample)))
+  # use mosaic to shuffle condition labels and calculate difference in mean AUC - repeat 10,000 times
+  null <- do(10000) * diff(mosaic::mean(sensitivity ~ shuffle(level), data = test_subsample))
+  n <- length(null[,1]) #number of shuffled calculations
+  # r = #replications at least as extreme as observed effect
+  r <- sum(abs(null[,1]) >= obs) #count number of shuffled differences as big or bigger than observed difference
+  # compute Monte Carlo p-value with correction (Davison & Hinkley, 1997)
+  p.value=(r+1)/(n+1)
+  return(p.value)
+}
+
+### MAIN #########################
+levels <- c("phylum","class","order","family","genus","otu","asv")
+levels_names <- c("Phylum","Class","Order","Family","Genus","OTU","ASV")
+
+sens_spec90 <- read_csv("analysis/sens_spec_by_level.csv") %>%
+  filter(specificity == 0.9) %>%
+  mutate(level = factor(level,levels=rev(levels),labels=rev(levels_names)))
+
+
+senspec_pvals <- expand_grid(x=1:length(levels_names),
+                             y=1:length(levels_names)) %>%
+  filter(x < y) %>%
+  mutate(level1 = levels_names[x],
+         level2 = levels_names[y]) %>%
+  select(-x, -y) %>%
+  group_by(level1, level2) %>%
+  summarize(p_value = perm_p_value_senspec(sens_spec90,level1,level2),
+            .groups="drop") %>%
+  write_csv("analysis/sens_spec90_pvals.csv")

### 2021-08-27 code/R/figures/sensitivity_at_90_specificity.R

@@ -1,0 +1,45 @@
+#generate a plot comparing the sensitivity at 90% sensitivity_at_90_specificity
+
+library(tidyverse)
+
+### VARIABLES #####################
+levels <- c("phylum","class","order","family","genus","otu","asv")
+levels_names <- c("Phylum","Class","Order","Family","Genus","OTU","ASV")
+colors <- c("#D04C44","#E7834E","#E6A444","#829E88","#83BCC3","#7989A4","#B29ABA")
+names(colors) <- levels
+
+colors2 <- colors
+names(colors2) <- levels_names
+
+sens_spec90 <- read_csv("analysis/sens_spec_by_level.csv") %>%
+  filter(specificity == 0.9) %>%
+  mutate(level = factor(level,levels=rev(levels),labels=rev(levels_names)))
+
+### PLOTS #####################
+
+#check stats
+stats <- read_csv("analysis/sens_spec90_pvals.csv") %>%
+  filter(p_value >= 0.05)
+if( nrow(stats) != 4){ stop("problem with stats)")}
+
+sens_spec90 %>%
+  ggplot(aes(x=level,y=sensitivity,color=level)) +
+    geom_jitter(width=0.2,alpha=0.5,size=2,height = 0) +
+    #stat_summary(fun.data=median_hilow, fun.args=0.5, geom="pointrange", color="black") +
+    stat_summary(fun = mean,
+                 fun.max = function(x) mean(x) + sd(x),
+                 fun.min = function(x) mean(x) - sd(x),
+                 #geom = "crossbar", width=0.6,
+                 geom = "pointrange",
+                 color="black") +
+    theme_bw() +
+    xlab("") + ylab("Sensitivity") +
+    theme(axis.text = element_text(size=12),
+          axis.title = element_text(size=12),
+          legend.position = "none",
+          panel.grid.major.y = element_blank()) +
+    coord_flip() +
+    scale_y_continuous(breaks=seq(0,1,0.1),labels = scales::percent_format(accuracy = 1)) +
+    scale_color_manual(values=rev(colors2)) +
+    annotate("text",x=levels_names,y=rep((max(sens_spec90$sensitivity)+0.05),length(levels_names)),label=c("W","W","X","Y","Y","Y","Z"))
+ggsave("analysis/figures/sensitivity_at_90_specificity.png",width = 4.5,height=5,units="in")

### 2021-08-27 code/R/figures/figure_1.R
code for paper figures

@@ -1,0 +1,14 @@
+library(png)
+library(grid)
+library(gridExtra)
+library(tidyverse)
+
+img1 <-  rasterGrob(as.raster(readPNG("analysis/figures/rf_AUC_all_level.png")), interpolate = FALSE)
+img2 <-  rasterGrob(as.raster(readPNG("analysis/figures/sensitivity_at_90_specificity.png")), interpolate = FALSE)
+grid.arrange(arrangeGrob(img1, left = textGrob("A", x = unit(1, "npc"), y = unit(0.85, "npc"), gp = gpar(fontface="bold"))),
+             arrangeGrob(img2, left = textGrob("B", x = unit(1, "npc"), y = unit(0.85, "npc"), gp = gpar(fontface="bold"))),
+             ncol = 2)
+g <- arrangeGrob(arrangeGrob(img1, left = textGrob("A", x = unit(1, "npc"), y = unit(0.95, "npc"), gp = gpar(fontface="bold"))),
+            arrangeGrob(img2, left = textGrob("B", x = unit(1, "npc"), y = unit(0.95, "npc"), gp = gpar(fontface="bold"))),
+            ncol = 2)
+ggsave("paper/figure_1.png",g,height=5,width=10)

### 2021-08-27 paper/manuscript.Rmd
adjustments from pat comments

@@
 library(kableExtra)
+library(schtools)
 ```
@@
 colonoscopy), many people do not comply with recommended screening guidelines [@garcía2011a] prompting a need for low cost, non-invasive detection methods.
-A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasias (neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e.
+A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasia (SRNs) consisting of pre-cancerous polyps (i.e.
 advanced adenomas) and carcinomas.
-Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on amplicon sequencing of the 16S rRNA gene {}.
+Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances of operational taxonomic unit (OTU) classifications based on amplicon sequencing of the 16S rRNA gene {}.
 However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown.
@@
 Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups.
-Topçuoğlu *et al* [@topçuolu2020] recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
-This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.
+Topçuoğlu *et al* [@topçuolu2020] recently demonstrated how to effectively apply machine learning (ML) methods to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
+This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to determine which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.
 
@@
 
-Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels.
-Using the taxonomic abundance data, we quantified how accurately samples could be classified as "normal" or "SRN" (i.e.
-advanced adenoma or carcinoma) using five machine learning methods with the Mikropml R package (methods).
+Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated abundance tables with mothur {} annotated to phylum, class, order, family, genus, OTU and ASV levels.
+Using the taxonomic abundance data, we quantified how reliably samples could be classified as "normal" or "SRN" (i.e.
+advanced adenoma or carcinoma) using five machine learning methods with the mikropml R package (methods).
 Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1).
-Random forest (RF) was consistently the top performer at most taxonomic levels.
+Regardless of the taxonomic level, random forest (RF) models consistently had the largest area under the receiver operating characteristic curve (AUROC).
 Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1A, Supplemental Figure 2).
-Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTUs (p \< 0.01), but borderline equivalent to family (p = `r round(rf_pvals["family","asv"],digits=2)`) and genus (p = `r round(rf_pvals["genus","asv"],digits=2)`) levels (Figure 1A).
-These results were consistent for ASVs generated with DADA2 [@callahan2016] (median AUROC: `r round(dada2_med_auc,digits=3)`).
+Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTUs (p \< 0.01), but comparable to family (p = `r round(rf_pvals["family","asv"],digits=2)`) and genus (p = `r round(rf_pvals["genus","asv"],digits=2)`) levels (Figure 1A).
 These results suggest that finer taxonomic resolution is not necessarily better for identifying individuals with SRNs based on microbiome composition.
@@
 To further compare the model performance across taxonomic levels we compared the sensitivity of the models at a specificity of 90%.
-The highest sensitivity values were observed for Family, Genus, and OTU level data (Figure 1B), consistent with the AUROC results.
-Phylum, Class, Order, and ASV sensitivity values were all significantly lower than Family, Genus, and OTU sensitivity values (Figure 1B).
-This analysis further supports that finer resolution is does not improve SRN detection.
+The highest sensitivity values were observed for family, genus, and OTU level data (Figure 1B), consistent with the AUROC results.
+phylum, class, order, and ASV sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (Figure 1B).
+This analysis further supports the observation that finer resolution is does not improve SRN detection.
 
@@
 phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls.
-At the level of genus/OTU data there is enough data and variation but at the ASV level, the data is too specific to individuals and doesn't overlap enough.
+At the level of genus or OTU there is enough data and variation but at the ASV level, the data is too specific to individuals and does not overlap enough.
 Examination of the prevalence of taxa in samples at each level supports this idea.
-A majority of taxa are present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels.
-The opposite is observed at the OTU and ASV level where 60% and 53% of taxa respectively are only present in less than 25% of samples (Supplemental Figure 3).
-Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods).
-As part of this step, features are removed that will not provide useful information to build the model.
-For example, perfectly correlated features provide the same information to build the model and thus can be collapsed.
-Additionally, features with zero or near-zero variance will not help the model differentiate groups and thus can be removed.
-Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing.
-At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).
+A majority of taxa were present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels.
+The opposite was observed at the OTU and ASV level where 60% and 53% of taxa respectively were only present in less than 25% of samples (Supplemental Figure 3).
+Of note, the ML pipeline includes a pre-processing step that occured prior to training and classifying the ML models.
+As part of this step, taxa were removed that would not provide useful information to build the model.
+For example, perfectly correlated taxa provide the same information to build the model and thus can be collapsed.
+Additionally, features with zero or near-zero variance across samples would not help the model differentiate groups and thus can be removed.
+Interestingly, despite starting with `r format_number(input %>% filter(level == "asv") %>% select(n_features) %>% pull())` ASVs, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing.
+At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` of the `r format_number(input %>% filter(level == "otu") %>% select(n_features) %>% pull())` OTUs (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).
 While the resolution provided by ASVs is useful in certain contexts {}, these results suggest that the resolution is too fine for use in machine learning classification of SRNs based on microbiome composition.
 
-A look into the most important taxa at each level for classifying samples reveals some nesting where several genera and their higher taxonomic classifications are in the top 10 most important taxa (Supplemental Figure 4).
-For example, the genus *Gemella* is an important taxa at the genus and OTU levels and its higher classifications are also important (*Firmicutes* > *Bacilli* > *Bacillales* > *Bacillales Incertae Sedis XI* > *Gemella*).
-*Fusobacterium* displays a similar pattern, except that the family level classification (*Fusobacteriaceae*) importance is ranked 16th out of 54 families, a little beyond the top 10.
-In the case of unclassified *Lachnospiraceae*, there are several OTUs with this label that are in the top 10, however at the Genus level this taxon is ranked lower in importance (21st out of 115 genera) suggesting there may be some benefit to separating different taxonomic groupings within *Lachnospiraceae*.
+A look into the most important taxa at each level for classifying samples revealed some nesting where several genera and their higher taxonomic classifications were in the top 10 most important taxa (Supplemental Figure 4).
+For example, the genus *Gemella* was an important taxon at the genus and OTU levels and its higher classifications were also important (*Firmicutes* > *Bacilli* > *Bacillales* > *Bacillales Incertae Sedis XI* > *Gemella*).
+*Fusobacterium* displayed a similar pattern, except that the family level classification (*Fusobacteriaceae*) importance was ranked 16th out of 54 families.
+In the case of unclassified *Lachnospiraceae*, there were several OTUs with this label that were in the top 10, however at the genus level this taxon was ranked lower in importance (21st out of 115 genera) suggesting there may be some benefit to separating different taxonomic groupings within *Lachnospiraceae*.
 
@@
 In general, we found that finer taxonomic resolution (e.g.
-OTU and ASV) does not add additional sensitivity to predicting SRNs based on microbiome composition.
-Additionally, at the ASV level the fine resolution actually impedes model performance due to the sparsity of shared taxa and leads to decreased model performance.
+OTU and ASV) did not add additional sensitivity to predicting SRNs based on microbiome composition.
+Additionally, at the ASV level the fine resolution actually impedes model performance due to the sparsity of shared taxa and led to decreased model performance.
 The tendancy for ASV level annotation to split single bacterial genomes into multiple taxa {Pat mSphere 2021} could also be a contributing factor to the sparsity of shared taxa.
-Overall, either Family, Genus, or OTU level taxonomy appear to perform equally for predicting subjects with SRNs based on the composition of the gut microbiome.
-A potential benefit of utilizing genus or family level data could be that it may allow for merging data generated from different 16S regions or sequencing platforms.
+Overall, either family, genus, or OTU level taxonomy appear to perform equally for predicting subjects with SRNs based on the composition of the gut microbiome.
+A potential benefit of utilizing genus or family level data could be that it may allow for merging data generated from different 16S rRNA gene regions or sequencing platforms.
 
@@
 
-***Data processing.*** Sequence data was processed with Mothur (1.44.3) using the SILVA reference database (v132) to produce count tables for phylum, class, order, family, genus, OTU, and ASV following the Schloss Lab MiSeq SOP described on the Mothur website {}.
+***Data processing.*** Sequence data was processed with mothur (1.44.3) using the SILVA reference database (v132) to produce count tables for phylum, class, order, family, genus, OTU, and ASV following the Schloss Lab MiSeq SOP described on the mothur website.
 ASV level data was also produced using DADA2 to ensure consistent results with a different pipeline.
 Data was processed following the DADA2 pipeline, but setting pool=TRUE to infer ASVs from the whole dataset rather than per sample.
-The resulting ASV table was subsampled for consistency with the Mothur data.
+The resulting ASV table was subsampled for consistency with the mothur data.
+These results were consistent for ASVs generated with DADA2 [@callahan2016] (median AUROC: `r round(dada2_med_auc,digits=3)`).
 
@@
 Each taxonomic model taxonomic level combination (e.g.
-Random Forest on genus) was run with 100 different seeds (1-100).
+random forest on genus) was run with 100 different seeds.
 Each seed split the data into a training (80%) and testing (20%) set, and output performance of the training and testing as area under the receiver operating curve (AUROC).
@@
 
-```{r fig1}
-#![](../exploratory/figures/rf_with_stats.png){width="50%"} ![](../exploratory/figures/rf_with_stats.png){width="50%"}
-# library(cowplot)
-# 
-# p1 <- ggdraw + draw_image("../exploratory/figures/rf_with_stats.png", width = 0.5) 
-# p2 <- ggdraw + draw_image("../exploratory/figures/rf_with_stats.png", width = 0.5, x = 0.5)
-# 
-# plot_grid(p1,p1,labels = c('A', 'B'), label_size = 12)
-
-library(png)
-library(grid)
-library(gridExtra)
-img1 <-  rasterGrob(as.raster(readPNG("../exploratory/figures/rf_with_stats.png")), interpolate = FALSE)
-img2 <-  rasterGrob(as.raster(readPNG("../exploratory/figures/sens_90spec.png")), interpolate = FALSE)
-grid.arrange(arrangeGrob(img1, left = textGrob("A)", x = unit(1, "npc"), y = unit(0.85, "npc"), gp = gpar(fontface="bold"))),
-             arrangeGrob(img2, left = textGrob("B)", x = unit(1, "npc"), y = unit(0.85, "npc"), gp = gpar(fontface="bold"))), 
-             ncol = 2)
-```
+![](figure_1.png)
 
@@
 
-input %>%
+read_csv("table_1.csv") %>%
   select(-n_samples) %>%
+  mutate(pct_kept=round((n_features_preproc/n_features)*100,digits=1)) %>%
   mutate(pct_kept=paste0(pct_kept," %"),
-         level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv"),
+         n_features = unlist(lapply(n_features,format_number)),
+         level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv" ),
                       labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV"))) %>%
@@
 
-![](../exploratory/figures/all_model_level.png)
+![](figure_s1.png)
 
@@
 
-![](../exploratory/figures/roc_by_level.png){height="50%"}
+![](figure_s2.png){height="50%"}
 
@@
 
-![](../exploratory/figures/bin_prevalence.png){height="45%"}\
+![](figure_s3.png){height="45%"}\
 **Supplemental Figure 3: Prevalence of Taxa in Samples.** Distribution of taxa prevalence in samples at each taxonomic level.
@@
 
-![](../exploratory/figures/top10_imp.png){height="85%"}\
+![](figure_s4.png){height="85%"}\
 **Supplemental Figure 4: Top 10 important taxa at each taxonomic level.** Summary of the 10 most important taxa for the random forest models at each taxonomic level based on the average decrease in AUC when the feature is permuted.

### 2021-08-27 paper/manuscript.tex
adjustments from pat comments

@@
 points to the gut microbiome as a promising target for non-invasive
-screening to detect screen relevant neoplasias (neoplasms?) (SRNs)
-consisting of pre-cancerous polyps (i.e.~advanced adenomas) and
-carcinomas. Efforts to realize the diagnostic potential of the gut
-microbiome in detecting SRNs have focused on machine learning (ML)
-methods using abundances from operational taxonomic unit (OTU)
-classifications based on amplicon sequencing of the 16S rRNA gene \{\}.
-However, whether this is the optimal taxonomic resolution for
-classifying SRNs from microbiome data is unknown. Additionally, recent
-work has pushed for the use of amplicon sequence variants (ASVs) to
-replace OTUs for marker-gene analysis because of the improved resolution
-with ASVs (\protect\hyperlink{ref-callahan2017}{3}). However, whether
-the additional resolution provided by ASVs is useful for ML
-classification is unclear \{\}. Since ML classification relies on
-consistent differences between groups, its possible that the resolution
-at the ASV level is too individualized to accurately differentiate
-groups. Topçuoğlu \emph{et al}
-(\protect\hyperlink{ref-topuxe7uolu2020}{4}) recently demonstrated
-effective application of machine learning (ML) to microbiome based
-classification problems and developed a framework for applying ML
-practices in a more reproducible way (mikropml). This analysis utilizes
-the reproducible framework developed by Topçuoğlu \emph{et al} to
-quantify which ML method and taxonomic level produce the best performing
-classifier for detecting SRNs from microbiome data.
+screening to detect screen relevant neoplasia (SRNs) consisting of
+pre-cancerous polyps (i.e.~advanced adenomas) and carcinomas. Efforts to
+realize the diagnostic potential of the gut microbiome in detecting SRNs
+have focused on machine learning (ML) methods using abundances of
+operational taxonomic unit (OTU) classifications based on amplicon
+sequencing of the 16S rRNA gene \{\}. However, whether this is the
+optimal taxonomic resolution for classifying SRNs from microbiome data
+is unknown. Additionally, recent work has pushed for the use of amplicon
+sequence variants (ASVs) to replace OTUs for marker-gene analysis
+because of the improved resolution with ASVs
+(\protect\hyperlink{ref-callahan2017}{3}). However, whether the
+additional resolution provided by ASVs is useful for ML classification
+is unclear \{\}. Since ML classification relies on consistent
+differences between groups, its possible that the resolution at the ASV
+level is too individualized to accurately differentiate groups.
+Topçuoğlu \emph{et al} (\protect\hyperlink{ref-topuxe7uolu2020}{4})
+recently demonstrated how to effectively apply machine learning (ML)
+methods to microbiome based classification problems and developed a
+framework for applying ML practices in a more reproducible way
+(mikropml). This analysis utilizes the reproducible framework developed
+by Topçuoğlu \emph{et al} to determine which ML method and taxonomic
+level produce the best performing classifier for detecting SRNs from
+microbiome data.
 
@@
 patients with SRNs and healthy controls, we generated abundance tables
-with Mothur \{\} annotated to Phylum, Class, Order, Family, Genus, OTU
+with mothur \{\} annotated to phylum, class, order, family, genus, OTU
 and ASV levels. Using the taxonomic abundance data, we quantified how
-accurately samples could be classified as ``normal'' or ``SRN''
+reliably samples could be classified as ``normal'' or ``SRN''
 (i.e.~advanced adenoma or carcinoma) using five machine learning methods
-with the Mikropml R package (methods). Across the five machine learning
+with the mikropml R package (methods). Across the five machine learning
 methods tested, model performance tended to increase with taxonomic
 level usually peaking around genus/OTU level before dropping off
-slightly with ASVs (Supplemental Figure 1). Random forest (RF) was
-consistently the top performer at most taxonomic levels. Within the RF
+slightly with ASVs (Supplemental Figure 1). Regardless of the taxonomic
+level, random forest (RF) models consistently had the largest area under
+the receiver operating characteristic curve (AUROC). Within the RF
 model, the highest AUROCs were observed for family (median AUROC:
@@
 Supplemental Figure 2). Performance with ASVs (median AUROC: 0.676) was
-significantly lower than OTUs (p \textless{} 0.01), but borderline
-equivalent to family (p = 0.06) and genus (p = 0.05) levels (Figure 1A).
-These results were consistent for ASVs generated with DADA2
-(\protect\hyperlink{ref-callahan2016}{5}) (median AUROC: 0.66). These
-results suggest that finer taxonomic resolution is not necessarily
-better for identifying individuals with SRNs based on microbiome
-composition.
+significantly lower than OTUs (p \textless{} 0.01), but comparable to
+family (p = 0.06) and genus (p = 0.05) levels (Figure 1A). These results
+suggest that finer taxonomic resolution is not necessarily better for
+identifying individuals with SRNs based on microbiome composition.
 
@@
 thresholds and can be misleading since models with the same AUC can have
-different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{6}).
+different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{5}).
 Depending on the intended implementation of the model, one may want to
@@
 compared the sensitivity of the models at a specificity of 90\%. The
-highest sensitivity values were observed for Family, Genus, and OTU
-level data (Figure 1B), consistent with the AUROC results. Phylum,
-Class, Order, and ASV sensitivity values were all significantly lower
-than Family, Genus, and OTU sensitivity values (Figure 1B). This
-analysis further supports that finer resolution is does not improve SRN
-detection.
+highest sensitivity values were observed for family, genus, and OTU
+level data (Figure 1B), consistent with the AUROC results. phylum,
+class, order, and ASV sensitivity values were all significantly lower
+than family, genus, and OTU sensitivity values (Figure 1B). This
+analysis further supports the observation that finer resolution is does
+not improve SRN detection.
 
@@
 overlap to reliably differentiate between cases and controls. At the
-level of genus/OTU data there is enough data and variation but at the
-ASV level, the data is too specific to individuals and doesn't overlap
+level of genus or OTU there is enough data and variation but at the ASV
+level, the data is too specific to individuals and does not overlap
 enough. Examination of the prevalence of taxa in samples at each level
-supports this idea. A majority of taxa are present in greater than 75\%
+supports this idea. A majority of taxa were present in greater than 75\%
 of samples at the phylum (67\% of taxa) and class (63\% of taxa) levels.
-The opposite is observed at the OTU and ASV level where 60\% and 53\% of
-taxa respectively are only present in less than 25\% of samples
+The opposite was observed at the OTU and ASV level where 60\% and 53\%
+of taxa respectively were only present in less than 25\% of samples
 (Supplemental Figure 3). Of note, the ML pipeline includes a
-pre-processing step that occurs prior to training and classifying the ML
-models (methods). As part of this step, features are removed that will
-not provide useful information to build the model. For example,
-perfectly correlated features provide the same information to build the
-model and thus can be collapsed. Additionally, features with zero or
-near-zero variance will not help the model differentiate groups and thus
-can be removed. Interestingly, despite starting with 104106 features at
-the ASV level, only 478 (0.5\%) remained after pre-processing. At the
-OTU level, 20079 of the 705 features (3.5\%) remained after
-preprocessing (Table 1). While the resolution provided by ASVs is useful
-in certain contexts \{\}, these results suggest that the resolution is
-too fine for use in machine learning classification of SRNs based on
-microbiome composition.
+pre-processing step that occured prior to training and classifying the
+ML models. As part of this step, taxa were removed that would not
+provide useful information to build the model. For example, perfectly
+correlated taxa provide the same information to build the model and thus
+can be collapsed. Additionally, features with zero or near-zero variance
+across samples would not help the model differentiate groups and thus
+can be removed. Interestingly, despite starting with 104,106 ASVs, only
+478 (0.5\%) remained after pre-processing. At the OTU level, 705 of the
+20,079 OTUs (3.5\%) remained after preprocessing (Table 1). While the
+resolution provided by ASVs is useful in certain contexts \{\}, these
+results suggest that the resolution is too fine for use in machine
+learning classification of SRNs based on microbiome composition.
 
 A look into the most important taxa at each level for classifying
-samples reveals some nesting where several genera and their higher
-taxonomic classifications are in the top 10 most important taxa
-(Supplemental Figure 4). For example, the genus \emph{Gemella} is an
-important taxa at the genus and OTU levels and its higher
-classifications are also important (\emph{Firmicutes} \textgreater{}
+samples revealed some nesting where several genera and their higher
+taxonomic classifications were in the top 10 most important taxa
+(Supplemental Figure 4). For example, the genus \emph{Gemella} was an
+important taxon at the genus and OTU levels and its higher
+classifications were also important (\emph{Firmicutes} \textgreater{}
 \emph{Bacilli} \textgreater{} \emph{Bacillales} \textgreater{}
 \emph{Bacillales Incertae Sedis XI} \textgreater{} \emph{Gemella}).
-\emph{Fusobacterium} displays a similar pattern, except that the family
-level classification (\emph{Fusobacteriaceae}) importance is ranked 16th
-out of 54 families, a little beyond the top 10. In the case of
-unclassified \emph{Lachnospiraceae}, there are several OTUs with this
-label that are in the top 10, however at the Genus level this taxon is
-ranked lower in importance (21st out of 115 genera) suggesting there may
-be some benefit to separating different taxonomic groupings within
+\emph{Fusobacterium} displayed a similar pattern, except that the family
+level classification (\emph{Fusobacteriaceae}) importance was ranked
+16th out of 54 families. In the case of unclassified
+\emph{Lachnospiraceae}, there were several OTUs with this label that
+were in the top 10, however at the genus level this taxon was ranked
+lower in importance (21st out of 115 genera) suggesting there may be
+some benefit to separating different taxonomic groupings within
 \emph{Lachnospiraceae}.
@@
 is warranted. In general, we found that finer taxonomic resolution
-(e.g.~OTU and ASV) does not add additional sensitivity to predicting
-SRNs based on microbiome composition. Additionally, at the ASV level the
-fine resolution actually impedes model performance due to the sparsity
-of shared taxa and leads to decreased model performance. The tendancy
-for ASV level annotation to split single bacterial genomes into multiple
-taxa \{Pat mSphere 2021\} could also be a contributing factor to the
-sparsity of shared taxa. Overall, either Family, Genus, or OTU level
-taxonomy appear to perform equally for predicting subjects with SRNs
-based on the composition of the gut microbiome. A potential benefit of
-utilizing genus or family level data could be that it may allow for
-merging data generated from different 16S regions or sequencing
-platforms.
+(e.g.~OTU and ASV) did not add additional sensitivity to predicting SRNs
+based on microbiome composition. Additionally, at the ASV level the fine
+resolution actually impedes model performance due to the sparsity of
+shared taxa and led to decreased model performance. The tendancy for ASV
+level annotation to split single bacterial genomes into multiple taxa
+\{Pat mSphere 2021\} could also be a contributing factor to the sparsity
+of shared taxa. Overall, either family, genus, or OTU level taxonomy
+appear to perform equally for predicting subjects with SRNs based on the
+composition of the gut microbiome. A potential benefit of utilizing
+genus or family level data could be that it may allow for merging data
+generated from different 16S rRNA gene regions or sequencing platforms.
 
@@
 
-\textbf{\emph{Data processing.}} Sequence data was processed with Mothur
+\textbf{\emph{Data processing.}} Sequence data was processed with mothur
 (1.44.3) using the SILVA reference database (v132) to produce count
 tables for phylum, class, order, family, genus, OTU, and ASV following
-the Schloss Lab MiSeq SOP described on the Mothur website \{\}. ASV
-level data was also produced using DADA2 to ensure consistent results
-with a different pipeline. Data was processed following the DADA2
-pipeline, but setting pool=TRUE to infer ASVs from the whole dataset
-rather than per sample. The resulting ASV table was subsampled for
-consistency with the Mothur data.
+the Schloss Lab MiSeq SOP described on the mothur website. ASV level
+data was also produced using DADA2 to ensure consistent results with a
+different pipeline. Data was processed following the DADA2 pipeline, but
+setting pool=TRUE to infer ASVs from the whole dataset rather than per
+sample. The resulting ASV table was subsampled for consistency with the
+mothur data. These results were consistent for ASVs generated with DADA2
+(\protect\hyperlink{ref-callahan2016}{6}) (median AUROC: 0.66).
 
@@
 peak performance was at the edge of the hyperparameter range. Each
-taxonomic model taxonomic level combination (e.g.~Random Forest on
-genus) was run with 100 different seeds (1-100). Each seed split the
-data into a training (80\%) and testing (20\%) set, and output
-performance of the training and testing as area under the receiver
-operating curve (AUROC).
+taxonomic model taxonomic level combination (e.g.~random forest on
+genus) was run with 100 different seeds. Each seed split the data into a
+training (80\%) and testing (20\%) set, and output performance of the
+training and testing as area under the receiver operating curve (AUROC).
 
@@
 
-\includegraphics{manuscript_files/figure-latex/fig1-1.pdf}
+\includegraphics{figure_1.png}
 
@@
 \makecell[c]{Number of Features \\ After Preprocessing} &
-\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\ \addlinespace
+\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\
 \midrule
 \endhead
-Phylum & 19 & 9 & 47.4 \% \\ \addlinespace
-Class & 36 & 19 & 52.8 \% \\ \addlinespace
-Order & 65 & 28 & 43.1 \% \\ \addlinespace
-Family & 124 & 54 & 43.5 \% \\ \addlinespace
-Genus & 316 & 115 & 36.4 \% \\ \addlinespace
-OTU & 20079 & 705 & 3.5 \% \\ \addlinespace
-ASV & 104106 & 478 & 0.5 \% \\ \addlinespace
+Phylum & 19 & 9 & 47.4 \% \\
+Class & 36 & 19 & 52.8 \% \\
+Order & 65 & 28 & 43.1 \% \\
+Family & 124 & 54 & 43.5 \% \\
+Genus & 316 & 115 & 36.4 \% \\
+OTU & 20,079 & 705 & 3.5 \% \\
+ASV & 104,106 & 478 & 0.5 \% \\
 \bottomrule
@@
 
-\includegraphics{../exploratory/figures/all_model_level.png}
+\includegraphics{figure_s1.png}
 
@@
 
-\includegraphics[width=\textwidth,height=0.5\textheight]{../exploratory/figures/roc_by_level.png}
+\includegraphics[width=\textwidth,height=0.5\textheight]{figure_s2.png}
 
@@
 
-\includegraphics[width=\textwidth,height=0.45\textheight]{../exploratory/figures/bin_prevalence.png}\\
+\includegraphics[width=\textwidth,height=0.45\textheight]{figure_s3.png}\\
 \textbf{Supplemental Figure 3: Prevalence of Taxa in Samples.}
@@
 
-\includegraphics[width=\textwidth,height=0.85\textheight]{../exploratory/figures/top10_imp.png}\\
+\includegraphics[width=\textwidth,height=0.85\textheight]{figure_s4.png}\\
 \textbf{Supplemental Figure 4: Top 10 important taxa at each taxonomic
@@
 
+\leavevmode\hypertarget{ref-lobo2008}{}%
+\CSLLeftMargin{5. }
+\CSLRightInline{\textbf{Lobo JM}, \textbf{Jiménez-Valverde A},
+\textbf{Real R}. 2008. AUC: a misleading measure of the performance of
+predictive distribution models. Global Ecology and Biogeography
+\textbf{17}:145--151.
+doi:\href{https://doi.org/10.1111/j.1466-8238.2007.00358.x}{10.1111/j.1466-8238.2007.00358.x}.}
+
 \leavevmode\hypertarget{ref-callahan2016}{}%
-\CSLLeftMargin{5. }
+\CSLLeftMargin{6. }
 \CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
@@
 
-\leavevmode\hypertarget{ref-lobo2008}{}%
-\CSLLeftMargin{6. }
-\CSLRightInline{\textbf{Lobo JM}, \textbf{Jiménez-Valverde A},
-\textbf{Real R}. 2008. AUC: a misleading measure of the performance of
-predictive distribution models. Global Ecology and Biogeography
-\textbf{17}:145--151.
-doi:\href{https://doi.org/10.1111/j.1466-8238.2007.00358.x}{10.1111/j.1466-8238.2007.00358.x}.}
-
 \end{CSLReferences}

### 2021-08-27 code/R/figures/averaged_ROC_curve.R
update average AUC on avg roc

@@ -1,5 +1,4 @@
 library(tidyverse)
-library(DescTools)
 
 ### VARIABLES #####################
 levels <- c("phylum","class","order","family","genus","otu","asv")
@@ -56,13 +55,9 @@ avg_sens %>%
 #ggsave("analysis/figures/average_roc_all.png",width=6)
 
 #plot faceted
-mean_auc <- sens_spec %>%
-  group_by(specificity,level) %>%
-  summarise(mean_sensitivity = mean(sensitivity),
-            sd_sensitivity = sd(sensitivity)) %>%
-  group_by(level) %>%
-  summarize(mean_AUC = DescTools::AUC(specificity,mean_sensitivity),
-            .groups="drop") %>%
+mean_auc <- read_csv("analysis/mean_AUC_by_model.csv")  %>% 
+  select(level,rf) %>%
+  rename(mean_AUC = rf) %>%
   mutate(level=factor(level,levels=levels,labels=levels_names))
 
 avg_sens %>%

### 2021-08-30 exploratory/exploratory.Rmd
draft to send to coauthors

@@
 library(pROC)
-
-#given a number, adds correct commas
-set_commas <- function(num){
-
-}
+library(schtools)
+
 ```
@@
 
+read_combined <- function(tax_level,model){
+  read_csv(paste0("../data/process/combined-",tax_level,"-",model,".csv"),
+           col_types = c(method = col_character(),.default=col_double()),
+           na = c("","NA","NaN")) %>% #some of the Pos/Neg Pred Values were NaN
+    mutate(level=tax_level) 
+}
+
 for( m in models ){
-  phylum <- read_csv(paste0("../data/process/combined-phylum-",m,".csv"),
-                     col_types = c(method = col_character(),.default=col_double())) %>% 
-    mutate(level="phylum")
-  class <- read_csv(paste0("../data/process/combined-class-",m,".csv"),
-                     col_types = c(method = col_character(),.default=col_double())) %>% 
-    mutate(level="class")
-  order <- read_csv(paste0("../data/process/combined-order-",m,".csv"),
-                     col_types = c(method = col_character(),.default=col_double())) %>% 
-    mutate(level="order")
-  family <- read_csv(paste0("../data/process/combined-family-",m,".csv"),
-                     col_types = c(method = col_character(),.default=col_double())) %>% 
-    mutate(level="family")
-  genus <- read_csv(paste0("../data/process/combined-genus-",m,".csv"),
-                     col_types = c(method = col_character(),.default=col_double())) %>%
-    mutate(level="genus")
-  otu <- read_csv(paste0("../data/process/combined-otu-",m,".csv"),
-                     col_types = c(method = col_character(),.default=col_double())) %>%
-    mutate(level="otu")
-  asv <- read_csv(paste0("../data/process/combined-asv-",m,".csv"),
-                     col_types = c(method = col_character(),.default=col_double())) %>%
-    mutate(level="asv")
-
-  model_df <- bind_rows(phylum,class,order,family,genus,otu,asv)
+  model_df <- map_dfr(levels, read_combined, model = m)
 
@@
   asv <- read_csv(paste0("../data/dada2/process/combined-dada2-",m,".csv"),
-                     col_types = c(method = col_character(),.default=col_double())) %>%
+                  col_types = c(method = col_character(),.default=col_double()),
+                  na = c("","NA","NaN")) %>%
     mutate(type="DADA2")
@@
       stat_summary(fun.data=median_hilow, fun.args=0.5, geom="pointrange", color=colors2[tax_level]) +
+      # stat_summary(fun = mean,geom = "pointrange",  color=colors2[tax_level],
+      #              fun.max = function(x) mean(x) + sd(x),
+      #              fun.min = function(x) mean(x) - sd(x)) +
       theme_bw() +

### 2021-08-30 paper/manuscript.Rmd
draft to send to coauthors

@@
 library(schtools)
+
+format_decimal <- function(x,digits){
+  format(round(x,digits = digits),nsmall=digits)
+}
 ```
@@
 
-Courtney R. Armour, Begüm D.Topçuoğlu, Andrea Garretto, Patrick D. Schloss ${^\dagger}$
+Courtney R. Armour${^1}$, Begüm D.Topçuoğlu, Andrea Garretto${^1}$, Patrick D. Schloss ${^1}$${^*}$
 
-\vspace{20mm}
+\vspace{5mm}
 
-${\dagger}$ To whom correspondence should be addressed:
+${^1}$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor
 
-\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
+\vspace{5mm}
 
-Department of Microbiology
-
-University of Michigan
-
-Ann Arbor, MI 48109
+${^*}$ Correspondence: Patrick D. Schloss \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
 
@@
 
-```{r median_AUC}
-table <- read_csv("../exploratory/figures/median_test_auc.csv",
+```{r mean_AUC}
+table <- read_csv("../analysis/mean_AUC_by_model.csv",
                   col_types = cols(level=col_character(),
@@
 
-rf_med_auc <- table %>% 
+rf_mean_auc <- table %>% 
   pull(rf)
 
-names(rf_med_auc) <- table %>% pull(level)
+names(rf_mean_auc) <- table %>% pull(level)
 ```
@@
                                            .default = col_double()))
-dada2_med_auc <- dada2_results %>% 
+dada2_mean_auc <- dada2_results %>% 
   pull(AUC) %>% 
-  median()
+  mean()
 
@@
 Regardless of the taxonomic level, random forest (RF) models consistently had the largest area under the receiver operating characteristic curve (AUROC).
-Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1A, Supplemental Figure 2).
-Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTUs (p \< 0.01), but comparable to family (p = `r round(rf_pvals["family","asv"],digits=2)`) and genus (p = `r round(rf_pvals["genus","asv"],digits=2)`) levels (Figure 1A).
-These results suggest that finer taxonomic resolution is not necessarily better for identifying individuals with SRNs based on microbiome composition.
+Within the RF model, the highest AUROCs were observed for family (mean AUROC: `r format_decimal(rf_mean_auc["family"],digits=3)`), genus (mean AUROC: `r format_decimal(rf_mean_auc["genus"],digits=3)`), and OTU (mean AUROC: `r format_decimal(rf_mean_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1A, Supplemental Figure 2).
+Performance with ASVs (mean AUROC: `r format_decimal(rf_mean_auc["asv"],digits=3)`) was significantly lower than OTUs (p \< 0.01), but comparable to family (p = `r format_decimal(rf_pvals["family","asv"],digits=2)`) and genus (p = `r format_decimal(rf_pvals["genus","asv"],digits=2)`) levels (Figure 1A).
+These results suggest that increased resolution improves model performance until the OTU level where further taxonomic resolution is not necessarily better for identifying individuals with SRNs based on microbiome composition.
 
-While comparing AUROC values between models is a useful way to assess the overall model performance, they summarize the performance across all thresholds and can be misleading since models with the same AUC can have different ROC curve shapes [@lobo2008].
+While comparing AUROC values between models is a useful way to assess the overall model performance, they summarize the performance across all thresholds and can be misleading since models with the same AUROC can have different ROC curve shapes [@lobo2008].
 Depending on the intended implementation of the model, one may want to optimize the true-positive rate (or sensitivity) over the false-positive rate (or 1-specificity), or vice versa.
@@
 The highest sensitivity values were observed for family, genus, and OTU level data (Figure 1B), consistent with the AUROC results.
-phylum, class, order, and ASV sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (Figure 1B).
+Phylum, class, order, and ASV sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (Figure 1B).
 This analysis further supports the observation that finer resolution is does not improve SRN detection.
+
+```{r prevalence}
+
+bin_prevalence <- function(prev,tax_level){
+  sub_data <- prev %>% 
+    filter(level==tax_level,kept==1)
+  hist <- hist(sub_data$total_prevalence,breaks = seq(0,1,0.1),plot=FALSE)
+
+  nOTU <- sub_data %>%
+    select(name) %>% 
+    n_distinct()
+
+  tibble(level=tax_level,
+         prev=hist$breaks[-1],
+         count=hist$counts,
+         pct=hist$counts/nOTU)
+}
+
+prev <- read_csv("../analysis/prevalence_by_level.csv")
+
+levels <- c("phylum","class","order","family","genus","otu","asv")
+
+pct_bins <- map_dfr(levels,bin_prevalence,prev=prev)
+
+#percent of taxa present in > 70% samples
+pct_above <- pct_bins %>% 
+  mutate(prev = round(prev,digits = 1)) %>% 
+  filter(prev > 0.7) %>% 
+  group_by(level)  %>% 
+  summarise(prev_above = round((sum(pct)) * 100)) %>% 
+  pivot_wider(names_from = level,values_from = prev_above)
+
+#percent of taxa present in < 20% samples
+pct_below <- pct_bins %>% 
+  mutate(prev = round(prev,digits = 1)) %>% 
+  filter(prev <= 0.2) %>% 
+  group_by(level)  %>% 
+  summarise(prev_above = round((sum(pct)) * 100)) %>% 
+  pivot_wider(names_from = level,values_from = prev_above)
+```
 
@@
 Examination of the prevalence of taxa in samples at each level supports this idea.
-A majority of taxa were present in greater than 75% of samples at the phylum (67% of taxa) and class (63% of taxa) levels.
-The opposite was observed at the OTU and ASV level where 60% and 53% of taxa respectively were only present in less than 25% of samples (Supplemental Figure 3).
-Of note, the ML pipeline includes a pre-processing step that occured prior to training and classifying the ML models.
+A majority of taxa were present in greater than 70% of samples at the phylum (`r pct_above$phylum`% of taxa) and class (`r pct_above$class`% of taxa) levels.
+The opposite was observed at the OTU and ASV level where `r pct_below$otu`% and `r pct_below$asv`% of taxa respectively were only present in 20% or less of the samples (Supplemental Figure 3).
+Of note, the ML pipeline includes a pre-processing step that occurred prior to training and classifying the ML models.
 As part of this step, taxa were removed that would not provide useful information to build the model.
@@
 The resulting ASV table was subsampled for consistency with the mothur data.
-These results were consistent for ASVs generated with DADA2 [@callahan2016] (median AUROC: `r round(dada2_med_auc,digits=3)`).
+These results were consistent for ASVs generated with DADA2 [@callahan2016] (mean AUROC: `r format_decimal(dada2_mean_auc,digits=3)`).
 
@@
 
-**Figure 1: Random Forest Model Performance.** **A)** Boxplots with points of area under the receiver operating characteristic curve (AUROC) values on the test dataset for 100 seeds predicting SRNs using a Random Forest model.
+**Figure 1: Random Forest Model Performance.** **A)** Strip plot of the area under the receiver operating characteristic curve (AUROC) values on the test dataset for 100 seeds predicting SRNs using a random forest model.
+Black points denote the mean and lines denote the standard deviation.
 Dashed line denotes AUROC of 0.5 which is equivalent to random classification.
 Significance between taxonomic levels was quantified by comparing the difference in mean AUROC and is denoted by letters A through E on the right side of the plot; taxonomic levels with the same letter are in the same significance group and are not significantly different from one another.
-**B)** Strip plot of the sensitivity at a specificity of 90% across the 100 model iterations.
-Black points denote the median and the lines denote the IQR.
-The letters W through Z denote the significance groups.
+**B)** Strip plot of the sensitivity at a specificity of 90% across the 100 model iterations for each taxonomic level.
+Black points denote the mean and the lines denote the standard deviation.
+The letters W through Z on the right side of the plot denote the significance groups.
 
@@
   select(-n_samples) %>%
-  mutate(pct_kept=round((n_features_preproc/n_features)*100,digits=1)) %>%
+  mutate(pct_kept=format_decimal((n_features_preproc/n_features)*100,digits=1)) %>%
   mutate(pct_kept=paste0(pct_kept," %"),
@@
 
-**Supplemental Figure 1: Model Performance across Taxonomy.** Boxplot of AUROC values on the test dataset for 100 seeds for each model type across all taxonomic levels.
+**Supplemental Figure 1: Model Performance across Taxonomy.** Boxplots of AUROC values on the test dataset split for 100 seeds for each model type across all taxonomic levels.
 
@@
 
-**Supplemental Figure 2: Averaged ROC curves** ROC curves averaged across the 100 iterations of the model. The shaded region represents the standard deviation form the mean.
+**Supplemental Figure 2: Averaged ROC curves** ROC curves with averaged true positive rate across the 100 iterations of the model.
+The shaded region represents the standard deviation form the mean.Dashed line represents an AUROC of 0.5, which is equivalent to random classification.
+The mean AUROC for each taxonomic level is printed on the bottom right of the plot.
 
@@
 ![](figure_s3.png){height="45%"}\
-**Supplemental Figure 3: Prevalence of Taxa in Samples.** Distribution of taxa prevalence in samples at each taxonomic level.
+**Supplemental Figure 3: Prevalence of Taxa in Samples.** Distribution of the prevalence of taxa across samples at each taxonomic level.
+Percent of samples is split into 10 groups where the first is taxa present in 0 to 10% of samples, then >10% to 20% of samples, and so on.
+The total number of taxa for each taxonomic level after preprocessing is in parenthesis next to the title of the plot (the name of the taxonomic level).
 
@@
 ![](figure_s4.png){height="85%"}\
-**Supplemental Figure 4: Top 10 important taxa at each taxonomic level.** Summary of the 10 most important taxa for the random forest models at each taxonomic level based on the average decrease in AUC when the feature is permuted.
+**Supplemental Figure 4: Top 10 important taxa at each taxonomic level.** Summary of the 10 most important taxa for the random forest models at each taxonomic level based on the average decrease in AUROC when the feature is permuted.
 

### 2021-08-30 paper/manuscript.tex
draft to send to coauthors

@@
 
-Courtney R. Armour, Begüm D.Topçuoğlu, Andrea Garretto, Patrick D.
-Schloss \({^\dagger}\)
+Courtney R. Armour\({^1}\), Begüm D.Topçuoğlu, Andrea Garretto\({^1}\),
+Patrick D. Schloss \({^1}\)\({^*}\)
 
-\vspace{20mm}
+\vspace{5mm}
 
-\({\dagger}\) To whom correspondence should be addressed:
+\({^1}\) Department of Microbiology and Immunology, University of
+Michigan, Ann Arbor
 
+\vspace{5mm}
+
+\({^*}\) Correspondence: Patrick D. Schloss
 \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
-
-Department of Microbiology
-
-University of Michigan
-
-Ann Arbor, MI 48109
 
@@
 the receiver operating characteristic curve (AUROC). Within the RF
-model, the highest AUROCs were observed for family (median AUROC:
-0.687), genus (median AUROC: 0.686), and OTU (median AUROC: 0.698) level
-data with no significant difference between the three (Figure 1A,
-Supplemental Figure 2). Performance with ASVs (median AUROC: 0.676) was
-significantly lower than OTUs (p \textless{} 0.01), but comparable to
-family (p = 0.06) and genus (p = 0.05) levels (Figure 1A). These results
-suggest that finer taxonomic resolution is not necessarily better for
+model, the highest AUROCs were observed for family (mean AUROC: 0.689),
+genus (mean AUROC: 0.690), and OTU (mean AUROC: 0.693) level data with
+no significant difference between the three (Figure 1A, Supplemental
+Figure 2). Performance with ASVs (mean AUROC: 0.676) was significantly
+lower than OTUs (p \textless{} 0.01), but comparable to family (p =
+0.06) and genus (p = 0.05) levels (Figure 1A). These results suggest
+that increased resolution improves model performance until the OTU level
+where further taxonomic resolution is not necessarily better for
 identifying individuals with SRNs based on microbiome composition.
@@
 the overall model performance, they summarize the performance across all
-thresholds and can be misleading since models with the same AUC can have
-different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{5}).
+thresholds and can be misleading since models with the same AUROC can
+have different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{5}).
 Depending on the intended implementation of the model, one may want to
@@
 highest sensitivity values were observed for family, genus, and OTU
-level data (Figure 1B), consistent with the AUROC results. phylum,
+level data (Figure 1B), consistent with the AUROC results. Phylum,
 class, order, and ASV sensitivity values were all significantly lower
@@
 enough. Examination of the prevalence of taxa in samples at each level
-supports this idea. A majority of taxa were present in greater than 75\%
+supports this idea. A majority of taxa were present in greater than 70\%
 of samples at the phylum (67\% of taxa) and class (63\% of taxa) levels.
-The opposite was observed at the OTU and ASV level where 60\% and 53\%
-of taxa respectively were only present in less than 25\% of samples
+The opposite was observed at the OTU and ASV level where 50\% and 41\%
+of taxa respectively were only present in 20\% or less of the samples
 (Supplemental Figure 3). Of note, the ML pipeline includes a
-pre-processing step that occured prior to training and classifying the
+pre-processing step that occurred prior to training and classifying the
 ML models. As part of this step, taxa were removed that would not
@@
 mothur data. These results were consistent for ASVs generated with DADA2
-(\protect\hyperlink{ref-callahan2016}{6}) (median AUROC: 0.66).
+(\protect\hyperlink{ref-callahan2016}{6}) (mean AUROC: 0.659).
 
@@
 
-\textbf{Figure 1: Random Forest Model Performance.} \textbf{A)} Boxplots
-with points of area under the receiver operating characteristic curve
+\textbf{Figure 1: Random Forest Model Performance.} \textbf{A)} Strip
+plot of the area under the receiver operating characteristic curve
 (AUROC) values on the test dataset for 100 seeds predicting SRNs using a
-Random Forest model. Dashed line denotes AUROC of 0.5 which is
-equivalent to random classification. Significance between taxonomic
-levels was quantified by comparing the difference in mean AUROC and is
-denoted by letters A through E on the right side of the plot; taxonomic
-levels with the same letter are in the same significance group and are
-not significantly different from one another. \textbf{B)} Strip plot of
-the sensitivity at a specificity of 90\% across the 100 model
-iterations. Black points denote the median and the lines denote the IQR.
-The letters W through Z denote the significance groups.
+random forest model. Black points denote the mean and lines denote the
+standard deviation. Dashed line denotes AUROC of 0.5 which is equivalent
+to random classification. Significance between taxonomic levels was
+quantified by comparing the difference in mean AUROC and is denoted by
+letters A through E on the right side of the plot; taxonomic levels with
+the same letter are in the same significance group and are not
+significantly different from one another. \textbf{B)} Strip plot of the
+sensitivity at a specificity of 90\% across the 100 model iterations for
+each taxonomic level. Black points denote the mean and the lines denote
+the standard deviation. The letters W through Z on the right side of the
+plot denote the significance groups.
 
@@
 \textbf{Supplemental Figure 1: Model Performance across Taxonomy.}
-Boxplot of AUROC values on the test dataset for 100 seeds for each model
-type across all taxonomic levels.
+Boxplots of AUROC values on the test dataset split for 100 seeds for
+each model type across all taxonomic levels.
 
@@
 
-\textbf{Supplemental Figure 2: Averaged ROC curves} ROC curves averaged
-across the 100 iterations of the model. The shaded region represents the
-standard deviation form the mean.
+\textbf{Supplemental Figure 2: Averaged ROC curves} ROC curves with
+averaged true positive rate across the 100 iterations of the model. The
+shaded region represents the standard deviation form the mean.Dashed
+line represents an AUROC of 0.5, which is equivalent to random
+classification. The mean AUROC for each taxonomic level is printed on
+the bottom right of the plot.
 
@@
 \textbf{Supplemental Figure 3: Prevalence of Taxa in Samples.}
-Distribution of taxa prevalence in samples at each taxonomic level.
+Distribution of the prevalence of taxa across samples at each taxonomic
+level. Percent of samples is split into 10 groups where the first is
+taxa present in 0 to 10\% of samples, then \textgreater10\% to 20\% of
+samples, and so on. The total number of taxa for each taxonomic level
+after preprocessing is in parenthesis next to the title of the plot (the
+name of the taxonomic level).
 
@@
 level.} Summary of the 10 most important taxa for the random forest
-models at each taxonomic level based on the average decrease in AUC when
-the feature is permuted.
+models at each taxonomic level based on the average decrease in AUROC
+when the feature is permuted.
 

### 2021-08-30 paper/manuscript.tex
update word document

@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-]{article}
-\usepackage{amsmath,amssymb}
+\documentclass[]{article}
+\usepackage[T1]{fontenc}
 \usepackage{lmodern}
+\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
   \fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use microtype if available
+\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
+\usepackage[margin=1.0in]{geometry}
+\usepackage{longtable,booktabs}
+\usepackage{graphicx}
+% Redefine \includegraphics so that, unless explicit options are
+% given, the image width will not exceed the width of the page.
+% Images get their normal width if they fit onto the page, but
+% are scaled down if they would overflow the margins.
 \makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
+\def\ScaleIfNeeded{%
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+\let\Oldincludegraphics\includegraphics
+{%
+ \catcode`\@=11\relax%
+ \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
+}%
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
 \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1.0in]{geometry}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{0}
+
+\author{}
+\date{\vspace{-2.5em}}
 \usepackage{helvet}
@@
 \usepackage{xcolor}
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\author{}
-\date{\vspace{-2.5em}}
 
@@
 
-\hypertarget{taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}{%
 \section{Taxonomic Resolution Matters for Microbiome-Based
 Classification of Colorectal
-Cancer}\label{taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}}
+Cancer}\label{taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}
 
@@
 
-Courtney R. Armour\({^1}\), Begüm D.Topçuoğlu, Andrea Garretto\({^1}\),
-Patrick D. Schloss \({^1}\)\({^*}\)
+Courtney R. Armour${^1}$, Begüm D.Topçuoğlu, Andrea Garretto${^1}$,
+Patrick D. Schloss ${^1}$${^*}$
 
@@
 
-\({^1}\) Department of Microbiology and Immunology, University of
+${^1}$ Department of Microbiology and Immunology, University of
 Michigan, Ann Arbor
@@
 
-\({^*}\) Correspondence: Patrick D. Schloss
+${^*}$ Correspondence: Patrick D. Schloss
 \href{mailto:pschloss@umich.edu}{pschloss@umich.edu}
@@
 Colorectal cancer is one of the most common cancers in men and women and
-a leading cause of cancer related deaths in the United States
-(\protect\hyperlink{ref-siegel2020}{1}). Early detection and treatment
-are essential to increase survival rates, but for a variety of reasons
-including the invasiveness and high cost of screening
-(i.e.~colonoscopy), many people do not comply with recommended screening
-guidelines (\protect\hyperlink{ref-garcuxeda2011a}{2}) prompting a need
-for low cost, non-invasive detection methods. A growing body of research
-points to the gut microbiome as a promising target for non-invasive
-screening to detect screen relevant neoplasia (SRNs) consisting of
-pre-cancerous polyps (i.e.~advanced adenomas) and carcinomas. Efforts to
-realize the diagnostic potential of the gut microbiome in detecting SRNs
-have focused on machine learning (ML) methods using abundances of
-operational taxonomic unit (OTU) classifications based on amplicon
-sequencing of the 16S rRNA gene \{\}. However, whether this is the
-optimal taxonomic resolution for classifying SRNs from microbiome data
-is unknown. Additionally, recent work has pushed for the use of amplicon
-sequence variants (ASVs) to replace OTUs for marker-gene analysis
-because of the improved resolution with ASVs
-(\protect\hyperlink{ref-callahan2017}{3}). However, whether the
-additional resolution provided by ASVs is useful for ML classification
-is unclear \{\}. Since ML classification relies on consistent
-differences between groups, its possible that the resolution at the ASV
-level is too individualized to accurately differentiate groups.
-Topçuoğlu \emph{et al} (\protect\hyperlink{ref-topuxe7uolu2020}{4})
-recently demonstrated how to effectively apply machine learning (ML)
-methods to microbiome based classification problems and developed a
+a leading cause of cancer related deaths in the United States (1). Early
+detection and treatment are essential to increase survival rates, but
+for a variety of reasons including the invasiveness and high cost of
+screening (i.e.~colonoscopy), many people do not comply with recommended
+screening guidelines (2) prompting a need for low cost, non-invasive
+detection methods. A growing body of research points to the gut
+microbiome as a promising target for non-invasive screening to detect
+screen relevant neoplasia (SRNs) consisting of pre-cancerous polyps
+(i.e.~advanced adenomas) and carcinomas. Efforts to realize the
+diagnostic potential of the gut microbiome in detecting SRNs have
+focused on machine learning (ML) methods using abundances of operational
+taxonomic unit (OTU) classifications based on amplicon sequencing of the
+16S rRNA gene \{\}. However, whether this is the optimal taxonomic
+resolution for classifying SRNs from microbiome data is unknown.
+Additionally, recent work has pushed for the use of amplicon sequence
+variants (ASVs) to replace OTUs for marker-gene analysis because of the
+improved resolution with ASVs (3). However, whether the additional
+resolution provided by ASVs is useful for ML classification is unclear
+\{\}. Since ML classification relies on consistent differences between
+groups, its possible that the resolution at the ASV level is too
+individualized to accurately differentiate groups. Topçuoğlu \emph{et
+al} (4) recently demonstrated how to effectively apply machine learning
+(ML) methods to microbiome based classification problems and developed a
 framework for applying ML practices in a more reproducible way
@@
 thresholds and can be misleading since models with the same AUROC can
-have different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{5}).
-Depending on the intended implementation of the model, one may want to
-optimize the true-positive rate (or sensitivity) over the false-positive
-rate (or 1-specificity), or vice versa. In this case, the optimal model
-will detect as many true positives (people with SRNs) as possible. To
-further compare the model performance across taxonomic levels we
-compared the sensitivity of the models at a specificity of 90\%. The
-highest sensitivity values were observed for family, genus, and OTU
-level data (Figure 1B), consistent with the AUROC results. Phylum,
-class, order, and ASV sensitivity values were all significantly lower
-than family, genus, and OTU sensitivity values (Figure 1B). This
-analysis further supports the observation that finer resolution is does
-not improve SRN detection.
+have different ROC curve shapes (5). Depending on the intended
+implementation of the model, one may want to optimize the true-positive
+rate (or sensitivity) over the false-positive rate (or 1-specificity),
+or vice versa. In this case, the optimal model will detect as many true
+positives (people with SRNs) as possible. To further compare the model
+performance across taxonomic levels we compared the sensitivity of the
+models at a specificity of 90\%. The highest sensitivity values were
+observed for family, genus, and OTU level data (Figure 1B), consistent
+with the AUROC results. Phylum, class, order, and ASV sensitivity values
+were all significantly lower than family, genus, and OTU sensitivity
+values (Figure 1B). This analysis further supports the observation that
+finer resolution is does not improve SRN detection.
 
@@
 
-\hypertarget{materials-and-methods}{%
-\subsection{Materials and Methods}\label{materials-and-methods}}
+\subsection{Materials and Methods}\label{materials-and-methods}
 
@@
 mothur data. These results were consistent for ASVs generated with DADA2
-(\protect\hyperlink{ref-callahan2016}{6}) (mean AUROC: 0.659).
+(6) (mean AUROC: 0.659).
 
@@
 
-\hypertarget{figures}{%
-\subsection{Figures}\label{figures}}
+\subsection{Figures}\label{figures}
 
+\begin{figure}[htbp]
+\centering
 \includegraphics{figure_1.png}
+\end{figure}
 
@@
 
-\hypertarget{tables}{%
-\subsection{Tables}\label{tables}}
+\subsection{Tables}\label{tables}
 
-\begin{longtable}[]{@{}lrrc@{}}
-\toprule
+\begin{longtable}[c]{@{}lrrc@{}}
+\toprule\addlinespace
 Taxonomic Level & Number of Features &
 \makecell[c]{Number of Features \\ After Preprocessing} &
-\makecell[c]{Percent of Features Kept \\ After Preprocessing} \\
-\midrule
-\endhead
-Phylum & 19 & 9 & 47.4 \% \\
-Class & 36 & 19 & 52.8 \% \\
-Order & 65 & 28 & 43.1 \% \\
-Family & 124 & 54 & 43.5 \% \\
-Genus & 316 & 115 & 36.4 \% \\
-OTU & 20,079 & 705 & 3.5 \% \\
-ASV & 104,106 & 478 & 0.5 \% \\
+\makecell[c]{Percent of Features Kept \\ After Preprocessing}
+\\\addlinespace
+\midrule\endhead
+Phylum & 19 & 9 & 47.4 \%
+\\\addlinespace
+Class & 36 & 19 & 52.8 \%
+\\\addlinespace
+Order & 65 & 28 & 43.1 \%
+\\\addlinespace
+Family & 124 & 54 & 43.5 \%
+\\\addlinespace
+Genus & 316 & 115 & 36.4 \%
+\\\addlinespace
+OTU & 20,079 & 705 & 3.5 \%
+\\\addlinespace
+ASV & 104,106 & 478 & 0.5 \%
+\\\addlinespace
 \bottomrule
@@
 
-\hypertarget{supplemental-figures}{%
-\subsection{Supplemental Figures}\label{supplemental-figures}}
+\subsection{Supplemental Figures}\label{supplemental-figures}
 
+\begin{figure}[htbp]
+\centering
 \includegraphics{figure_s1.png}
+\end{figure}
 
@@
 
-\includegraphics[width=\textwidth,height=0.5\textheight]{figure_s2.png}
+\includegraphics{figure_s2.png}\{height=``50\%''\}
 
@@
 
-\includegraphics[width=\textwidth,height=0.45\textheight]{figure_s3.png}\\
-\textbf{Supplemental Figure 3: Prevalence of Taxa in Samples.}
-Distribution of the prevalence of taxa across samples at each taxonomic
-level. Percent of samples is split into 10 groups where the first is
-taxa present in 0 to 10\% of samples, then \textgreater10\% to 20\% of
-samples, and so on. The total number of taxa for each taxonomic level
-after preprocessing is in parenthesis next to the title of the plot (the
-name of the taxonomic level).
+\includegraphics{figure_s3.png}\{height=``45\%''\}\\\textbf{Supplemental
+Figure 3: Prevalence of Taxa in Samples.} Distribution of the prevalence
+of taxa across samples at each taxonomic level. Percent of samples is
+split into 10 groups where the first is taxa present in 0 to 10\% of
+samples, then \textgreater{}10\% to 20\% of samples, and so on. The
+total number of taxa for each taxonomic level after preprocessing is in
+parenthesis next to the title of the plot (the name of the taxonomic
+level).
 
@@
 
-\includegraphics[width=\textwidth,height=0.85\textheight]{figure_s4.png}\\
-\textbf{Supplemental Figure 4: Top 10 important taxa at each taxonomic
-level.} Summary of the 10 most important taxa for the random forest
-models at each taxonomic level based on the average decrease in AUROC
-when the feature is permuted.
+\includegraphics{figure_s4.png}\{height=``85\%''\}\\\textbf{Supplemental
+Figure 4: Top 10 important taxa at each taxonomic level.} Summary of the
+10 most important taxa for the random forest models at each taxonomic
+level based on the average decrease in AUROC when the feature is
+permuted.
 
@@
 
-\hypertarget{references}{%
-\subsection*{References}\label{references}}
-\addcontentsline{toc}{subsection}{References}
+\subsection{References}\label{references}
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{1}
-\leavevmode\hypertarget{ref-siegel2020}{}%
-\CSLLeftMargin{1. }
-\CSLRightInline{\textbf{Siegel RL}, \textbf{Miller KD}, \textbf{Sauer
-AG}, \textbf{Fedewa SA}, \textbf{Butterly LF}, \textbf{Anderson JC},
-\textbf{Cercek A}, \textbf{Smith RA}, \textbf{Jemal A}. 2020. Colorectal
-cancer statistics, 2020. CA: A Cancer Journal for Clinicians
-\textbf{70}:145--164. doi:\url{https://doi.org/10.3322/caac.21601}.}
+1. \textbf{{Siegel} R}, \textbf{{Miller} K}, \textbf{{Sauer} A},
+\textbf{{Fedewa} S}, \textbf{{Butterly} L}, \textbf{{Anderson} J},
+\textbf{{Cercek} A}, \textbf{{Smith} R}, \textbf{{Jemal} A}. 2020.
+Colorectal cancer statistics, 2020. CA: A Cancer Journal for Clinicians
+\textbf{70}:145--164.
+doi:\href{http://dx.doi.org/https://doi.org/10.3322/caac.21601}{https://doi.org/10.3322/caac.21601}.
 
-\leavevmode\hypertarget{ref-garcuxeda2011a}{}%
-\CSLLeftMargin{2. }
-\CSLRightInline{\textbf{García G}, \textbf{Z A}. 2011. Factors
-Influencing Colorectal Cancer Screening Participation. Gastroenterology
-Research and Practice \textbf{2012}:e483417.
-doi:\href{https://doi.org/10.1155/2012/483417}{10.1155/2012/483417}.}
+2. \textbf{{García} G}, \textbf{{Z} A}. 2011. Factors influencing
+colorectal cancer screening participation. Gastroenterology Research and
+Practice \textbf{2012}:e483417.
+doi:\href{http://dx.doi.org/10.1155/2012/483417}{10.1155/2012/483417}.
 
-\leavevmode\hypertarget{ref-callahan2017}{}%
-\CSLLeftMargin{3. }
-\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
-\textbf{Holmes SP}. 2017. Exact sequence variants should replace
-operational taxonomic units in marker-gene data analysis. The ISME
-Journal \textbf{11}:2639--2643.
-doi:\href{https://doi.org/10.1038/ismej.2017.119}{10.1038/ismej.2017.119}.}
+3. \textbf{{Callahan} B}, \textbf{{McMurdie} P}, \textbf{{Holmes} S}.
+2017. Exact sequence variants should replace operational taxonomic units
+in marker-gene data analysis. The ISME Journal \textbf{11}:2639--2643.
+doi:\href{http://dx.doi.org/10.1038/ismej.2017.119}{10.1038/ismej.2017.119}.
 
-\leavevmode\hypertarget{ref-topuxe7uolu2020}{}%
-\CSLLeftMargin{4. }
-\CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
-\textbf{Ruffin MT}, \textbf{Wiens J}, \textbf{Schloss PD}. 2020. A
-Framework for Effective Application of Machine Learning to
-Microbiome-Based Classification Problems. mBio \textbf{11}.
-doi:\href{https://doi.org/10.1128/mBio.00434-20}{10.1128/mBio.00434-20}.}
+4. \textbf{{Topçuo{ğ}lu} B}, \textbf{{Lesniak} N}, \textbf{{Ruffin} M},
+\textbf{{Wiens} J}, \textbf{{Schloss} P}. 2020. A framework for
+effective application of machine learning to microbiome-based
+classification problems. mBio \textbf{11}.
+doi:\href{http://dx.doi.org/10.1128/mBio.00434-20}{10.1128/mBio.00434-20}.
 
-\leavevmode\hypertarget{ref-lobo2008}{}%
-\CSLLeftMargin{5. }
-\CSLRightInline{\textbf{Lobo JM}, \textbf{Jiménez-Valverde A},
-\textbf{Real R}. 2008. AUC: a misleading measure of the performance of
-predictive distribution models. Global Ecology and Biogeography
+5. \textbf{{Lobo} J}, \textbf{{Jiménez-Valverde} A}, \textbf{{Real} R}.
+2008. AUC: a misleading measure of the performance of predictive
+distribution models. Global Ecology and Biogeography
 \textbf{17}:145--151.
-doi:\href{https://doi.org/10.1111/j.1466-8238.2007.00358.x}{10.1111/j.1466-8238.2007.00358.x}.}
+doi:\href{http://dx.doi.org/10.1111/j.1466-8238.2007.00358.x}{10.1111/j.1466-8238.2007.00358.x}.
 
-\leavevmode\hypertarget{ref-callahan2016}{}%
-\CSLLeftMargin{6. }
-\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
-\textbf{Rosen MJ}, \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes
-SP}. 2016. DADA2: High-resolution sample inference from Illumina
-amplicon data. Nature Methods \textbf{13}:581--583.
-doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.}
-
-\end{CSLReferences}
+6. \textbf{{Callahan} B}, \textbf{{McMurdie} P}, \textbf{{Rosen} M},
+\textbf{{Han} A}, \textbf{{Johnson} A}, \textbf{{Holmes} S}. 2016.
+DADA2: High-resolution sample inference from illumina amplicon data.
+Nature Methods \textbf{13}:581--583.
+doi:\href{http://dx.doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.
 

### 2021-08-31 paper/manuscript.Rmd
final draft for review

@@
 output:
-  pdf_document:
-    keep_tex: true
-  word_document: 
-    reference_docx: word_formatting.docx
+  rmarkdown::pdf_document:
+    keep_tex: yes
 geometry: margin=1.0in

### 2021-09-21 paper/manuscript.Rmd
manuscript edits

@@
 
-Courtney R. Armour${^1}$, Begüm D.Topçuoğlu, Andrea Garretto${^1}$, Patrick D. Schloss ${^1}$${^*}$
+Courtney R. Armour${^1}$, Begüm D. Topçuoğlu$^{1,2}$, Andrea Garretto${^1}$, Patrick D. Schloss ${^1}$${^*}$
 
@@
 
-${^1}$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor
+${^1}$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor MI
+
+${^2}$ Exploratory Science Center, Merck & Co., Inc., Cambridge, Massachusetts, USA
 
@@
 
-**observation format - max 1200 words, 2 figures, 25 ref**
+**observation format**
 
@@
 
-## Abstract (max 250 words)
+### Abstract  
+Colorectal cancer is a common and deadly disease in the United States accounting for over 50,000 deaths in 2020.
+This progressive disease is highly preventable with early detection and treatment, but many people do not comply with the recommended screening guidelines.
+The gut microbiome has emerged as a promising target for non-invasive detection of colorectal cancer.
+Most microbiome-based classification efforts utilize taxonomic abundance data from operational taxonomic units (OTUs) or amplicon sequence variants (ASVs).
+However, it is unknown which taxonomic resolution is optimal for microbiome-based classification of colorectal cancer.
+To address this question, we used a reproducible machine learning framework to quantify classification performance of models based on data annotated to phylum, class, order, family, genus, OTU, and ASV levels.
+We found that model performance increased with increasing taxonomic resolution, up to the family level where performance was equal among family, genus, and OTU data before decreasing at the ASV level.
+These results are a step towards optimizing a microbiome based diagnostic tool.
 
-## Importance (max 150 words)
+### Importance  
+Despite being highly preventable, colorectal cancer remains a leading cause of cancer related death in the United States.
+Low-cost, non-invasive detection methods could greatly improve our ability to identify and treat early stages of disease.
+While the microbiome has shown promise as a diagnostic tool, there are numerous details to refine before a diagnostic could be widely deployed.
+Among these are the need for standardization of sample collection, storage, processing, and analysis.
+We address one aspect of data analysis by examining which taxonomic level provides the optimal resolution for microbiome-based classification of colorectal cancer.
+This is an important step to inform the development of microbiome-based diagnostic tools.
 
@@
 Colorectal cancer is one of the most common cancers in men and women and a leading cause of cancer related deaths in the United States [@siegel2020].
-Early detection and treatment are essential to increase survival rates, but for a variety of reasons including the invasiveness and high cost of screening (i.e.
-colonoscopy), many people do not comply with recommended screening guidelines [@garcía2011a] prompting a need for low cost, non-invasive detection methods.
-A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasia (SRNs) consisting of pre-cancerous polyps (i.e.
-advanced adenomas) and carcinomas.
-Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances of operational taxonomic unit (OTU) classifications based on amplicon sequencing of the 16S rRNA gene {}.
-However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown.
-Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs [@callahan2017].
-However, whether the additional resolution provided by ASVs is useful for ML classification is unclear {}.
-Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups.
+Early detection and treatment are essential to increase survival rates, but for reasons such as invasiveness and high screening costs (i.e.
+colonoscopy), many people do not comply with recommended screening guidelines [@garcía2011].
+This prompts a need for low cost, non-invasive detection methods.
+A growing body of research points to the gut microbiome as a promising target for non-invasive detection of screen relevant neoplasia (SRNs) consisting of advanced adenomas and carcinomas [@zackular2014; @zeller2014].
+The diagnostic potential of the gut microbiome in detecting SRNs has been explored through machine learning (ML) methods using abundances of operational taxonomic unit (OTU) classifications based on amplicon sequencing of the 16S rRNA gene [@zackular2014].
+Recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs [@callahan2017].
+However, it is unclear if OTUs are the optimal taxonomic resolution for classifying SRNs from microbiome data or whether the additional resolution provided by ASVs is useful for ML classification.
 Topçuoğlu *et al* [@topçuolu2020] recently demonstrated how to effectively apply machine learning (ML) methods to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
@@
 
-Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated abundance tables with mothur {} annotated to phylum, class, order, family, genus, OTU and ASV levels.
-Using the taxonomic abundance data, we quantified how reliably samples could be classified as "normal" or "SRN" (i.e.
-advanced adenoma or carcinoma) using five machine learning methods with the mikropml R package (methods).
-Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1).
+Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated taxonomic abundance tables with mothur [@schloss2009] annotated to phylum, class, order, family, genus, OTU and ASV levels.
+Using the taxonomic abundance data and the mikropml R package [@topçuolu2021], we quantified how reliably samples could be classified as SRN or normal using five machine learning methods including random forest, L2-regularized logistic regression, decision tree, gradient boosted trees (XGBoost), and support vector machine with radial basis kernel (SVM radial).
+Across the five machine learning methods tested, model performance increased with increasing taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1).
 Regardless of the taxonomic level, random forest (RF) models consistently had the largest area under the receiver operating characteristic curve (AUROC).
@@
 Performance with ASVs (mean AUROC: `r format_decimal(rf_mean_auc["asv"],digits=3)`) was significantly lower than OTUs (p \< 0.01), but comparable to family (p = `r format_decimal(rf_pvals["family","asv"],digits=2)`) and genus (p = `r format_decimal(rf_pvals["genus","asv"],digits=2)`) levels (Figure 1A).
-These results suggest that increased resolution improves model performance until the OTU level where further taxonomic resolution is not necessarily better for identifying individuals with SRNs based on microbiome composition.
+These results suggest that increased resolution improves model performance up to the OTU level where further taxonomic resolution is not necessarily better for identifying individuals with SRNs based on microbiome composition.
 
 While comparing AUROC values between models is a useful way to assess the overall model performance, they summarize the performance across all thresholds and can be misleading since models with the same AUROC can have different ROC curve shapes [@lobo2008].
-Depending on the intended implementation of the model, one may want to optimize the true-positive rate (or sensitivity) over the false-positive rate (or 1-specificity), or vice versa.
+Depending on the intended implementation of the model, one may want to optimize the true-positive rate (sensitivity) over the false-positive rate (1-specificity) or vice versa.
 In this case, the optimal model will detect as many true positives (people with SRNs) as possible.
@@
 Phylum, class, order, and ASV sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (Figure 1B).
-This analysis further supports the observation that finer resolution is does not improve SRN detection.
+This analysis further supports the observation that finer resolution does not improve SRN detection.
 
@@
 A majority of taxa were present in greater than 70% of samples at the phylum (`r pct_above$phylum`% of taxa) and class (`r pct_above$class`% of taxa) levels.
-The opposite was observed at the OTU and ASV level where `r pct_below$otu`% and `r pct_below$asv`% of taxa respectively were only present in 20% or less of the samples (Supplemental Figure 3).
+The opposite was observed at the OTU and ASV level where `r pct_below$otu`% and `r pct_below$asv`% of taxa, respectively, were only present in 20% or less of the samples (Supplemental Figure 3).
 Of note, the ML pipeline includes a pre-processing step that occurred prior to training and classifying the ML models.
-As part of this step, taxa were removed that would not provide useful information to build the model.
 For example, perfectly correlated taxa provide the same information to build the model and thus can be collapsed.
-Additionally, features with zero or near-zero variance across samples would not help the model differentiate groups and thus can be removed.
+Additionally, features with zero or near-zero variance across samples were removed.
 Interestingly, despite starting with `r format_number(input %>% filter(level == "asv") %>% select(n_features) %>% pull())` ASVs, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing.
 At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` of the `r format_number(input %>% filter(level == "otu") %>% select(n_features) %>% pull())` OTUs (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).
-While the resolution provided by ASVs is useful in certain contexts {}, these results suggest that the resolution is too fine for use in machine learning classification of SRNs based on microbiome composition.
+While the resolution provided by ASVs is useful in certain contexts[@eren2013; @eren2015], these results suggest that the resolution is too fine for use in machine learning classification of SRNs based on microbiome composition.
 
 A look into the most important taxa at each level for classifying samples revealed some nesting where several genera and their higher taxonomic classifications were in the top 10 most important taxa (Supplemental Figure 4).
-For example, the genus *Gemella* was an important taxon at the genus and OTU levels and its higher classifications were also important (*Firmicutes* > *Bacilli* > *Bacillales* > *Bacillales Incertae Sedis XI* > *Gemella*).
+For example, the genus *Gemella* was an important taxon at the genus and OTU levels and its higher classifications were also important (*Firmicutes* \> *Bacilli* \> *Bacillales* \> *Bacillales Incertae Sedis XI* \> *Gemella*).
 *Fusobacterium* displayed a similar pattern, except that the family level classification (*Fusobacteriaceae*) importance was ranked 16th out of 54 families.
@@
 OTU and ASV) did not add additional sensitivity to predicting SRNs based on microbiome composition.
-Additionally, at the ASV level the fine resolution actually impedes model performance due to the sparsity of shared taxa and led to decreased model performance.
-The tendancy for ASV level annotation to split single bacterial genomes into multiple taxa {Pat mSphere 2021} could also be a contributing factor to the sparsity of shared taxa.
+Family, genus, and OTU level data all performed equally.
+At the ASV level the fine resolution actually impedes model performance due to the sparsity of shared taxa and led to decreased model performance.
+The tendency for ASV level annotation to split single bacterial genomes into multiple taxa [@schloss] could also be a contributing factor to the sparsity of shared taxa.
 Overall, either family, genus, or OTU level taxonomy appear to perform equally for predicting subjects with SRNs based on the composition of the gut microbiome.
@@
 
-## Materials and Methods
+### Materials and Methods
 
-***Dataset.*** Raw 16S rRNA gene amplicon sequence data isolated from human gut samples {Baxter} was downloaded from NCBI SRA (accession \#).
+***Dataset.*** Raw 16S rRNA gene amplicon sequence data isolated from human gut samples [@baxter2016] was downloaded from NCBI Sequence Read Archive (SRP062005).
 This dataset contains stool samples from 490 subjects.
@@
 
-***Data processing.*** Sequence data was processed with mothur (1.44.3) using the SILVA reference database (v132) to produce count tables for phylum, class, order, family, genus, OTU, and ASV following the Schloss Lab MiSeq SOP described on the mothur website.
-ASV level data was also produced using DADA2 to ensure consistent results with a different pipeline.
+***Data processing.*** Sequence data was processed with mothur (1.44.3) [@schloss2009] using the SILVA reference database (v132) [@quast2013] to produce count tables for phylum, class, order, family, genus, OTU, and ASV following the Schloss Lab MiSeq SOP described on the mothur website (<https://mothur.org/wiki/miseq_sop/>).
+ASV level data was also produced using DADA2 [@callahan2016] to ensure consistent results with a different pipeline.
 Data was processed following the DADA2 pipeline, but setting pool=TRUE to infer ASVs from the whole dataset rather than per sample.
 The resulting ASV table was subsampled for consistency with the mothur data.
-These results were consistent for ASVs generated with DADA2 [@callahan2016] (mean AUROC: `r format_decimal(dada2_mean_auc,digits=3)`).
+The results were consistent for ASVs generated with DADA2 (mean AUROC: `r format_decimal(dada2_mean_auc,digits=3)`).
 
-***Machine Learning.*** Machine learning models were run with the R package mikropml (v0.0.2) {} to predict the diagnosis category (normal vs SRN) of each sample.
+***Machine Learning.*** Machine learning models were run with the R package mikropml (v0.0.2) [@topçuolu2021] to predict the diagnosis category (normal vs SRN) of each sample.
 Data was preprocessed to normalize values (scale/center), remove values with zero or near-zero variance, and collapse collinear features using default parameters.
@@
 
-To compare performance between taxonomic levels and models, P values were calculated as previously described {begum}.
-To compare sensitivity at 90% specificity, probabilities on the test dataset were collected for each seed and used to construct ROC curves (R pROC::roc).
-From the ROC curves The sensitivity at a specificity of 90% was pulled for each seed.
+To compare performance between taxonomic levels and models, P values were calculated as previously described [@topçuolu2020].
+To compare sensitivity at 90% specificity, probabilities on the test dataset were collected for each seed and used to calculate sensitivity for specificity values ranging from 0 to 1 in 0.01 increments.
+The sensitivity at a specificity of 90% was pulled for each seed.
+The averaged ROC curves were plotted by taking the average and standard deviation of the sensitivity at each specificity value.
 An optional output from the mikropml package is the permuted feature importance which is quantified by iteratively permuting each feature in the model and assessing the change in model performance.
 Features are presumed to be important if the performance of the model, measured by the AUROC, decreases when that feature is permuted.
-Ranking of feature importance was determined by ordering the features based on the change in AUROC where features with a larger decrease in AUROC are ranked higher in importance.
+Ranking of feature importance was determined by ordering the features based on the average change in AUROC across the 100 seeds where features with a larger decrease in AUROC are ranked higher in importance.
 
-To quantify prevalence of the features, the number of samples with non-zero abundance was divided by the total number of samples resulting in values ranging from 0 to 1 where 0 indicates the feature is not found in any samples, 0.5 indicates the feature is found in half of the samples, an 1 indicating the feature is found in all of the samples.
+To quantify prevalence of the features, the number of samples with non-zero abundance was divided by the total number of samples resulting in values ranging from 0 to 1 where 0 indicates the feature is not found in any samples, 0.5 indicates the feature is found in half of the samples, and 1 indicates the feature is found in all of the samples.
 
@@
 
-## Acknowledgements
+### Acknowledgements
+
+This work was supported through a grant from the NIH (R01CA215574).
 
@@
 
-## Figures
+### References
+
+::: {#refs}
+:::
+
+\newpage
+
+### Figures
 
@@
 
-## Tables
+### Tables
 
@@
 
-## Supplemental Figures
+### Supplemental Figures
 
@@
 
-**Supplemental Figure 1: Model Performance across Taxonomy.** Boxplots of AUROC values on the test dataset split for 100 seeds for each model type across all taxonomic levels.
+**Supplemental Figure 1: Model Performance across Taxonomy.** Boxplots of AUROC values from predicting whether samples came from subjects with screen relevant neoplasias (i.e.
+advanced adenoma or cancer) or healthy controls across five machine learning methods including random forest, L2-regularized logistic regression (logistic regression), decision tree, gradient boosted trees (XGBoost), and support vector machine with radial basis kernel (SVM radial).
+Due to the random split of data into training and testing sets, each model was run across 100 seeds to account for variation in training/test datasplits.
 
@@
 
-**Supplemental Figure 2: Averaged ROC curves** ROC curves with averaged true positive rate across the 100 iterations of the model.
-The shaded region represents the standard deviation form the mean.Dashed line represents an AUROC of 0.5, which is equivalent to random classification.
+**Supplemental Figure 2: Averaged ROC curves.** ROC curves with averaged true positive rate (or sensitivity) across the 100 iterations of the random forest model.
+The shaded region represents the standard deviation form the mean.
+Dashed line represents an AUROC of 0.5, which is equivalent to random classification.
 The mean AUROC for each taxonomic level is printed on the bottom right of the plot.
@@
 **Supplemental Figure 3: Prevalence of Taxa in Samples.** Distribution of the prevalence of taxa across samples at each taxonomic level.
-Percent of samples is split into 10 groups where the first is taxa present in 0 to 10% of samples, then >10% to 20% of samples, and so on.
-The total number of taxa for each taxonomic level after preprocessing is in parenthesis next to the title of the plot (the name of the taxonomic level).
+Percent of samples is split into 10 groups where the first is for taxa present in 0 to 10% of samples, then \>10% to 20% of samples, and so on.
+The total number of taxa for each taxonomic level after preprocessing is in parenthesis next to the title of the plot (or the name of the taxonomic level).
 
@@
 **Supplemental Figure 4: Top 10 important taxa at each taxonomic level.** Summary of the 10 most important taxa for the random forest models at each taxonomic level based on the average decrease in AUROC when the feature is permuted.
-
-\newpage
-
-## References
+Dot represents the mean decrease in AUROC and the lines extending from the dot represent the standard deviation from the mean.

### 2021-09-21 paper/manuscript.tex
manuscript edits

--- 2021-08-30 paper/manuscript.tex
@@
 
-Courtney R. Armour\({^1}\), Begüm D.Topçuoğlu, Andrea Garretto\({^1}\),
-Patrick D. Schloss \({^1}\)\({^*}\)
+Courtney R. Armour\({^1}\), Begüm D. Topçuoğlu\(^{1,2}\), Andrea
+Garretto\({^1}\), Patrick D. Schloss \({^1}\)\({^*}\)
 
@@
 \({^1}\) Department of Microbiology and Immunology, University of
-Michigan, Ann Arbor
+Michigan, Ann Arbor MI
+
+\({^2}\) Exploratory Science Center, Merck \& Co., Inc., Cambridge,
+Massachusetts, USA~
 
@@
 
-\textbf{observation format - max 1200 words, 2 figures, 25 ref}
+\textbf{observation format}
 
@@
 
-\hypertarget{abstract-max-250-words}{%
-\subsection{Abstract (max 250 words)}\label{abstract-max-250-words}}
+\hypertarget{abstract}{%
+\subsubsection{Abstract}\label{abstract}}
 
-\hypertarget{importance-max-150-words}{%
-\subsection{Importance (max 150 words)}\label{importance-max-150-words}}
+Colorectal cancer is a common and deadly disease in the United States
+accounting for over 50,000 deaths in 2020. This progressive disease is
+highly preventable with early detection and treatment, but many people
+do not comply with the recommended screening guidelines. The gut
+microbiome has emerged as a promising target for non-invasive detection
+of colorectal cancer. Most microbiome-based classification efforts
+utilize taxonomic abundance data from operational taxonomic units (OTUs)
+or amplicon sequence variants (ASVs). However, it is unknown which
+taxonomic resolution is optimal for microbiome-based classification of
+colorectal cancer. To address this question, we used a reproducible
+machine learning framework to quantify classification performance of
+models based on data annotated to phylum, class, order, family, genus,
+OTU, and ASV levels. We found that model performance increased with
+increasing taxonomic resolution, up to the family level where
+performance was equal among family, genus, and OTU data before
+decreasing at the ASV level. These results are a step towards optimizing
+a microbiome based diagnostic tool.
+
+\hypertarget{importance}{%
+\subsubsection{Importance}\label{importance}}
+
+Despite being highly preventable, colorectal cancer remains a leading
+cause of cancer related death in the United States. Low-cost,
+non-invasive detection methods could greatly improve our ability to
+identify and treat early stages of disease. While the microbiome has
+shown promise as a diagnostic tool, there are numerous details to refine
+before a diagnostic could be widely deployed. Among these are the need
+for standardization of sample collection, storage, processing, and
+analysis. We address one aspect of data analysis by examining which
+taxonomic level provides the optimal resolution for microbiome-based
+classification of colorectal cancer. This is an important step to inform
+the development of microbiome-based diagnostic tools.
 
@@
 (\protect\hyperlink{ref-siegel2020}{1}). Early detection and treatment
-are essential to increase survival rates, but for a variety of reasons
-including the invasiveness and high cost of screening
-(i.e.~colonoscopy), many people do not comply with recommended screening
-guidelines (\protect\hyperlink{ref-garcuxeda2011a}{2}) prompting a need
-for low cost, non-invasive detection methods. A growing body of research
-points to the gut microbiome as a promising target for non-invasive
-screening to detect screen relevant neoplasia (SRNs) consisting of
-pre-cancerous polyps (i.e.~advanced adenomas) and carcinomas. Efforts to
-realize the diagnostic potential of the gut microbiome in detecting SRNs
-have focused on machine learning (ML) methods using abundances of
-operational taxonomic unit (OTU) classifications based on amplicon
-sequencing of the 16S rRNA gene \{\}. However, whether this is the
-optimal taxonomic resolution for classifying SRNs from microbiome data
-is unknown. Additionally, recent work has pushed for the use of amplicon
-sequence variants (ASVs) to replace OTUs for marker-gene analysis
-because of the improved resolution with ASVs
-(\protect\hyperlink{ref-callahan2017}{3}). However, whether the
-additional resolution provided by ASVs is useful for ML classification
-is unclear \{\}. Since ML classification relies on consistent
-differences between groups, its possible that the resolution at the ASV
-level is too individualized to accurately differentiate groups.
-Topçuoğlu \emph{et al} (\protect\hyperlink{ref-topuxe7uolu2020}{4})
-recently demonstrated how to effectively apply machine learning (ML)
-methods to microbiome based classification problems and developed a
-framework for applying ML practices in a more reproducible way
-(mikropml). This analysis utilizes the reproducible framework developed
-by Topçuoğlu \emph{et al} to determine which ML method and taxonomic
-level produce the best performing classifier for detecting SRNs from
-microbiome data.
+are essential to increase survival rates, but for reasons such as
+invasiveness and high screening costs (i.e.~colonoscopy), many people do
+not comply with recommended screening guidelines
+(\protect\hyperlink{ref-garcuxeda2011}{2}). This prompts a need for low
+cost, non-invasive detection methods. A growing body of research points
+to the gut microbiome as a promising target for non-invasive detection
+of screen relevant neoplasia (SRNs) consisting of advanced adenomas and
+carcinomas (\protect\hyperlink{ref-zackular2014}{3},
+\protect\hyperlink{ref-zeller2014}{4}). The diagnostic potential of the
+gut microbiome in detecting SRNs has been explored through machine
+learning (ML) methods using abundances of operational taxonomic unit
+(OTU) classifications based on amplicon sequencing of the 16S rRNA gene
+(\protect\hyperlink{ref-zackular2014}{3}). Recent work has pushed for
+the use of amplicon sequence variants (ASVs) to replace OTUs for
+marker-gene analysis because of the improved resolution with ASVs
+(\protect\hyperlink{ref-callahan2017}{5}). However, it is unclear if
+OTUs are the optimal taxonomic resolution for classifying SRNs from
+microbiome data or whether the additional resolution provided by ASVs is
+useful for ML classification. Topçuoğlu \emph{et al}
+(\protect\hyperlink{ref-topuxe7uolu2020}{6}) recently demonstrated how
+to effectively apply machine learning (ML) methods to microbiome based
+classification problems and developed a framework for applying ML
+practices in a more reproducible way (mikropml). This analysis utilizes
+the reproducible framework developed by Topçuoğlu \emph{et al} to
+determine which ML method and taxonomic level produce the best
+performing classifier for detecting SRNs from microbiome data.
 
 Utilizing publicly available 16S rRNA sequence data from stool of
-patients with SRNs and healthy controls, we generated abundance tables
-with mothur \{\} annotated to phylum, class, order, family, genus, OTU
-and ASV levels. Using the taxonomic abundance data, we quantified how
-reliably samples could be classified as ``normal'' or ``SRN''
-(i.e.~advanced adenoma or carcinoma) using five machine learning methods
-with the mikropml R package (methods). Across the five machine learning
-methods tested, model performance tended to increase with taxonomic
-level usually peaking around genus/OTU level before dropping off
-slightly with ASVs (Supplemental Figure 1). Regardless of the taxonomic
-level, random forest (RF) models consistently had the largest area under
-the receiver operating characteristic curve (AUROC). Within the RF
-model, the highest AUROCs were observed for family (mean AUROC: 0.689),
-genus (mean AUROC: 0.690), and OTU (mean AUROC: 0.693) level data with
-no significant difference between the three (Figure 1A, Supplemental
-Figure 2). Performance with ASVs (mean AUROC: 0.676) was significantly
-lower than OTUs (p \textless{} 0.01), but comparable to family (p =
-0.06) and genus (p = 0.05) levels (Figure 1A). These results suggest
-that increased resolution improves model performance until the OTU level
-where further taxonomic resolution is not necessarily better for
-identifying individuals with SRNs based on microbiome composition.
+patients with SRNs and healthy controls, we generated taxonomic
+abundance tables with mothur (\protect\hyperlink{ref-schloss2009}{7})
+annotated to phylum, class, order, family, genus, OTU and ASV levels.
+Using the taxonomic abundance data and the mikropml R package
+(\protect\hyperlink{ref-topuxe7uolu2021}{8}), we quantified how reliably
+samples could be classified as SRN or normal using five machine learning
+methods including random forest, L2-regularized logistic regression,
+decision tree, gradient boosted trees (XGBoost), and support vector
+machine with radial basis kernel (SVM radial). Across the five machine
+learning methods tested, model performance increased with increasing
+taxonomic level usually peaking around genus/OTU level before dropping
+off slightly with ASVs (Supplemental Figure 1). Regardless of the
+taxonomic level, random forest (RF) models consistently had the largest
+area under the receiver operating characteristic curve (AUROC). Within
+the RF model, the highest AUROCs were observed for family (mean AUROC:
+0.689), genus (mean AUROC: 0.690), and OTU (mean AUROC: 0.693) level
+data with no significant difference between the three (Figure 1A,
+Supplemental Figure 2). Performance with ASVs (mean AUROC: 0.676) was
+significantly lower than OTUs (p \textless{} 0.01), but comparable to
+family (p = 0.06) and genus (p = 0.05) levels (Figure 1A). These results
+suggest that increased resolution improves model performance up to the
+OTU level where further taxonomic resolution is not necessarily better
+for identifying individuals with SRNs based on microbiome composition.
 
@@
 thresholds and can be misleading since models with the same AUROC can
-have different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{5}).
+have different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{9}).
 Depending on the intended implementation of the model, one may want to
-optimize the true-positive rate (or sensitivity) over the false-positive
-rate (or 1-specificity), or vice versa. In this case, the optimal model
-will detect as many true positives (people with SRNs) as possible. To
-further compare the model performance across taxonomic levels we
-compared the sensitivity of the models at a specificity of 90\%. The
-highest sensitivity values were observed for family, genus, and OTU
-level data (Figure 1B), consistent with the AUROC results. Phylum,
-class, order, and ASV sensitivity values were all significantly lower
-than family, genus, and OTU sensitivity values (Figure 1B). This
-analysis further supports the observation that finer resolution is does
-not improve SRN detection.
+optimize the true-positive rate (sensitivity) over the false-positive
+rate (1-specificity) or vice versa. In this case, the optimal model will
+detect as many true positives (people with SRNs) as possible. To further
+compare the model performance across taxonomic levels we compared the
+sensitivity of the models at a specificity of 90\%. The highest
+sensitivity values were observed for family, genus, and OTU level data
+(Figure 1B), consistent with the AUROC results. Phylum, class, order,
+and ASV sensitivity values were all significantly lower than family,
+genus, and OTU sensitivity values (Figure 1B). This analysis further
+supports the observation that finer resolution does not improve SRN
+detection.
 
@@
 The opposite was observed at the OTU and ASV level where 50\% and 41\%
-of taxa respectively were only present in 20\% or less of the samples
+of taxa, respectively, were only present in 20\% or less of the samples
 (Supplemental Figure 3). Of note, the ML pipeline includes a
 pre-processing step that occurred prior to training and classifying the
-ML models. As part of this step, taxa were removed that would not
-provide useful information to build the model. For example, perfectly
-correlated taxa provide the same information to build the model and thus
-can be collapsed. Additionally, features with zero or near-zero variance
-across samples would not help the model differentiate groups and thus
-can be removed. Interestingly, despite starting with 104,106 ASVs, only
-478 (0.5\%) remained after pre-processing. At the OTU level, 705 of the
-20,079 OTUs (3.5\%) remained after preprocessing (Table 1). While the
-resolution provided by ASVs is useful in certain contexts \{\}, these
-results suggest that the resolution is too fine for use in machine
-learning classification of SRNs based on microbiome composition.
+ML models. For example, perfectly correlated taxa provide the same
+information to build the model and thus can be collapsed. Additionally,
+features with zero or near-zero variance across samples were removed.
+Interestingly, despite starting with 104,106 ASVs, only 478 (0.5\%)
+remained after pre-processing. At the OTU level, 705 of the 20,079 OTUs
+(3.5\%) remained after preprocessing (Table 1). While the resolution
+provided by ASVs is useful in certain
+contexts(\protect\hyperlink{ref-eren2013}{10},
+\protect\hyperlink{ref-eren2015}{11}), these results suggest that the
+resolution is too fine for use in machine learning classification of
+SRNs based on microbiome composition.
 
@@
 (e.g.~OTU and ASV) did not add additional sensitivity to predicting SRNs
-based on microbiome composition. Additionally, at the ASV level the fine
-resolution actually impedes model performance due to the sparsity of
-shared taxa and led to decreased model performance. The tendancy for ASV
-level annotation to split single bacterial genomes into multiple taxa
-\{Pat mSphere 2021\} could also be a contributing factor to the sparsity
-of shared taxa. Overall, either family, genus, or OTU level taxonomy
-appear to perform equally for predicting subjects with SRNs based on the
-composition of the gut microbiome. A potential benefit of utilizing
-genus or family level data could be that it may allow for merging data
-generated from different 16S rRNA gene regions or sequencing platforms.
+based on microbiome composition. Family, genus, and OTU level data all
+performed equally. At the ASV level the fine resolution actually impedes
+model performance due to the sparsity of shared taxa and led to
+decreased model performance. The tendency for ASV level annotation to
+split single bacterial genomes into multiple taxa
+(\protect\hyperlink{ref-schloss}{12}) could also be a contributing
+factor to the sparsity of shared taxa. Overall, either family, genus, or
+OTU level taxonomy appear to perform equally for predicting subjects
+with SRNs based on the composition of the gut microbiome. A potential
+benefit of utilizing genus or family level data could be that it may
+allow for merging data generated from different 16S rRNA gene regions or
+sequencing platforms.
 
 \hypertarget{materials-and-methods}{%
-\subsection{Materials and Methods}\label{materials-and-methods}}
+\subsubsection{Materials and Methods}\label{materials-and-methods}}
 
 \textbf{\emph{Dataset.}} Raw 16S rRNA gene amplicon sequence data
-isolated from human gut samples \{Baxter\} was downloaded from NCBI SRA
-(accession \#). This dataset contains stool samples from 490 subjects.
-Based on the available metadata, samples categorized as normal, high
-risk normal, or adenoma were labeled ``normal'' for this analysis and
-samples categorized as advanced adenoma or carcinoma were labeled as
-``screen relevant neoplasia'' (SRN). This resulted in a total of 261
-``normal'' samples and 229 ``SRN'' samples.
+isolated from human gut samples (\protect\hyperlink{ref-baxter2016}{13})
+was downloaded from NCBI Sequence Read Archive (SRP062005). This dataset
+contains stool samples from 490 subjects. Based on the available
+metadata, samples categorized as normal, high risk normal, or adenoma
+were labeled ``normal'' for this analysis and samples categorized as
+advanced adenoma or carcinoma were labeled as ``screen relevant
+neoplasia'' (SRN). This resulted in a total of 261 ``normal'' samples
+and 229 ``SRN'' samples.
 
 \textbf{\emph{Data processing.}} Sequence data was processed with mothur
-(1.44.3) using the SILVA reference database (v132) to produce count
-tables for phylum, class, order, family, genus, OTU, and ASV following
-the Schloss Lab MiSeq SOP described on the mothur website. ASV level
-data was also produced using DADA2 to ensure consistent results with a
-different pipeline. Data was processed following the DADA2 pipeline, but
-setting pool=TRUE to infer ASVs from the whole dataset rather than per
-sample. The resulting ASV table was subsampled for consistency with the
-mothur data. These results were consistent for ASVs generated with DADA2
-(\protect\hyperlink{ref-callahan2016}{6}) (mean AUROC: 0.659).
+(1.44.3) (\protect\hyperlink{ref-schloss2009}{7}) using the SILVA
+reference database (v132) (\protect\hyperlink{ref-quast2013}{14}) to
+produce count tables for phylum, class, order, family, genus, OTU, and
+ASV following the Schloss Lab MiSeq SOP described on the mothur website
+(\url{https://mothur.org/wiki/miseq_sop/}). ASV level data was also
+produced using DADA2 (\protect\hyperlink{ref-callahan2016}{15}) to
+ensure consistent results with a different pipeline. Data was processed
+following the DADA2 pipeline, but setting pool=TRUE to infer ASVs from
+the whole dataset rather than per sample. The resulting ASV table was
+subsampled for consistency with the mothur data. The results were
+consistent for ASVs generated with DADA2 (mean AUROC: 0.659).
 
 \textbf{\emph{Machine Learning.}} Machine learning models were run with
-the R package mikropml (v0.0.2) \{\} to predict the diagnosis category
-(normal vs SRN) of each sample. Data was preprocessed to normalize
-values (scale/center), remove values with zero or near-zero variance,
-and collapse collinear features using default parameters. Initially the
-models were run with default hyperparameters, but were expanded if the
-peak performance was at the edge of the hyperparameter range. Each
-taxonomic model taxonomic level combination (e.g.~random forest on
-genus) was run with 100 different seeds. Each seed split the data into a
-training (80\%) and testing (20\%) set, and output performance of the
-training and testing as area under the receiver operating curve (AUROC).
+the R package mikropml (v0.0.2)
+(\protect\hyperlink{ref-topuxe7uolu2021}{8}) to predict the diagnosis
+category (normal vs SRN) of each sample. Data was preprocessed to
+normalize values (scale/center), remove values with zero or near-zero
+variance, and collapse collinear features using default parameters.
+Initially the models were run with default hyperparameters, but were
+expanded if the peak performance was at the edge of the hyperparameter
+range. Each taxonomic model taxonomic level combination (e.g.~random
+forest on genus) was run with 100 different seeds. Each seed split the
+data into a training (80\%) and testing (20\%) set, and output
+performance of the training and testing as area under the receiver
+operating curve (AUROC).
 
 To compare performance between taxonomic levels and models, P values
-were calculated as previously described \{begum\}. To compare
-sensitivity at 90\% specificity, probabilities on the test dataset were
-collected for each seed and used to construct ROC curves (R pROC::roc).
-From the ROC curves The sensitivity at a specificity of 90\% was pulled
-for each seed. An optional output from the mikropml package is the
+were calculated as previously described
+(\protect\hyperlink{ref-topuxe7uolu2020}{6}). To compare sensitivity at
+90\% specificity, probabilities on the test dataset were collected for
+each seed and used to calculate sensitivity for specificity values
+ranging from 0 to 1 in 0.01 increments. The sensitivity at a specificity
+of 90\% was pulled for each seed. The averaged ROC curves were plotted
+by taking the average and standard deviation of the sensitivity at each
+specificity value. An optional output from the mikropml package is the
 permuted feature importance which is quantified by iteratively permuting
@@
 of feature importance was determined by ordering the features based on
-the change in AUROC where features with a larger decrease in AUROC are
-ranked higher in importance.
+the average change in AUROC across the 100 seeds where features with a
+larger decrease in AUROC are ranked higher in importance.
 
@@
 in any samples, 0.5 indicates the feature is found in half of the
-samples, an 1 indicating the feature is found in all of the samples.
+samples, and 1 indicates the feature is found in all of the samples.
 
@@
 \hypertarget{acknowledgements}{%
-\subsection{Acknowledgements}\label{acknowledgements}}
+\subsubsection{Acknowledgements}\label{acknowledgements}}
+
+This work was supported through a grant from the NIH (R01CA215574).
+
+\newpage
+
+\hypertarget{references}{%
+\subsubsection{References}\label{references}}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{0}{0}
+\leavevmode\hypertarget{ref-siegel2020}{}%
+\CSLLeftMargin{1. }
+\CSLRightInline{\textbf{Siegel RL}, \textbf{Miller KD}, \textbf{Sauer
+AG}, \textbf{Fedewa SA}, \textbf{Butterly LF}, \textbf{Anderson JC},
+\textbf{Cercek A}, \textbf{Smith RA}, \textbf{Jemal A}. 2020. Colorectal
+cancer statistics, 2020. CA: A Cancer Journal for Clinicians
+\textbf{70}:145--164. doi:\url{https://doi.org/10.3322/caac.21601}.}
+
+\leavevmode\hypertarget{ref-garcuxeda2011}{}%
+\CSLLeftMargin{2. }
+\CSLRightInline{\textbf{García G}, \textbf{Z A}. 2011. Factors
+Influencing Colorectal Cancer Screening Participation. Gastroenterology
+Research and Practice \textbf{2012}:e483417.
+doi:\href{https://doi.org/10.1155/2012/483417}{10.1155/2012/483417}.}
+
+\leavevmode\hypertarget{ref-zackular2014}{}%
+\CSLLeftMargin{3. }
+\CSLRightInline{\textbf{Zackular JP}, \textbf{Rogers MAM},
+\textbf{Ruffin MT}, \textbf{Schloss PD}. 2014. The Human Gut Microbiome
+as a Screening Tool for Colorectal Cancer. Cancer Prevention Research
+\textbf{7}:1112--1121.
+doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.}
+
+\leavevmode\hypertarget{ref-zeller2014}{}%
+\CSLLeftMargin{4. }
+\CSLRightInline{\textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
+\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
+\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
+\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
+\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Nhieu JTV},
+\textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
+\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Doeberitz M von K},
+\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
+for early-stage detection of colorectal cancer. Molecular Systems
+Biology \textbf{10}:766.
+doi:\href{https://doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.}
+
+\leavevmode\hypertarget{ref-callahan2017}{}%
+\CSLLeftMargin{5. }
+\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
+\textbf{Holmes SP}. 2017. Exact sequence variants should replace
+operational taxonomic units in marker-gene data analysis. The ISME
+Journal \textbf{11}:2639--2643.
+doi:\href{https://doi.org/10.1038/ismej.2017.119}{10.1038/ismej.2017.119}.}
+
+\leavevmode\hypertarget{ref-topuxe7uolu2020}{}%
+\CSLLeftMargin{6. }
+\CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+\textbf{Ruffin MT}, \textbf{Wiens J}, \textbf{Schloss PD}. 2020. A
+Framework for Effective Application of Machine Learning to
+Microbiome-Based Classification Problems. mBio \textbf{11}.
+doi:\href{https://doi.org/10.1128/mBio.00434-20}{10.1128/mBio.00434-20}.}
+
+\leavevmode\hypertarget{ref-schloss2009}{}%
+\CSLLeftMargin{7. }
+\CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
+\textbf{Ryabin T}, \textbf{Hall JR}, \textbf{Hartmann M},
+\textbf{Hollister EB}, \textbf{Lesniewski RA}, \textbf{Oakley BB},
+\textbf{Parks DH}, \textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres
+B}, \textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}.
+2009. Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{https://doi.org/10.1128/AEM.01541-09}{10.1128/AEM.01541-09}.}
+
+\leavevmode\hypertarget{ref-topuxe7uolu2021}{}%
+\CSLLeftMargin{8. }
+\CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lapp Z}, \textbf{Sovacool
+KL}, \textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss PD}. 2021.
+mikropml: User-Friendly R Package for Supervised Machine Learning
+Pipelines. Journal of Open Source Software \textbf{6}:3073.
+doi:\href{https://doi.org/10.21105/joss.03073}{10.21105/joss.03073}.}
+
+\leavevmode\hypertarget{ref-lobo2008}{}%
+\CSLLeftMargin{9. }
+\CSLRightInline{\textbf{Lobo JM}, \textbf{Jiménez-Valverde A},
+\textbf{Real R}. 2008. AUC: a misleading measure of the performance of
+predictive distribution models. Global Ecology and Biogeography
+\textbf{17}:145--151.
+doi:\href{https://doi.org/10.1111/j.1466-8238.2007.00358.x}{10.1111/j.1466-8238.2007.00358.x}.}
+
+\leavevmode\hypertarget{ref-eren2013}{}%
+\CSLLeftMargin{10. }
+\CSLRightInline{\textbf{Eren AM}, \textbf{Maignien L}, \textbf{Sul WJ},
+\textbf{Murphy LG}, \textbf{Grim SL}, \textbf{Morrison HG},
+\textbf{Sogin ML}. 2013. Oligotyping: differentiating between closely
+related microbial taxa using 16S rRNA gene data. Methods in Ecology and
+Evolution \textbf{4}:1111--1119.
+doi:\href{https://doi.org/10.1111/2041-210X.12114}{10.1111/2041-210X.12114}.}
+
+\leavevmode\hypertarget{ref-eren2015}{}%
+\CSLLeftMargin{11. }
+\CSLRightInline{\textbf{Eren AM}, \textbf{Morrison HG}, \textbf{Lescault
+PJ}, \textbf{Reveillaud J}, \textbf{Vineis JH}, \textbf{Sogin ML}. 2015.
+Minimum entropy decomposition: Unsupervised oligotyping for sensitive
+partitioning of high-throughput marker gene sequences. The ISME Journal
+\textbf{9}:968--979.
+doi:\href{https://doi.org/10.1038/ismej.2014.195}{10.1038/ismej.2014.195}.}
+
+\leavevmode\hypertarget{ref-schloss}{}%
+\CSLLeftMargin{12. }
+\CSLRightInline{\textbf{Schloss PD}. Amplicon sequence variants
+artificially split bacterial genomes into separate clusters. mSphere
+\textbf{6}:e00191--21.
+doi:\href{https://doi.org/10.1128/mSphere.00191-21}{10.1128/mSphere.00191-21}.}
+
+\leavevmode\hypertarget{ref-baxter2016}{}%
+\CSLLeftMargin{13. }
+\CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
+MAM}, \textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.}
+
+\leavevmode\hypertarget{ref-quast2013}{}%
+\CSLLeftMargin{14. }
+\CSLRightInline{\textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
+\textbf{Gerken J}, \textbf{Schweer T}, \textbf{Yarza P}, \textbf{Peplies
+J}, \textbf{Glöckner FO}. 2013. The SILVA ribosomal RNA gene database
+project: Improved data processing and web-based tools. Nucleic Acids
+Research \textbf{41}:D590--D596.
+doi:\href{https://doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.}
+
+\leavevmode\hypertarget{ref-callahan2016}{}%
+\CSLLeftMargin{15. }
+\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
+\textbf{Rosen MJ}, \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes
+SP}. 2016. DADA2: High-resolution sample inference from Illumina
+amplicon data. Nature Methods \textbf{13}:581--583.
+doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.}
+
+\end{CSLReferences}
 
@@
 \hypertarget{figures}{%
-\subsection{Figures}\label{figures}}
+\subsubsection{Figures}\label{figures}}
 
@@
 \hypertarget{tables}{%
-\subsection{Tables}\label{tables}}
+\subsubsection{Tables}\label{tables}}
 
@@
 \hypertarget{supplemental-figures}{%
-\subsection{Supplemental Figures}\label{supplemental-figures}}
+\subsubsection{Supplemental Figures}\label{supplemental-figures}}
 
@@
 \textbf{Supplemental Figure 1: Model Performance across Taxonomy.}
-Boxplots of AUROC values on the test dataset split for 100 seeds for
-each model type across all taxonomic levels.
+Boxplots of AUROC values from predicting whether samples came from
+subjects with screen relevant neoplasias (i.e.~advanced adenoma or
+cancer) or healthy controls across five machine learning methods
+including random forest, L2-regularized logistic regression (logistic
+regression), decision tree, gradient boosted trees (XGBoost), and
+support vector machine with radial basis kernel (SVM radial). Due to the
+random split of data into training and testing sets, each model was run
+across 100 seeds to account for variation in training/test datasplits.
 
@@
 
-\textbf{Supplemental Figure 2: Averaged ROC curves} ROC curves with
-averaged true positive rate across the 100 iterations of the model. The
-shaded region represents the standard deviation form the mean.Dashed
-line represents an AUROC of 0.5, which is equivalent to random
-classification. The mean AUROC for each taxonomic level is printed on
-the bottom right of the plot.
+\textbf{Supplemental Figure 2: Averaged ROC curves.} ROC curves with
+averaged true positive rate (or sensitivity) across the 100 iterations
+of the random forest model. The shaded region represents the standard
+deviation form the mean. Dashed line represents an AUROC of 0.5, which
+is equivalent to random classification. The mean AUROC for each
+taxonomic level is printed on the bottom right of the plot.
 
@@
 Distribution of the prevalence of taxa across samples at each taxonomic
-level. Percent of samples is split into 10 groups where the first is
+level. Percent of samples is split into 10 groups where the first is for
 taxa present in 0 to 10\% of samples, then \textgreater10\% to 20\% of
 samples, and so on. The total number of taxa for each taxonomic level
-after preprocessing is in parenthesis next to the title of the plot (the
-name of the taxonomic level).
+after preprocessing is in parenthesis next to the title of the plot (or
+the name of the taxonomic level).
 
@@
 models at each taxonomic level based on the average decrease in AUROC
-when the feature is permuted.
-
-\newpage
-
-\hypertarget{references}{%
-\subsection*{References}\label{references}}
-\addcontentsline{toc}{subsection}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{1}
-\leavevmode\hypertarget{ref-siegel2020}{}%
-\CSLLeftMargin{1. }
-\CSLRightInline{\textbf{Siegel RL}, \textbf{Miller KD}, \textbf{Sauer
-AG}, \textbf{Fedewa SA}, \textbf{Butterly LF}, \textbf{Anderson JC},
-\textbf{Cercek A}, \textbf{Smith RA}, \textbf{Jemal A}. 2020. Colorectal
-cancer statistics, 2020. CA: A Cancer Journal for Clinicians
-\textbf{70}:145--164. doi:\url{https://doi.org/10.3322/caac.21601}.}
-
-\leavevmode\hypertarget{ref-garcuxeda2011a}{}%
-\CSLLeftMargin{2. }
-\CSLRightInline{\textbf{García G}, \textbf{Z A}. 2011. Factors
-Influencing Colorectal Cancer Screening Participation. Gastroenterology
-Research and Practice \textbf{2012}:e483417.
-doi:\href{https://doi.org/10.1155/2012/483417}{10.1155/2012/483417}.}
-
-\leavevmode\hypertarget{ref-callahan2017}{}%
-\CSLLeftMargin{3. }
-\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
-\textbf{Holmes SP}. 2017. Exact sequence variants should replace
-operational taxonomic units in marker-gene data analysis. The ISME
-Journal \textbf{11}:2639--2643.
-doi:\href{https://doi.org/10.1038/ismej.2017.119}{10.1038/ismej.2017.119}.}
-
-\leavevmode\hypertarget{ref-topuxe7uolu2020}{}%
-\CSLLeftMargin{4. }
-\CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
-\textbf{Ruffin MT}, \textbf{Wiens J}, \textbf{Schloss PD}. 2020. A
-Framework for Effective Application of Machine Learning to
-Microbiome-Based Classification Problems. mBio \textbf{11}.
-doi:\href{https://doi.org/10.1128/mBio.00434-20}{10.1128/mBio.00434-20}.}
-
-\leavevmode\hypertarget{ref-lobo2008}{}%
-\CSLLeftMargin{5. }
-\CSLRightInline{\textbf{Lobo JM}, \textbf{Jiménez-Valverde A},
-\textbf{Real R}. 2008. AUC: a misleading measure of the performance of
-predictive distribution models. Global Ecology and Biogeography
-\textbf{17}:145--151.
-doi:\href{https://doi.org/10.1111/j.1466-8238.2007.00358.x}{10.1111/j.1466-8238.2007.00358.x}.}
-
-\leavevmode\hypertarget{ref-callahan2016}{}%
-\CSLLeftMargin{6. }
-\CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
-\textbf{Rosen MJ}, \textbf{Han AW}, \textbf{Johnson AJA}, \textbf{Holmes
-SP}. 2016. DADA2: High-resolution sample inference from Illumina
-amplicon data. Nature Methods \textbf{13}:581--583.
-doi:\href{https://doi.org/10.1038/nmeth.3869}{10.1038/nmeth.3869}.}
-
-\end{CSLReferences}
+when the feature is permuted. Dot represents the mean decrease in AUROC
+and the lines extending from the dot represent the standard deviation
+from the mean.
 

### 2021-09-21 paper/manuscript.tex
update figure s2/s4 AUC to AUROC

--- 2021-08-30 paper/manuscript.tex
@@
 
-Courtney R. Armour${^1}$, Begüm D.Topçuoğlu, Andrea Garretto${^1}$,
-Patrick D. Schloss ${^1}$${^*}$
+Courtney R. Armour${^1}$, Begüm D. Topçuoğlu$^{1,2}$, Andrea
+Garretto${^1}$, Patrick D. Schloss ${^1}$${^*}$
 
@@
 ${^1}$ Department of Microbiology and Immunology, University of
-Michigan, Ann Arbor
+Michigan, Ann Arbor MI
+
+${^2}$ Exploratory Science Center, Merck \& Co., Inc., Cambridge,
+Massachusetts, USA~
 
@@
 
-\subsection{Abstract (max 250 words)}\label{abstract-max-250-words}
+\subsubsection{Abstract}\label{abstract}
 
-\subsection{Importance (max 150 words)}\label{importance-max-150-words}
+Colorectal cancer is a common and deadly disease in the United States
+accounting for over 50,000 deaths in 2020. This progressive disease is
+highly preventable with early detection and treatment, but many people
+do not comply with the recommended screening guidelines. The gut
+microbiome has emerged as a promising target for non-invasive detection
+of colorectal cancer. Most microbiome-based classification efforts
+utilize taxonomic abundance data from operational taxonomic units (OTUs)
+or amplicon sequence variants (ASVs). However, it is unknown which
+taxonomic resolution is optimal for microbiome-based classification of
+colorectal cancer. To address this question, we used a reproducible
+machine learning framework to quantify classification performance of
+models based on data annotated to phylum, class, order, family, genus,
+OTU, and ASV levels. We found that model performance increased with
+increasing taxonomic resolution, up to the family level where
+performance was equal among family, genus, and OTU data before
+decreasing at the ASV level. These results are a step towards optimizing
+a microbiome based diagnostic tool.
+
+\subsubsection{Importance}\label{importance}
+
+Despite being highly preventable, colorectal cancer remains a leading
+cause of cancer related death in the United States. Low-cost,
+non-invasive detection methods could greatly improve our ability to
+identify and treat early stages of disease. While the microbiome has
+shown promise as a diagnostic tool, there are numerous details to refine
+before a diagnostic could be widely deployed. Among these are the need
+for standardization of sample collection, storage, processing, and
+analysis. We address one aspect of data analysis by examining which
+taxonomic level provides the optimal resolution for microbiome-based
+classification of colorectal cancer. This is an important step to inform
+the development of microbiome-based diagnostic tools.
 
@@
 detection and treatment are essential to increase survival rates, but
-for a variety of reasons including the invasiveness and high cost of
-screening (i.e.~colonoscopy), many people do not comply with recommended
-screening guidelines (2) prompting a need for low cost, non-invasive
-detection methods. A growing body of research points to the gut
-microbiome as a promising target for non-invasive screening to detect
-screen relevant neoplasia (SRNs) consisting of pre-cancerous polyps
-(i.e.~advanced adenomas) and carcinomas. Efforts to realize the
-diagnostic potential of the gut microbiome in detecting SRNs have
-focused on machine learning (ML) methods using abundances of operational
-taxonomic unit (OTU) classifications based on amplicon sequencing of the
-16S rRNA gene \{\}. However, whether this is the optimal taxonomic
-resolution for classifying SRNs from microbiome data is unknown.
-Additionally, recent work has pushed for the use of amplicon sequence
-variants (ASVs) to replace OTUs for marker-gene analysis because of the
-improved resolution with ASVs (3). However, whether the additional
-resolution provided by ASVs is useful for ML classification is unclear
-\{\}. Since ML classification relies on consistent differences between
-groups, its possible that the resolution at the ASV level is too
-individualized to accurately differentiate groups. Topçuoğlu \emph{et
-al} (4) recently demonstrated how to effectively apply machine learning
-(ML) methods to microbiome based classification problems and developed a
+for reasons such as invasiveness and high screening costs
+(i.e.~colonoscopy), many people do not comply with recommended screening
+guidelines (2). This prompts a need for low cost, non-invasive detection
+methods. A growing body of research points to the gut microbiome as a
+promising target for non-invasive detection of screen relevant neoplasia
+(SRNs) consisting of advanced adenomas and carcinomas (3, 4). The
+diagnostic potential of the gut microbiome in detecting SRNs has been
+explored through machine learning (ML) methods using abundances of
+operational taxonomic unit (OTU) classifications based on amplicon
+sequencing of the 16S rRNA gene (3). Recent work has pushed for the use
+of amplicon sequence variants (ASVs) to replace OTUs for marker-gene
+analysis because of the improved resolution with ASVs (5). However, it
+is unclear if OTUs are the optimal taxonomic resolution for classifying
+SRNs from microbiome data or whether the additional resolution provided
+by ASVs is useful for ML classification. Topçuoğlu \emph{et al} (6)
+recently demonstrated how to effectively apply machine learning (ML)
+methods to microbiome based classification problems and developed a
 framework for applying ML practices in a more reproducible way
@@
 Utilizing publicly available 16S rRNA sequence data from stool of
-patients with SRNs and healthy controls, we generated abundance tables
-with mothur \{\} annotated to phylum, class, order, family, genus, OTU
-and ASV levels. Using the taxonomic abundance data, we quantified how
-reliably samples could be classified as ``normal'' or ``SRN''
-(i.e.~advanced adenoma or carcinoma) using five machine learning methods
-with the mikropml R package (methods). Across the five machine learning
-methods tested, model performance tended to increase with taxonomic
+patients with SRNs and healthy controls, we generated taxonomic
+abundance tables with mothur (7) annotated to phylum, class, order,
+family, genus, OTU and ASV levels. Using the taxonomic abundance data
+and the mikropml R package (8), we quantified how reliably samples could
+be classified as SRN or normal using five machine learning methods
+including random forest, L2-regularized logistic regression, decision
+tree, gradient boosted trees (XGBoost), and support vector machine with
+radial basis kernel (SVM radial). Across the five machine learning
+methods tested, model performance increased with increasing taxonomic
 level usually peaking around genus/OTU level before dropping off
@@
 0.06) and genus (p = 0.05) levels (Figure 1A). These results suggest
-that increased resolution improves model performance until the OTU level
+that increased resolution improves model performance up to the OTU level
 where further taxonomic resolution is not necessarily better for
@@
 thresholds and can be misleading since models with the same AUROC can
-have different ROC curve shapes (5). Depending on the intended
+have different ROC curve shapes (9). Depending on the intended
 implementation of the model, one may want to optimize the true-positive
-rate (or sensitivity) over the false-positive rate (or 1-specificity),
-or vice versa. In this case, the optimal model will detect as many true
+rate (sensitivity) over the false-positive rate (1-specificity) or vice
+versa. In this case, the optimal model will detect as many true
 positives (people with SRNs) as possible. To further compare the model
@@
 values (Figure 1B). This analysis further supports the observation that
-finer resolution is does not improve SRN detection.
+finer resolution does not improve SRN detection.
 
@@
 The opposite was observed at the OTU and ASV level where 50\% and 41\%
-of taxa respectively were only present in 20\% or less of the samples
+of taxa, respectively, were only present in 20\% or less of the samples
 (Supplemental Figure 3). Of note, the ML pipeline includes a
 pre-processing step that occurred prior to training and classifying the
-ML models. As part of this step, taxa were removed that would not
-provide useful information to build the model. For example, perfectly
-correlated taxa provide the same information to build the model and thus
-can be collapsed. Additionally, features with zero or near-zero variance
-across samples would not help the model differentiate groups and thus
-can be removed. Interestingly, despite starting with 104,106 ASVs, only
-478 (0.5\%) remained after pre-processing. At the OTU level, 705 of the
-20,079 OTUs (3.5\%) remained after preprocessing (Table 1). While the
-resolution provided by ASVs is useful in certain contexts \{\}, these
-results suggest that the resolution is too fine for use in machine
-learning classification of SRNs based on microbiome composition.
+ML models. For example, perfectly correlated taxa provide the same
+information to build the model and thus can be collapsed. Additionally,
+features with zero or near-zero variance across samples were removed.
+Interestingly, despite starting with 104,106 ASVs, only 478 (0.5\%)
+remained after pre-processing. At the OTU level, 705 of the 20,079 OTUs
+(3.5\%) remained after preprocessing (Table 1). While the resolution
+provided by ASVs is useful in certain contexts(10, 11), these results
+suggest that the resolution is too fine for use in machine learning
+classification of SRNs based on microbiome composition.
 
@@
 (e.g.~OTU and ASV) did not add additional sensitivity to predicting SRNs
-based on microbiome composition. Additionally, at the ASV level the fine
-resolution actually impedes model performance due to the sparsity of
-shared taxa and led to decreased model performance. The tendancy for ASV
-level annotation to split single bacterial genomes into multiple taxa
-\{Pat mSphere 2021\} could also be a contributing factor to the sparsity
-of shared taxa. Overall, either family, genus, or OTU level taxonomy
-appear to perform equally for predicting subjects with SRNs based on the
-composition of the gut microbiome. A potential benefit of utilizing
-genus or family level data could be that it may allow for merging data
-generated from different 16S rRNA gene regions or sequencing platforms.
+based on microbiome composition. Family, genus, and OTU level data all
+performed equally. At the ASV level the fine resolution actually impedes
+model performance due to the sparsity of shared taxa and led to
+decreased model performance. The tendency for ASV level annotation to
+split single bacterial genomes into multiple taxa (12) could also be a
+contributing factor to the sparsity of shared taxa. Overall, either
+family, genus, or OTU level taxonomy appear to perform equally for
+predicting subjects with SRNs based on the composition of the gut
+microbiome. A potential benefit of utilizing genus or family level data
+could be that it may allow for merging data generated from different 16S
+rRNA gene regions or sequencing platforms.
 
-\subsection{Materials and Methods}\label{materials-and-methods}
+\subsubsection{Materials and Methods}\label{materials-and-methods}
 
 \textbf{\emph{Dataset.}} Raw 16S rRNA gene amplicon sequence data
-isolated from human gut samples \{Baxter\} was downloaded from NCBI SRA
-(accession \#). This dataset contains stool samples from 490 subjects.
-Based on the available metadata, samples categorized as normal, high
-risk normal, or adenoma were labeled ``normal'' for this analysis and
-samples categorized as advanced adenoma or carcinoma were labeled as
-``screen relevant neoplasia'' (SRN). This resulted in a total of 261
-``normal'' samples and 229 ``SRN'' samples.
+isolated from human gut samples (13) was downloaded from NCBI Sequence
+Read Archive (SRP062005). This dataset contains stool samples from 490
+subjects. Based on the available metadata, samples categorized as
+normal, high risk normal, or adenoma were labeled ``normal'' for this
+analysis and samples categorized as advanced adenoma or carcinoma were
+labeled as ``screen relevant neoplasia'' (SRN). This resulted in a total
+of 261 ``normal'' samples and 229 ``SRN'' samples.
 
 \textbf{\emph{Data processing.}} Sequence data was processed with mothur
-(1.44.3) using the SILVA reference database (v132) to produce count
-tables for phylum, class, order, family, genus, OTU, and ASV following
-the Schloss Lab MiSeq SOP described on the mothur website. ASV level
-data was also produced using DADA2 to ensure consistent results with a
-different pipeline. Data was processed following the DADA2 pipeline, but
-setting pool=TRUE to infer ASVs from the whole dataset rather than per
-sample. The resulting ASV table was subsampled for consistency with the
-mothur data. These results were consistent for ASVs generated with DADA2
-(6) (mean AUROC: 0.659).
+(1.44.3) (7) using the SILVA reference database (v132) (14) to produce
+count tables for phylum, class, order, family, genus, OTU, and ASV
+following the Schloss Lab MiSeq SOP described on the mothur website
+(\url{https://mothur.org/wiki/miseq_sop/}). ASV level data was also
+produced using DADA2 (15) to ensure consistent results with a different
+pipeline. Data was processed following the DADA2 pipeline, but setting
+pool=TRUE to infer ASVs from the whole dataset rather than per sample.
+The resulting ASV table was subsampled for consistency with the mothur
+data. The results were consistent for ASVs generated with DADA2 (mean
+AUROC: 0.659).
 
 \textbf{\emph{Machine Learning.}} Machine learning models were run with
-the R package mikropml (v0.0.2) \{\} to predict the diagnosis category
+the R package mikropml (v0.0.2) (8) to predict the diagnosis category
 (normal vs SRN) of each sample. Data was preprocessed to normalize
@@
 To compare performance between taxonomic levels and models, P values
-were calculated as previously described \{begum\}. To compare
-sensitivity at 90\% specificity, probabilities on the test dataset were
-collected for each seed and used to construct ROC curves (R pROC::roc).
-From the ROC curves The sensitivity at a specificity of 90\% was pulled
-for each seed. An optional output from the mikropml package is the
+were calculated as previously described (6). To compare sensitivity at
+90\% specificity, probabilities on the test dataset were collected for
+each seed and used to calculate sensitivity for specificity values
+ranging from 0 to 1 in 0.01 increments. The sensitivity at a specificity
+of 90\% was pulled for each seed. The averaged ROC curves were plotted
+by taking the average and standard deviation of the sensitivity at each
+specificity value. An optional output from the mikropml package is the
 permuted feature importance which is quantified by iteratively permuting
@@
 
-\subsection{Acknowledgements}\label{acknowledgements}
+\subsubsection{Acknowledgements}\label{acknowledgements}
+
+This work was supported through a grant from the NIH (R01CA215574).
 
@@
 
-\subsection{Figures}\label{figures}
+\subsubsection{References}\label{references}
+
+::: \{\#refs\} :::
+
+\newpage
+
+\subsubsection{Figures}\label{figures}
 
@@
 
-\subsection{Tables}\label{tables}
+\subsubsection{Tables}\label{tables}
 
@@
 
-\subsection{Supplemental Figures}\label{supplemental-figures}
+\subsubsection{Supplemental Figures}\label{supplemental-figures}
 
@@
 of taxa across samples at each taxonomic level. Percent of samples is
-split into 10 groups where the first is taxa present in 0 to 10\% of
+split into 10 groups where the first is for taxa present in 0 to 10\% of
 samples, then \textgreater{}10\% to 20\% of samples, and so on. The
 total number of taxa for each taxonomic level after preprocessing is in
-parenthesis next to the title of the plot (the name of the taxonomic
+parenthesis next to the title of the plot (or the name of the taxonomic
 level).
@@
 level based on the average decrease in AUROC when the feature is
-permuted.
-
-\newpage
-
-\subsection{References}\label{references}
+permuted. Dot represents the mean decrease in AUROC and the lines
+extending from the dot represent the standard deviation from the mean.
 
@@
 
-3. \textbf{{Callahan} B}, \textbf{{McMurdie} P}, \textbf{{Holmes} S}.
+3. \textbf{{Zackular} J}, \textbf{{Rogers} M}, \textbf{{Ruffin} M},
+\textbf{{Schloss} P}. 2014. The human gut microbiome as a screening tool
+for colorectal cancer. Cancer Prevention Research \textbf{7}:1112--1121.
+doi:\href{http://dx.doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.
+
+4. \textbf{{Zeller} G}, \textbf{{Tap} J}, \textbf{{Voigt} A},
+\textbf{{Sunagawa} S}, \textbf{{Kultima} J}, \textbf{{Costea} P},
+\textbf{{Amiot} A}, \textbf{{Böhm} J}, \textbf{{Brunetti} F},
+\textbf{{Habermann} N}, \textbf{{Hercog} R}, \textbf{{Koch} M},
+\textbf{{Luciani} A}, \textbf{{Mende} D}, \textbf{{Schneider} M},
+\textbf{{Schrotz-King} P}, \textbf{{Tournigand} C}, \textbf{{Nhieu} J},
+\textbf{{Yamada} T}, \textbf{{Zimmermann} J}, \textbf{{Benes} V},
+\textbf{{Kloor} M}, \textbf{{Ulrich} C}, \textbf{{Doeberitz} M},
+\textbf{{Sobhani} I}, \textbf{{Bork} P}. 2014. Potential of fecal
+microbiota for early-stage detection of colorectal cancer. Molecular
+Systems Biology \textbf{10}:766.
+doi:\href{http://dx.doi.org/10.15252/msb.20145645}{10.15252/msb.20145645}.
+
+5. \textbf{{Callahan} B}, \textbf{{McMurdie} P}, \textbf{{Holmes} S}.
 2017. Exact sequence variants should replace operational taxonomic units
@@
 
-4. \textbf{{Topçuo{ğ}lu} B}, \textbf{{Lesniak} N}, \textbf{{Ruffin} M},
+6. \textbf{{Topçuo{ğ}lu} B}, \textbf{{Lesniak} N}, \textbf{{Ruffin} M},
 \textbf{{Wiens} J}, \textbf{{Schloss} P}. 2020. A framework for
@@
 
-5. \textbf{{Lobo} J}, \textbf{{Jiménez-Valverde} A}, \textbf{{Real} R}.
+7. \textbf{{Schloss} P}, \textbf{{Westcott} S}, \textbf{{Ryabin} T},
+\textbf{{Hall} J}, \textbf{{Hartmann} M}, \textbf{{Hollister} E},
+\textbf{{Lesniewski} R}, \textbf{{Oakley} B}, \textbf{{Parks} D},
+\textbf{{Robinson} C}, \textbf{{Sahl} J}, \textbf{{Stres} B},
+\textbf{{Thallinger} G}, \textbf{{Van Horn} D}, \textbf{{Weber} C}.
+2009. Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{http://dx.doi.org/10.1128/AEM.01541-09}{10.1128/AEM.01541-09}.
+
+8. \textbf{{Topçuo{ğ}lu} B}, \textbf{{Lapp} Z}, \textbf{{Sovacool} K},
+\textbf{{Snitkin} E}, \textbf{{Wiens} J}, \textbf{{Schloss} P}. 2021.
+mikropml: User-friendly r package for supervised machine learning
+pipelines. Journal of Open Source Software \textbf{6}:3073.
+doi:\href{http://dx.doi.org/10.21105/joss.03073}{10.21105/joss.03073}.
+
+9. \textbf{{Lobo} J}, \textbf{{Jiménez-Valverde} A}, \textbf{{Real} R}.
 2008. AUC: a misleading measure of the performance of predictive
@@
 
-6. \textbf{{Callahan} B}, \textbf{{McMurdie} P}, \textbf{{Rosen} M},
+10. \textbf{{Eren} A}, \textbf{{Maignien} L}, \textbf{{Sul} W},
+\textbf{{Murphy} L}, \textbf{{Grim} S}, \textbf{{Morrison} H},
+\textbf{{Sogin} M}. 2013. Oligotyping: differentiating between closely
+related microbial taxa using 16S rRNA gene data. Methods in Ecology and
+Evolution \textbf{4}:1111--1119.
+doi:\href{http://dx.doi.org/10.1111/2041-210X.12114}{10.1111/2041-210X.12114}.
+
+11. \textbf{{Eren} A}, \textbf{{Morrison} H}, \textbf{{Lescault} P},
+\textbf{{Reveillaud} J}, \textbf{{Vineis} J}, \textbf{{Sogin} M}. 2015.
+Minimum entropy decomposition: Unsupervised oligotyping for sensitive
+partitioning of high-throughput marker gene sequences. The ISME Journal
+\textbf{9}:968--979.
+doi:\href{http://dx.doi.org/10.1038/ismej.2014.195}{10.1038/ismej.2014.195}.
+
+12. \textbf{{Schloss} P}. Amplicon sequence variants artificially split
+bacterial genomes into separate clusters. mSphere \textbf{6}:e00191--21.
+doi:\href{http://dx.doi.org/10.1128/mSphere.00191-21}{10.1128/mSphere.00191-21}.
+
+13. \textbf{{Baxter} N}, \textbf{{Ruffin} M}, \textbf{{Rogers} M},
+\textbf{{Schloss} P}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\href{http://dx.doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
+
+14. \textbf{{Quast} C}, \textbf{{Pruesse} E}, \textbf{{Yilmaz} P},
+\textbf{{Gerken} J}, \textbf{{Schweer} T}, \textbf{{Yarza} P},
+\textbf{{Peplies} J}, \textbf{{Glöckner} F}. 2013. The sILVA ribosomal
+rNA gene database project: improved data processing and web-based tools.
+Nucleic Acids Research \textbf{41}:D590--D596.
+doi:\href{http://dx.doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.
+
+15. \textbf{{Callahan} B}, \textbf{{McMurdie} P}, \textbf{{Rosen} M},
 \textbf{{Han} A}, \textbf{{Johnson} A}, \textbf{{Holmes} S}. 2016.

### 2021-09-21 code/R/figures/averaged_ROC_curve.R

@@ -55,7 +55,7 @@ avg_sens %>%
 #ggsave("analysis/figures/average_roc_all.png",width=6)
 
 #plot faceted
-mean_auc <- read_csv("analysis/mean_AUC_by_model.csv")  %>% 
+mean_auc <- read_csv("analysis/mean_AUC_by_model.csv")  %>%
   select(level,rf) %>%
   rename(mean_AUC = rf) %>%
   mutate(level=factor(level,levels=levels,labels=levels_names))
@@ -82,8 +82,8 @@ avg_sens %>%
           panel.spacing.x = unit(4, "mm")) +
     guides(color = guide_legend(nrow = 1,title=""),
            fill = guide_legend(nrow = 1,title="")) +
-    geom_text(aes(x = .7, y = .2,label = paste0("mean AUC\n",format(round(mean_AUC,digits=3),nsmall=3))),
-              size=3.5) +
+    geom_text(aes(x = .7, y = .2,label = paste0("mean AUROC\n",format(round(mean_AUC,digits=3),nsmall=3))),
+              size=3) +
     scale_fill_manual(values=colors2) +
     scale_color_manual(values=colors2)
 ggsave("analysis/figures/average_roc_by_level.png",width=7,height=5)

### 2021-09-21 code/R/figures/top_10_important_tax.R

@@ -49,7 +49,7 @@ plot_top_imp <- function(top_data,tax_level){
             axis.text.y = element_markdown(),
             plot.caption = element_markdown(lineheight = 1.2)) +
       ggtitle(tax_level) +
-      xlab("Decrease in AUC")
+      xlab("Decrease in AUROC")
   }else{
     top_data %>%
       filter(level == tax_level) %>%
@@ -65,7 +65,7 @@ plot_top_imp <- function(top_data,tax_level){
             axis.text.y = element_markdown(),
             plot.caption = element_markdown(lineheight = 1.2)) +
       ggtitle(tax_level) +
-      xlab("Decrease in AUC")
+      xlab("Decrease in AUROC")
   }
 }
 

### 2021-09-27 paper/manuscript.Rmd
final draft
abstract,importance,refs

@@
 # Taxonomic Resolution Matters for Microbiome-Based Classification of Colorectal Cancer
+
+alternative titles:  
+High resolution taxonomy impedes microbiome-based classification of colorectal cancer  
+High resolution ASVs impede microbiome-based classification of colorectal cancer  
+Mid-level taxonomic resolution is optimal for microbiome based classification of colorectal cancer  
+Too much of a good thing: high resolution ASVs do not improve microbiome-based classification of colorectal cancer  
+More is not always better: high resolution ASVs do not improve microbiome-based classification of colorectal cancer
 
@@
 While comparing AUROC values between models is a useful way to assess the overall model performance, they summarize the performance across all thresholds and can be misleading since models with the same AUROC can have different ROC curve shapes [@lobo2008].
-Depending on the intended implementation of the model, one may want to optimize the true-positive rate (sensitivity) over the false-positive rate (1-specificity) or vice versa.
+Depending on the intended implementation of the model, one may want to optimize the sensitivity over the specificity or vice versa.
 In this case, the optimal model will detect as many true positives (people with SRNs) as possible.
@@
 The tendency for ASV level annotation to split single bacterial genomes into multiple taxa [@schloss] could also be a contributing factor to the sparsity of shared taxa.
+Additionally, these results indicate that there are not specific individual bacterial strains that are useful to resolve SRNs, rather sets of closely related bacterial taxa.
 Overall, either family, genus, or OTU level taxonomy appear to perform equally for predicting subjects with SRNs based on the composition of the gut microbiome.
@@
 
-All code is available at: **TODO: link to code**
+All code is available at: https://github.com/SchlossLab/Armour_Resolution_XXXX_2021
 

### 2021-09-27 paper/manuscript.tex
final draft
abstract,importance,refs

--- 2021-09-21 paper/manuscript.tex
@@
 
+alternative titles:\\
+High resolution taxonomy impedes microbiome-based classification of
+colorectal cancer\\
+High resolution ASVs impede microbiome-based classification of
+colorectal cancer\\
+Mid-level taxonomic resolution is optimal for microbiome based
+classification of colorectal cancer\\
+Too much of a good thing: high resolution ASVs do not improve
+microbiome-based classification of colorectal cancer\\
+More is not always better: high resolution ASVs do not improve
+microbiome-based classification of colorectal cancer
+
 \vspace{10mm}
@@
 Depending on the intended implementation of the model, one may want to
-optimize the true-positive rate (sensitivity) over the false-positive
-rate (1-specificity) or vice versa. In this case, the optimal model will
-detect as many true positives (people with SRNs) as possible. To further
-compare the model performance across taxonomic levels we compared the
-sensitivity of the models at a specificity of 90\%. The highest
-sensitivity values were observed for family, genus, and OTU level data
-(Figure 1B), consistent with the AUROC results. Phylum, class, order,
-and ASV sensitivity values were all significantly lower than family,
-genus, and OTU sensitivity values (Figure 1B). This analysis further
-supports the observation that finer resolution does not improve SRN
-detection.
+optimize the sensitivity over the specificity or vice versa. In this
+case, the optimal model will detect as many true positives (people with
+SRNs) as possible. To further compare the model performance across
+taxonomic levels we compared the sensitivity of the models at a
+specificity of 90\%. The highest sensitivity values were observed for
+family, genus, and OTU level data (Figure 1B), consistent with the AUROC
+results. Phylum, class, order, and ASV sensitivity values were all
+significantly lower than family, genus, and OTU sensitivity values
+(Figure 1B). This analysis further supports the observation that finer
+resolution does not improve SRN detection.
 
@@
 (\protect\hyperlink{ref-schloss}{12}) could also be a contributing
-factor to the sparsity of shared taxa. Overall, either family, genus, or
-OTU level taxonomy appear to perform equally for predicting subjects
-with SRNs based on the composition of the gut microbiome. A potential
-benefit of utilizing genus or family level data could be that it may
-allow for merging data generated from different 16S rRNA gene regions or
-sequencing platforms.
+factor to the sparsity of shared taxa. Additionally, these results
+indicate that there are not specific individual bacterial strains that
+are useful to resolve SRNs, rather sets of closely related bacterial
+taxa. Overall, either family, genus, or OTU level taxonomy appear to
+perform equally for predicting subjects with SRNs based on the
+composition of the gut microbiome. A potential benefit of utilizing
+genus or family level data could be that it may allow for merging data
+generated from different 16S rRNA gene regions or sequencing platforms.
 
@@
 
-All code is available at: \textbf{TODO: link to code}
+All code is available at:
+\url{https://github.com/SchlossLab/Armour_Resolution_XXXX_2021}
 

### 2021-10-19 code/R/figures/prevalence.R
update look of figS3

@@ -59,12 +59,15 @@ pct_bins %>%
   scale_x_continuous(expand=c(0,0),breaks = seq(0,100,10)) +
   xlab("Percent of Samples") +
   ylab("Percent of Taxa") +
-  theme_bw() +
+  theme_classic() +
   theme(legend.position = "none",
         axis.text = element_text(size=11),
         axis.title = element_text(size=14),
         panel.grid.major.x = element_blank(),
-        panel.grid.minor.x = element_blank()) +
+        panel.grid.minor.x = element_blank(),
+        panel.border = element_rect(colour = "black", fill=NA,size=0.75),
+        panel.spacing.x = unit(4, "mm"),
+        strip.text = element_text(size=12)) +
   geom_text(aes(label = paste0((round(pct,digits=2)*100),"%")),nudge_y=0.08) +
   scale_fill_manual(values=pal)
 ggsave("analysis/figures/prevalence.png",width=10)

### 2021-10-19 paper/manuscript.Rmd
final paper edits

--- 2021-09-21 paper/manuscript.Rmd
@@
 bibliography: references.bib
-link-citations: true
 csl: msystems.csl
@@
 
-# Taxonomic Resolution Matters for Microbiome-Based Classification of Colorectal Cancer
+# A Goldilocks Principle for the Gut Microbiome: Taxonomic Resolution Matters for Microbiome-Based Classification of Colorectal Cancer
 
@@
 
-Courtney R. Armour${^1}$, Begüm D. Topçuoğlu$^{1,2}$, Andrea Garretto${^1}$, Patrick D. Schloss ${^1}$${^*}$
+Courtney R. Armour${^1}$, Begüm D. Topçuoğlu$^{1,\#}$, Andrea Garretto${^1}$, Patrick D. Schloss ${^1}$${^*}$
 
@@
 
-${^2}$ Exploratory Science Center, Merck & Co., Inc., Cambridge, Massachusetts, USA
+${^\#}$ Current Affiliation: Exploratory Science Center, Merck & Co., Inc., Cambridge, Massachusetts, USA
 
@@
 
+```{r mean_AUC}
+table <- read_csv("../analysis/mean_AUC_by_model.csv",
+                  col_types = cols(level=col_character(),
+                                   .default=col_double()))
+
+rf_mean_auc <- table %>%
+  select(level,rf) %>%
+  pivot_wider(names_from=level,values_from=rf)
+
+```
+
 ### Abstract
@@
 The gut microbiome has emerged as a promising target for non-invasive detection of colorectal cancer.
-Most microbiome-based classification efforts utilize taxonomic abundance data from operational taxonomic units (OTUs) or amplicon sequence variants (ASVs).
+Most microbiome-based classification efforts utilize taxonomic abundance data from operational taxonomic units (OTUs) or amplicon sequence variants (ASVs) with the goal of increasing taxonomic resolution.
 However, it is unknown which taxonomic resolution is optimal for microbiome-based classification of colorectal cancer.
 To address this question, we used a reproducible machine learning framework to quantify classification performance of models based on data annotated to phylum, class, order, family, genus, OTU, and ASV levels.
-We found that model performance increased with increasing taxonomic resolution, up to the family level where performance was equal among family, genus, and OTU data before decreasing at the ASV level.
-These results are a step towards optimizing a microbiome based diagnostic tool.
+We found that model performance increased with increasing taxonomic resolution, up to the family level where performance was equal (p > 0.05) among family (mean AUC: `r format_decimal(rf_mean_auc$family,digits=3)`), genus (mean AUC: `r format_decimal(rf_mean_auc$genus,digits=3)`), and OTU (mean AUC: `r format_decimal(rf_mean_auc$otu,digits=3)`) levels before decreasing at the ASV level (p < 0.05, mean AUC: `r format_decimal(rf_mean_auc$asv,digits=3)`).
+These results demonstrate a trade-off between taxonomic resolution and prediction performance, where coarse taxonomic resolution (e.g.
+phylum) is not distinct enough but fine resolution (e.g.
+ASV) is to individualized to accurately classify samples.
+Similar to the story of Goldilocks and the three bears, mid-range resolution (i.e.
+family, genus, and OTU) is just right for optimal prediction of colorectal cancer from microbiome data.
+(238 words)
 
@@
 Low-cost, non-invasive detection methods could greatly improve our ability to identify and treat early stages of disease.
-While the microbiome has shown promise as a diagnostic tool, there are numerous details to refine before a diagnostic could be widely deployed.
-Among these are the need for standardization of sample collection, storage, processing, and analysis.
-We address one aspect of data analysis by examining which taxonomic level provides the optimal resolution for microbiome-based classification of colorectal cancer.
-This is an important step to inform the development of microbiome-based diagnostic tools.
+The microbiome has shown promise as a resource for detection of colorectal cancer.
+Research on the gut microbiome tends to focus on improving our ability to profile species and strain level taxonomic resolution.
+However, we found that finer resolution impedes the ability to predict colorectal cancer based on the gut microbiome.
+These results highlight the need for consideration of the appropriate taxonomic resolution for microbiome analyses and that finer resolution is not always more informative.
+(110 words)
 
@@
 However, it is unclear if OTUs are the optimal taxonomic resolution for classifying SRNs from microbiome data or whether the additional resolution provided by ASVs is useful for ML classification.
-Topçuoğlu *et al* [@topçuolu2020] recently demonstrated how to effectively apply machine learning (ML) methods to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml).
+Topçuoğlu *et al* [@topçuolu2020] recently demonstrated how to effectively apply machine learning (ML) methods to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way.
 This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to determine which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.
 
-```{r mean_AUC}
-table <- read_csv("../analysis/mean_AUC_by_model.csv",
-                  col_types = cols(level=col_character(),
-                                   .default=col_double()))
+```{r mean_sens_90spec}
+mean_sens_spec90 <- read_csv("../analysis/sens_spec_by_level.csv") %>% 
+  filter(specificity == 0.9) %>% 
+  group_by(level) %>% 
+  summarise(mean_sens = mean(sensitivity)) %>% 
+  pivot_wider(names_from = level, values_from = mean_sens)
 
-rf_mean_auc <- table %>%
-  pull(rf)
-
-names(rf_mean_auc) <- table %>% pull(level)
+sens_spec90_pvals <- read_csv("../analysis/sens_spec90_pvals.csv")
 ```
@@
 Regardless of the taxonomic level, random forest (RF) models consistently had the largest area under the receiver operating characteristic curve (AUROC).
-Within the RF model, the highest AUROCs were observed for family (mean AUROC: `r format_decimal(rf_mean_auc["family"],digits=3)`), genus (mean AUROC: `r format_decimal(rf_mean_auc["genus"],digits=3)`), and OTU (mean AUROC: `r format_decimal(rf_mean_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1A, Supplemental Figure 2).
-Performance with ASVs (mean AUROC: `r format_decimal(rf_mean_auc["asv"],digits=3)`) was significantly lower than OTUs (p \< 0.01), but comparable to family (p = `r format_decimal(rf_pvals["family","asv"],digits=2)`) and genus (p = `r format_decimal(rf_pvals["genus","asv"],digits=2)`) levels (Figure 1A).
+Within the RF model, the highest AUROCs were observed for family (mean AUROC: `r format_decimal(rf_mean_auc$family,digits=3)`), genus (mean AUROC: `r format_decimal(rf_mean_auc$genus,digits=3)`), and OTU (mean AUROC: `r format_decimal(rf_mean_auc$otu,digits=3)`) level data with no significant difference between the three (p \> 0.05, Figure 1A, Supplemental Figure 2).
+Performance with ASVs (mean AUROC: `r format_decimal(rf_mean_auc$asv,digits=3)`) was significantly lower than OTUs (p \< 0.01), but comparable to family (p = `r format_decimal(rf_pvals["family","asv"],digits=2)`) and genus (p = `r format_decimal(rf_pvals["genus","asv"],digits=2)`) levels (Figure 1A).
 These results suggest that increased resolution improves model performance up to the OTU level where further taxonomic resolution is not necessarily better for identifying individuals with SRNs based on microbiome composition.
@@
 While comparing AUROC values between models is a useful way to assess the overall model performance, they summarize the performance across all thresholds and can be misleading since models with the same AUROC can have different ROC curve shapes [@lobo2008].
-Depending on the intended implementation of the model, one may want to optimize the true-positive rate (sensitivity) over the false-positive rate (1-specificity) or vice versa.
+Depending on the intended implementation of the model, one may want to optimize the sensitivity over the specificity or vice versa.
 In this case, the optimal model will detect as many true positives (people with SRNs) as possible.
 To further compare the model performance across taxonomic levels we compared the sensitivity of the models at a specificity of 90%.
-The highest sensitivity values were observed for family, genus, and OTU level data (Figure 1B), consistent with the AUROC results.
-Phylum, class, order, and ASV sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (Figure 1B).
+The highest sensitivity values were observed for family (mean sensitivity: `r format_decimal(mean_sens_spec90$family,digits=2)`), genus (mean sensitivity: `r format_decimal(mean_sens_spec90$genus,digits=2)`), and OTU (mean sensitivity: `r format_decimal(mean_sens_spec90$otu,digits=2)`) level data (p > 0.05, Figure 1B), consistent with the AUROC results.
+Phylum (mean sensitivity: `r format_decimal(mean_sens_spec90$phylum,digits=2)`), class (mean sensitivity: `r format_decimal(mean_sens_spec90$class,digits=2)`), order (mean sensitivity: `r format_decimal(mean_sens_spec90$order,digits=2)`), and ASV (mean sensitivity: `r format_decimal(mean_sens_spec90$asv,digits=2)`) sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (p < 0.05,Figure 1B).
 This analysis further supports the observation that finer resolution does not improve SRN detection.
@@
 
-One hypothesis for the observation that model performance increases from phylum to OTU level then drops slightly at ASV level is that at higher taxonomic levels (e.g.
+One hypothesis for the observation that model performance increases from phylum to OTU level then drops at the ASV level is that at higher taxonomic levels (e.g.
 phylum) there are too few taxa and too much overlap to reliably differentiate between cases and controls.
@@
 
-These results demonstrate that consideration of the appropriate taxonomic resolution for utilizing the microbiome as a predictive tool is warranted.
+These results demonstrate a Goldilocks effect such that consideration of the appropriate taxonomic resolution for utilizing the microbiome as a predictive tool is warranted.
 In general, we found that finer taxonomic resolution (e.g.
-OTU and ASV) did not add additional sensitivity to predicting SRNs based on microbiome composition.
+ASV) did not add additional sensitivity to predicting SRNs based on microbiome composition.
 Family, genus, and OTU level data all performed equally.
-At the ASV level the fine resolution actually impedes model performance due to the sparsity of shared taxa and led to decreased model performance.
+At the ASV level the fine resolution actually impeded model performance due to the sparsity of shared taxa and led to decreased model performance.
 The tendency for ASV level annotation to split single bacterial genomes into multiple taxa [@schloss] could also be a contributing factor to the sparsity of shared taxa.
+Additionally, these results indicate that there are not specific individual bacterial strains that are useful to resolve SRNs, rather sets of closely related bacterial taxa.
 Overall, either family, genus, or OTU level taxonomy appear to perform equally for predicting subjects with SRNs based on the composition of the gut microbiome.
@@
 The resulting ASV table was subsampled for consistency with the mothur data.
-The results were consistent for ASVs generated with DADA2 (mean AUROC: `r format_decimal(dada2_mean_auc,digits=3)`).
+The DADA2 generated ASVs performed worse than the mothur generated ASVs (DADA2 ASV mean AUROC: `r format_decimal(dada2_mean_auc,digits=3)`, p < 0.05).
 
@@
 
-All code is available at: **TODO: link to code**
+All code is available at: <https://github.com/SchlossLab/Armour_Resolution_XXXX_2021>
 
@@
 
-::: {#refs}
-:::
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+\noindent
+
+<div id="refs"></div>
+
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
 
@@
 
-![](figure_s3.png){height="45%"}\
+![](figure_s3.png){height="45%"}  
 **Supplemental Figure 3: Prevalence of Taxa in Samples.** Distribution of the prevalence of taxa across samples at each taxonomic level.
@@
 
-![](figure_s4.png){height="85%"}\
+![](figure_s4.png){height="85%"}
 **Supplemental Figure 4: Top 10 important taxa at each taxonomic level.** Summary of the 10 most important taxa for the random forest models at each taxonomic level based on the average decrease in AUROC when the feature is permuted.

### 2021-10-19 paper/manuscript.tex
final paper edits

@@
 
-\hypertarget{taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}{%
-\section{Taxonomic Resolution Matters for Microbiome-Based
-Classification of Colorectal
-Cancer}\label{taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}}
-
-alternative titles:\\
-High resolution taxonomy impedes microbiome-based classification of
-colorectal cancer\\
-High resolution ASVs impede microbiome-based classification of
-colorectal cancer\\
-Mid-level taxonomic resolution is optimal for microbiome based
-classification of colorectal cancer\\
-Too much of a good thing: high resolution ASVs do not improve
-microbiome-based classification of colorectal cancer\\
-More is not always better: high resolution ASVs do not improve
-microbiome-based classification of colorectal cancer
+\hypertarget{a-goldilocks-principle-for-the-gut-microbiome-taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}{%
+\section{A Goldilocks Principle for the Gut Microbiome: Taxonomic
+Resolution Matters for Microbiome-Based Classification of Colorectal
+Cancer}\label{a-goldilocks-principle-for-the-gut-microbiome-taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}}
 
@@
 
-Courtney R. Armour\({^1}\), Begüm D. Topçuoğlu\(^{1,2}\), Andrea
+Courtney R. Armour\({^1}\), Begüm D. Topçuoğlu\(^{1,\#}\), Andrea
 Garretto\({^1}\), Patrick D. Schloss \({^1}\)\({^*}\)
@@
 
-\({^2}\) Exploratory Science Center, Merck \& Co., Inc., Cambridge,
-Massachusetts, USA~
+\({^\#}\) Current Affiliation: Exploratory Science Center, Merck \& Co.,
+Inc., Cambridge, Massachusetts, USA~
 
@@
 utilize taxonomic abundance data from operational taxonomic units (OTUs)
-or amplicon sequence variants (ASVs). However, it is unknown which
-taxonomic resolution is optimal for microbiome-based classification of
-colorectal cancer. To address this question, we used a reproducible
-machine learning framework to quantify classification performance of
-models based on data annotated to phylum, class, order, family, genus,
-OTU, and ASV levels. We found that model performance increased with
-increasing taxonomic resolution, up to the family level where
-performance was equal among family, genus, and OTU data before
-decreasing at the ASV level. These results are a step towards optimizing
-a microbiome based diagnostic tool.
+or amplicon sequence variants (ASVs) with the goal of increasing
+taxonomic resolution. However, it is unknown which taxonomic resolution
+is optimal for microbiome-based classification of colorectal cancer. To
+address this question, we used a reproducible machine learning framework
+to quantify classification performance of models based on data annotated
+to phylum, class, order, family, genus, OTU, and ASV levels. We found
+that model performance increased with increasing taxonomic resolution,
+up to the family level where performance was equal (p \textgreater{}
+0.05) among family (mean AUC: 0.689), genus (mean AUC: 0.690), and OTU
+(mean AUC: 0.693) levels before decreasing at the ASV level (p
+\textless{} 0.05, mean AUC: 0.676). These results demonstrate a
+trade-off between taxonomic resolution and prediction performance, where
+coarse taxonomic resolution (e.g.~phylum) is not distinct enough but
+fine resolution (e.g.~ASV) is to individualized to accurately classify
+samples. Similar to the story of Goldilocks and the three bears,
+mid-range resolution (i.e.~family, genus, and OTU) is just right for
+optimal prediction of colorectal cancer from microbiome data.\\
+(238 words)
 
@@
 non-invasive detection methods could greatly improve our ability to
-identify and treat early stages of disease. While the microbiome has
-shown promise as a diagnostic tool, there are numerous details to refine
-before a diagnostic could be widely deployed. Among these are the need
-for standardization of sample collection, storage, processing, and
-analysis. We address one aspect of data analysis by examining which
-taxonomic level provides the optimal resolution for microbiome-based
-classification of colorectal cancer. This is an important step to inform
-the development of microbiome-based diagnostic tools.
+identify and treat early stages of disease. The microbiome has shown
+promise as a resource for detection of colorectal cancer. Research on
+the gut microbiome tends to focus on improving our ability to profile
+species and strain level taxonomic resolution. However, we found that
+finer resolution impedes the ability to predict colorectal cancer based
+on the gut microbiome. These results highlight the need for
+consideration of the appropriate taxonomic resolution for microbiome
+analyses and that finer resolution is not always more informative.\\
+(110 words)
 
@@
 Colorectal cancer is one of the most common cancers in men and women and
-a leading cause of cancer related deaths in the United States
-(\protect\hyperlink{ref-siegel2020}{1}). Early detection and treatment
-are essential to increase survival rates, but for reasons such as
-invasiveness and high screening costs (i.e.~colonoscopy), many people do
-not comply with recommended screening guidelines
-(\protect\hyperlink{ref-garcuxeda2011}{2}). This prompts a need for low
-cost, non-invasive detection methods. A growing body of research points
-to the gut microbiome as a promising target for non-invasive detection
-of screen relevant neoplasia (SRNs) consisting of advanced adenomas and
-carcinomas (\protect\hyperlink{ref-zackular2014}{3},
-\protect\hyperlink{ref-zeller2014}{4}). The diagnostic potential of the
-gut microbiome in detecting SRNs has been explored through machine
-learning (ML) methods using abundances of operational taxonomic unit
-(OTU) classifications based on amplicon sequencing of the 16S rRNA gene
-(\protect\hyperlink{ref-zackular2014}{3}). Recent work has pushed for
-the use of amplicon sequence variants (ASVs) to replace OTUs for
-marker-gene analysis because of the improved resolution with ASVs
-(\protect\hyperlink{ref-callahan2017}{5}). However, it is unclear if
-OTUs are the optimal taxonomic resolution for classifying SRNs from
-microbiome data or whether the additional resolution provided by ASVs is
-useful for ML classification. Topçuoğlu \emph{et al}
-(\protect\hyperlink{ref-topuxe7uolu2020}{6}) recently demonstrated how
-to effectively apply machine learning (ML) methods to microbiome based
-classification problems and developed a framework for applying ML
-practices in a more reproducible way (mikropml). This analysis utilizes
-the reproducible framework developed by Topçuoğlu \emph{et al} to
-determine which ML method and taxonomic level produce the best
-performing classifier for detecting SRNs from microbiome data.
+a leading cause of cancer related deaths in the United States (1). Early
+detection and treatment are essential to increase survival rates, but
+for reasons such as invasiveness and high screening costs
+(i.e.~colonoscopy), many people do not comply with recommended screening
+guidelines (2). This prompts a need for low cost, non-invasive detection
+methods. A growing body of research points to the gut microbiome as a
+promising target for non-invasive detection of screen relevant neoplasia
+(SRNs) consisting of advanced adenomas and carcinomas (3, 4). The
+diagnostic potential of the gut microbiome in detecting SRNs has been
+explored through machine learning (ML) methods using abundances of
+operational taxonomic unit (OTU) classifications based on amplicon
+sequencing of the 16S rRNA gene (3). Recent work has pushed for the use
+of amplicon sequence variants (ASVs) to replace OTUs for marker-gene
+analysis because of the improved resolution with ASVs (5). However, it
+is unclear if OTUs are the optimal taxonomic resolution for classifying
+SRNs from microbiome data or whether the additional resolution provided
+by ASVs is useful for ML classification. Topçuoğlu \emph{et al} (6)
+recently demonstrated how to effectively apply machine learning (ML)
+methods to microbiome based classification problems and developed a
+framework for applying ML practices in a more reproducible way. This
+analysis utilizes the reproducible framework developed by Topçuoğlu
+\emph{et al} to determine which ML method and taxonomic level produce
+the best performing classifier for detecting SRNs from microbiome data.
 
@@
 patients with SRNs and healthy controls, we generated taxonomic
-abundance tables with mothur (\protect\hyperlink{ref-schloss2009}{7})
-annotated to phylum, class, order, family, genus, OTU and ASV levels.
-Using the taxonomic abundance data and the mikropml R package
-(\protect\hyperlink{ref-topuxe7uolu2021}{8}), we quantified how reliably
-samples could be classified as SRN or normal using five machine learning
-methods including random forest, L2-regularized logistic regression,
-decision tree, gradient boosted trees (XGBoost), and support vector
-machine with radial basis kernel (SVM radial). Across the five machine
-learning methods tested, model performance increased with increasing
-taxonomic level usually peaking around genus/OTU level before dropping
-off slightly with ASVs (Supplemental Figure 1). Regardless of the
-taxonomic level, random forest (RF) models consistently had the largest
-area under the receiver operating characteristic curve (AUROC). Within
-the RF model, the highest AUROCs were observed for family (mean AUROC:
-0.689), genus (mean AUROC: 0.690), and OTU (mean AUROC: 0.693) level
-data with no significant difference between the three (Figure 1A,
-Supplemental Figure 2). Performance with ASVs (mean AUROC: 0.676) was
-significantly lower than OTUs (p \textless{} 0.01), but comparable to
-family (p = 0.06) and genus (p = 0.05) levels (Figure 1A). These results
-suggest that increased resolution improves model performance up to the
-OTU level where further taxonomic resolution is not necessarily better
-for identifying individuals with SRNs based on microbiome composition.
+abundance tables with mothur (7) annotated to phylum, class, order,
+family, genus, OTU and ASV levels. Using the taxonomic abundance data
+and the mikropml R package (8), we quantified how reliably samples could
+be classified as SRN or normal using five machine learning methods
+including random forest, L2-regularized logistic regression, decision
+tree, gradient boosted trees (XGBoost), and support vector machine with
+radial basis kernel (SVM radial). Across the five machine learning
+methods tested, model performance increased with increasing taxonomic
+level usually peaking around genus/OTU level before dropping off
+slightly with ASVs (Supplemental Figure 1). Regardless of the taxonomic
+level, random forest (RF) models consistently had the largest area under
+the receiver operating characteristic curve (AUROC). Within the RF
+model, the highest AUROCs were observed for family (mean AUROC: 0.689),
+genus (mean AUROC: 0.690), and OTU (mean AUROC: 0.693) level data with
+no significant difference between the three (p \textgreater{} 0.05,
+Figure 1A, Supplemental Figure 2). Performance with ASVs (mean AUROC:
+0.676) was significantly lower than OTUs (p \textless{} 0.01), but
+comparable to family (p = 0.06) and genus (p = 0.05) levels (Figure 1A).
+These results suggest that increased resolution improves model
+performance up to the OTU level where further taxonomic resolution is
+not necessarily better for identifying individuals with SRNs based on
+microbiome composition.
 
@@
 thresholds and can be misleading since models with the same AUROC can
-have different ROC curve shapes (\protect\hyperlink{ref-lobo2008}{9}).
-Depending on the intended implementation of the model, one may want to
-optimize the sensitivity over the specificity or vice versa. In this
-case, the optimal model will detect as many true positives (people with
-SRNs) as possible. To further compare the model performance across
-taxonomic levels we compared the sensitivity of the models at a
-specificity of 90\%. The highest sensitivity values were observed for
-family, genus, and OTU level data (Figure 1B), consistent with the AUROC
-results. Phylum, class, order, and ASV sensitivity values were all
-significantly lower than family, genus, and OTU sensitivity values
-(Figure 1B). This analysis further supports the observation that finer
-resolution does not improve SRN detection.
+have different ROC curve shapes (9). Depending on the intended
+implementation of the model, one may want to optimize the sensitivity
+over the specificity or vice versa. In this case, the optimal model will
+detect as many true positives (people with SRNs) as possible. To further
+compare the model performance across taxonomic levels we compared the
+sensitivity of the models at a specificity of 90\%. The highest
+sensitivity values were observed for family (mean sensitivity: 0.38),
+genus (mean sensitivity: 0.39), and OTU (mean sensitivity: 0.37) level
+data (p \textgreater{} 0.05, Figure 1B), consistent with the AUROC
+results. Phylum (mean sensitivity: 0.21), class (mean sensitivity:
+0.22), order (mean sensitivity: 0.28), and ASV (mean sensitivity: 0.32)
+sensitivity values were all significantly lower than family, genus, and
+OTU sensitivity values (p \textless{} 0.05,Figure 1B). This analysis
+further supports the observation that finer resolution does not improve
+SRN detection.
 
 One hypothesis for the observation that model performance increases from
-phylum to OTU level then drops slightly at ASV level is that at higher
+phylum to OTU level then drops at the ASV level is that at higher
 taxonomic levels (e.g.~phylum) there are too few taxa and too much
@@
 (3.5\%) remained after preprocessing (Table 1). While the resolution
-provided by ASVs is useful in certain
-contexts(\protect\hyperlink{ref-eren2013}{10},
-\protect\hyperlink{ref-eren2015}{11}), these results suggest that the
-resolution is too fine for use in machine learning classification of
-SRNs based on microbiome composition.
+provided by ASVs is useful in certain contexts(10, 11), these results
+suggest that the resolution is too fine for use in machine learning
+classification of SRNs based on microbiome composition.
 
@@
 
-These results demonstrate that consideration of the appropriate
-taxonomic resolution for utilizing the microbiome as a predictive tool
-is warranted. In general, we found that finer taxonomic resolution
-(e.g.~OTU and ASV) did not add additional sensitivity to predicting SRNs
-based on microbiome composition. Family, genus, and OTU level data all
-performed equally. At the ASV level the fine resolution actually impedes
-model performance due to the sparsity of shared taxa and led to
+These results demonstrate a Goldilocks effect such that consideration of
+the appropriate taxonomic resolution for utilizing the microbiome as a
+predictive tool is warranted. In general, we found that finer taxonomic
+resolution (e.g.~ASV) did not add additional sensitivity to predicting
+SRNs based on microbiome composition. Family, genus, and OTU level data
+all performed equally. At the ASV level the fine resolution actually
+impeded model performance due to the sparsity of shared taxa and led to
 decreased model performance. The tendency for ASV level annotation to
-split single bacterial genomes into multiple taxa
-(\protect\hyperlink{ref-schloss}{12}) could also be a contributing
-factor to the sparsity of shared taxa. Additionally, these results
-indicate that there are not specific individual bacterial strains that
-are useful to resolve SRNs, rather sets of closely related bacterial
-taxa. Overall, either family, genus, or OTU level taxonomy appear to
-perform equally for predicting subjects with SRNs based on the
+split single bacterial genomes into multiple taxa (12) could also be a
+contributing factor to the sparsity of shared taxa. Additionally, these
+results indicate that there are not specific individual bacterial
+strains that are useful to resolve SRNs, rather sets of closely related
+bacterial taxa. Overall, either family, genus, or OTU level taxonomy
+appear to perform equally for predicting subjects with SRNs based on the
 composition of the gut microbiome. A potential benefit of utilizing
@@
 \textbf{\emph{Dataset.}} Raw 16S rRNA gene amplicon sequence data
-isolated from human gut samples (\protect\hyperlink{ref-baxter2016}{13})
-was downloaded from NCBI Sequence Read Archive (SRP062005). This dataset
-contains stool samples from 490 subjects. Based on the available
-metadata, samples categorized as normal, high risk normal, or adenoma
-were labeled ``normal'' for this analysis and samples categorized as
-advanced adenoma or carcinoma were labeled as ``screen relevant
-neoplasia'' (SRN). This resulted in a total of 261 ``normal'' samples
-and 229 ``SRN'' samples.
+isolated from human gut samples (13) was downloaded from NCBI Sequence
+Read Archive (SRP062005). This dataset contains stool samples from 490
+subjects. Based on the available metadata, samples categorized as
+normal, high risk normal, or adenoma were labeled ``normal'' for this
+analysis and samples categorized as advanced adenoma or carcinoma were
+labeled as ``screen relevant neoplasia'' (SRN). This resulted in a total
+of 261 ``normal'' samples and 229 ``SRN'' samples.
 
 \textbf{\emph{Data processing.}} Sequence data was processed with mothur
-(1.44.3) (\protect\hyperlink{ref-schloss2009}{7}) using the SILVA
-reference database (v132) (\protect\hyperlink{ref-quast2013}{14}) to
-produce count tables for phylum, class, order, family, genus, OTU, and
-ASV following the Schloss Lab MiSeq SOP described on the mothur website
+(1.44.3) (7) using the SILVA reference database (v132) (14) to produce
+count tables for phylum, class, order, family, genus, OTU, and ASV
+following the Schloss Lab MiSeq SOP described on the mothur website
 (\url{https://mothur.org/wiki/miseq_sop/}). ASV level data was also
-produced using DADA2 (\protect\hyperlink{ref-callahan2016}{15}) to
-ensure consistent results with a different pipeline. Data was processed
-following the DADA2 pipeline, but setting pool=TRUE to infer ASVs from
-the whole dataset rather than per sample. The resulting ASV table was
-subsampled for consistency with the mothur data. The results were
-consistent for ASVs generated with DADA2 (mean AUROC: 0.659).
+produced using DADA2 (15) to ensure consistent results with a different
+pipeline. Data was processed following the DADA2 pipeline, but setting
+pool=TRUE to infer ASVs from the whole dataset rather than per sample.
+The resulting ASV table was subsampled for consistency with the mothur
+data. The DADA2 generated ASVs performed worse than the mothur generated
+ASVs (DADA2 ASV mean AUROC: 0.659, p \textless{} 0.05).
 
 \textbf{\emph{Machine Learning.}} Machine learning models were run with
-the R package mikropml (v0.0.2)
-(\protect\hyperlink{ref-topuxe7uolu2021}{8}) to predict the diagnosis
-category (normal vs SRN) of each sample. Data was preprocessed to
-normalize values (scale/center), remove values with zero or near-zero
-variance, and collapse collinear features using default parameters.
-Initially the models were run with default hyperparameters, but were
-expanded if the peak performance was at the edge of the hyperparameter
-range. Each taxonomic model taxonomic level combination (e.g.~random
-forest on genus) was run with 100 different seeds. Each seed split the
-data into a training (80\%) and testing (20\%) set, and output
-performance of the training and testing as area under the receiver
-operating curve (AUROC).
+the R package mikropml (v0.0.2) (8) to predict the diagnosis category
+(normal vs SRN) of each sample. Data was preprocessed to normalize
+values (scale/center), remove values with zero or near-zero variance,
+and collapse collinear features using default parameters. Initially the
+models were run with default hyperparameters, but were expanded if the
+peak performance was at the edge of the hyperparameter range. Each
+taxonomic model taxonomic level combination (e.g.~random forest on
+genus) was run with 100 different seeds. Each seed split the data into a
+training (80\%) and testing (20\%) set, and output performance of the
+training and testing as area under the receiver operating curve (AUROC).
 
 To compare performance between taxonomic levels and models, P values
-were calculated as previously described
-(\protect\hyperlink{ref-topuxe7uolu2020}{6}). To compare sensitivity at
+were calculated as previously described (6). To compare sensitivity at
 90\% specificity, probabilities on the test dataset were collected for
@@
 \subsubsection{References}\label{references}}
+
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+
+\noindent
 
@@
 
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
+
 \newpage
@@
 
-\includegraphics[width=\textwidth,height=0.85\textheight]{figure_s4.png}\\
+\includegraphics[width=\textwidth,height=0.85\textheight]{figure_s4.png}
 \textbf{Supplemental Figure 4: Top 10 important taxa at each taxonomic

### 2021-10-20 code/R/figures/top_10_important_tax.R
change Otu to OTU

@@ -109,7 +109,7 @@ top_10_data <- inner_join(rf_imp_tax,top_10_tax,by=c("level","OTU","Taxonomy")) 
   mutate(name = str_replace(name,"\\*(.*)\\?(.*)\\*","\\*\\1\\* \\2")) %>%
   mutate(name = gsub("\\?"," ",name)) %>%
   mutate(name = gsub("_"," ",name)) %>%
-  mutate(OTUname = str_replace(OTU,"Otu0*([0-9]*)","Otu \\1")) %>%
+  mutate(OTUname = str_replace(OTU,"Otu0*([0-9]*)","OTU \\1")) %>%
   mutate(name = gsub(" $","",name)) %>%
   mutate(name = glue("{name}({OTUname})"))
 

### 2021-10-20 paper/manuscript.Rmd
remove hyphens, add running title, update figS4

@@
   - \usepackage{multirow}
+  - \usepackage[none]{hyphenat}
 ---
@@
 # A Goldilocks Principle for the Gut Microbiome: Taxonomic Resolution Matters for Microbiome-Based Classification of Colorectal Cancer
+
+\vspace{10mm}
+
+**Running title:** A Goldilocks Principle for the Gut Microbiome
 
@@
 family, genus, and OTU) is just right for optimal prediction of colorectal cancer from microbiome data.
-(238 words)
 
@@
 These results highlight the need for consideration of the appropriate taxonomic resolution for microbiome analyses and that finer resolution is not always more informative.
-(110 words)
 

### 2021-10-20 paper/manuscript.tex
remove hyphens, add running title, update figS4

@@
 \usepackage{multirow}
+\usepackage[none]{hyphenat}
 \usepackage{booktabs}
@@
 Cancer}\label{a-goldilocks-principle-for-the-gut-microbiome-taxonomic-resolution-matters-for-microbiome-based-classification-of-colorectal-cancer}}
+
+\vspace{10mm}
+
+\textbf{Running title:} A Goldilocks Principle for the Gut Microbiome
 
@@
 mid-range resolution (i.e.~family, genus, and OTU) is just right for
-optimal prediction of colorectal cancer from microbiome data.\\
-(238 words)
+optimal prediction of colorectal cancer from microbiome data.
 
@@
 consideration of the appropriate taxonomic resolution for microbiome
-analyses and that finer resolution is not always more informative.\\
-(110 words)
+analyses and that finer resolution is not always more informative.
 

### 2021-11-19 paper/manuscript.Rmd
update Begum affiliation
change equally to similarly lines 106,111

@@
 
-${^\#}$ Current Affiliation: Exploratory Science Center, Merck & Co., Inc., Cambridge, Massachusetts, USA
+${^\#}$ Current Affiliation: Bristol Myers Squibb, Summit, New Jersey, USA
 
@@
 ASV) did not add additional sensitivity to predicting SRNs based on microbiome composition.
-Family, genus, and OTU level data all performed equally.
+Family, genus, and OTU level data all performed similarly.
 At the ASV level the fine resolution actually impeded model performance due to the sparsity of shared taxa and led to decreased model performance.
@@
 Additionally, these results indicate that there are not specific individual bacterial strains that are useful to resolve SRNs, rather sets of closely related bacterial taxa.
-Overall, either family, genus, or OTU level taxonomy appear to perform equally for predicting subjects with SRNs based on the composition of the gut microbiome.
+Overall, either family, genus, or OTU level taxonomy appear to perform similarly for predicting subjects with SRNs based on the composition of the gut microbiome.
 A potential benefit of utilizing genus or family level data could be that it may allow for merging data generated from different 16S rRNA gene regions or sequencing platforms.

### 2021-11-19 paper/manuscript.Rmd
change figure s3 legend
adjust figure S4 legend

@@
 Percent of samples is split into 10 groups where the first is for taxa present in 0 to 10% of samples, then \>10% to 20% of samples, and so on.
-The total number of taxa for each taxonomic level after preprocessing is in parenthesis next to the title of the plot (or the name of the taxonomic level).
+The total number of taxa for each taxonomic level after preprocessing is in parenthesis next to the title of the plot.
 
@@
 Dot represents the mean decrease in AUROC and the lines extending from the dot represent the standard deviation from the mean.
+Taxonomic annotations are only reliably assigned to the genus level, therefore the OTU and ASV plots have genus labels.

### 2021-11-19 code/R/figures/AUC_all_model_all_level.R
change format from png to tiff

@@ -51,4 +51,6 @@ df_all %>%
         axis.text.x = element_text(angle = 0),
         axis.text = element_text(size=14)) +
   scale_fill_manual(values=pal2,name="Model")
-ggsave("analysis/figures/AUC_all_model_all_level.png",width = 7,height=4,units="in")
+ggsave("analysis/figures/AUC_all_model_all_level.tiff",
+       width = 6.87,height=3.5,units="in",
+       compression = "lzw")

### 2021-11-19 code/R/figures/averaged_ROC_curve.R

@@ -86,4 +86,6 @@ avg_sens %>%
               size=3) +
     scale_fill_manual(values=colors2) +
     scale_color_manual(values=colors2)
-ggsave("analysis/figures/average_roc_by_level.png",width=7,height=5)
+ggsave("analysis/figures/average_roc_by_level.tiff",
+       width=6.87,height=5,
+       compression = "lzw")

### 2021-11-19 code/R/figures/figure_1.R

@@ -1,14 +1,138 @@
-library(png)
-library(grid)
-library(gridExtra)
-library(tidyverse)
+# library(png)
+# library(grid)
+# library(gridExtra)
+# library(tidyverse)
 
-img1 <-  rasterGrob(as.raster(readPNG("analysis/figures/rf_AUC_all_level.png")), interpolate = FALSE)
-img2 <-  rasterGrob(as.raster(readPNG("analysis/figures/sensitivity_at_90_specificity.png")), interpolate = FALSE)
-grid.arrange(arrangeGrob(img1, left = textGrob("A", x = unit(1, "npc"), y = unit(0.85, "npc"), gp = gpar(fontface="bold"))),
-             arrangeGrob(img2, left = textGrob("B", x = unit(1, "npc"), y = unit(0.85, "npc"), gp = gpar(fontface="bold"))),
-             ncol = 2)
-g <- arrangeGrob(arrangeGrob(img1, left = textGrob("A", x = unit(1, "npc"), y = unit(0.95, "npc"), gp = gpar(fontface="bold"))),
-            arrangeGrob(img2, left = textGrob("B", x = unit(1, "npc"), y = unit(0.95, "npc"), gp = gpar(fontface="bold"))),
-            ncol = 2)
-ggsave("paper/figure_1.png",g,height=5,width=10)
+# img1 <-  rasterGrob(as.raster(readPNG("analysis/figures/rf_AUC_all_level.png")), interpolate = FALSE)
+# img2 <-  rasterGrob(as.raster(readPNG("analysis/figures/sensitivity_at_90_specificity.png")), interpolate = FALSE)
+# g <- grid.arrange(arrangeGrob(img1, left = textGrob("A", x = unit(1, "npc"), y = unit(0.95, "npc"), gp = gpar(fontface="bold"))),
+#              arrangeGrob(img2, left = textGrob("B", x = unit(1, "npc"), y = unit(0.95, "npc"), gp = gpar(fontface="bold"))),
+#              ncol = 2)
+# # g <- arrangeGrob(arrangeGrob(img1, left = textGrob("A", x = unit(1.2, "npc"), y = unit(0.95, "npc"), gp = gpar(fontface="bold"))),
+# #             arrangeGrob(img2, left = textGrob("B", x = unit(1, "npc"), y = unit(0.95, "npc"), gp = gpar(fontface="bold"))),
+# #             ncol = 2)
+# ggsave(filename = "paper/figure_1.tiff",
[…]

### 2021-11-19 code/R/figures/sensitivity_at_90_specificity.R

@@ -42,4 +42,5 @@ sens_spec90 %>%
     scale_y_continuous(breaks=seq(0,1,0.1),labels = scales::percent_format(accuracy = 1)) +
     scale_color_manual(values=rev(colors2)) +
     annotate("text",x=levels_names,y=rep((max(sens_spec90$sensitivity)+0.05),length(levels_names)),label=c("W","W","X","Y","Y","Y","Z"))
-ggsave("analysis/figures/sensitivity_at_90_specificity.png",width = 4.5,height=5,units="in")
+ggsave("analysis/figures/sensitivity_at_90_specificity.png",
+       width = 4.5,height=5)

### 2021-11-19 code/R/figures/top_10_important_tax.R

@@ -116,4 +116,7 @@ top_10_data <- inner_join(rf_imp_tax,top_10_tax,by=c("level","OTU","Taxonomy")) 
 plotlist <- map(levels_names,plot_top_imp,top_data=top_10_data)
 
 plot_grid(plotlist = plotlist,ncol = 2,align = "v")
-ggsave("analysis/figures/top_10_important_tax.png",width = 9, height = 11)
+ggsave("analysis/figures/top_10_important_tax.tiff",
+       scale=1.25,
+       height = 9,width = 6.87,units="in",
+       compression = "lzw")

### 2021-11-19 paper/manuscript.Rmd
undo change equally to similarly

@@
 ASV) did not add additional sensitivity to predicting SRNs based on microbiome composition.
-Family, genus, and OTU level data all performed similarly.
-At the ASV level the fine resolution actually impeded model performance due to the sparsity of shared taxa and led to decreased model performance.
+Family, genus, and OTU level data all performed equally At the ASV level the fine resolution actually impeded model performance due to the sparsity of shared taxa and led to decreased model performance.
 The tendency for ASV level annotation to split single bacterial genomes into multiple taxa [@schloss] could also be a contributing factor to the sparsity of shared taxa.
 Additionally, these results indicate that there are not specific individual bacterial strains that are useful to resolve SRNs, rather sets of closely related bacterial taxa.
-Overall, either family, genus, or OTU level taxonomy appear to perform similarly for predicting subjects with SRNs based on the composition of the gut microbiome.
+Overall, either family, genus, or OTU level taxonomy appear to perform equally for predicting subjects with SRNs based on the composition of the gut microbiome.
 A potential benefit of utilizing genus or family level data could be that it may allow for merging data generated from different 16S rRNA gene regions or sequencing platforms.

### 2021-11-19 paper/manuscript.tex
undo change equally to similarly

@@
 
-\({^\#}\) Current Affiliation: Exploratory Science Center, Merck \& Co.,
-Inc., Cambridge, Massachusetts, USA~
+\({^\#}\) Current Affiliation: Bristol Myers Squibb, Summit, New Jersey,
+USA~
 
@@
 SRNs based on microbiome composition. Family, genus, and OTU level data
-all performed equally. At the ASV level the fine resolution actually
+all performed equally At the ASV level the fine resolution actually
 impeded model performance due to the sparsity of shared taxa and led to
@@
 samples, and so on. The total number of taxa for each taxonomic level
-after preprocessing is in parenthesis next to the title of the plot (or
-the name of the taxonomic level).
+after preprocessing is in parenthesis next to the title of the plot.
 
@@
 and the lines extending from the dot represent the standard deviation
-from the mean.
+from the mean. Taxonomic annotations are only reliably assigned to the
+genus level, therefore the OTU and ASV plots have genus labels.
 

### 2021-11-19 paper/manuscript.Rmd
remove figures from pdf

@@
 
-### Figures
-
-![](figure_1.png)
-
-**Figure 1: Random Forest Model Performance.** **A)** Strip plot of the area under the receiver operating characteristic curve (AUROC) values on the test dataset for 100 seeds predicting SRNs using a random forest model.
-Black points denote the mean and lines denote the standard deviation.
-Dashed line denotes AUROC of 0.5 which is equivalent to random classification.
-Significance between taxonomic levels was quantified by comparing the difference in mean AUROC and is denoted by letters A through E on the right side of the plot; taxonomic levels with the same letter are in the same significance group and are not significantly different from one another.
-**B)** Strip plot of the sensitivity at a specificity of 90% across the 100 model iterations for each taxonomic level.
-Black points denote the mean and the lines denote the standard deviation.
-The letters W through Z on the right side of the plot denote the significance groups.
-
-\newpage
-
 ### Tables
@@
 
-### Supplemental Figures
+### Figure Legends
 
-![](figure_s1.png)
+**Figure 1: Random Forest Model Performance.** **A)** Strip plot of the area under the receiver operating characteristic curve (AUROC) values on the test dataset for 100 seeds predicting SRNs using a random forest model.
+Black points denote the mean and lines denote the standard deviation.
+Dashed line denotes AUROC of 0.5 which is equivalent to random classification.
+Significance between taxonomic levels was quantified by comparing the difference in mean AUROC and is denoted by letters A through E on the right side of the plot; taxonomic levels with the same letter are in the same significance group and are not significantly different from one another.
+**B)** Strip plot of the sensitivity at a specificity of 90% across the 100 model iterations for each taxonomic level.
+Black points denote the mean and the lines denote the standard deviation.
+The letters W through Z on the right side of the plot denote the significance groups.
+
+\newpage
+
+### Supplemental Figure Legends
 
@@
 Due to the random split of data into training and testing sets, each model was run across 100 seeds to account for variation in training/test datasplits.
-
-\newpage
-
-![](figure_s2.png){height="50%"}
 
@@
 
-\newpage
-
-![](figure_s3.png){height="45%"}  
 **Supplemental Figure 3: Prevalence of Taxa in Samples.** Distribution of the prevalence of taxa across samples at each taxonomic level.
@@
 
-\newpage
-
-![](figure_s4.png){height="85%"}
 **Supplemental Figure 4: Top 10 important taxa at each taxonomic level.** Summary of the 10 most important taxa for the random forest models at each taxonomic level based on the average decrease in AUROC when the feature is permuted.
 Dot represents the mean decrease in AUROC and the lines extending from the dot represent the standard deviation from the mean.
-Taxonomic annotations are only reliably assigned to the genus level, therefore the OTU and ASV plots have genus labels.

### 2021-11-19 paper/manuscript.Rmd
add goldilocks citation

@@
 ### Abstract
+
 Colorectal cancer is a common and deadly disease in the United States accounting for over 50,000 deaths in 2020.
@@
 To address this question, we used a reproducible machine learning framework to quantify classification performance of models based on data annotated to phylum, class, order, family, genus, OTU, and ASV levels.
-We found that model performance increased with increasing taxonomic resolution, up to the family level where performance was equal (p > 0.05) among family (mean AUC: `r format_decimal(rf_mean_auc$family,digits=3)`), genus (mean AUC: `r format_decimal(rf_mean_auc$genus,digits=3)`), and OTU (mean AUC: `r format_decimal(rf_mean_auc$otu,digits=3)`) levels before decreasing at the ASV level (p < 0.05, mean AUC: `r format_decimal(rf_mean_auc$asv,digits=3)`).
+We found that model performance increased with increasing taxonomic resolution, up to the family level where performance was equal (p \> 0.05) among family (mean AUC: `r format_decimal(rf_mean_auc$family,digits=3)`), genus (mean AUC: `r format_decimal(rf_mean_auc$genus,digits=3)`), and OTU (mean AUC: `r format_decimal(rf_mean_auc$otu,digits=3)`) levels before decreasing at the ASV level (p \< 0.05, mean AUC: `r format_decimal(rf_mean_auc$asv,digits=3)`).
 These results demonstrate a trade-off between taxonomic resolution and prediction performance, where coarse taxonomic resolution (e.g.
@@
 ASV) is to individualized to accurately classify samples.
-Similar to the story of Goldilocks and the three bears, mid-range resolution (i.e.
+Similar to the story of Goldilocks and the three bears[@cauley1981], mid-range resolution (i.e.
 family, genus, and OTU) is just right for optimal prediction of colorectal cancer from microbiome data.
@@
 ### Importance
+
 Despite being highly preventable, colorectal cancer remains a leading cause of cancer related death in the United States.
@@
 To further compare the model performance across taxonomic levels we compared the sensitivity of the models at a specificity of 90%.
-The highest sensitivity values were observed for family (mean sensitivity: `r format_decimal(mean_sens_spec90$family,digits=2)`), genus (mean sensitivity: `r format_decimal(mean_sens_spec90$genus,digits=2)`), and OTU (mean sensitivity: `r format_decimal(mean_sens_spec90$otu,digits=2)`) level data (p > 0.05, Figure 1B), consistent with the AUROC results.
-Phylum (mean sensitivity: `r format_decimal(mean_sens_spec90$phylum,digits=2)`), class (mean sensitivity: `r format_decimal(mean_sens_spec90$class,digits=2)`), order (mean sensitivity: `r format_decimal(mean_sens_spec90$order,digits=2)`), and ASV (mean sensitivity: `r format_decimal(mean_sens_spec90$asv,digits=2)`) sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (p < 0.05,Figure 1B).
+The highest sensitivity values were observed for family (mean sensitivity: `r format_decimal(mean_sens_spec90$family,digits=2)`), genus (mean sensitivity: `r format_decimal(mean_sens_spec90$genus,digits=2)`), and OTU (mean sensitivity: `r format_decimal(mean_sens_spec90$otu,digits=2)`) level data (p \> 0.05, Figure 1B), consistent with the AUROC results.
+Phylum (mean sensitivity: `r format_decimal(mean_sens_spec90$phylum,digits=2)`), class (mean sensitivity: `r format_decimal(mean_sens_spec90$class,digits=2)`), order (mean sensitivity: `r format_decimal(mean_sens_spec90$order,digits=2)`), and ASV (mean sensitivity: `r format_decimal(mean_sens_spec90$asv,digits=2)`) sensitivity values were all significantly lower than family, genus, and OTU sensitivity values (p \< 0.05,Figure 1B).
 This analysis further supports the observation that finer resolution does not improve SRN detection.
@@
 The resulting ASV table was subsampled for consistency with the mothur data.
-The DADA2 generated ASVs performed worse than the mothur generated ASVs (DADA2 ASV mean AUROC: `r format_decimal(dada2_mean_auc,digits=3)`, p < 0.05).
+The DADA2 generated ASVs performed worse than the mothur generated ASVs (DADA2 ASV mean AUROC: `r format_decimal(dada2_mean_auc,digits=3)`, p \< 0.05).
 
@@
 
+```{=tex}
 \setlength{\parindent}{-0.25in}
 \setlength{\leftskip}{0.25in}
+```
 \noindent
 
-<div id="refs"></div>
+::: {#refs}
+:::
 
+```{=tex}
 \setlength{\parindent}{0in}
 \setlength{\leftskip}{0in}
-
+```
 \newpage

### 2021-11-19 paper/manuscript.tex
remove figures from pdf
add goldilocks citation

@@
 fine resolution (e.g.~ASV) is to individualized to accurately classify
-samples. Similar to the story of Goldilocks and the three bears,
+samples. Similar to the story of Goldilocks and the three bears(1),
 mid-range resolution (i.e.~family, genus, and OTU) is just right for
@@
 Colorectal cancer is one of the most common cancers in men and women and
-a leading cause of cancer related deaths in the United States (1). Early
+a leading cause of cancer related deaths in the United States (2). Early
 detection and treatment are essential to increase survival rates, but
@@
 (i.e.~colonoscopy), many people do not comply with recommended screening
-guidelines (2). This prompts a need for low cost, non-invasive detection
+guidelines (3). This prompts a need for low cost, non-invasive detection
 methods. A growing body of research points to the gut microbiome as a
 promising target for non-invasive detection of screen relevant neoplasia
-(SRNs) consisting of advanced adenomas and carcinomas (3, 4). The
+(SRNs) consisting of advanced adenomas and carcinomas (4, 5). The
 diagnostic potential of the gut microbiome in detecting SRNs has been
@@
 operational taxonomic unit (OTU) classifications based on amplicon
-sequencing of the 16S rRNA gene (3). Recent work has pushed for the use
+sequencing of the 16S rRNA gene (4). Recent work has pushed for the use
 of amplicon sequence variants (ASVs) to replace OTUs for marker-gene
-analysis because of the improved resolution with ASVs (5). However, it
+analysis because of the improved resolution with ASVs (6). However, it
 is unclear if OTUs are the optimal taxonomic resolution for classifying
 SRNs from microbiome data or whether the additional resolution provided
-by ASVs is useful for ML classification. Topçuoğlu \emph{et al} (6)
+by ASVs is useful for ML classification. Topçuoğlu \emph{et al} (7)
 recently demonstrated how to effectively apply machine learning (ML)
@@
 patients with SRNs and healthy controls, we generated taxonomic
-abundance tables with mothur (7) annotated to phylum, class, order,
+abundance tables with mothur (8) annotated to phylum, class, order,
 family, genus, OTU and ASV levels. Using the taxonomic abundance data
-and the mikropml R package (8), we quantified how reliably samples could
+and the mikropml R package (9), we quantified how reliably samples could
 be classified as SRN or normal using five machine learning methods
@@
 thresholds and can be misleading since models with the same AUROC can
-have different ROC curve shapes (9). Depending on the intended
+have different ROC curve shapes (10). Depending on the intended
 implementation of the model, one may want to optimize the sensitivity
@@
 (3.5\%) remained after preprocessing (Table 1). While the resolution
-provided by ASVs is useful in certain contexts(10, 11), these results
+provided by ASVs is useful in certain contexts(11, 12), these results
 suggest that the resolution is too fine for use in machine learning
@@
 decreased model performance. The tendency for ASV level annotation to
-split single bacterial genomes into multiple taxa (12) could also be a
+split single bacterial genomes into multiple taxa (13) could also be a
 contributing factor to the sparsity of shared taxa. Additionally, these
@@
 \textbf{\emph{Dataset.}} Raw 16S rRNA gene amplicon sequence data
-isolated from human gut samples (13) was downloaded from NCBI Sequence
+isolated from human gut samples (14) was downloaded from NCBI Sequence
 Read Archive (SRP062005). This dataset contains stool samples from 490
@@
 \textbf{\emph{Data processing.}} Sequence data was processed with mothur
-(1.44.3) (7) using the SILVA reference database (v132) (14) to produce
+(1.44.3) (8) using the SILVA reference database (v132) (15) to produce
 count tables for phylum, class, order, family, genus, OTU, and ASV
@@
 (\url{https://mothur.org/wiki/miseq_sop/}). ASV level data was also
-produced using DADA2 (15) to ensure consistent results with a different
+produced using DADA2 (16) to ensure consistent results with a different
 pipeline. Data was processed following the DADA2 pipeline, but setting
@@
 \textbf{\emph{Machine Learning.}} Machine learning models were run with
-the R package mikropml (v0.0.2) (8) to predict the diagnosis category
+the R package mikropml (v0.0.2) (9) to predict the diagnosis category
 (normal vs SRN) of each sample. Data was preprocessed to normalize
@@
 To compare performance between taxonomic levels and models, P values
-were calculated as previously described (6). To compare sensitivity at
+were calculated as previously described (7). To compare sensitivity at
 90\% specificity, probabilities on the test dataset were collected for
@@
 \begin{CSLReferences}{0}{0}
+\leavevmode\hypertarget{ref-cauley1981}{}%
+\CSLLeftMargin{1. }
+\CSLRightInline{\textbf{Cauley LB}. 1981. Goldilocks and the three
+bears. New York : Putnam, 1981.}
+
 \leavevmode\hypertarget{ref-siegel2020}{}%
-\CSLLeftMargin{1. }
+\CSLLeftMargin{2. }
 \CSLRightInline{\textbf{Siegel RL}, \textbf{Miller KD}, \textbf{Sauer
@@
 \leavevmode\hypertarget{ref-garcuxeda2011}{}%
-\CSLLeftMargin{2. }
+\CSLLeftMargin{3. }
 \CSLRightInline{\textbf{García G}, \textbf{Z A}. 2011. Factors
@@
 \leavevmode\hypertarget{ref-zackular2014}{}%
-\CSLLeftMargin{3. }
+\CSLLeftMargin{4. }
 \CSLRightInline{\textbf{Zackular JP}, \textbf{Rogers MAM},
@@
 \leavevmode\hypertarget{ref-zeller2014}{}%
-\CSLLeftMargin{4. }
+\CSLLeftMargin{5. }
 \CSLRightInline{\textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
@@
 \leavevmode\hypertarget{ref-callahan2017}{}%
-\CSLLeftMargin{5. }
+\CSLLeftMargin{6. }
 \CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
@@
 \leavevmode\hypertarget{ref-topuxe7uolu2020}{}%
-\CSLLeftMargin{6. }
+\CSLLeftMargin{7. }
 \CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
@@
 \leavevmode\hypertarget{ref-schloss2009}{}%
-\CSLLeftMargin{7. }
+\CSLLeftMargin{8. }
 \CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
@@
 \leavevmode\hypertarget{ref-topuxe7uolu2021}{}%
-\CSLLeftMargin{8. }
+\CSLLeftMargin{9. }
 \CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lapp Z}, \textbf{Sovacool
@@
 \leavevmode\hypertarget{ref-lobo2008}{}%
-\CSLLeftMargin{9. }
+\CSLLeftMargin{10. }
 \CSLRightInline{\textbf{Lobo JM}, \textbf{Jiménez-Valverde A},
@@
 \leavevmode\hypertarget{ref-eren2013}{}%
-\CSLLeftMargin{10. }
+\CSLLeftMargin{11. }
 \CSLRightInline{\textbf{Eren AM}, \textbf{Maignien L}, \textbf{Sul WJ},
@@
 \leavevmode\hypertarget{ref-eren2015}{}%
-\CSLLeftMargin{11. }
+\CSLLeftMargin{12. }
 \CSLRightInline{\textbf{Eren AM}, \textbf{Morrison HG}, \textbf{Lescault
@@
 \leavevmode\hypertarget{ref-schloss}{}%
-\CSLLeftMargin{12. }
+\CSLLeftMargin{13. }
 \CSLRightInline{\textbf{Schloss PD}. Amplicon sequence variants
@@
 \leavevmode\hypertarget{ref-baxter2016}{}%
-\CSLLeftMargin{13. }
+\CSLLeftMargin{14. }
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
@@
 \leavevmode\hypertarget{ref-quast2013}{}%
-\CSLLeftMargin{14. }
+\CSLLeftMargin{15. }
 \CSLRightInline{\textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
@@
 \leavevmode\hypertarget{ref-callahan2016}{}%
-\CSLLeftMargin{15. }
+\CSLLeftMargin{16. }
 \CSLRightInline{\textbf{Callahan BJ}, \textbf{McMurdie PJ},
@@
 \setlength{\leftskip}{0in}
-
-\newpage
-
-\hypertarget{figures}{%
-\subsubsection{Figures}\label{figures}}
-
-\includegraphics{figure_1.png}
-
-\textbf{Figure 1: Random Forest Model Performance.} \textbf{A)} Strip
-plot of the area under the receiver operating characteristic curve
-(AUROC) values on the test dataset for 100 seeds predicting SRNs using a
-random forest model. Black points denote the mean and lines denote the
-standard deviation. Dashed line denotes AUROC of 0.5 which is equivalent
-to random classification. Significance between taxonomic levels was
-quantified by comparing the difference in mean AUROC and is denoted by
-letters A through E on the right side of the plot; taxonomic levels with
-the same letter are in the same significance group and are not
-significantly different from one another. \textbf{B)} Strip plot of the
-sensitivity at a specificity of 90\% across the 100 model iterations for
-each taxonomic level. Black points denote the mean and the lines denote
-the standard deviation. The letters W through Z on the right side of the
-plot denote the significance groups.
 
@@
 
-\hypertarget{supplemental-figures}{%
-\subsubsection{Supplemental Figures}\label{supplemental-figures}}
+\hypertarget{figure-legends}{%
+\subsubsection{Figure Legends}\label{figure-legends}}
 
-\includegraphics{figure_s1.png}
+\textbf{Figure 1: Random Forest Model Performance.} \textbf{A)} Strip
+plot of the area under the receiver operating characteristic curve
+(AUROC) values on the test dataset for 100 seeds predicting SRNs using a
+random forest model. Black points denote the mean and lines denote the
+standard deviation. Dashed line denotes AUROC of 0.5 which is equivalent
+to random classification. Significance between taxonomic levels was
+quantified by comparing the difference in mean AUROC and is denoted by
+letters A through E on the right side of the plot; taxonomic levels with
+the same letter are in the same significance group and are not
+significantly different from one another. \textbf{B)} Strip plot of the
+sensitivity at a specificity of 90\% across the 100 model iterations for
+each taxonomic level. Black points denote the mean and the lines denote
+the standard deviation. The letters W through Z on the right side of the
+plot denote the significance groups.
+
+\newpage
+
+\hypertarget{supplemental-figure-legends}{%
+\subsubsection{Supplemental Figure
+Legends}\label{supplemental-figure-legends}}
 
@@
 
-\newpage
-
-\includegraphics[width=\textwidth,height=0.5\textheight]{figure_s2.png}
-
 \textbf{Supplemental Figure 2: Averaged ROC curves.} ROC curves with
@@
 
-\newpage
-
-\includegraphics[width=\textwidth,height=0.45\textheight]{figure_s3.png}\\
 \textbf{Supplemental Figure 3: Prevalence of Taxa in Samples.}
@@
 
-\newpage
-
-\includegraphics[width=\textwidth,height=0.85\textheight]{figure_s4.png}
 \textbf{Supplemental Figure 4: Top 10 important taxa at each taxonomic
@@
 and the lines extending from the dot represent the standard deviation
-from the mean. Taxonomic annotations are only reliably assigned to the
-genus level, therefore the OTU and ASV plots have genus labels.
+from the mean.
 
